### 2022-02-28 Paper_Draft1/Draft1.Rmd

---
title: Full title of the paper (Capitalized)
author:
  - name: Amrita Acharya
    affil: 1,2,\ddagger,*
    orcid: 0000-0003-3293-2315
  - name: Dianne Caravela
    affil: 2, \dagger, \ddagger
  - name: Eunice Kim
    affil: 3
  - name: Emma Kornberg
    affil: 4
  - name: Elisabeth Nesmith
    affil: 5
affiliation:
  - num: 1
    address: |
      Smith College
    email: aarcharya@smith.edu
  - num: 2
    address: |
      Smith College
    email: dcaravela@smith.edu
# firstnote to eighthnote
firstnote: |
  Current address: Updated affiliation
secondnote: |
  These authors contributed equally to this work.
correspondence: |
  leutnant@fh-muenster.de; Tel.: +XX-000-00-0000.
journal: water
type: article
status: submit
bibliography: mybibfile.bib
appendix: appendix.tex
simplesummary: |
  A Simple summary goes here.
abstract: |
  A single paragraph of about 200 words maximum. For research articles, 
  abstracts should give a pertinent overview of the work. We strongly encourage
  authors to use the following style of structured abstracts, but without 
  headings: 1) Background: Place the question addressed in a broad context and
  highlight the purpose of the study; 2) Methods: Describe briefly the main
  methods or treatments applied; 3) Results: Summarize the article's main 
  findings; and 4) Conclusion: Indicate the main conclusions or interpretations. 
  The abstract should be an objective representation of the article, it must not 
  contain results which are not presented and substantiated in the main text and 
  should not exaggerate the main conclusions.
keywords: |
  keyword 1; keyword 2; keyword 3 (list three to ten pertinent keywords specific 
  to the article, yet reasonably common within the subject discipline.).
acknowledgement: |
  All sources of funding of the study should be disclosed. Please clearly 
  indicate grants that you have received in support of your research work. 
  Clearly state if you received funds for covering the costs to publish in open 
  access.
authorcontributions: |
  For research articles with several authors, a short paragraph specifying their 
  individual contributions must be provided. The following statements should be 
  used ``X.X. and Y.Y. conceive and designed the experiments; X.X. performed the 
  experiments; X.X. and Y.Y. analyzed the data; W.W. contributed 
  reagents/materials/analysis tools; Y.Y. wrote the paper.'' Authorship must be
  limited to those who have contributed substantially to the work reported.
conflictsofinterest: |
  Declare conflicts of interest or state 'The authors declare no conflict of 
  interest.' Authors must identify and declare any personal circumstances or
  interest that may be perceived as inappropriately influencing the
  representation or interpretation of reported research results. Any role of the
  funding sponsors in the design of the study; in the collection, analyses or 
  interpretation of data in the writing of the manuscript, or in the decision to 
  publish the results must be declared in this section. If there is no role, 
  please state 'The founding sponsors had no role in the design of the study; 
  in the collection, analyses, or interpretation of data; in the writing of the 
  manuscript, an in the decision to publish the results'.
sampleavailability: |
  Samples of the compounds ...... are available from the authors.
abbreviations:
  - short: MDPI
    long: Multidisciplinary Digital Publishing Institute
  - short: DOAJ
    long: Directory of open access journals
  - short: TLA
    long: Three letter acronym
  - short: LD 
    long: linear dichroism
output: rticles::mdpi_article
---

# Version

This Rmd-skeleton uses the mdpi Latex template published 2019/02. 
However, the official template gets more frequently updated than the 'rticles'
package. Therefore, please make sure prior to paper submission, that you're 
using the most recent .cls, .tex and .bst files 
(available [here](http://www.mdpi.com/authors/latex)).

# Introduction

A recidivist is a term to describe criminals who re-offend.

# Data

The data we are using for this project is the Correctional Offender Management Profiling for Alternative Sanctions (COMPAS) Recidivism Risk Scores dataset from AIF360. Northpointe created COMPAS as a commercial risk assessment tool to score defendants on their potential recidivism and violent recidivism risk. The COMPAS dataset we are using has 6,167 rows and each row represents an arrest charge for a defendant. We also have data on the defendant's age, race, sex, what they were charged with, and whether or not the defendant ultimately recidivated.

# Results

This section may be divided by subheadings. It should provide a concise and precise description of the experimental results, their interpretation as well as the experimental conclusions that can be drawn.

## Subsection Heading Here

Subsection text here.

### Subsubsection Heading Here

Bulleted lists look like this:

* First bullet
* Second bullet
* Third bullet

Numbered lists can be added as follows:

1. First item
2. Second item
3. Third item

The text continues here.

All figures and tables should be cited in the main text as Figure 1, Table 1, etc.

\begin{figure}[H]
\centering
\includegraphics[width=3 cm]{logo-mdpi}
\caption{This is a figure, Schemes follow the same formatting. If there are multiple panels, they should be listed as: (\textbf{a}) Description of what is contained in the first panel. (\textbf{b}) Description of what is contained in the second panel. Figures should be placed in the main text near to the first time they are cited. A caption on a single line should be centered.}
\end{figure}

\begin{table}[H]
\caption{This is a table caption. Tables should be placed in the main text near to the first time they are cited.}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{ccc}
\toprule
\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}\\
\midrule
entry 1		& data			& data\\
entry 2		& data			& data\\
\bottomrule
\end{tabular}
\end{table}

This is an example of an equation:

\begin{equation}
\mathbb{S}
\end{equation}
<!-- If the documentclass option "submit" is chosen, please insert a blank line before and after any math environment (equation and eqnarray environments). This ensures correct linenumbering. The blank line should be removed when the documentclass option is changed to "accept" because the text following an equation should not be a new paragraph. -->

<!-- Please punctuate equations as regular text. Theorem-type environments (including propositions, lemmas, corollaries etc.) can be formatted as follows: -->
Example of a theorem:
\begin{Theorem}
Example text of a theorem.
\end{Theorem}

The text continues here. Proofs must be formatted as follows:

Example of a proof:
\begin{proof}[Proof of Theorem 1]
Text of the proof. Note that the phrase `of Theorem 1' is optional if it is clear which theorem is being referred to.
\end{proof}
The text continues here.

# Discussion

Authors should discuss the results and how they can be interpreted in perspective of previous studies and of the working hypotheses. The findings and their implications should be discussed in the broadest context possible. Future research directions may also be highlighted.

# Conclusion

This section is not mandatory, but can be added to the manuscript if the discussion is unusually long or complex.

# Patents

This section is not mandatory, but may be added if there are patents resulting from the work reported in this manuscript.

### 2022-02-28 Paper_Draft1/Draft1.tex

%  LaTeX support: latex@mdpi.com
%  In case you need support, please attach all files that are necessary for compiling as well as the log file, and specify the details of your LaTeX setup (which operating system and LaTeX version / tools you are using).

%=================================================================
\documentclass[water,article,submit,moreauthors,pdftex]{mdpi}

% If you would like to post an early version of this manuscript as a preprint, you may use preprint as the journal and change 'submit' to 'accept'. The document class line would be, e.g., \documentclass[preprints,article,accept,moreauthors,pdftex]{mdpi}. This is especially recommended for submission to arXiv, where line numbers should be removed before posting. For preprints.org, the editorial staff will make this change immediately prior to posting.

%% Some pieces required from the pandoc template
\setlist[itemize]{leftmargin=*,labelsep=5.8mm}
\setlist[enumerate]{leftmargin=*,labelsep=4.9mm}

%--------------------
% Class Options:
%--------------------
%----------
% journal
%----------
% Choose between the following MDPI journals:
% acoustics, actuators, addictions, admsci, aerospace, agriculture, agriengineering, agronomy, algorithms, animals, antibiotics, antibodies, antioxidants, applsci, arts, asc, asi, atmosphere, atoms, axioms, batteries, bdcc, behavsci , beverages, bioengineering, biology, biomedicines, biomimetics, biomolecules, biosensors, brainsci , buildings, cancers, carbon , catalysts, cells, ceramics, challenges, chemengineering, chemistry, chemosensors, children, cleantechnol, climate, clockssleep, cmd, coatings, colloids, computation, computers, condensedmatter, cosmetics, cryptography, crystals, dairy, data, dentistry, designs , diagnostics, diseases, diversity, drones, econometrics, economies, education, electrochem, electronics, energies, entropy, environments, epigenomes, est, fermentation, fibers, fire, fishes, fluids, foods, forecasting, forests, fractalfract, futureinternet, futurephys, galaxies, games, gastrointestdisord, gels, genealogy, genes, geohazards, geosciences, geriatrics, hazardousmatters, healthcare, heritage, highthroughput, horticulturae, humanities, hydrology, ijerph, ijfs, ijgi, ijms, ijns, ijtpp, informatics, information, infrastructures, inorganics, insects, instruments, inventions, iot, j, jcdd, jcm, jcp, jcs, jdb, jfb, jfmk, jimaging, jintelligence, jlpea, jmmp, jmse, jnt, jof, joitmc, jpm, jrfm, jsan, land, languages, laws, life, literature, logistics, lubricants, machines, magnetochemistry, make, marinedrugs, materials, mathematics, mca, medicina, medicines, medsci, membranes, metabolites, metals, microarrays, micromachines, microorganisms, minerals, modelling, molbank, molecules, mps, mti, nanomaterials, ncrna, neuroglia, nitrogen, notspecified, nutrients, ohbm, particles, pathogens, pharmaceuticals, pharmaceutics, pharmacy, philosophies, photonics, physics, plants, plasma, polymers, polysaccharides, preprints , proceedings, processes, proteomes, psych, publications, quantumrep, quaternary, qubs, reactions, recycling, religions, remotesensing, reports, resources, risks, robotics, safety, sci, scipharm, sensors, separations, sexes, signals, sinusitis, smartcities, sna, societies, socsci, soilsystems, sports, standards, stats, surfaces, surgeries, sustainability, symmetry, systems, technologies, test, toxics, toxins, tropicalmed, universe, urbansci, vaccines, vehicles, vetsci, vibration, viruses, vision, water, wem, wevj

%---------
% article
%---------
% The default type of manuscript is "article", but can be replaced by:
% abstract, addendum, article, benchmark, book, bookreview, briefreport, casereport, changes, comment, commentary, communication, conceptpaper, conferenceproceedings, correction, conferencereport, expressionofconcern, extendedabstract, meetingreport, creative, datadescriptor, discussion, editorial, essay, erratum, hypothesis, interestingimages, letter, meetingreport, newbookreceived, obituary, opinion, projectreport, reply, retraction, review, perspective, protocol, shortnote, supfile, technicalnote, viewpoint
% supfile = supplementary materials

%----------
% submit
%----------
% The class option "submit" will be changed to "accept" by the Editorial Office when the paper is accepted. This will only make changes to the frontpage (e.g., the logo of the journal will get visible), the headings, and the copyright information. Also, line numbering will be removed. Journal info and pagination for accepted papers will also be assigned by the Editorial Office.

%------------------
% moreauthors
%------------------
% If there is only one author the class option oneauthor should be used. Otherwise use the class option moreauthors.

%---------
% pdftex
%---------
% The option pdftex is for use with pdfLaTeX. If eps figures are used, remove the option pdftex and use LaTeX and dvi2pdf.

%=================================================================
\firstpage{1}
\makeatletter
\setcounter{page}{\@firstpage}
\makeatother
\pubvolume{xx}
\issuenum{1}
\articlenumber{5}
\pubyear{2019}
\copyrightyear{2019}
%\externaleditor{Academic Editor: name}
\history{Received: date; Accepted: date; Published: date}
\updates{yes} % If there is an update available, un-comment this line

%% MDPI internal command: uncomment if new journal that already uses continuous page numbers
%\continuouspages{yes}

%------------------------------------------------------------------
% The following line should be uncommented if the LaTeX file is uploaded to arXiv.org
%\pdfoutput=1

%=================================================================
% Add packages and commands here. The following packages are loaded in our class file: fontenc, calc, indentfirst, fancyhdr, graphicx, lastpage, ifthen, lineno, float, amsmath, setspace, enumitem, mathpazo, booktabs, titlesec, etoolbox, amsthm, hyphenat, natbib, hyperref, footmisc, geometry, caption, url, mdframed, tabto, soul, multirow, microtype, tikz

%=================================================================
%% Please use the following mathematics environments: Theorem, Lemma, Corollary, Proposition, Characterization, Property, Problem, Example, ExamplesandDefinitions, Hypothesis, Remark, Definition
%% For proofs, please use the proof environment (the amsthm package is loaded by the MDPI class).

%=================================================================
% Full title of the paper (Capitalized)
\Title{Full title of the paper (Capitalized)}

% Authors, for the paper (add full first names)
\Author{Amrita
Acharya$^{1,2,\ddagger,*}$\href{https://orcid.org/0000-0003-3293-2315}{\orcidicon}, Dianne
Caravela$^{2, \dagger, \ddagger}$, Eunice Kim$^{3}$, Emma
Kornberg$^{4}$, Elisabeth Nesmith$^{5}$}

% Authors, for metadata in PDF
\AuthorNames{Amrita Acharya, Dianne Caravela, Eunice Kim, Emma
Kornberg, Elisabeth Nesmith}

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad Smith
College; \href{mailto:aarcharya@smith.edu}{\nolinkurl{aarcharya@smith.edu}}\\
$^{2}$ \quad Smith
College; \href{mailto:dcaravela@smith.edu}{\nolinkurl{dcaravela@smith.edu}}\\
}
% Contact information of the corresponding author
\corres{Correspondence: \href{mailto:leutnant@fh-muenster.de}{\nolinkurl{leutnant@fh-muenster.de}};
Tel.: +XX-000-00-0000.}

% Current address and/or shared authorship
\firstnote{Current address: Updated affiliation}
\secondnote{These authors contributed equally to this work.}

% The commands \thirdnote{} till \eighthnote{} are available for further notes

% Simple summary
\simplesumm{A Simple summary goes here.}

% Abstract (Do not insert blank lines, i.e. \\)
\abstract{A single paragraph of about 200 words maximum. For research
articles, abstracts should give a pertinent overview of the work. We
strongly encourage authors to use the following style of structured
abstracts, but without headings: 1) Background: Place the question
addressed in a broad context and highlight the purpose of the study; 2)
Methods: Describe briefly the main methods or treatments applied; 3)
Results: Summarize the article's main findings; and 4) Conclusion:
Indicate the main conclusions or interpretations. The abstract should be
an objective representation of the article, it must not contain results
which are not presented and substantiated in the main text and should
not exaggerate the main conclusions.}

% Keywords
\keyword{keyword 1; keyword 2; keyword 3 (list three to ten pertinent
keywords specific to the article, yet reasonably common within the
subject discipline.).}

% The fields PACS, MSC, and JEL may be left empty or commented out if not applicable
%\PACS{J0101}
%\MSC{}
%\JEL{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Diversity
%\LSID{\url{http://}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Applied Sciences:
%\featuredapplication{Authors are encouraged to provide a concise description of the specific application or a potential application of the work. This section is not mandatory.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Data:
%\dataset{DOI number or link to the deposited data set in cases where the data set is published or set to be published separately. If the data set is submitted and will be published as a supplement to this paper in the journal Data, this field will be filled by the editors of the journal. In this case, please make sure to submit the data set as a supplement when entering your manuscript into our manuscript editorial system.}

%\datasetlicense{license under which the data set is made available (CC0, CC-BY, CC-BY-SA, CC-BY-NC, etc.)}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Toxins
%\keycontribution{The breakthroughs or highlights of the manuscript. Authors can write one or two sentences to describe the most important part of the paper.}

%\setcounter{secnumdepth}{4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\hypertarget{version}{%
\section{Version}\label{version}}

This Rmd-skeleton uses the mdpi Latex template published 2019/02.
However, the official template gets more frequently updated than the
`rticles' package. Therefore, please make sure prior to paper
submission, that you're using the most recent .cls, .tex and .bst files
(available \href{http://www.mdpi.com/authors/latex}{here}).

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

A recidivist is a term to describe criminals who re-offend.

\hypertarget{data}{%
\section{Data}\label{data}}

The data we are using for this project is the Correctional Offender
Management Profiling for Alternative Sanctions (COMPAS) Recidivism Risk
Scores dataset from AIF360. Northpointe created COMPAS as a commercial
risk assessment tool to score defendants on their potential recidivism
and violent recidivism risk. The COMPAS dataset we are using has 6,167
rows and each row represents an arrest charge for a defendant. We also
have data on the defendant's age, race, sex, what they were charged
with, and whether or not the defendant ultimately recidivated.

\hypertarget{results}{%
\section{Results}\label{results}}

This section may be divided by subheadings. It should provide a concise
and precise description of the experimental results, their
interpretation as well as the experimental conclusions that can be
drawn.

\hypertarget{subsection-heading-here}{%
\subsection{Subsection Heading Here}\label{subsection-heading-here}}

Subsection text here.

\hypertarget{subsubsection-heading-here}{%
\subsubsection{Subsubsection Heading
Here}\label{subsubsection-heading-here}}

Bulleted lists look like this:

\begin{itemize}
\tightlist
\item
  First bullet
\item
  Second bullet
\item
  Third bullet
\end{itemize}

Numbered lists can be added as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  First item
\item
  Second item
\item
  Third item
\end{enumerate}

The text continues here.

All figures and tables should be cited in the main text as Figure 1,
Table 1, etc.

\begin{figure}[H]
\centering
\includegraphics[width=3 cm]{logo-mdpi}
\caption{This is a figure, Schemes follow the same formatting. If there are multiple panels, they should be listed as: (\textbf{a}) Description of what is contained in the first panel. (\textbf{b}) Description of what is contained in the second panel. Figures should be placed in the main text near to the first time they are cited. A caption on a single line should be centered.}
\end{figure}

\begin{table}[H]
\caption{This is a table caption. Tables should be placed in the main text near to the first time they are cited.}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{ccc}
\toprule
\textbf{Title 1}    & \textbf{Title 2}  & \textbf{Title 3}\\
\midrule
entry 1     & data          & data\\
entry 2     & data          & data\\
\bottomrule
\end{tabular}
\end{table}

This is an example of an equation:

\begin{equation}
\mathbb{S}
\end{equation}

Example of a theorem:

\begin{Theorem}
Example text of a theorem.
\end{Theorem}

The text continues here. Proofs must be formatted as follows:

Example of a proof:

\begin{proof}[Proof of Theorem 1]
Text of the proof. Note that the phrase `of Theorem 1' is optional if it is clear which theorem is being referred to.
\end{proof}

The text continues here.

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

Authors should discuss the results and how they can be interpreted in
perspective of previous studies and of the working hypotheses. The
findings and their implications should be discussed in the broadest
context possible. Future research directions may also be highlighted.

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

This section is not mandatory, but can be added to the manuscript if the
discussion is unusually long or complex.

\hypertarget{patents}{%
\section{Patents}\label{patents}}

This section is not mandatory, but may be added if there are patents
resulting from the work reported in this manuscript.

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% %% optional
% \supplementary{The following are available online at www.mdpi.com/link, Figure S1: title, Table S1: title, Video S1: title.}
%
% % Only for the journal Methods and Protocols:
% % If you wish to submit a video article, please do so with any other supplementary material.
% % \supplementary{The following are available at www.mdpi.com/link: Figure S1: title, Table S1: title, Video S1: title. A supporting video article is available at doi: link.}

\vspace{6pt}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\acknowledgments{All sources of funding of the study should be
disclosed. Please clearly indicate grants that you have received in
support of your research work. Clearly state if you received funds for
covering the costs to publish in open access.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\authorcontributions{For research articles with several authors, a short
paragraph specifying their individual contributions must be provided.
The following statements should be used ``X.X. and Y.Y. conceive and
designed the experiments; X.X. performed the experiments; X.X. and Y.Y.
analyzed the data; W.W. contributed reagents/materials/analysis tools;
Y.Y. wrote the paper.'\,' Authorship must be limited to those who have
contributed substantially to the work reported.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\conflictsofinterest{Declare conflicts of interest or state `The authors
declare no conflict of interest.' Authors must identify and declare any
personal circumstances or interest that may be perceived as
inappropriately influencing the representation or interpretation of
reported research results. Any role of the funding sponsors in the
design of the study; in the collection, analyses or interpretation of
data in the writing of the manuscript, or in the decision to publish the
results must be declared in this section. If there is no role, please
state `The founding sponsors had no role in the design of the study; in
the collection, analyses, or interpretation of data; in the writing of
the manuscript, an in the decision to publish the results'.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional
\abbreviations{The following abbreviations are used in this manuscript:\\

\noindent
\begin{tabular}{@{}ll}
MDPI & Multidisciplinary Digital Publishing Institute \\
DOAJ & Directory of open access journals \\
TLA & Three letter acronym \\
LD & linear dichroism \\
\end{tabular}}

\input{"appendix.tex"}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Citations and References in Supplementary files are permitted provided that they also appear in the reference list here.

%=====================================
% References, variant A: internal bibliography
%=====================================
%\reftitle{References}
%\begin{thebibliography}{999}
% Reference 1
%\bibitem[Author1(year)]{ref-journal}
%Author1, T. The title of the cited article. {\em Journal Abbreviation} {\bf 2008}, {\em 10}, 142--149.
% Reference 2
%\bibitem[Author2(year)]{ref-book}
%Author2, L. The title of the cited contribution. In {\em The Book Title}; Editor1, F., Editor2, A., Eds.; Publishing House: City, Country, 2007; pp. 32--58.
%\end{thebibliography}

% The following MDPI journals use author-date citation: Arts, Econometrics, Economies, Genealogy, Humanities, IJFS, JRFM, Laws, Religions, Risks, Social Sciences. For those journals, please follow the formatting guidelines on http://www.mdpi.com/authors/references
% To cite two works by the same author: \citeauthor{ref-journal-1a} (\citeyear{ref-journal-1a}, \citeyear{ref-journal-1b}). This produces: Whittaker (1967, 1975)
% To cite two works by the same author with specific pages: \citeauthor{ref-journal-3a} (\citeyear{ref-journal-3a}, p. 328; \citeyear{ref-journal-3b}, p.475). This produces: Wong (1999, p. 328; 2000, p. 475)

%=====================================
% References, variant B: external bibliography
%=====================================
\reftitle{References}
\externalbibliography{yes}
\bibliography{mybibfile.bib}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional
\sampleavailability{Samples of the compounds \ldots\ldots{} are
available from the authors.}

%% for journal Sci
%\reviewreports{\\
%Reviewer 1 comments and authors’ response\\
%Reviewer 2 comments and authors’ response\\
%Reviewer 3 comments and authors’ response
%}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

### 2022-03-02 Paper_Draft1/Draft1.Rmd

@@
     affil: 1,2,\ddagger,*
-    orcid: 0000-0003-3293-2315
   - name: Dianne Caravela
@@
     address: |
+      Statistical and Data Sciences
       Smith College
-    email: aarcharya@smith.edu
+      Northampton, MA 01063
+    email: aacharya@smith.edu
   - num: 2
@@
 
+See Figure \ref{fig:random-plot} below.
+
+```{r random-plot, fig.cap="This is a plot.", fig.align='center'}
+plot(rnorm(10), rnorm(10))
+```
+
 The text continues here.
 
-All figures and tables should be cited in the main text as Figure 1, Table 1, etc.
+All figures and tables should be cited in the main text as Figure \ref{fig:mdpi-logo}, Table 1, etc.
 
-\begin{figure}[H]
-\centering
-\includegraphics[width=3 cm]{logo-mdpi}
-\caption{This is a figure, Schemes follow the same formatting.
-If there are multiple panels, they should be listed as: (\textbf{a}) Description of what is contained in the first panel.
-(\textbf{b}) Description of what is contained in the second panel.
+```{r mdpi-logo, fig.align='center', echo=FALSE, fig.cap="This is a figure, Schemes follow the same formatting.
+If there are multiple panels, they should be listed as: (\\textbf{a}) Description of what is contained in the first panel.
+(\\textbf{b}) Description of what is contained in the second panel.
 Figures should be placed in the main text near to the first time they are cited.
-A caption on a single line should be centered.}
-\end{figure}
+A caption on a single line should be centered."}
 
-\begin{table}[H]
-\caption{This is a table caption. Tables should be placed in the main text near to the first time they are cited.}
-\centering
-%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
-\begin{tabular}{ccc}
-\toprule
-\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}\\
-\midrule
-entry 1		& data			& data\\
-entry 2		& data			& data\\
-\bottomrule
-\end{tabular}
-\end{table}
+# Ben's tutorial on how to include figures with R code chunk
+knitr::include_graphics("logo-mdpi.pdf")
+```
+
+Please see Table \ref{tab:mytable}.
+
+```{r}
+x <- tibble::tribble(~`Title 1`, ~`Title 2`, ~`Title 3`,
+"entry 1", "data", "data",
+"entry 2", "data", "data"
+)
+
+knitr::kable(x, caption = "This is a table caption. Tables should be placed in the main text near to the first time they are cited.\\label{tab:mytable}")
+```
 
@@
 
-# Patents
+# Bibliography
+Citing book reference @gujer_systems_2008
+Citing reference with parentheses [@leutnant_stormwater_2016]
 
-This section is not mandatory, but may be added if there are patents resulting from the work reported in this manuscript.
+@hardin2015data advocated for more data science in statistics curricula.

### 2022-03-02 Paper_Draft1/Draft1.Rmd
BibTex Round1 Citations

--- 2022-02-28 Paper_Draft1/Draft1.Rmd
@@
 bibliography: mybibfile.bib
+csl: ../resources/chicago-author-date.csl
 appendix: appendix.tex

### 2022-03-03 Paper_Draft1/Draft1.tex

@@
 % Authors, for the paper (add full first names)
-\Author{Amrita
-Acharya$^{1,2,\ddagger,*}$\href{https://orcid.org/0000-0003-3293-2315}{\orcidicon}, Dianne
-Caravela$^{2, \dagger, \ddagger}$, Eunice Kim$^{3}$, Emma
-Kornberg$^{4}$, Elisabeth Nesmith$^{5}$}
+\Author{Amrita Acharya$^{1,2,\ddagger,*}$, Dianne Caravela$^{2, \dagger,
+\ddagger}$, Eunice Kim$^{3}$, Emma Kornberg$^{4}$, Elisabeth
+Nesmith$^{5}$}
 
@@
 \address{%
-$^{1}$ \quad Smith
-College; \href{mailto:aarcharya@smith.edu}{\nolinkurl{aarcharya@smith.edu}}\\
+$^{1}$ \quad Statistical and Data Sciences Smith College Northampton, MA
+01063; \href{mailto:aacharya@smith.edu}{\nolinkurl{aacharya@smith.edu}}\\
 $^{2}$ \quad Smith
@@
 
+% Pandoc syntax highlighting
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\usepackage{framed}
+\definecolor{shadecolor}{RGB}{248,248,248}
+\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
+\newcommand{\ExtensionTok}[1]{#1}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{#1}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\RegionMarkerTok}[1]{#1}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+
 % tightlist command for lists without linebreak
@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
+% From pandoc table feature
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
 
@@
 
+See Figure \ref{fig:random-plot} below.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{plot}\NormalTok{(}\FunctionTok{rnorm}\NormalTok{(}\DecValTok{10}\NormalTok{), }\FunctionTok{rnorm}\NormalTok{(}\DecValTok{10}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{figure}
+
+{\centering \includegraphics{Draft1_files/figure-latex/random-plot-1}
+
+}
+
+\caption{This is a plot.}\label{fig:random-plot}
+\end{figure}
+
 The text continues here.
 
-All figures and tables should be cited in the main text as Figure 1,
-Table 1, etc.
+All figures and tables should be cited in the main text as Figure
+\ref{fig:mdpi-logo}, Table 1, etc.
 
-\begin{figure}[H]
-\centering
-\includegraphics[width=3 cm]{logo-mdpi}
+\begin{figure}
+
+{\centering \includegraphics{logo-mdpi}
+
+}
+
 \caption{This is a figure, Schemes follow the same formatting.
@@
 Figures should be placed in the main text near to the first time they are cited.
-A caption on a single line should be centered.}
+A caption on a single line should be centered.}\label{fig:mdpi-logo}
 \end{figure}
 
-\begin{table}[H]
-\caption{This is a table caption. Tables should be placed in the main text near to the first time they are cited.}
-\centering
-%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
-\begin{tabular}{ccc}
+Please see Table \ref{tab:mytable}.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{x }\OtherTok{\textless{}{-}}\NormalTok{ tibble}\SpecialCharTok{::}\FunctionTok{tribble}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\StringTok{\textasciigrave{}}\AttributeTok{Title 1}\StringTok{\textasciigrave{}}\NormalTok{, }\SpecialCharTok{\textasciitilde{}}\StringTok{\textasciigrave{}}\AttributeTok{Title 2}\StringTok{\textasciigrave{}}\NormalTok{, }\SpecialCharTok{\textasciitilde{}}\StringTok{\textasciigrave{}}\AttributeTok{Title 3}\StringTok{\textasciigrave{}}\NormalTok{,}
+\StringTok{"entry 1"}\NormalTok{, }\StringTok{"data"}\NormalTok{, }\StringTok{"data"}\NormalTok{,}
+\StringTok{"entry 2"}\NormalTok{, }\StringTok{"data"}\NormalTok{, }\StringTok{"data"}
+\NormalTok{)}
+
+\NormalTok{knitr}\SpecialCharTok{::}\FunctionTok{kable}\NormalTok{(x, }\AttributeTok{caption =} \StringTok{"This is a table caption.
+Tables should be placed in the main text near to the first time they are cited.}\SpecialCharTok{\textbackslash{}\textbackslash{}}\StringTok{label\{tab:mytable\}"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{longtable}[]{@{}lll@{}}
+\caption{This is a table caption. Tables should be placed in the main
+text near to the first time they are
+cited.\label{tab:mytable}}\tabularnewline
 \toprule
-\textbf{Title 1}    & \textbf{Title 2}  & \textbf{Title 3}\\
+Title 1 & Title 2 & Title 3 \\
 \midrule
-entry 1     & data          & data\\
-entry 2     & data          & data\\
+\endfirsthead
+\toprule
+Title 1 & Title 2 & Title 3 \\
+\midrule
+\endhead
+entry 1 & data & data \\
+entry 2 & data & data \\
 \bottomrule
-\end{tabular}
-\end{table}
+\end{longtable}
 
@@
 
-\hypertarget{patents}{%
-\section{Patents}\label{patents}}
+\hypertarget{bibliography}{%
+\section{Bibliography}\label{bibliography}}
 
-This section is not mandatory, but may be added if there are patents
-resulting from the work reported in this manuscript.
+Citing book reference \citet{gujer_systems_2008} Citing reference with
+parentheses \citep{leutnant_stormwater_2016}
+
+\citet{hardin2015data} advocated for more data science in statistics
+curricula.
 

### 2022-03-06 Paper_Draft1/Draft1.Rmd
adding to bibliography section

@@
 # Bibliography
-Citing book reference @gujer_systems_2008
-Citing reference with parentheses [@leutnant_stormwater_2016]
 
-@hardin2015data advocated for more data science in statistics curricula.
+@angwin2016machine
+@hardin2015data
+@larson2016we
+@gebru2021datasheets
+@noauthor_aif360datasetscompasdataset_2018
+@noauthor_propublicacompas-analysis_2022
+@larson2016technical
+@bao2021s
+@barocas2017fairness
+@knight2020automated
+@kypraiou_what_2021
+@equivant_response_2018
+@baumer2017texts
+@james2013introduction
+@noauthor_ai_nodate
+@noauthor_trusted-ai_360
+@vartan_1st_2022
+@noauthor_womenthetable_nodate

### 2022-03-06 Paper_Draft1/Draft1.Rmd
update bib

--- 2022-03-03 Paper_Draft1/Draft1.Rmd
@@
 # Bibliography
-Citing book reference @gujer_systems_2008
-Citing reference with parentheses [@leutnant_stormwater_2016]
 
-@hardin2015data advocated for more data science in statistics curricula.
+@angwin2016machine
+@bao2021s
+@barocas2017fairness
+@baumer2017texts
+@equivant_response_2018
+@gebru2021datasheets
+@hardin2015data
+@james2013introduction
+@knight2020automated
+@kypraiou_what_2021
+@larson2016technical
+@larson2016we
+@noauthor_ai_nodate
+@noauthor_aif360datasetscompasdataset_2018
+@noauthor_trusted-ai_360
+@noauthor_propublicacompas-analysis_2022
+@noauthor_womenthetable_nodate
+@vartan_1st_2022

### 2022-03-07 Paper_Draft1/Draft1.tex
knit test

@@
 \toprule
-Title 1 & Title 2 & Title 3 \\
+Title 1 & Title 2 & Title 3\tabularnewline
 \midrule
@@
 \toprule
-Title 1 & Title 2 & Title 3 \\
+Title 1 & Title 2 & Title 3\tabularnewline
 \midrule
 \endhead
-entry 1 & data & data \\
-entry 2 & data & data \\
+entry 1 & data & data\tabularnewline
+entry 2 & data & data\tabularnewline
 \bottomrule
@@
 
-Citing book reference \citet{gujer_systems_2008} Citing reference with
-parentheses \citep{leutnant_stormwater_2016}
-
-\citet{hardin2015data} advocated for more data science in statistics
-curricula.
+\citet{angwin2016machine} \citet{bao2021s} \citet{barocas2017fairness}
+\citet{baumer2017texts} \citet{equivant_response_2018}
+\citet{gebru2021datasheets} \citet{hardin2015data}
+\citet{james2013introduction} \citet{knight2020automated}
+\citet{kypraiou_what_2021} \citet{larson2016technical}
+\citet{larson2016we} \citet{noauthor_ai_nodate}
+\citet{noauthor_aif360datasetscompasdataset_2018}
+\citet{noauthor_trusted-ai_360}
+\citet{noauthor_propublicacompas-analysis_2022}
+\citet{noauthor_womenthetable_nodate} \citet{vartan_1st_2022}
 

### 2022-03-07 Paper_Draft1/Draft1.Rmd
adding text and figures to draft

@@
 author:
+  -author:
   - name: Amrita Acharya
-    affil: 1,2,\ddagger,*
+    affil: 1, 2
   - name: Dianne Caravela
-    affil: 2, \dagger, \ddagger
+    affil: 1, 3
   - name: Eunice Kim
-    affil: 3
+    affil: 1, 4
   - name: Emma Kornberg
-    affil: 4
+    affil: 1, 5
   - name: Elisabeth Nesmith
-    affil: 5
+    affil: 1, 6
 affiliation:
@@
       Northampton, MA 01063
+  - num: 2
     email: aacharya@smith.edu
-  - num: 2
-    address: |
-      Smith College
+  - num: 3
     email: dcaravela@smith.edu
+  - num: 4
+    email: ekim89@smith.edu
+  - num: 5
+    email: ekornberg@smith.edu
+  - num: 6
+    email: enesmith@smith.edu
 # firstnote to eighthnote
@@
 appendix: appendix.tex
-simplesummary: |
-  A Simple summary goes here.
+
 abstract: |
-  A single paragraph of about 200 words maximum. For research articles, 
-  abstracts should give a pertinent overview of the work. We strongly encourage
-  authors to use the following style of structured abstracts, but without 
-  headings: 1) Background: Place the question addressed in a broad context and
-  highlight the purpose of the study; 2) Methods: Describe briefly the main
-  methods or treatments applied; 3) Results: Summarize the article's main 
-  findings; and 4) Conclusion: Indicate the main conclusions or interpretations. 
-  The abstract should be an objective representation of the article, it must not 
-  contain results which are not presented and substantiated in the main text and 
-  should not exaggerate the main conclusions.
-keywords: |
-  keyword 1; keyword 2; keyword 3 (list three to ten pertinent keywords specific 
-  to the article, yet reasonably common within the subject discipline.).
-acknowledgement: |
-  All sources of funding of the study should be disclosed. Please clearly 
-  indicate grants that you have received in support of your research work. 
-  Clearly state if you received funds for covering the costs to publish in open 
-  access.
-authorcontributions: |
-  For research articles with several authors, a short paragraph specifying their 
-  individual contributions must be provided. The following statements should be 
-  used ``X.X. and Y.Y. conceive and designed the experiments; X.X. performed the 
-  experiments; X.X. and Y.Y. analyzed the data; W.W. contributed 
-  reagents/materials/analysis tools; Y.Y. wrote the paper.'' Authorship must be
-  limited to those who have contributed substantially to the work reported.
-conflictsofinterest: |
-  Declare conflicts of interest or state 'The authors declare no conflict of 
-  interest.' Authors must identify and declare any personal circumstances or
-  interest that may be perceived as inappropriately influencing the
-  representation or interpretation of reported research results. Any role of the
-  funding sponsors in the design of the study; in the collection, analyses or 
-  interpretation of data in the writing of the manuscript, or in the decision to 
-  publish the results must be declared in this section. If there is no role, 
-  please state 'The founding sponsors had no role in the design of the study; 
-  in the collection, analyses, or interpretation of data; in the writing of the 
-  manuscript, an in the decision to publish the results'.
-sampleavailability: |
-  Samples of the compounds ...... are available from the authors.
-abbreviations:
-  - short: MDPI
-    long: Multidisciplinary Digital Publishing Institute
-  - short: DOAJ
-    long: Directory of open access journals
-  - short: TLA
-    long: Three letter acronym
-  - short: LD 
-    long: linear dichroism
+  A variety of disciplines use risk assessment instruments to help humans make data-driven decisions.
+Northpointe, a software company, created a risk assessment instrument known as the Correctional Offender Management Profiling for Alternative Sanctions (COMPAS).
+COMPAS uses various behavioral and psychological metrics related to recidivism to assist justice systems in assessing a defendant’s potential recidivism risk.
+ProPublica published an article which concludes that the biases in the criminal justice system are reflected in the COMPAS software.
+Using a human rights framework adopted from the organization Women at the Table, we use various debiasing algorithms to analyze both sides of the argument between Northpointe and ProPublica and determine the level and extent of racial bias in the COMPAS algorithm.
 output: rticles::mdpi_article
@@
 
-A recidivist is a term to describe criminals who re-offend.
+Women @ the Table, the sponsor organization for this project, is “a growing, global gender equality & democracy CSO based in Geneva, Switzerland focused on advancing feminist systems change by using the prism of technology, innovation & AI exercising leverage points in technology, the economy, sustainability & democratic governance” [@noauthor_womenthetable_nodate].
+We have been asked to collaborate on their AI & Equality [@noauthor_ai_nodate] initiative, tasked with debiasing the COMPAS algorithm [@noauthor_aif360datasetscompasdataset_2018] and producing a corresponding data story that will be added to their library.
+
+The Correctional Offender Management Profiling for Alternative Sanctions (COMPAS) algorithm was created by the private for-profit company Northpointe, also known by its parent company Equivant [@equivant_faq_nodate], to predict defendants’ risk of recidivism.
+It generates a score that classifies defendants’ risk of recidivism as either low, medium, or high [@angwin2016machine].
+Jurisdictions across the United States use the COMPAS risk assessment instrument.
+In 2016, ProPublica published a piece that analyzed the methods and algorithms used by Northpointe to uncover racial biases in defendants’ scores [@angwin2016machine].
+Their analysis looks at the distribution of decile COMPAS scores among Black and white defendants.
+ProPublica concludes, after using a statistical parity metric on false positives or false negatives, that the algorithm is racially biased [@larson2016we].
+Northpointe denies the allegations of racial bias, conducting their own analyses based on different statistical parity metrics [@equivant_response_2018].
+Since then, the two parties have had several exchanges, maintaining their original arguments.
+Currently, the two parties’ disagreement focuses on pretrial Risk Assessment Instruments (RAIs).
+ProPublica maintains that there are biases in the outcome values, protected attributes, and covariates during Northpointe’s data processing phase.
+ProPublica accounts for these biases in their analyses, and Northepointe’s response, highlights how ProPublica did not account for base rates of recidivism in their analysis, which are important to understand initial percentages without the presence of other information.
+
+Our project builds on Women at the Table’s various debiasing algorithms to conduct our own analyses on the COMPAS dataset.
+Based on this analysis, we employ a human rights framework to contribute to the ongoing ProPublica and Northpointe debate and investigate whether or to what extent there is racial bias in the COMPAS algorithm.
+With a solid understanding of the two sides, we aim to pinpoint the shortcomings of both arguments and correct them in our analyses.
+We will use various machine learning algorithms including logistic regression, cross validation, and lasso and ridge techniques to choose models.
+We will summarize our results using the JupyterNotebook framework from Women at the Table, to be used by members of the organization to teach in a workshop setting.
+We hope that our findings will highlight the importance of checking statistical analyses using varied methods and contribute to the ongoing discussion of the effects of machine biases in the justice system.
 
@@
 
-The data we are using for this project is the Correctional Offender Management Profiling for Alternative Sanctions (COMPAS) Recidivism Risk Scores dataset from AIF360.
-Northpointe created COMPAS as a commercial risk assessment tool to score defendants on their potential recidivism and violent recidivism risk.
-The COMPAS dataset we are using has 6,167 rows and each row represents an arrest charge for a defendant.
-We also have data on the defendant's age, race, sex, what they were charged with, and whether or not the defendant ultimately recidivated.
+The data we are using for this project is the COMPAS General Recidivism Risk Scores dataset from ProPublica used in their original analysis [@larson_propublicacompas-analysiscompas-scores-two-yearscsv_2022].
+The COMPAS dataset we are using is from AI Fairness 360 (AIF360) toolkit [@noauthor_trusted-ai_360] which does the same initial preprocessing as ProPublica.
+The raw data has 6,167 rows and each row represents an arrest charge for a defendant.
+ProPublica’s COMPAS data includes the defendant's age, race, sex, what they were charged with, and whether or not the defendant ultimately recidivated within a two-year period after their arrest.
+For the purposes of our project, which endeavors to evaluate the differing effects of the COMPAS algorithm on white defendants and Black defendants, we have filtered the data to only include individuals whose race is listed as Caucasian or African-American.
+Our data therefore has 5,723 rows (Figure \ref{fig:table snip}), with the below distributions of race (Figure \ref{fig:race plot}), age (Figure \ref{fig:age plot}), and two year recidivism rate (Figures \ref{fig:recid plot} & \ref{fig:recid race plot}).
+
+```{r table snip, fig.align='center', fig.cap= 'This table is a snippet the data set we will be using.'}
+knitr::include_graphics("/Users/elisabethnesmith/womenatthetable/images/table_snippet.png")
+```
+
+```{r race plot, fig.align='center', fig.cap= 'This plot shows the distribution of defendant races.'}
+knitr::include_graphics("/Users/elisabethnesmith/womenatthetable/images/race_bar_plot.png")
+```
+
+```{r age plot, fig.align='center', fig.cap= 'This plot shows the distribution of defendant ages by race.'}
+knitr::include_graphics("/Users/elisabethnesmith/womenatthetable/images/race_age_plot.png")
+```
+
+```{r recid plot, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes.'}
+knitr::include_graphics("/Users/elisabethnesmith/womenatthetable/images/recid_bar_plot.png")
+```
+
+```{r recid race plot, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes by race.'}
+knitr::include_graphics("/Users/elisabethnesmith/womenatthetable/images/race_recid_plot.png")
+```
 

### 2022-03-07 Paper_Draft1/Draft1.tex
adding text and figures to draft

@@
 % Authors, for the paper (add full first names)
-\Author{Amrita Acharya$^{1,2,\ddagger,*}$, Dianne Caravela$^{2, \dagger,
-\ddagger}$, Eunice Kim$^{3}$, Emma Kornberg$^{4}$, Elisabeth
-Nesmith$^{5}$}
+\Author{$^{}$}
 
 % Authors, for metadata in PDF
-\AuthorNames{Amrita Acharya, Dianne Caravela, Eunice Kim, Emma
-Kornberg, Elisabeth Nesmith}
+\AuthorNames{}
 
@@
 \address{%
-$^{1}$ \quad Statistical and Data Sciences Smith College Northampton, MA
-01063; \href{mailto:aacharya@smith.edu}{\nolinkurl{aacharya@smith.edu}}\\
-$^{2}$ \quad Smith
-College; \href{mailto:dcaravela@smith.edu}{\nolinkurl{dcaravela@smith.edu}}\\
+$^{1}$ \quad Statistical and Data Sciences Smith College Northampton, MA 01063; \\
+$^{2}$ \quad ; \href{mailto:aacharya@smith.edu}{\nolinkurl{aacharya@smith.edu}}\\
+$^{3}$ \quad ; \href{mailto:dcaravela@smith.edu}{\nolinkurl{dcaravela@smith.edu}}\\
+$^{4}$ \quad ; \href{mailto:ekim89@smith.edu}{\nolinkurl{ekim89@smith.edu}}\\
+$^{5}$ \quad ; \href{mailto:ekornberg@smith.edu}{\nolinkurl{ekornberg@smith.edu}}\\
+$^{6}$ \quad ; \href{mailto:enesmith@smith.edu}{\nolinkurl{enesmith@smith.edu}}\\
 }
@@
 % Simple summary
-\simplesumm{A Simple summary goes here.}
 
 % Abstract (Do not insert blank lines, i.e. \\)
-\abstract{A single paragraph of about 200 words maximum. For research
-articles, abstracts should give a pertinent overview of the work. We
-strongly encourage authors to use the following style of structured
-abstracts, but without headings: 1) Background: Place the question
-addressed in a broad context and highlight the purpose of the study; 2)
-Methods: Describe briefly the main methods or treatments applied; 3)
-Results: Summarize the article's main findings; and 4) Conclusion:
-Indicate the main conclusions or interpretations. The abstract should be
-an objective representation of the article, it must not contain results
-which are not presented and substantiated in the main text and should
-not exaggerate the main conclusions.}
+\abstract{A variety of disciplines use risk assessment instruments to help humans
+make data-driven decisions. Northpointe, a software company, created a
+risk assessment instrument known as the Correctional Offender Management
+Profiling for Alternative Sanctions (COMPAS). COMPAS uses various
+behavioral and psychological metrics related to recidivism to assist
+justice systems in assessing a defendant's potential recidivism risk.
+ProPublica published an article which concludes that the biases in the
+criminal justice system are reflected in the COMPAS software. Using a
+human rights framework adopted from the organization Women at the Table,
+we use various debiasing algorithms to analyze both sides of the
+argument between Northpointe and ProPublica and determine the level and
+extent of racial bias in the COMPAS algorithm.}
 
 % Keywords
-\keyword{keyword 1; keyword 2; keyword 3 (list three to ten pertinent
-keywords specific to the article, yet reasonably common within the
-subject discipline.).}
 
@@
 
-A recidivist is a term to describe criminals who re-offend.
+Women @ the Table, the sponsor organization for this project, is ``a
+growing, global gender equality \& democracy CSO based in Geneva,
+Switzerland focused on advancing feminist systems change by using the
+prism of technology, innovation \& AI exercising leverage points in
+technology, the economy, sustainability \& democratic governance''
+\citep{noauthor_womenthetable_nodate}. We have been asked to collaborate
+on their AI \& Equality \citep{noauthor_ai_nodate} initiative, tasked
+with debiasing the COMPAS algorithm
+\citep{noauthor_aif360datasetscompasdataset_2018} and producing a
+corresponding data story that will be added to their library.
+
+The Correctional Offender Management Profiling for Alternative Sanctions
+(COMPAS) algorithm was created by the private for-profit company
+Northpointe, also known by its parent company Equivant
+\citep{equivant_faq_nodate}, to predict defendants' risk of recidivism.
+It generates a score that classifies defendants' risk of recidivism as
+either low, medium, or high \citep{angwin2016machine}. Jurisdictions
+across the United States use the COMPAS risk assessment instrument. In
+2016, ProPublica published a piece that analyzed the methods and
+algorithms used by Northpointe to uncover racial biases in defendants'
+scores \citep{angwin2016machine}. Their analysis looks at the
+distribution of decile COMPAS scores among Black and white defendants.
+ProPublica concludes, after using a statistical parity metric on false
+positives or false negatives, that the algorithm is racially biased
+\citep{larson2016we}. Northpointe denies the allegations of racial bias,
+conducting their own analyses based on different statistical parity
+metrics \citep{equivant_response_2018}. Since then, the two parties have
+had several exchanges, maintaining their original arguments. Currently,
+the two parties' disagreement focuses on pretrial Risk Assessment
+Instruments (RAIs). ProPublica maintains that there are biases in the
+outcome values, protected attributes, and covariates during
+Northpointe's data processing phase. ProPublica accounts for these
+biases in their analyses, and Northepointe's response, highlights how
+ProPublica did not account for base rates of recidivism in their
+analysis, which are important to understand initial percentages without
+the presence of other information.
+
+Our project builds on Women at the Table's various debiasing algorithms
+to conduct our own analyses on the COMPAS dataset. Based on this
+analysis, we employ a human rights framework to contribute to the
+ongoing ProPublica and Northpointe debate and investigate whether or to
+what extent there is racial bias in the COMPAS algorithm. With a solid
+understanding of the two sides, we aim to pinpoint the shortcomings of
+both arguments and correct them in our analyses. We will use various
+machine learning algorithms including logistic regression, cross
+validation, and lasso and ridge techniques to choose models. We will
+summarize our results using the JupyterNotebook framework from Women at
+the Table, to be used by members of the organization to teach in a
+workshop setting. We hope that our findings will highlight the
+importance of checking statistical analyses using varied methods and
+contribute to the ongoing discussion of the effects of machine biases in
+the justice system.
 
@@
 
-The data we are using for this project is the Correctional Offender
-Management Profiling for Alternative Sanctions (COMPAS) Recidivism Risk
-Scores dataset from AIF360. Northpointe created COMPAS as a commercial
-risk assessment tool to score defendants on their potential recidivism
-and violent recidivism risk. The COMPAS dataset we are using has 6,167
-rows and each row represents an arrest charge for a defendant. We also
-have data on the defendant's age, race, sex, what they were charged
-with, and whether or not the defendant ultimately recidivated.
+The data we are using for this project is the COMPAS General Recidivism
+Risk Scores dataset from ProPublica used in their original analysis
+\citep{larson_propublicacompas-analysiscompas-scores-two-yearscsv_2022}.
+The COMPAS dataset we are using is from AI Fairness 360 (AIF360) toolkit
+\citep{noauthor_trusted-ai_360} which does the same initial
+preprocessing as ProPublica. The raw data has 6,167 rows and each row
+represents an arrest charge for a defendant. ProPublica's COMPAS data
+includes the defendant's age, race, sex, what they were charged with,
+and whether or not the defendant ultimately recidivated within a
+two-year period after their arrest. For the purposes of our project,
+which endeavors to evaluate the differing effects of the COMPAS
+algorithm on white defendants and Black defendants, we have filtered the
+data to only include individuals whose race is listed as Caucasian or
+African-American. Our data therefore has 5,723 rows (Figure
+\ref{fig:table snip}), with the below distributions of race (Figure
+\ref{fig:race plot}), age (Figure \ref{fig:age plot}), and two year
+recidivism rate (Figures \ref{fig:recid plot} \&
+\ref{fig:recid race plot}).
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{knitr}\OperatorTok{::}\KeywordTok{include_graphics}\NormalTok{(}\StringTok{"/Users/elisabethnesmith/womenatthetable/images/table_snippet.png"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{figure}
+
+{\centering \includegraphics[width=27.78in]{/Users/elisabethnesmith/womenatthetable/images/table_snippet}
+
+}
+
+\caption{This table is a snippet the data set we will be using.}\label{fig:table snip}
+\end{figure}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{knitr}\OperatorTok{::}\KeywordTok{include_graphics}\NormalTok{(}\StringTok{"/Users/elisabethnesmith/womenatthetable/images/race_bar_plot.png"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{figure}
+
+{\centering \includegraphics[width=12.69in]{/Users/elisabethnesmith/womenatthetable/images/race_bar_plot}
+
+}
+
+\caption{This plot shows the distribution of defendant races.}\label{fig:race plot}
+\end{figure}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{knitr}\OperatorTok{::}\KeywordTok{include_graphics}\NormalTok{(}\StringTok{"/Users/elisabethnesmith/womenatthetable/images/race_age_plot.png"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{figure}
+
+{\centering \includegraphics[width=11.61in]{/Users/elisabethnesmith/womenatthetable/images/race_age_plot}
+
+}
+
+\caption{This plot shows the distribution of defendant ages by race.}\label{fig:age plot}
+\end{figure}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{knitr}\OperatorTok{::}\KeywordTok{include_graphics}\NormalTok{(}\StringTok{"/Users/elisabethnesmith/womenatthetable/images/recid_bar_plot.png"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{figure}
+
+{\centering \includegraphics[width=11.28in]{/Users/elisabethnesmith/womenatthetable/images/recid_bar_plot}
+
+}
+
+\caption{This plot shows the distribution of two year recidivism outcomes.}\label{fig:recid plot}
+\end{figure}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{knitr}\OperatorTok{::}\KeywordTok{include_graphics}\NormalTok{(}\StringTok{"/Users/elisabethnesmith/womenatthetable/images/race_recid_plot.png"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{figure}
+
+{\centering \includegraphics[width=11.31in]{/Users/elisabethnesmith/womenatthetable/images/race_recid_plot}
+
+}
+
+\caption{This plot shows the distribution of two year recidivism outcomes by race.}\label{fig:recid race plot}
+\end{figure}
 
@@
 \begin{Highlighting}[]
-\FunctionTok{plot}\NormalTok{(}\FunctionTok{rnorm}\NormalTok{(}\DecValTok{10}\NormalTok{), }\FunctionTok{rnorm}\NormalTok{(}\DecValTok{10}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(}\KeywordTok{rnorm}\NormalTok{(}\DecValTok{10}\NormalTok{), }\KeywordTok{rnorm}\NormalTok{(}\DecValTok{10}\NormalTok{))}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\NormalTok{x }\OtherTok{\textless{}{-}}\NormalTok{ tibble}\SpecialCharTok{::}\FunctionTok{tribble}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\StringTok{\textasciigrave{}}\AttributeTok{Title 1}\StringTok{\textasciigrave{}}\NormalTok{, }\SpecialCharTok{\textasciitilde{}}\StringTok{\textasciigrave{}}\AttributeTok{Title 2}\StringTok{\textasciigrave{}}\NormalTok{, }\SpecialCharTok{\textasciitilde{}}\StringTok{\textasciigrave{}}\AttributeTok{Title 3}\StringTok{\textasciigrave{}}\NormalTok{,}
+\NormalTok{x <-}\StringTok{ }\NormalTok{tibble}\OperatorTok{::}\KeywordTok{tribble}\NormalTok{(}\OperatorTok{~}\StringTok{`}\DataTypeTok{Title 1}\StringTok{`}\NormalTok{, }\OperatorTok{~}\StringTok{`}\DataTypeTok{Title 2}\StringTok{`}\NormalTok{, }\OperatorTok{~}\StringTok{`}\DataTypeTok{Title 3}\StringTok{`}\NormalTok{,}
 \StringTok{"entry 1"}\NormalTok{, }\StringTok{"data"}\NormalTok{, }\StringTok{"data"}\NormalTok{,}
@@
 
-\NormalTok{knitr}\SpecialCharTok{::}\FunctionTok{kable}\NormalTok{(x, }\AttributeTok{caption =} \StringTok{"This is a table caption.
-Tables should be placed in the main text near to the first time they are cited.}\SpecialCharTok{\textbackslash{}\textbackslash{}}\StringTok{label\{tab:mytable\}"}\NormalTok{)}
+\NormalTok{knitr}\OperatorTok{::}\KeywordTok{kable}\NormalTok{(x, }\DataTypeTok{caption =} \StringTok{"This is a table caption.
+Tables should be placed in the main text near to the first time they are cited.}\CharTok{\textbackslash{}\textbackslash{}}\StringTok{label\{tab:mytable\}"}\NormalTok{)}
 \end{Highlighting}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\acknowledgments{All sources of funding of the study should be
-disclosed. Please clearly indicate grants that you have received in
-support of your research work. Clearly state if you received funds for
-covering the costs to publish in open access.}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\authorcontributions{For research articles with several authors, a short
-paragraph specifying their individual contributions must be provided.
-The following statements should be used ``X.X. and Y.Y. conceive and
-designed the experiments; X.X. performed the experiments; X.X. and Y.Y.
-analyzed the data; W.W. contributed reagents/materials/analysis tools;
-Y.Y. wrote the paper.'\,' Authorship must be limited to those who have
-contributed substantially to the work reported.}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\conflictsofinterest{Declare conflicts of interest or state `The authors
-declare no conflict of interest.' Authors must identify and declare any
-personal circumstances or interest that may be perceived as
-inappropriately influencing the representation or interpretation of
-reported research results. Any role of the funding sponsors in the
-design of the study; in the collection, analyses or interpretation of
-data in the writing of the manuscript, or in the decision to publish the
-results must be declared in this section. If there is no role, please
-state `The founding sponsors had no role in the design of the study; in
-the collection, analyses, or interpretation of data; in the writing of
-the manuscript, an in the decision to publish the results'.}
 
@@
 %% optional
-\abbreviations{The following abbreviations are used in this manuscript:\\
-
-\noindent
-\begin{tabular}{@{}ll}
-MDPI & Multidisciplinary Digital Publishing Institute \\
-DOAJ & Directory of open access journals \\
-TLA & Three letter acronym \\
-LD & linear dichroism \\
-\end{tabular}}
 
@@
 %% optional
-\sampleavailability{Samples of the compounds \ldots\ldots{} are
-available from the authors.}
 

### 2022-03-07 Paper_Draft1/Draft1.Rmd
fixing file paths

@@
 
-```{r table snip, fig.align='center', fig.cap= 'This table is a snippet the data set we will be using.'}
-knitr::include_graphics("/Users/elisabethnesmith/womenatthetable/images/table_snippet.png")
+```{r table snip, echo = FALSE, fig.align='center', fig.cap= 'This table is a snippet the data set we will be using.', out.width= "400px", out.height= "200px"}
+knitr::include_graphics("../images/table_snippet.png")
 ```
 
-```{r race plot, fig.align='center', fig.cap= 'This plot shows the distribution of defendant races.'}
-knitr::include_graphics("/Users/elisabethnesmith/womenatthetable/images/race_bar_plot.png")
+```{r race plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of defendant races.', out.width= "400px", out.height= "200px"}
+knitr::include_graphics("../images/race_bar_plot.png")
 ```
 
-```{r age plot, fig.align='center', fig.cap= 'This plot shows the distribution of defendant ages by race.'}
-knitr::include_graphics("/Users/elisabethnesmith/womenatthetable/images/race_age_plot.png")
+```{r age plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of defendant ages by race.', out.width= "400px", out.height= "200px"}
+knitr::include_graphics("../images/race_age_plot.png")
 ```
 
-```{r recid plot, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes.'}
-knitr::include_graphics("/Users/elisabethnesmith/womenatthetable/images/recid_bar_plot.png")
+```{r recid plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes.', out.width= "400px", out.height= "200px"}
+knitr::include_graphics("../images/recid_bar_plot.png")
 ```
 
-```{r recid race plot, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes by race.'}
-knitr::include_graphics("/Users/elisabethnesmith/womenatthetable/images/race_recid_plot.png")
+```{r recid race plot,echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes by race.', out.width= "400px", out.height= "200px"}
+knitr::include_graphics("../images/race_recid_plot.png")
 ```
@@
 3. Third item
-
-See Figure \ref{fig:random-plot} below.
-
-```{r random-plot, fig.cap="This is a plot.", fig.align='center'}
-plot(rnorm(10), rnorm(10))
-```
 

### 2022-03-07 Paper_Draft1/Draft1.tex
fixing file paths

@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{knitr}\OperatorTok{::}\KeywordTok{include_graphics}\NormalTok{(}\StringTok{"/Users/elisabethnesmith/womenatthetable/images/table_snippet.png"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
 \begin{figure}
 
-{\centering \includegraphics[width=27.78in]{/Users/elisabethnesmith/womenatthetable/images/table_snippet}
+{\centering \includegraphics[width=400px,height=200px]{../images/table_snippet}
 
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{knitr}\OperatorTok{::}\KeywordTok{include_graphics}\NormalTok{(}\StringTok{"/Users/elisabethnesmith/womenatthetable/images/race_bar_plot.png"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
 \begin{figure}
 
-{\centering \includegraphics[width=12.69in]{/Users/elisabethnesmith/womenatthetable/images/race_bar_plot}
+{\centering \includegraphics[width=400px,height=200px]{../images/race_bar_plot}
 
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{knitr}\OperatorTok{::}\KeywordTok{include_graphics}\NormalTok{(}\StringTok{"/Users/elisabethnesmith/womenatthetable/images/race_age_plot.png"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
 \begin{figure}
 
-{\centering \includegraphics[width=11.61in]{/Users/elisabethnesmith/womenatthetable/images/race_age_plot}
+{\centering \includegraphics[width=400px,height=200px]{../images/race_age_plot}
 
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{knitr}\OperatorTok{::}\KeywordTok{include_graphics}\NormalTok{(}\StringTok{"/Users/elisabethnesmith/womenatthetable/images/recid_bar_plot.png"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
 \begin{figure}
 
-{\centering \includegraphics[width=11.28in]{/Users/elisabethnesmith/womenatthetable/images/recid_bar_plot}
+{\centering \includegraphics[width=400px,height=200px]{../images/recid_bar_plot}
 
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{knitr}\OperatorTok{::}\KeywordTok{include_graphics}\NormalTok{(}\StringTok{"/Users/elisabethnesmith/womenatthetable/images/race_recid_plot.png"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
 \begin{figure}
 
-{\centering \includegraphics[width=11.31in]{/Users/elisabethnesmith/womenatthetable/images/race_recid_plot}
+{\centering \includegraphics[width=400px,height=200px]{../images/race_recid_plot}
 
@@
 \end{enumerate}
-
-See Figure \ref{fig:random-plot} below.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{plot}\NormalTok{(}\KeywordTok{rnorm}\NormalTok{(}\DecValTok{10}\NormalTok{), }\KeywordTok{rnorm}\NormalTok{(}\DecValTok{10}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{figure}
-
-{\centering \includegraphics{Draft1_files/figure-latex/random-plot-1}
-
-}
-
-\caption{This is a plot.}\label{fig:random-plot}
-\end{figure}
 

### 2022-03-07 Paper_Draft1/Draft1.tex
adding title, fixing authors formatting

@@
 % Full title of the paper (Capitalized)
-\Title{Full title of the paper (Capitalized)}
+\Title{It's COMPAS-licated Can We Find Fairness in the COMPAS Risk
+Assessment Algorithm?}
 
 % Authors, for the paper (add full first names)
-\Author{$^{}$}
+\Author{Amrita Acharya$^{1, 2}$, Dianne Caravela$^{1, 3}$, Eunice
+Kim$^{1, 4}$, Emma Kornberg$^{1, 5}$, Elisabeth Nesmith$^{1, 6}$}
 
 % Authors, for metadata in PDF
-\AuthorNames{}
+\AuthorNames{Amrita Acharya, Dianne Caravela, Eunice Kim, Emma
+Kornberg, Elisabeth Nesmith}
 
@@
 \address{%
-$^{1}$ \quad Statistical and Data Sciences Smith College Northampton, MA 01063; \\
+$^{1}$ \quad Statistical and Data Sciences Smith College Northampton, MA
+01063; \\
 $^{2}$ \quad ; \href{mailto:aacharya@smith.edu}{\nolinkurl{aacharya@smith.edu}}\\
@@
 % Current address and/or shared authorship
-\firstnote{Current address: Updated affiliation}
-\secondnote{These authors contributed equally to this work.}
 
@@
 % Abstract (Do not insert blank lines, i.e. \\)
-\abstract{A variety of disciplines use risk assessment instruments to help humans
-make data-driven decisions. Northpointe, a software company, created a
-risk assessment instrument known as the Correctional Offender Management
-Profiling for Alternative Sanctions (COMPAS). COMPAS uses various
-behavioral and psychological metrics related to recidivism to assist
-justice systems in assessing a defendant's potential recidivism risk.
-ProPublica published an article which concludes that the biases in the
-criminal justice system are reflected in the COMPAS software. Using a
-human rights framework adopted from the organization Women at the Table,
-we use various debiasing algorithms to analyze both sides of the
+\abstract{A variety of disciplines use risk assessment instruments to
+help humans make data-driven decisions. Northpointe, a software company,
+created a risk assessment instrument known as the Correctional Offender
+Management Profiling for Alternative Sanctions (COMPAS). COMPAS uses
+various behavioral and psychological metrics related to recidivism to
+assist justice systems in assessing a defendant's potential recidivism
+risk. ProPublica published an article which concludes that the biases in
+the criminal justice system are reflected in the COMPAS software. Using
+a human rights framework adopted from the organization Women at the
+Table, we use various debiasing algorithms to analyze both sides of the
 argument between Northpointe and ProPublica and determine the level and
@@
 \begin{Highlighting}[]
-\NormalTok{x <-}\StringTok{ }\NormalTok{tibble}\OperatorTok{::}\KeywordTok{tribble}\NormalTok{(}\OperatorTok{~}\StringTok{`}\DataTypeTok{Title 1}\StringTok{`}\NormalTok{, }\OperatorTok{~}\StringTok{`}\DataTypeTok{Title 2}\StringTok{`}\NormalTok{, }\OperatorTok{~}\StringTok{`}\DataTypeTok{Title 3}\StringTok{`}\NormalTok{,}
+\NormalTok{x }\OtherTok{\textless{}{-}}\NormalTok{ tibble}\SpecialCharTok{::}\FunctionTok{tribble}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\StringTok{\textasciigrave{}}\AttributeTok{Title 1}\StringTok{\textasciigrave{}}\NormalTok{, }\SpecialCharTok{\textasciitilde{}}\StringTok{\textasciigrave{}}\AttributeTok{Title 2}\StringTok{\textasciigrave{}}\NormalTok{, }\SpecialCharTok{\textasciitilde{}}\StringTok{\textasciigrave{}}\AttributeTok{Title 3}\StringTok{\textasciigrave{}}\NormalTok{,}
 \StringTok{"entry 1"}\NormalTok{, }\StringTok{"data"}\NormalTok{, }\StringTok{"data"}\NormalTok{,}
@@
 
-\NormalTok{knitr}\OperatorTok{::}\KeywordTok{kable}\NormalTok{(x, }\DataTypeTok{caption =} \StringTok{"This is a table caption.
-Tables should be placed in the main text near to the first time they are cited.}\CharTok{\textbackslash{}\textbackslash{}}\StringTok{label\{tab:mytable\}"}\NormalTok{)}
+\NormalTok{knitr}\SpecialCharTok{::}\FunctionTok{kable}\NormalTok{(x, }\AttributeTok{caption =} \StringTok{"This is a table caption.
+Tables should be placed in the main text near to the first time they are cited.}\SpecialCharTok{\textbackslash{}\textbackslash{}}\StringTok{label\{tab:mytable\}"}\NormalTok{)}
 \end{Highlighting}
@@
 \toprule
-Title 1 & Title 2 & Title 3\tabularnewline
+Title 1 & Title 2 & Title 3 \\
 \midrule
@@
 \toprule
-Title 1 & Title 2 & Title 3\tabularnewline
+Title 1 & Title 2 & Title 3 \\
 \midrule
 \endhead
-entry 1 & data & data\tabularnewline
-entry 2 & data & data\tabularnewline
+entry 1 & data & data \\
+entry 2 & data & data \\
 \bottomrule

### 2022-03-07 Paper_Draft1/Draft1.Rmd

@@
 ---
-title: Full title of the paper (Capitalized)
+title: It’s COMPAS-licated Can We Find Fairness in the COMPAS Risk Assessment Algorithm?
 author:
-  -author:
   - name: Amrita Acharya
@@
     email: enesmith@smith.edu
-# firstnote to eighthnote
-firstnote: |
-  Current address: Updated affiliation
-secondnote: |
-  These authors contributed equally to this work.
+
 correspondence: |

### 2022-03-07 Paper_Draft1/Draft1.tex
adding more URLs to bib

@@
 % Full title of the paper (Capitalized)
-\Title{It's COMPAS-licated Can We Find Fairness in the COMPAS Risk
+\Title{It's COMPAS-licated -- Can We Find Fairness in the COMPAS Risk
 Assessment Algorithm?}

### 2022-03-07 Paper_Draft1/Draft1.Rmd
adding more URLs to bib
note = {\url{}}

@@
 ---
-title: It’s COMPAS-licated Can We Find Fairness in the COMPAS Risk Assessment Algorithm?
+title: It’s COMPAS-licated -- Can We Find Fairness in the COMPAS Risk Assessment Algorithm?
 author:
@@
 # Bibliography
-
-@angwin2016machine
-@bao2021s
-@barocas2017fairness
-@baumer2017texts
-@equivant_response_2018
-@gebru2021datasheets
-@hardin2015data
-@james2013introduction
-@knight2020automated
-@kypraiou_what_2021
-@larson2016technical
-@larson2016we
-@noauthor_ai_nodate
-@noauthor_aif360datasetscompasdataset_2018
-@noauthor_trusted-ai_360
-@noauthor_propublicacompas-analysis_2022
-@noauthor_womenthetable_nodate
-@vartan_1st_2022

### 2022-03-08 Paper_Draft1/Draft1.Rmd

@@
 ---
-title: It’s COMPAS-licated -- Can We Find Fairness in the COMPAS Risk Assessment Algorithm?
+title: Does the COMPAS needle always point towards equity? Finding Fairness in the COMPAS Risk Assessment Algorithm
 author:

### 2022-03-08 Paper_Draft1/Draft1.tex

@@
 % Full title of the paper (Capitalized)
-\Title{It's COMPAS-licated -- Can We Find Fairness in the COMPAS Risk
-Assessment Algorithm?}
+\Title{Does the COMPAS needle always point towards equity? Finding
+Fairness in the COMPAS Risk Assessment Algorithm}
 
@@
 conducting their own analyses based on different statistical parity
-metrics \citep{equivant_response_2018}. Since then, the two parties have
-had several exchanges, maintaining their original arguments. Currently,
+metrics \citep{equivant_response_2018}. Since then, ProPublica has
+responded, maintaining their original argument and Northpointe has yet
+to respond (though in 2017, they adopted the name Equivant). Currently,
 the two parties' disagreement focuses on pretrial Risk Assessment
@@
 
-\hypertarget{bibliography}{%
-\section{Bibliography}\label{bibliography}}
-
-\citet{angwin2016machine} \citet{bao2021s} \citet{barocas2017fairness}
-\citet{baumer2017texts} \citet{equivant_response_2018}
-\citet{gebru2021datasheets} \citet{hardin2015data}
-\citet{james2013introduction} \citet{knight2020automated}
-\citet{kypraiou_what_2021} \citet{larson2016technical}
-\citet{larson2016we} \citet{noauthor_ai_nodate}
-\citet{noauthor_aif360datasetscompasdataset_2018}
-\citet{noauthor_trusted-ai_360}
-\citet{noauthor_propublicacompas-analysis_2022}
-\citet{noauthor_womenthetable_nodate} \citet{vartan_1st_2022}
-
 % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2022-03-23 race compas analysis.Rmd
knitting rmd

---
title: "COMPAS Race Analysis"
author: "Elisabeth Nesmith"
date: "3/8/2022"
output: html_document
---

```{r setup, include=FALSE}
library(reticulate)
```

```{python}
import matplotlib.pyplot as plt
import matplotlib.patches as patches
import numpy as np
import pandas as pd
import seaborn as sns

import matplotlib.pyplot as plt
import matplotlib.patches as mpatches

from sklearn.compose import make_column_transformer
from sklearn.linear_model import LogisticRegression
from sklearn.linear_model import LogisticRegressionCV

from sklearn.model_selection import GridSearchCV, train_test_split
from sklearn.preprocessing import OneHotEncoder

from aif360.datasets import CompasDataset
from aif360.datasets import BinaryLabelDataset
from aif360.sklearn.datasets import fetch_compas
from aif360.algorithms.inprocessing import MetaFairClassifier

from aif360.sklearn.preprocessing import ReweighingMeta
from aif360.sklearn.inprocessing import AdversarialDebiasing
from aif360.sklearn.postprocessing import CalibratedEqualizedOdds, PostProcessingMeta
from aif360.sklearn.metrics import disparate_impact_ratio, average_odds_error, generalized_fpr, average_odds_difference
from aif360.sklearn.metrics import generalized_fnr, difference, statistical_parity_difference, equal_opportunity_difference, generalized_entropy_error

from IPython.display import Markdown, display
```

```{python}
## load in data
## In machine learning tasks, specifically with supervised learning, you have features and labels. 
## The features are the descriptive attributes (they are defined as X), and the label (y) is what you're attempting to predict or forecast

X, y = fetch_compas()
print(f'There are {X.shape[0]} entries and {X.shape[1]} features')
X.head()
```

```{python}
## because our analysis is mainly focusing on how the algorithm treats white and Black people differently, we are
## dropping the rows of data where race != Caucasian or African American
X_new = X[(X.race == "Caucasian") | (X.race == "African-American")]
print(f'There are {X_new.shape[0]} entries and {X_new.shape[1]} features')
X_new.head()
```

```{python}
## drop unused race categories
# list of categories to be removed
X_new["race"] = X_new["race"].cat.remove_unused_categories()
```

```{python}
y_new = y[(y.index.get_level_values(2) == "Caucasian") | (y.index.get_level_values(2) == "African-American")]
y_new.head()
```

```{python}
# Function for visualising the confusion matrix and other statistics
# https://github.com/DTrimarchi10/confusion_matrix/blob/master/cf_matrix.py

def make_confusion_matrix(cf_matrix, model):
  group_names = ["True Negative","False Positive","False Negative","True Positive"]
  group_counts = ["{0:0.0f}".format(value) for value in
                  cf_matrix.flatten()]
  group_percentages = ["{0:.2%}".format(value) for value in
                      cf_matrix.flatten()/np.sum(cf_matrix)]

  group_labels = ["{}\n".format(value) for value in group_names]
  group_counts = ["{0:0.0f}\n".format(value) for value in cf_matrix.flatten()]
  group_percentages = ["{0:.2%}".format(value) for value in cf_matrix.flatten()/np.sum(cf_matrix)]

  box_labels = [f"{v1}{v2}{v3}".strip() for v1, v2, v3 in zip(group_labels,group_counts,group_percentages)]
  box_labels = np.asarray(box_labels).reshape(cf_matrix.shape[0],cf_matrix.shape[1])

  # add more statistics
  accuracy  = np.trace(cf_matrix) / float(np.sum(cf_matrix))
  precision = cf_matrix[1,1] / sum(cf_matrix[:,1])
  recall    = cf_matrix[1,1] / sum(cf_matrix[1,:])
  f1_score  = 2*precision*recall / (precision + recall)
  stats_text = "\n\nAccuracy={:0.3f}\nPrecision={:0.3f}\nRecall={:0.3f}\nF1 Score={:0.3f}".format(
      accuracy,precision,recall,f1_score)

  categories=["Survived", "Recidivated"]
  sns.heatmap(cf_matrix,annot=box_labels,fmt="",cmap='Purples',xticklabels=categories,yticklabels=categories)

  plt.ylabel('True label')
  plt.xlabel('Predicted label' + stats_text)
  plt.title(f"Confusion matrix and statistics for the {model} model");

## defining function for displaying metrics of training and test data by sex 
def metrics_per_group(y_test, y_pred):
	# y true per group
	y_test_white = y_test.loc[y_test.index.get_level_values(2) == 1]
	y_test_black = y_test.loc[y_test.index.get_level_values(2) == 0]

	# y_pred per group
	y_pred_white = y_pred[y_test.index.get_level_values(2) == 1]
	y_pred_black = y_pred[y_test.index.get_level_values(2) == 0]

	# metrics
	scores = []
	scores.append(accuracy_score(y_test, y_pred))
	scores.append(recall_score(y_test, y_pred))
	scores.append(precision_score(y_test, y_pred))

	scores.append(accuracy_score(y_test_black, y_pred_black))
	scores.append(recall_score(y_test_black, y_pred_black))
	scores.append(precision_score(y_test_black, y_pred_black))

	scores.append(accuracy_score(y_test_white, y_pred_white))
	scores.append(recall_score(y_test_white, y_pred_white))
	scores.append(precision_score(y_test_white, y_pred_white))

	attribute = ["all"]*3 + ["black"] *3 + ["white"] *3
	metric = ["accuracy", "recall", "precision"] * 3

	# dictionary of lists 
	dict = {'race': attribute, 'metrics': metric, 'score': scores}

	df = pd.DataFrame(dict)

	sns.barplot(x = "metrics", y = "score", hue = "race", data = df, palette = ['#dfcd1a', '#9d0677', '#236c48'])
	plt.title("Performance metrics by groups")

def plot_fair_metrics(fair_metrics_mitigated, model): 
  cols = ['statistical_parity_difference', 'equal_opportunity_difference', 'generalized_entropy']
  obj_fairness = [[0,0,1]]

  # row for objectives    
  fair_metrics = pd.DataFrame(data=obj_fairness, index=['objective'], columns=cols)

  # row for baseline model
  fair_metrics.loc['Baseline Model'] = [stat_par_diff, eq_opp_diff, gen_entr_error]

  # row for mitigated bias
  fair_metrics.loc[model] = fair_metrics_mitigated

  metrics_len = len(cols)

  fig, ax = plt.subplots(figsize=(20,4), ncols=metrics_len, nrows=1)

  plt.subplots_adjust(
      left    =  0.125, 
      bottom  =  0.1, 
      right   =  0.9, 
      top     =  0.9, 
      wspace  =  .5, 
      hspace  =  1.1
  )

  y_title_margin = 1.2

  plt.suptitle("Fairness metrics", y = 1.09, fontsize=20)
  sns.set(style="dark")

  cols = fair_metrics.columns.values
  obj = fair_metrics.loc['objective']
  size_rect = [0.2,0.2,0.2,0.4,0.25]
  rect = [-0.1,-0.1,-0.1,0.8,0]
  bottom = [-1,-1,-1,0,0]
  top = [1,1,1,2,1]
  bound = [[-0.1,0.1],[-0.1,0.1],[-0.1,0.1],[0.8,1.2],[0,0.25]]

  for i in range(0,metrics_len):
      plt.subplot(1, metrics_len, i+1)
      ax = sns.barplot(x=fair_metrics.index[1:len(fair_metrics)], y=fair_metrics.iloc[1:len(fair_metrics)][cols[i]])

      for j in range(0,len(fair_metrics)-1):
          a, val = ax.patches[j], fair_metrics.iloc[j+1][cols[i]]
          marg = -0.2 if val < 0 else 0.1
          ax.text(a.get_x()+a.get_width()/5, a.get_y()+a.get_height()+marg, round(val, 3), fontsize=15,color='black')

      plt.ylim(bottom[i], top[i])
      plt.setp(ax.patches, linewidth=0)
      ax.add_patch(patches.Rectangle((-5,rect[i]), 10, size_rect[i], alpha=0.3, facecolor="green", linewidth=1, linestyle='solid'))
      plt.axhline(obj[i], color='black', alpha=0.3)
      plt.title(cols[i])
      ax.set_ylabel('')    
      ax.set_xlabel('')
```

```{python}
X_new.index = pd.MultiIndex.from_arrays(X_new.index.codes, names=X_new.index.names)
y_new.index = pd.MultiIndex.from_arrays(y_new.index.codes, names=y_new.index.names)
# 0 is African American, 2 is Caucasian
```

```{python}
# set caucasian equal to 1 instead of 2
X_new = X_new.rename(index={2: 1}, level='race')
X_new
```

```{python}
# set target class to 0/1 
y_new = pd.Series(y_new.factorize(sort=True)[0], index=y_new.index)
# set caucasian equal to 1 instead of 2
y_new = y_new.rename(index={2: 1}, level='race')
```

```{python}
y_new
```

# Exploratory Data Analysis
```{python}
df_viz = X_new.copy()
df_viz['race'] = X_new['race'].replace({1.0: 'Caucasian', 0.0: 'African-American'})
df_viz['two_year_recid'] = y_new.replace({1:'Recidivated', 0: 'Survived'})
df_viz.index = df_viz.index.droplevel('race')

purple = '#9d0677'
green = '#30875c'
workshop_palette = [purple, green]
df_viz.head()
```

```{python}
# barplot of recividism
sns.countplot(x='two_year_recid', data=df_viz, palette=workshop_palette)
plt.title('Two Year Recidivism Rate')
```

```{python}
# barplot of race
sns.countplot(y="race", data=df_viz, palette=["#30875c"])
plt.title('Race Distribution (White and Black individuals only)')
```

```{python}
# age distribution by race
ax = sns.kdeplot(x="age", hue="race", data=df_viz);
kdeline = ax.lines[0]
mean_fem = df_viz.groupby('race').age.median()[0]
xs = kdeline.get_xdata()
ys = kdeline.get_ydata()
height_fem = np.interp(mean_fem, xs, ys)
ax.vlines(mean_fem, 0, height_fem, color=purple, ls=':')

mean_m = df_viz.groupby('race').age.median()[1]
height_m = np.interp(mean_m, xs, ys)
ax.vlines(mean_m, 0, height_m, color=green, ls=':')

print(df_viz.groupby('race').age.median())

plt.title('Distribution of Age by Race')
```

```{python}
# recividism by race
by_sex = sns.countplot(x="race", hue="two_year_recid", data=df_viz, palette=workshop_palette)

plt.title('Two Year Recidivism by Race')
plt.show()
```

# Baseline Metrics

```{python}
X_train, X_test, y_train, y_test = train_test_split(X_new, y_new, random_state=1234567)

# one-hoy encode the categorical features
data_preproc = make_column_transformer(
        (OneHotEncoder(sparse=False, handle_unknown='ignore'), X_train.dtypes == 'category'))

X_train = pd.DataFrame(data_preproc.fit_transform(X_train), index=X_train.index)
X_test = pd.DataFrame(data_preproc.transform(X_test), index=X_test.index)

# to save the information for the column names
pd.get_dummies(X_new).head()
```

```{python}
from aif360.sklearn.preprocessing import ReweighingMeta, Reweighing
from sklearn.metrics import accuracy_score, confusion_matrix, plot_roc_curve, recall_score, precision_score

lr = LogisticRegressionCV(solver='lbfgs')
reg = lr.fit(X_train, y_train)
y_pred = reg.predict(X_test)

acc_base = accuracy_score(y_test, y_pred)
print(f'[Baseline] The test accuracy of the algorithm is: {acc_base: .2%}')
```

```{python}
##Look at 2nd column of matrix for precision 
##2nd row of matrix for recall 
cf_matrix = confusion_matrix(y_test, y_pred)
make_confusion_matrix(cf_matrix, "[Baseline]")
plt.show()
```

```{python}
metrics_per_group(y_test, y_pred)
plt.show()
```

In the Black population, the model correctly identifies 70% of people who recidivated. In the white population, the model correctly identifies 20% of people who recidivated.

### (Group fairness) Statistical Parity Difference

Computed as the difference of the rate of favorable outcomes (in this case, did not recidivate) received by the unprivileged group to the privileged group. It essentially equalizes the outcomes across the protected and non-protected groups.
The ideal value of this metric is 0. Fairness for this metric is between -0.1 and 0.1. A negative value means there is higher benefit for the privileged group (in this case, white defendants).

```{python}
stat_par_diff = statistical_parity_difference(y_test, prot_attr='race', pos_label = 0)

print(f'[Baseline] The statistical parity difference is {stat_par_diff: .2}')
```

### Generalized entropy error

Generalized entropy index is proposed as a unified individual and group fairness measure. It measures the inequality in benefit allocation for individuals.
A value of 0 implies perfect fairness.
Fairness is indicated by lower scores, higher scores are problematic.

```{python}
gen_entr_error = generalized_entropy_error(y_test, y_pred, alpha=1, pos_label = 0)

print(f'[Baseline] The generalized entropy error is {gen_entr_error: .2}')
```

### (Group fairness) Equal Opportunity Difference

This metric is computed as the difference of true positive rates between the unprivileged and the privileged groups. The true positive rate is the ratio of true positives to the total number of actual positives for a given group.
The ideal value is 0. A value of < 0 implies higher benefit for the privileged group and a value > 0 implies higher benefit for the unprivileged group.
Fairness for this metric is between -0.1 and 0.1

```{python}
eq_opp_diff = equal_opportunity_difference(y_test, y_pred, prot_attr='race', pos_label = 0)

print(f'[Baseline] The equal opportunity difference is {eq_opp_diff: .2}')
```

```{python}
plot_fair_metrics([0, 0, 0], '')
plt.show()
```

# Pre-Processing Approach
```{python}
lr = LogisticRegressionCV(solver='lbfgs')
rew = ReweighingMeta(estimator=lr, reweigher=Reweighing('race'))
rew.fit(X_train, y_train)
y_pred_REW = rew.predict(X_test)
```

```{python}
acc_REW = accuracy_score(y_test, y_pred_REW)
print(f'[Reweighting] The test accuracy of the algorithm is: {acc_REW: .2%}')
```

```{python}
cf_matrix = confusion_matrix(y_test, y_pred_REW)
make_confusion_matrix(cf_matrix, "[Reweighting]")
plt.show()
```

```{python}
metrics_per_group(y_test, y_pred_REW)
plt.show()
```

```{python}
stat_par_diff_RW = statistical_parity_difference(y_test, y_pred_REW, prot_attr='race', pos_label = 0)

print(f'[Reweighting] The statistical parity difference is {stat_par_diff_RW: .2}')
```

This is a large improvement over our baseline model, which was -0.14. It still implies a slight benefit for white individuals, but it is in the range of -0.1 and 0.1.

```{python}
eq_opp_diff_RW = equal_opportunity_difference(y_test, y_pred_REW, prot_attr='race', pos_label = 0)

print(f'[Reweighting] The equal opportunity difference is {eq_opp_diff_RW: .2}')
```

This is also an improvement from the baseline, as it is now between -0.1 and 0.1. It also now implies a slight benefit for Black defendants.

```{python}
gen_entr_error_RW = generalized_entropy_error(y_test, y_pred_REW, alpha=1, pos_label = 0)

print(f'[Reweighting] The generalized entropy error is {gen_entr_error_RW: .2}')
```

This is only a very slight improvement from 0.24 to 0.23.

```{python}
plot_fair_metrics([stat_par_diff_RW, eq_opp_diff_RW, gen_entr_error_RW], 'Reweighting')
plt.show()
```

# In-Processing

```{python}
df_train = X_train.copy()
df_train["two_year_recid"] = y_train
df_train.rename(columns={6:'race'}, inplace=True)
df_train = df_train.reset_index(drop=True)

df_test = X_test.copy()
df_test["two_year_recid"] = y_test
df_test = df_test.reset_index(drop=True)
df_test.rename(columns={6:'race'}, inplace=True)
```

```{python}
train_BLD = BinaryLabelDataset(favorable_label='0',
                                unfavorable_label='1',
                                df=df_train,
                                label_names=['two_year_recid'],
                                protected_attribute_names=['race'])

test_BLD = BinaryLabelDataset(favorable_label='0',
                                unfavorable_label='1',
                                df=df_test,
                                label_names=['two_year_recid'],
                                protected_attribute_names=['race'])
```

```{python}
# meta = MetaFairClassifier(tau=0.8, sensitive_attr='race', type='fdr', seed=1234567)
# 
# meta.fit(train_BLD)
# y_pred_META = meta.predict(test_BLD)
# y_pred_META = y_pred_META.labels.flatten()
```

```{python}
# acc_AD = accuracy_score(y_test, y_pred_META)
# print(f'[ Meta-Classification] The test accuracy of the algorithm is: {acc_AD: .2%}')
```

# Post-Processing

```{python}
from aif360.sklearn.postprocessing import CalibratedEqualizedOdds, PostProcessingMeta

pp = CalibratedEqualizedOdds('race', cost_constraint='fnr', random_state=1234567)
ceo = PostProcessingMeta(estimator=lr, postprocessor=pp, random_state=1234567)
ceo.fit(X_train, y_train)
y_pred_CEO = ceo.predict(X_test)
y_proba_CEO = ceo.predict_proba(X_test)
```

```{python}
acc_CEO = accuracy_score(y_test, y_pred_CEO)
print(f'[Calibrated Equalized Odds] The test accuracy of the algorithm is: {acc_CEO: .2%}')
```

```{python}
cf_matrix = confusion_matrix(y_test, y_pred_CEO)
make_confusion_matrix(cf_matrix, "[Calibrated Equalized Odds]")
plt.show()
```

```{python}
stat_par_diff_CEO = statistical_parity_difference(y_test, y_pred_CEO, prot_attr='race', pos_label = 0)

print(f'[Calibrated Equalized Odds] The statistical parity difference is {stat_par_diff_CEO: .2}')
```

```{python}
eq_opp_diff_CEO = equal_opportunity_difference(y_test, y_pred_CEO, prot_attr='race', pos_label = 0)

print(f'[Calibrated Equalized Odds] The equal opportunity difference is {eq_opp_diff_CEO: .2}')
```

```{python}
gen_entr_error_CEO = generalized_entropy_error(y_test, y_pred_CEO, alpha=1, pos_label = 0)

print(f'[Calibrated Equalized Odds] The generalized entropy error is {gen_entr_error_CEO: .2}')
```

```{python}
plot_fair_metrics([stat_par_diff_CEO, eq_opp_diff_CEO, gen_entr_error_CEO], 'Calibrated Equalized Odds')
plt.show()
```

### 2022-03-28 race compas analysis.Rmd

@@
 ```{python}
-# meta = MetaFairClassifier(tau=0.8, sensitive_attr='race', type='fdr', seed=1234567)
-# 
-# meta.fit(train_BLD)
-# y_pred_META = meta.predict(test_BLD)
-# y_pred_META = y_pred_META.labels.flatten()
+meta = MetaFairClassifier(tau=0.8, sensitive_attr='race', type='fdr', seed=1234567)
+
+meta.fit(train_BLD)
+y_pred_META = meta.predict(test_BLD)
+y_pred_META = y_pred_META.labels.flatten()
 ```
@@
 ```{python}
-# acc_AD = accuracy_score(y_test, y_pred_META)
-# print(f'[ Meta-Classification] The test accuracy of the algorithm is: {acc_AD: .2%}')
+acc_AD = accuracy_score(y_test, y_pred_META)
+print(f'[ Meta-Classification] The test accuracy of the algorithm is: {acc_AD: .2%}')
 ```

### 2022-03-28 race compas analysis.Rmd
adding table of race and recidivism

--- 2022-03-23 race compas analysis.Rmd
@@
 ```{python}
-# recividism by race
+# recidivism by race
 by_sex = sns.countplot(x="race", hue="two_year_recid", data=df_viz, palette=workshop_palette)
@@
 plt.show()
+```
+
+```{python}
+# table of recidivism by race
+pd.crosstab(index = df_viz["race"], columns = df_viz["two_year_recid"])
 ```

### 2022-03-30 race compas analysis.Rmd
testing fairness metrics code

@@
 print(f'[Baseline] The statistical parity difference is {stat_par_diff: .2}')
+
+disp_impact_ratio = disparate_impact_ratio(y_test, y_pred, prot_attr='race', pos_label = 0)
+
+print(f'[Baseline] The disparate impact ratio is {disp_impact_ratio: .2}')
+# disparate impact ratio without y_pred attribute was 0.77
+
+avg_odds_error = average_odds_error(y_test, y_pred, prot_attr='race', pos_label = 0)
+
+print(f'[Baseline] The average odds error is {avg_odds_error: .2}')
+
+avg_odds_diff = average_odds_difference(y_test, y_pred, prot_attr='race', pos_label = 0)
+
+print(f'[Baseline] The average odds difference is {avg_odds_diff: .2}')
 ```
@@
 print(f'[Baseline] The equal opportunity difference is {eq_opp_diff: .2}')
+
+#bt_grp_gen_entr_error = between_group_generalized_entropy_error(y_test, y_pred, alpha=1, pos_label = 0)
+
+#print(f'[Baseline] The between group generalized entropy error is {bt_grp_gen_entr_error: .2}')
+
+#gen_entr_index = generalized_entropy_index(x_test, alpha=0, pos_label = 0)
+
+#print(f'[Baseline] The generalized entropy index is {gen_entr_index: .2}')
+
+#consistency_score = consistency_score(x_test, y_test, n_neighbors=5, pos_label = 0)
+
+#print(f'[Baseline] The consistency score is {consistency_score: .2}')
 ```

### 2022-04-05 race compas analysis fairness metrics.Rmd
label chunks

--- 2022-03-30 race compas analysis.Rmd
@@
 ---
-title: "COMPAS Race Analysis"
+title: "COMPAS Race Analysis and Fairness Metrics"
 author: "Elisabeth Nesmith"
@@
 
-```{r setup, include=FALSE}
+# Set Up
+
+```{r py setup, include=FALSE}
 library(reticulate)
@@
 
-```{python}
+```{python import, include= FALSE}
 import matplotlib.pyplot as plt
@@
 
-```{python}
+# Data
+
+```{python fetch compas}
 ## load in data
@@
 
-```{python}
+## Filter/configure race
+
+```{python filter race}
 ## because our analysis is mainly focusing on how the algorithm treats white and Black people differently, we are
@@
 
-```{python}
+```{python drop other races}
 ## drop unused race categories
@@
 
-```{python}
+```{python configure races}
 y_new = y[(y.index.get_level_values(2) == "Caucasian") | (y.index.get_level_values(2) == "African-American")]
@@
 
-```{python}
+```{python initial confusion matrix}
 # Function for visualising the confusion matrix and other statistics
@@
 
-```{python}
+```{python new race index}
 X_new.index = pd.MultiIndex.from_arrays(X_new.index.codes, names=X_new.index.names)
@@
 
-```{python}
+```{python re-index race}
 # set caucasian equal to 1 instead of 2
@@
 
-```{python}
+```{python target class}
 # set target class to 0/1 
@@
 
-```{python}
+```{python y new race}
 y_new
@@
 # Exploratory Data Analysis
-```{python}
+```{python exploratory visualization set up}
 df_viz = X_new.copy()
@@
 
-```{python}
+```{python exploratory recidivism bar plot}
 # barplot of recividism
@@
 
-```{python}
+```{python exploratory race bar plot}
 # barplot of race
@@
 
-```{python}
+```{python exploratory age by race plot}
 # age distribution by race
@@
 
-```{python}
+```{python exploratory recidivism by race plot}
 # recidivism by race
@@
 
-```{python}
+```{python exploratory recidivism by race table}
 # table of recidivism by race
@@
 
-```{python}
+```{python baseline train and test}
 X_train, X_test, y_train, y_test = train_test_split(X_new, y_new, random_state=1234567)
@@
 
-```{python}
+```{python baseline set up, warning = FALSE, message = FALSE}
 from aif360.sklearn.preprocessing import ReweighingMeta, Reweighing
@@
 
-```{python}
+```{python baseline confusion matrix}
 ##Look at 2nd column of matrix for precision 
@@
 
-```{python}
+```{python baseline metrics per group}
 metrics_per_group(y_test, y_pred)
@@
 
-```{python}
+```{python group fairnesss statistical parity difference}
 stat_par_diff = statistical_parity_difference(y_test, prot_attr='race', pos_label = 0)
@@
 
-```{python}
+```{python group fairnesss generalized entropy error}
 gen_entr_error = generalized_entropy_error(y_test, y_pred, alpha=1, pos_label = 0)
@@
 
-```{python}
+```{python group fairnesss equal opportunity difference}
 eq_opp_diff = equal_opportunity_difference(y_test, y_pred, prot_attr='race', pos_label = 0)
@@
 
-```{python}
+```{python group fairness metrics plot}
 plot_fair_metrics([0, 0, 0], '')
@@
 # Pre-Processing Approach
-```{python}
+```{python pre-processing set up, warning = FALSE, message = FALSE}
 lr = LogisticRegressionCV(solver='lbfgs')
@@
 
-```{python}
+```{python pre-processing accuracy score} 
 acc_REW = accuracy_score(y_test, y_pred_REW)
@@
 
-```{python}
+```{python pre-processing confusion matrix}
 cf_matrix = confusion_matrix(y_test, y_pred_REW)
@@
 
-```{python}
+```{python pre-processing metrics per group}
 metrics_per_group(y_test, y_pred_REW)
@@
 
-```{python}
+```{python pre-processing statistical parity difference}
 stat_par_diff_RW = statistical_parity_difference(y_test, y_pred_REW, prot_attr='race', pos_label = 0)
@@
 
-```{python}
+```{python pre-processing equal opportunity difference}
 eq_opp_diff_RW = equal_opportunity_difference(y_test, y_pred_REW, prot_attr='race', pos_label = 0)
@@
 
-```{python}
+```{python pre-processing generalized entropy error}
 gen_entr_error_RW = generalized_entropy_error(y_test, y_pred_REW, alpha=1, pos_label = 0)
@@
 
-```{python}
+```{python pre-processing fairness metrics plot}
 plot_fair_metrics([stat_par_diff_RW, eq_opp_diff_RW, gen_entr_error_RW], 'Reweighting')
@@
 
-```{python}
+```{python in-processing set up}
 df_train = X_train.copy()
@@
 
-```{python}
+```{python in-processing training}
 train_BLD = BinaryLabelDataset(favorable_label='0',
@@
 
-```{python}
+```{python in-processing meta}
 # meta = MetaFairClassifier(tau=0.8, sensitive_attr='race', type='fdr', seed=1234567)
@@
 
-```{python}
+```{python in-processing accuracy score}
 # acc_AD = accuracy_score(y_test, y_pred_META)
@@
 
-```{python}
+```{python post-processing calibrated equal odds set up, warning = FALSE, message = FALSE}
 from aif360.sklearn.postprocessing import CalibratedEqualizedOdds, PostProcessingMeta
@@
 
-```{python}
+```{python post-processing accuracy score}
 acc_CEO = accuracy_score(y_test, y_pred_CEO)
@@
 
-```{python}
+```{python post-processing calibrated equal odds}
 cf_matrix = confusion_matrix(y_test, y_pred_CEO)
@@
 
-```{python}
+```{python post-processing stat parity difference}
 stat_par_diff_CEO = statistical_parity_difference(y_test, y_pred_CEO, prot_attr='race', pos_label = 0)
@@
 
-```{python}
+```{python post-processing equal opportunity diff}
 eq_opp_diff_CEO = equal_opportunity_difference(y_test, y_pred_CEO, prot_attr='race', pos_label = 0)
@@
 
-```{python}
+```{python post-processing generalized entropy error}
 gen_entr_error_CEO = generalized_entropy_error(y_test, y_pred_CEO, alpha=1, pos_label = 0)
@@
 
-```{python}
+```{python post-processing fairness metrics plot}
 plot_fair_metrics([stat_par_diff_CEO, eq_opp_diff_CEO, gen_entr_error_CEO], 'Calibrated Equalized Odds')

### 2022-04-05 race-compas-analysis.Rmd
label chunks

--- 2022-04-05 race compas analysis fairness metrics.Rmd
@@
 ---
-title: "COMPAS Race Analysis and Fairness Metrics"
+title: "COMPAS Race Analysis"
 author: "Elisabeth Nesmith"
@@
 
-```{python baseline set up, warning = FALSE, message = FALSE}
+```{python baseline set up}
 from aif360.sklearn.preprocessing import ReweighingMeta, Reweighing
@@
 # Pre-Processing Approach
-```{python pre-processing set up, warning = FALSE, message = FALSE}
+```{python pre-processing set up}
 lr = LogisticRegressionCV(solver='lbfgs')
@@
 
-```{python post-processing calibrated equal odds set up, warning = FALSE, message = FALSE}
+```{python post-processing calibrated equal odds set up}
 from aif360.sklearn.postprocessing import CalibratedEqualizedOdds, PostProcessingMeta

### 2022-04-05 Paper_Draft1/Draft2.Rmd

--- 2022-03-07 Paper_Draft1/Draft1.Rmd
@@
 ---
-
-# Version
-
-This Rmd-skeleton uses the mdpi Latex template published 2019/02. 
-However, the official template gets more frequently updated than the 'rticles'
-package. Therefore, please make sure prior to paper submission, that you're 
-using the most recent .cls, .tex and .bst files 
-(available [here](http://www.mdpi.com/authors/latex)).
 
@@
 
-# Results
-
-This section may be divided by subheadings.
-It should provide a concise and precise description of the experimental results, their interpretation as well as the experimental conclusions that can be drawn.
-
-## Subsection Heading Here
-
-Subsection text here.
-
-### Subsubsection Heading Here
-
-Bulleted lists look like this:
-
-* First bullet
-* Second bullet
-* Third bullet
-
-Numbered lists can be added as follows:
-
-1. First item
-2. Second item
-3. Third item
-
-The text continues here.
-
-All figures and tables should be cited in the main text as Figure \ref{fig:mdpi-logo}, Table 1, etc.
-
-```{r mdpi-logo, fig.align='center', echo=FALSE, fig.cap="This is a figure, Schemes follow the same formatting.
-If there are multiple panels, they should be listed as: (\\textbf{a}) Description of what is contained in the first panel.
-(\\textbf{b}) Description of what is contained in the second panel.
-Figures should be placed in the main text near to the first time they are cited.
-A caption on a single line should be centered."}
-
-# Ben's tutorial on how to include figures with R code chunk
-knitr::include_graphics("logo-mdpi.pdf")
-```
-
-Please see Table \ref{tab:mytable}.
-
-```{r}
-x <- tibble::tribble(~`Title 1`, ~`Title 2`, ~`Title 3`,
-"entry 1", "data", "data",
-"entry 2", "data", "data"
-)
-
-knitr::kable(x, caption = "This is a table caption. Tables should be placed in the main text near to the first time they are cited.\\label{tab:mytable}")
-```
-
-This is an example of an equation:
-
-\begin{equation}
-\mathbb{S}
-\end{equation}
-<!-- If the documentclass option "submit" is chosen, please insert a blank line before and after any math environment (equation and eqnarray environments).
-This ensures correct linenumbering.
-The blank line should be removed when the documentclass option is changed to "accept" because the text following an equation should not be a new paragraph.
--->
-
-<!-- Please punctuate equations as regular text. Theorem-type environments (including propositions, lemmas, corollaries etc.) can be formatted as follows: -->
-Example of a theorem:
-\begin{Theorem}
-Example text of a theorem.
-\end{Theorem}
-
-The text continues here. Proofs must be formatted as follows:
-
-Example of a proof:
-\begin{proof}[Proof of Theorem 1]
-Text of the proof. Note that the phrase `of Theorem 1' is optional if it is clear which theorem is being referred to.
-\end{proof}
-The text continues here.
-
-# Discussion
-
-Authors should discuss the results and how they can be interpreted in perspective of previous studies and of the working hypotheses.
-The findings and their implications should be discussed in the broadest context possible.
-Future research directions may also be highlighted.
-
-# Conclusion
-
-This section is not mandatory, but can be added to the manuscript if the discussion is unusually long or complex.
-
 # Bibliography

### 2022-04-05 race compas analysis.Rmd
echo = false

@@
 
-```{python}
+```{python baseline set up, warning = FALSE, message = FALSE, echo = FALSE}
 from aif360.sklearn.preprocessing import ReweighingMeta, Reweighing
@@
 # Pre-Processing Approach
-```{python}
+```{python pre-processing set up, warning = FALSE, message = FALSE, echo = FALSE}
 lr = LogisticRegressionCV(solver='lbfgs')
@@
 
-```{python}
+```{python post-processing calibrated equal odds set up, warning = FALSE, message = FALSE, echo = FALSE}
 from aif360.sklearn.postprocessing import CalibratedEqualizedOdds, PostProcessingMeta

### 2022-04-05 final_metrics_analysis.Rmd

--- 2022-04-05 race compas analysis fairness metrics.Rmd
@@
 
-```{python baseline set up, warning = FALSE, message = FALSE}
+```{python baseline set up, include = FALSE}
 from aif360.sklearn.preprocessing import ReweighingMeta, Reweighing
@@
 
-### (Group fairness) Statistical Parity Difference
+## Group Fairness Metrics
+
+### Statistical Parity Difference
 
@@
 print(f'[Baseline] The statistical parity difference is {stat_par_diff: .2}')
+```
 
+### Disparate Impact Ratio
+
+```{python group fairnesss disparate impact ratio}
 disp_impact_ratio = disparate_impact_ratio(y_test, y_pred, prot_attr='race', pos_label = 0)
@@
 # disparate impact ratio without y_pred attribute was 0.77
+```
 
-avg_odds_error = average_odds_error(y_test, y_pred, prot_attr='race', pos_label = 0)
+### Average Odds Difference
 
-print(f'[Baseline] The average odds error is {avg_odds_error: .2}')
-
+```{python group fairnesss average odds difference}
 avg_odds_diff = average_odds_difference(y_test, y_pred, prot_attr='race', pos_label = 0)
@@
 ```
+
+### Equal Opportunity Difference
+
+This metric is computed as the difference of true positive rates between the unprivileged and the privileged groups.
+The true positive rate is the ratio of true positives to the total number of actual positives for a given group.
+The ideal value is 0. A value of < 0 implies higher benefit for the privileged group and a value > 0 implies higher benefit for the unprivileged group.
+Fairness for this metric is between -0.1 and 0.1
+
+```{python group fairnesss equal opportunity difference}
+eq_opp_diff = equal_opportunity_difference(y_test, y_pred, prot_attr='race', pos_label = 0)
+
+print(f'[Baseline] The equal opportunity difference is {eq_opp_diff: .2}')
+```
+
+## Individual Fairness Metrics
 
@@
 print(f'[Baseline] The generalized entropy error is {gen_entr_error: .2}')
-```
-
-### (Group fairness) Equal Opportunity Difference
-
-This metric is computed as the difference of true positive rates between the unprivileged and the privileged groups.
-The true positive rate is the ratio of true positives to the total number of actual positives for a given group.
-The ideal value is 0. A value of < 0 implies higher benefit for the privileged group and a value > 0 implies higher benefit for the unprivileged group.
-Fairness for this metric is between -0.1 and 0.1
-
-```{python group fairnesss equal opportunity difference}
-eq_opp_diff = equal_opportunity_difference(y_test, y_pred, prot_attr='race', pos_label = 0)
-
-print(f'[Baseline] The equal opportunity difference is {eq_opp_diff: .2}')
 
@@
 # Pre-Processing Approach
-```{python pre-processing set up, warning = FALSE, message = FALSE}
+```{python pre-processing set up, include = FALSE}
 lr = LogisticRegressionCV(solver='lbfgs')
@@
 
-```{python post-processing calibrated equal odds set up, warning = FALSE, message = FALSE}
+```{python post-processing calibrated equal odds set up, include = FALSE}
 from aif360.sklearn.postprocessing import CalibratedEqualizedOdds, PostProcessingMeta

### 2022-04-05 final_metrics_analysis.Rmd

@@
 from aif360.sklearn.postprocessing import CalibratedEqualizedOdds, PostProcessingMeta
-from aif360.sklearn.metrics import disparate_impact_ratio, average_odds_error, generalized_fpr, average_odds_difference
-from aif360.sklearn.metrics import generalized_fnr, difference, statistical_parity_difference, equal_opportunity_difference, generalized_entropy_error
+
+from aif360.sklearn.metrics import disparate_impact_ratio, average_odds_difference, statistical_parity_difference, equal_opportunity_difference
+from aif360.sklearn.metrics import generalized_entropy_error, generalized_entropy_index, theil_index, consistency_score
 
@@
 
+**add brief definition of group fairness**
+
 ### Statistical Parity Difference
@@
 ### Disparate Impact Ratio
+
+This metric is the ratio of how often the favorable outcome occurs in one group versus the other.
+In the case of recidivism, this is the ratio of how many white defendants are predicted to not recidivate compared to how many black defendants are predicted to not recidivate.
+
+A value of 1 means that the ratio is exactly 1:1.
+Less than 1 means the privileged group (white defendants) benefits, while a value greater than 1 means the unprivileged group (Black defendants) benefits.
+According to AI Fairness 360, a ratio between .8 to 1.25 is considered fair.
+
+The disparate impact ratio is calculated with the formula:
 
@@
 print(f'[Baseline] The generalized entropy error is {gen_entr_error: .2}')
-
-#bt_grp_gen_entr_error = between_group_generalized_entropy_error(y_test, y_pred, alpha=1, pos_label = 0)
-
-#print(f'[Baseline] The between group generalized entropy error is {bt_grp_gen_entr_error: .2}')
 

### 2022-04-07 final_metrics_analysis.Rmd
adding in fairness definitions to rmd

@@
 
-**add brief definition of group fairness**
+With group fairness metrics, different groups should receive similar treatments or outcomes.
+In this case of recidivism and race, this means that Black defendants should have similar rates of predicted recidivism as white defendants.
 
@@
 
-Computed as the difference of the rate of favorable outcomes (in this case, did not recidivate) received by the unprivileged group to the privileged group.
-It essentially equalizes the outcomes across the protected and non-protected groups.
+Statistical Parity difference is computed as the difference in the rate of favorable outcomes (in this case, did not recidivate) received by the unprivileged group to the privileged group.
+It essentially equalizes the outcomes across the privileged and non-privileged groups.
 The ideal value of this metric is 0. Fairness for this metric is between -0.1 and 0.1. A negative value means there is higher benefit for the privileged group (in this case, white defendants).
+
+$P(\hat{Y}=1|D=Unprivileged) - P(\hat{Y}=1|D=Privileged)$
 
@@
 ```
+
+Because the statistical parity difference is negative and not within -0.1 and 0.1, the algorithm unfairly benefits white defendants.
 
@@
 
+$\frac{P(\hat{Y}=1|D=Unprivileged)}{P(\hat{Y}=1|D=Privileged)}$
+
+where Y is the favorable outcome of a defendant's predicted two year recidivism and D is the race of the defendant.
+In our case, the favorable outcome is survived, or not recidivated, which is Y = 0.
+In terms of race, 0 = Black and 1 = White.
+So to calculate the disparate impact ratio on this data, the formula is:
+
+$\frac{P(\hat{Y}=0|D=0)}{P(\hat{Y}=0|D=1)}$
+
+Say we have a data test set of 200 defendants, 125 Black and 75 white.
+Of the 125 Black defendants, 44% (55) were predicted to not recidivate.
+Of the 75 white defendants, 66% (50) were predicted to not recidivate.
+
+The disparate impact ratio would therefore be
+
+$\frac{0.44}{0.66} = 0.667$
+
+Because this value is less than 1, it benefits white defendants. Additionally, because it is below 0.8, it is considered "unfair."
+
 ```{python group fairnesss disparate impact ratio}
-disp_impact_ratio = disparate_impact_ratio(y_test, y_pred, prot_attr='race', pos_label = 0)
+disp_impact_ratio = disparate_impact_ratio(y_test, y_pred, prot_attr='race', priv_group = 1, pos_label = 0)
 
 print(f'[Baseline] The disparate impact ratio is {disp_impact_ratio: .2}')
-# disparate impact ratio without y_pred attribute was 0.77
 ```
 
+Since the actual disparate impact ratio for our test data is 0.47 and below 0.8, the algorithm unfairly benefits white defendants i.e. disproportionately predicts them to not recidivate.
+
 ### Average Odds Difference
+
+This metric returns the average difference in false positive rate and true positive rate for the privileged and unprivileged groups.
+A value of 0 is considered fair, and a value below 0 implies benefit for the privileged group.
+Equality of odds is achieved in the case of recidivism when the proportion of people who were predicted to recidivate and did recidivate is equal (true positive rate) for both black and white defendants AND the proportion of people who were predicted to recidivate and did not recidivate (false positive rate) is equal for both black and white defendants.
+
+$\frac{(FPR_{D = unprivileged} - FPR_{D = privileged}) + (TPR_{D = unprivileged} - TPR_{D = privileged})}{2}$
 
@@
 
+Since the average odds difference is below 0 by a substantial amount, white defendants are unfairly benefitted by the algorithm.
+
 ### Equal Opportunity Difference
@@
 The ideal value is 0. A value of < 0 implies higher benefit for the privileged group and a value > 0 implies higher benefit for the unprivileged group.
-Fairness for this metric is between -0.1 and 0.1
+Fairness for this metric is between -0.1 and 0.1.
+
+$TPR_{D = Unprivileged} - TPR_{D = Privileged}$
 
@@
 ```
+
+Since the equal opportunity difference is below 0 and beyond -0.1, the algorithm unfairly benefits white defendants.
 

### 2022-04-08 final_metrics_analysis.Rmd

@@
 ---
-title: "COMPAS Race Analysis"
-author: "Elisabeth Nesmith"
+title: "Analyzing Fairness: Race, Recidivism, and the COMPAS Algorithm"
+author: "Elisabeth Nesmith, Eunice Kim, Emma Kornberg, Amrita Acharya, Dianne Caravela"
 date: "3/8/2022"

### 2022-04-11 final_metrics_analysis.Rmd

@@
 from aif360.sklearn.metrics import generalized_entropy_error, generalized_entropy_index, theil_index, consistency_score
+
+from aif360.algorithms.preprocessing import DisparateImpactRemover
+from BlackBoxAuditing.repairers.GeneralRepairer import Repairer
 
@@
 
+```{python pre-processing disparate impact remover}
+#lr = LogisticRegressionCV(solver='lbfgs')
+#rew = ReweighingMeta(estimator=lr, reweigher=Reweighing('race'))
+#rew.fit(X_train, y_train)
+#y_pred_REW = rew.predict(X_test)
+
+diremover = DisparateImpactRemover(repair_level=1.0, sensitive_attribute="race")
+#diremover.fit(X_train, y_train)
+#y_pred_diremover = diremover.predict(X_test)
+
+diremover_rp = diremover.fit_transform(X_test)
+# AttributeError: 'DataFrame' object has no attribute 'features' (probem with .fit_transform function)
+# .fit_transform() angry because of X_test...
+
+```
+
 # In-Processing

### 2022-04-12 final_metrics_analysis.Rmd
Amrita's changes

--- 2022-04-07 final_metrics_analysis.Rmd
@@
 library(reticulate)
+use_condaenv("anaconda3")
 ```
@@
 from aif360.sklearn.postprocessing import CalibratedEqualizedOdds, PostProcessingMeta
-
-from aif360.sklearn.metrics import disparate_impact_ratio, average_odds_difference, statistical_parity_difference, equal_opportunity_difference
-from aif360.sklearn.metrics import generalized_entropy_error, generalized_entropy_index, theil_index, consistency_score
+from aif360.sklearn.metrics import disparate_impact_ratio, average_odds_error, generalized_fpr, average_odds_difference
+from aif360.sklearn.metrics import generalized_fnr, difference, statistical_parity_difference, equal_opportunity_difference, generalized_entropy_error
 
@@
 
-```{python baseline set up, include = FALSE}
+```{python baseline set up, warning = FALSE, message = FALSE}
 from aif360.sklearn.preprocessing import ReweighingMeta, Reweighing
@@
 
-## Group Fairness Metrics
+### (Group fairness) Statistical Parity Difference
 
-With group fairness metrics, different groups should receive similar treatments or outcomes.
-In this case of recidivism and race, this means that Black defendants should have similar rates of predicted recidivism as white defendants.
-
-### Statistical Parity Difference
-
-Statistical Parity difference is computed as the difference in the rate of favorable outcomes (in this case, did not recidivate) received by the unprivileged group to the privileged group.
-It essentially equalizes the outcomes across the privileged and non-privileged groups.
+Computed as the difference of the rate of favorable outcomes (in this case, did not recidivate) received by the unprivileged group to the privileged group.
+It essentially equalizes the outcomes across the protected and non-protected groups.
 The ideal value of this metric is 0. Fairness for this metric is between -0.1 and 0.1. A negative value means there is higher benefit for the privileged group (in this case, white defendants).
-
-$P(\hat{Y}=1|D=Unprivileged) - P(\hat{Y}=1|D=Privileged)$
 
@@
 print(f'[Baseline] The statistical parity difference is {stat_par_diff: .2}')
-```
 
-Because the statistical parity difference is negative and not within -0.1 and 0.1, the algorithm unfairly benefits white defendants.
-
-### Disparate Impact Ratio
-
-This metric is the ratio of how often the favorable outcome occurs in one group versus the other.
-In the case of recidivism, this is the ratio of how many white defendants are predicted to not recidivate compared to how many black defendants are predicted to not recidivate.
-
-A value of 1 means that the ratio is exactly 1:1.
-Less than 1 means the privileged group (white defendants) benefits, while a value greater than 1 means the unprivileged group (Black defendants) benefits.
-According to AI Fairness 360, a ratio between .8 to 1.25 is considered fair.
-
-The disparate impact ratio is calculated with the formula:
-
-$\frac{P(\hat{Y}=1|D=Unprivileged)}{P(\hat{Y}=1|D=Privileged)}$
-
-where Y is the favorable outcome of a defendant's predicted two year recidivism and D is the race of the defendant.
-In our case, the favorable outcome is survived, or not recidivated, which is Y = 0.
-In terms of race, 0 = Black and 1 = White.
-So to calculate the disparate impact ratio on this data, the formula is:
-
-$\frac{P(\hat{Y}=0|D=0)}{P(\hat{Y}=0|D=1)}$
-
-Say we have a data test set of 200 defendants, 125 Black and 75 white.
-Of the 125 Black defendants, 44% (55) were predicted to not recidivate.
-Of the 75 white defendants, 66% (50) were predicted to not recidivate.
-
-The disparate impact ratio would therefore be
-
-$\frac{0.44}{0.66} = 0.667$
-
-Because this value is less than 1, it benefits white defendants. Additionally, because it is below 0.8, it is considered "unfair."
-
-```{python group fairnesss disparate impact ratio}
-disp_impact_ratio = disparate_impact_ratio(y_test, y_pred, prot_attr='race', priv_group = 1, pos_label = 0)
+disp_impact_ratio = disparate_impact_ratio(y_test, y_pred, prot_attr='race', pos_label = 0)
 
 print(f'[Baseline] The disparate impact ratio is {disp_impact_ratio: .2}')
-```
+# disparate impact ratio without y_pred attribute was 0.77
 
-Since the actual disparate impact ratio for our test data is 0.47 and below 0.8, the algorithm unfairly benefits white defendants i.e. disproportionately predicts them to not recidivate.
+avg_odds_error = average_odds_error(y_test, y_pred, prot_attr='race', pos_label = 0)
 
-### Average Odds Difference
+print(f'[Baseline] The average odds error is {avg_odds_error: .2}')
 
-This metric returns the average difference in false positive rate and true positive rate for the privileged and unprivileged groups.
-A value of 0 is considered fair, and a value below 0 implies benefit for the privileged group.
-Equality of odds is achieved in the case of recidivism when the proportion of people who were predicted to recidivate and did recidivate is equal (true positive rate) for both black and white defendants AND the proportion of people who were predicted to recidivate and did not recidivate (false positive rate) is equal for both black and white defendants.
-
-$\frac{(FPR_{D = unprivileged} - FPR_{D = privileged}) + (TPR_{D = unprivileged} - TPR_{D = privileged})}{2}$
-
-```{python group fairnesss average odds difference}
 avg_odds_diff = average_odds_difference(y_test, y_pred, prot_attr='race', pos_label = 0)
@@
 ```
-
-Since the average odds difference is below 0 by a substantial amount, white defendants are unfairly benefitted by the algorithm.
-
-### Equal Opportunity Difference
-
-This metric is computed as the difference of true positive rates between the unprivileged and the privileged groups.
-The true positive rate is the ratio of true positives to the total number of actual positives for a given group.
-The ideal value is 0. A value of < 0 implies higher benefit for the privileged group and a value > 0 implies higher benefit for the unprivileged group.
-Fairness for this metric is between -0.1 and 0.1.
-
-$TPR_{D = Unprivileged} - TPR_{D = Privileged}$
-
-```{python group fairnesss equal opportunity difference}
-eq_opp_diff = equal_opportunity_difference(y_test, y_pred, prot_attr='race', pos_label = 0)
-
-print(f'[Baseline] The equal opportunity difference is {eq_opp_diff: .2}')
-```
-
-Since the equal opportunity difference is below 0 and beyond -0.1, the algorithm unfairly benefits white defendants.
-
-## Individual Fairness Metrics
 
@@
 print(f'[Baseline] The generalized entropy error is {gen_entr_error: .2}')
+```
+
+### (Group fairness) Equal Opportunity Difference
+
+This metric is computed as the difference of true positive rates between the unprivileged and the privileged groups.
+The true positive rate is the ratio of true positives to the total number of actual positives for a given group.
+The ideal value is 0. A value of < 0 implies higher benefit for the privileged group and a value > 0 implies higher benefit for the unprivileged group.
+Fairness for this metric is between -0.1 and 0.1
+
+```{python group fairnesss equal opportunity difference}
+eq_opp_diff = equal_opportunity_difference(y_test, y_pred, prot_attr='race', pos_label = 0)
+
+print(f'[Baseline] The equal opportunity difference is {eq_opp_diff: .2}')
+
+#bt_grp_gen_entr_error = between_group_generalized_entropy_error(y_test, y_pred, alpha=1, pos_label = 0)
+
+#print(f'[Baseline] The between group generalized entropy error is {bt_grp_gen_entr_error: .2}')
 
@@
 # Pre-Processing Approach
-```{python pre-processing set up, include = FALSE}
+```{python pre-processing set up, warning = FALSE, message = FALSE}
 lr = LogisticRegressionCV(solver='lbfgs')
@@
 
-```{python post-processing calibrated equal odds set up, include = FALSE}
+```{python post-processing calibrated equal odds set up, warning = FALSE, message = FALSE}
 from aif360.sklearn.postprocessing import CalibratedEqualizedOdds, PostProcessingMeta

### 2022-04-12 final_metrics_analysis.Rmd

@@
 from aif360.sklearn.metrics import generalized_fnr, difference, statistical_parity_difference, equal_opportunity_difference, generalized_entropy_error
+
+from aif360.algorithms.preprocessing import DisparateImpactRemover
+from BlackBoxAuditing.repairers.GeneralRepairer import Repairer
 

### 2022-04-13 final_metrics_analysis.Rmd
get rid of blackboxauditing
updating metrics plot function

--- 2022-04-11 final_metrics_analysis.Rmd
@@
 from aif360.sklearn.metrics import disparate_impact_ratio, average_odds_difference, statistical_parity_difference, equal_opportunity_difference
-from aif360.sklearn.metrics import generalized_entropy_error, generalized_entropy_index, theil_index, consistency_score
 
@@
 def plot_fair_metrics(fair_metrics_mitigated, model): 
-  cols = ['statistical_parity_difference', 'equal_opportunity_difference', 'generalized_entropy']
-  obj_fairness = [[0,0,1]]
+  cols = ['statistical_parity_difference','equal_opportunity_difference','average_odds_difference','disparate_impact_ratio']
+  obj_fairness = [[0,0,0,1]]
 
@@
   # row for baseline model
-  fair_metrics.loc['Baseline Model'] = [stat_par_diff, eq_opp_diff, gen_entr_error]
+  fair_metrics.loc['Baseline Model'] = [stat_par_diff, eq_opp_diff, avg_odds_diff, disp_impact_ratio]
 
@@
   top = [1,1,1,2,1]
-  bound = [[-0.1,0.1],[-0.1,0.1],[-0.1,0.1],[0.8,1.2],[0,0.25]]
+  bound = [[-0.1,0.1],[-0.1,0.1],[-0.1,0.1],[0.8,1.25]]
 
@@
 
-## Individual Fairness Metrics
-
-### Generalized entropy error
-
-Generalized entropy index is proposed as a unified individual and group fairness measure. It measures the inequality in benefit allocation for individuals.
-A value of 0 implies perfect fairness.
-Fairness is indicated by lower scores, higher scores are problematic.
-
-```{python group fairnesss generalized entropy error}
-gen_entr_error = generalized_entropy_error(y_test, y_pred, alpha=1, pos_label = 0)
-
-print(f'[Baseline] The generalized entropy error is {gen_entr_error: .2}')
-
-#gen_entr_index = generalized_entropy_index(x_test, alpha=0, pos_label = 0)
-
-#print(f'[Baseline] The generalized entropy index is {gen_entr_index: .2}')
-
-#consistency_score = consistency_score(x_test, y_test, n_neighbors=5, pos_label = 0)
-
-#print(f'[Baseline] The consistency score is {consistency_score: .2}')
-```
-
 ```{python group fairness metrics plot}
-plot_fair_metrics([0, 0, 0], '')
+plot_fair_metrics([0, 0, 0, 0], '')
 plt.show()
@@
 
-# Pre-Processing Approach
+# Pre-Processing Approaches
+
+## Reweighing
+
 ```{python pre-processing set up, include = FALSE}
@@
 
+### Statistical Parity Difference
+
 ```{python pre-processing statistical parity difference}
@@
 This is a large improvement over our baseline model, which was -0.14. It still implies a slight benefit for white individuals, but it is in the range of -0.1 and 0.1.
+
+### Equal Opportunity Difference
 
@@
 
-```{python pre-processing generalized entropy error}
-gen_entr_error_RW = generalized_entropy_error(y_test, y_pred_REW, alpha=1, pos_label = 0)
+### Average Odds Difference
 
-print(f'[Reweighting] The generalized entropy error is {gen_entr_error_RW: .2}')
+```{python pre-processing average odds difference}
+avg_odds_diff_RW = average_odds_difference(y_test, y_pred_REW, prot_attr='race', pos_label = 0)
+
+print(f'[Reweighting] The average odds difference is {avg_odds_diff_RW: .2}')
 ```
 
-This is only a very slight improvement from 0.24 to 0.23.
+This is also a vast improvement from -0.44 to 0.014. Since it is above 1 and below 0.1, it implies a slight benefit for Black defendants, but not an unfair benefit.
+
+### Disparate Impact Ratio
+```{python pre_processing disparate impact ratio}
+disp_impact_ratio_RW = disparate_impact_ratio(y_test, y_pred_REW, prot_attr='race', priv_group = 1, pos_label = 0)
+
+print(f'[Reweighting] The disparate impact ratio is {disp_impact_ratio_RW: .2}')
+```
+
+The reweighting approach also improves the disparate impact ratio, bringing the value almost to 1, which conveys absolute fairness.
 
 ```{python pre-processing fairness metrics plot}
-plot_fair_metrics([stat_par_diff_RW, eq_opp_diff_RW, gen_entr_error_RW], 'Reweighting')
+plot_fair_metrics([stat_par_diff_RW, eq_opp_diff_RW, avg_odds_diff_RW, disp_impact_ratio_RW], 'Reweighting')
 plt.show()
@@
 
-diremover = DisparateImpactRemover(repair_level=1.0, sensitive_attribute="race")
+# diremover = DisparateImpactRemover(repair_level=1.0, sensitive_attribute="race")
 #diremover.fit(X_train, y_train)
@@
 
-diremover_rp = diremover.fit_transform(X_test)
+# diremover_rp = diremover.fit_transform(X_test)
 # AttributeError: 'DataFrame' object has no attribute 'features' (probem with .fit_transform function)

### 2022-04-13 final_metrics_analysis.Rmd

@@
 
-from aif360.algorithms.preprocessing import DisparateImpactRemover
-from BlackBoxAuditing.repairers.GeneralRepairer import Repairer
+# from aif360.algorithms.preprocessing import DisparateImpactRemover
+# from BlackBoxAuditing.repairers.GeneralRepairer import Repairer
 
@@
   obj = fair_metrics.loc['objective']
-  size_rect = [0.2,0.2,0.2,0.4,0.25]
-  rect = [-0.1,-0.1,-0.1,0.8,0]
-  bottom = [-1,-1,-1,0,0]
-  top = [1,1,1,2,1]
+  size_rect = [0.2,0.2,0.2,0.4]
+  rect = [-0.1,-0.1,-0.1,0.8]
+  bottom = [-1,-1,-1,0]
+  top = [1,1,1,2]
   bound = [[-0.1,0.1],[-0.1,0.1],[-0.1,0.1],[0.8,1.25]]
@@
 
+### Plot Fairness Metrics (Baseline)
+
 ```{python group fairness metrics plot}
@@
 
+### Plot Fairness Metrics (Reweighting)
+
 ```{python pre-processing fairness metrics plot}
@@
 # .fit_transform() angry because of X_test... 
-
 ```
@@
 
+### Statistical Parity Difference
+
 ```{python post-processing stat parity difference}
@@
 ```
+
+### Equal Opportunity Difference
 
@@
 
-```{python post-processing generalized entropy error}
-gen_entr_error_CEO = generalized_entropy_error(y_test, y_pred_CEO, alpha=1, pos_label = 0)
+### Average Odds Difference
 
-print(f'[Calibrated Equalized Odds] The generalized entropy error is {gen_entr_error_CEO: .2}')
+```{python post-processing average odds difference}
+avg_odds_diff_CEO = average_odds_difference(y_test, y_pred_CEO, prot_attr='race', pos_label = 0)
+
+print(f'[Calibrated Equalized Odds] The average odds difference is {avg_odds_diff_CEO: .2}')
 ```
 
+### Disparate Impact Ratio
+
+```{python post_processing disparate impact ratio}
+disp_impact_ratio_CEO = disparate_impact_ratio(y_test, y_pred_CEO, prot_attr='race', priv_group = 1, pos_label = 0)
+
+print(f'[Calibrated Equalized Odds] The disparate impact ratio is {disp_impact_ratio_CEO: .2}')
+```
+
+### Plot Fairness Metrics (Calibrated Equalized Odds)
+
 ```{python post-processing fairness metrics plot}
-plot_fair_metrics([stat_par_diff_CEO, eq_opp_diff_CEO, gen_entr_error_CEO], 'Calibrated Equalized Odds')
+plot_fair_metrics([stat_par_diff_CEO, eq_opp_diff_CEO, avg_odds_diff_CEO, disp_impact_ratio_CEO], 'Calibrated Equalized Odds')
 plt.show()

### 2022-04-13 final_metrics_analysis.Rmd
Dianne's correct conda env

--- 2022-04-13 final_metrics_analysis.Rmd
@@
 library(reticulate)
+use_condaenv("base")
 ```
@@
 from aif360.sklearn.metrics import disparate_impact_ratio, average_odds_difference, statistical_parity_difference, equal_opportunity_difference
+from aif360.sklearn.metrics import generalized_entropy_error, generalized_entropy_index, theil_index, consistency_score
 
 from aif360.algorithms.preprocessing import DisparateImpactRemover
-from BlackBoxAuditing.repairers.GeneralRepairer import Repairer
+#from BlackBoxAuditing.repairers.GeneralRepairer import Repairer
 
@@
 def plot_fair_metrics(fair_metrics_mitigated, model): 
-  cols = ['statistical_parity_difference','equal_opportunity_difference','average_odds_difference','disparate_impact_ratio']
-  obj_fairness = [[0,0,0,1]]
+  cols = ['statistical_parity_difference', 'equal_opportunity_difference', 'generalized_entropy']
+  obj_fairness = [[0,0,1]]
 
@@
   # row for baseline model
-  fair_metrics.loc['Baseline Model'] = [stat_par_diff, eq_opp_diff, avg_odds_diff, disp_impact_ratio]
+  fair_metrics.loc['Baseline Model'] = [stat_par_diff, eq_opp_diff, gen_entr_error]
 
@@
   top = [1,1,1,2,1]
-  bound = [[-0.1,0.1],[-0.1,0.1],[-0.1,0.1],[0.8,1.25]]
+  bound = [[-0.1,0.1],[-0.1,0.1],[-0.1,0.1],[0.8,1.2],[0,0.25]]
 
@@
 
+## Individual Fairness Metrics
+
+### Generalized entropy error
+
+Generalized entropy index is proposed as a unified individual and group fairness measure. It measures the inequality in benefit allocation for individuals.
+A value of 0 implies perfect fairness.
+Fairness is indicated by lower scores, higher scores are problematic.
+
+```{python group fairnesss generalized entropy error}
+gen_entr_error = generalized_entropy_error(y_test, y_pred, alpha=1, pos_label = 0)
+
+print(f'[Baseline] The generalized entropy error is {gen_entr_error: .2}')
+
+#gen_entr_index = generalized_entropy_index(x_test, alpha=0, pos_label = 0)
+
+#print(f'[Baseline] The generalized entropy index is {gen_entr_index: .2}')
+
+#consistency_score = consistency_score(x_test, y_test, n_neighbors=5, pos_label = 0)
+
+#print(f'[Baseline] The consistency score is {consistency_score: .2}')
+```
+
 ```{python group fairness metrics plot}
-plot_fair_metrics([0, 0, 0, 0], '')
+plot_fair_metrics([0, 0, 0], '')
 plt.show()
@@
 
-# Pre-Processing Approaches
-
-## Reweighing
-
+# Pre-Processing Approach
 ```{python pre-processing set up, include = FALSE}
@@
 
-### Statistical Parity Difference
-
 ```{python pre-processing statistical parity difference}
@@
 This is a large improvement over our baseline model, which was -0.14. It still implies a slight benefit for white individuals, but it is in the range of -0.1 and 0.1.
-
-### Equal Opportunity Difference
 
@@
 
-### Average Odds Difference
+```{python pre-processing generalized entropy error}
+gen_entr_error_RW = generalized_entropy_error(y_test, y_pred_REW, alpha=1, pos_label = 0)
 
-```{python pre-processing average odds difference}
-avg_odds_diff_RW = average_odds_difference(y_test, y_pred_REW, prot_attr='race', pos_label = 0)
-
-print(f'[Reweighting] The average odds difference is {avg_odds_diff_RW: .2}')
+print(f'[Reweighting] The generalized entropy error is {gen_entr_error_RW: .2}')
 ```
 
-This is also a vast improvement from -0.44 to 0.014. Since it is above 1 and below 0.1, it implies a slight benefit for Black defendants, but not an unfair benefit.
-
-### Disparate Impact Ratio
-```{python pre_processing disparate impact ratio}
-disp_impact_ratio_RW = disparate_impact_ratio(y_test, y_pred_REW, prot_attr='race', priv_group = 1, pos_label = 0)
-
-print(f'[Reweighting] The disparate impact ratio is {disp_impact_ratio_RW: .2}')
-```
-
-The reweighting approach also improves the disparate impact ratio, bringing the value almost to 1, which conveys absolute fairness.
+This is only a very slight improvement from 0.24 to 0.23.
 
 ```{python pre-processing fairness metrics plot}
-plot_fair_metrics([stat_par_diff_RW, eq_opp_diff_RW, avg_odds_diff_RW, disp_impact_ratio_RW], 'Reweighting')
+plot_fair_metrics([stat_par_diff_RW, eq_opp_diff_RW, gen_entr_error_RW], 'Reweighting')
 plt.show()
@@
 
-# diremover = DisparateImpactRemover(repair_level=1.0, sensitive_attribute="race")
+diremover = DisparateImpactRemover(repair_level=1.0, sensitive_attribute="race")
 #diremover.fit(X_train, y_train)
@@
 
-# diremover_rp = diremover.fit_transform(X_test)
+diremover_rp = diremover.fit_transform(X_test)
 # AttributeError: 'DataFrame' object has no attribute 'features' (probem with .fit_transform function)

### 2022-04-13 final_metrics_analysis.Rmd

--- 2022-04-13 final_metrics_analysis.Rmd
@@
 # age distribution by race
-ax = sns.kdeplot(x="age", hue="race", data=df_viz);
-kdeline = ax.lines[0]
-mean_fem = df_viz.groupby('race').age.median()[0]
-xs = kdeline.get_xdata()
-ys = kdeline.get_ydata()
-height_fem = np.interp(mean_fem, xs, ys)
-ax.vlines(mean_fem, 0, height_fem, color=purple, ls=':')
+#ax = sns.kdeplot(x="age", hue="race", data=df_viz);
+#kdeline = ax.lines[0]
+#mean_fem = df_viz.groupby('race').age.median()[0]
+#xs = kdeline.get_xdata()
+#ys = kdeline.get_ydata()
+#height_fem = np.interp(mean_fem, xs, ys)
+#ax.vlines(mean_fem, 0, height_fem, color=purple, ls=':')
 
-mean_m = df_viz.groupby('race').age.median()[1]
-height_m = np.interp(mean_m, xs, ys)
-ax.vlines(mean_m, 0, height_m, color=green, ls=':')
+#mean_m = df_viz.groupby('race').age.median()[1]
+#height_m = np.interp(mean_m, xs, ys)
+#ax.vlines(mean_m, 0, height_m, color=green, ls=':')
 
-print(df_viz.groupby('race').age.median())
+#print(df_viz.groupby('race').age.median())
 
-plt.title('Distribution of Age by Race')
+#plt.title('Distribution of Age by Race')
 ```

### 2022-04-13 Paper_Draft2/Draft2.tex
draft 2

--- 2022-03-08 Paper_Draft1/Draft1.tex
@@
 % Full title of the paper (Capitalized)
-\Title{Does the COMPAS needle always point towards equity? Finding
-Fairness in the COMPAS Risk Assessment Algorithm}
+\Title{Does the COMPAS Needle Always Point Towards Equity? Finding
+Fairness in the COMPAS Risk Assessment Algorithm':' A Case Study}
 
@@
 % Contact information of the corresponding author
-\corres{Correspondence: \href{mailto:leutnant@fh-muenster.de}{\nolinkurl{leutnant@fh-muenster.de}};
-Tel.: +XX-000-00-0000.}
+\corres{Correspondence: }
 
@@
 help humans make data-driven decisions. Northpointe, a software company,
-created a risk assessment instrument known as the Correctional Offender
-Management Profiling for Alternative Sanctions (COMPAS). COMPAS uses
-various behavioral and psychological metrics related to recidivism to
-assist justice systems in assessing a defendant's potential recidivism
-risk. ProPublica published an article which concludes that the biases in
-the criminal justice system are reflected in the COMPAS software. Using
-a human rights framework adopted from the organization Women at the
-Table, we use various debiasing algorithms to analyze both sides of the
-argument between Northpointe and ProPublica and determine the level and
-extent of racial bias in the COMPAS algorithm.}
+created an algorithmic risk assessment instrument known as the
+Correctional Offender Management Profiling for Alternative Sanctions
+(COMPAS). COMPAS uses various behavioral and psychological metrics
+related to recidivism to assist justice systems in assessing a
+defendant's potential recidivism risk. Angwin et al.~published a
+ProPublica article in which they conclude that the racial biases in the
+criminal justice system are reflected in the COMPAS recidivism risk
+scores. In response, Dieterich et al.~published a rebuttal on behalf of
+Northpointe defending the COMPAS algorithm and refuting Angwin et al.'s
+allegation of racial bias. Using a human rights framework adopted from
+the organizations Women at the Table and AI Fairness 360, we use
+debiasing algorithms and fairness metrics to analyze the argument
+between Northpointe and ProPublica and determine whether and to what
+extent there is racial bias in the COMPAS algorithm.}
 
@@
 
-% Pandoc syntax highlighting
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{248,248,248}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{#1}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-
 % tightlist command for lists without linebreak
@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-
-% From pandoc table feature
-\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
 
@@
 
-\hypertarget{version}{%
-\section{Version}\label{version}}
-
-This Rmd-skeleton uses the mdpi Latex template published 2019/02.
-However, the official template gets more frequently updated than the
-`rticles' package. Therefore, please make sure prior to paper
-submission, that you're using the most recent .cls, .tex and .bst files
-(available \href{http://www.mdpi.com/authors/latex}{here}).
-
 \hypertarget{introduction}{%
@@
 
-Women @ the Table, the sponsor organization for this project, is ``a
-growing, global gender equality \& democracy CSO based in Geneva,
-Switzerland focused on advancing feminist systems change by using the
-prism of technology, innovation \& AI exercising leverage points in
-technology, the economy, sustainability \& democratic governance''
-\citep{noauthor_womenthetable_nodate}. We have been asked to collaborate
-on their AI \& Equality \citep{noauthor_ai_nodate} initiative, tasked
-with debiasing the COMPAS algorithm
-\citep{noauthor_aif360datasetscompasdataset_2018} and producing a
-corresponding data story that will be added to their library.
+The Correctional Offender Management Profiling for Alternative Sanctions
+(COMPAS) algorithm was created by the private, for-profit company
+Northpointe (now known by its parent company
+\href{https://www.equivant.com/faq/}{Equivant}), to predict defendants'
+risk of recidivism. It generates a decile score that classifies
+defendants' risk of recidivism as either low, medium, or high
+\citep{angwin2016machine}. Jurisdictions across the United States use
+the COMPAS risk assessment instrument, including but not limited to the
+New York \citep{NYDOCCS2019}, Massachusetts, Michigan, California
+\citep{Jackson2020Setting}, and Wisconsin \citep{WisconsinDOC}
+Departments of Corrections. Due to the proprietary nature of the COMPAS
+algorithm, it is unknown how exactly these recidivism risk scores are
+calculated; however, a sample COMPAS Risk Assessment Survey has been
+made publicly available, revealing the algorithm's input information.
+This survey has been widely critiqued for using proxy variables for race
+that do not explicitly factor in a defendant's race but heavily imply
+it, allowing Northpointe to claim that their algorithm is free of racial
+bias. For example, the COMPAS risk assessment survey asks screeners to
+speculate if a defendant might be affiliated with a gang. It also asks
+if a defendant has any friends or family members who have been crime
+victims \citep{Angwin2016Sample}. Although these questions do not
+directly ask about race, they do not take into account the pervasive
+nature of systemic racism that infiltrates every aspect of the lives of
+marginalized people, thereby indirectly asking about race. Angwin et
+al.~published a 2016 piece in the news publication ProPublica that
+analyzes the methods and algorithms used by Northpointe in their COMPAS
+risk score assessment algorithm and uncovers the racial biases in
+defendants' scores \citep{angwin2016machine}. Their research finds that
+``the algorithm {[}is{]} somewhat more accurate than a coin flip,'' a
+worrisome level of accuracy given the potential impact its
+determinations may have on real people's lives. Angwin et
+al.~specifically investigate the distribution of COMPAS scores by decile
+among Black and white defendants. They write: ``The analysis also
+{[}shows{]} that even when controlling for prior crimes, future
+recidivism, age, and gender, black defendants {[}are{]} 45 percent more
+likely to be assigned higher risk scores than white defendants''
+\citep{larson2016we}. After examining the fairness metric statistical
+parity difference, Angwin et al.~conclude that the algorithm is racially
+biased \citep{larson2016we}.
 
-The Correctional Offender Management Profiling for Alternative Sanctions
-(COMPAS) algorithm was created by the private for-profit company
-Northpointe, also known by its parent company Equivant
-\citep{equivant_faq_nodate}, to predict defendants' risk of recidivism.
-It generates a score that classifies defendants' risk of recidivism as
-either low, medium, or high \citep{angwin2016machine}. Jurisdictions
-across the United States use the COMPAS risk assessment instrument. In
-2016, ProPublica published a piece that analyzed the methods and
-algorithms used by Northpointe to uncover racial biases in defendants'
-scores \citep{angwin2016machine}. Their analysis looks at the
-distribution of decile COMPAS scores among Black and white defendants.
-ProPublica concludes, after using a statistical parity metric on false
-positives or false negatives, that the algorithm is racially biased
-\citep{larson2016we}. Northpointe denies the allegations of racial bias,
-conducting their own analyses based on different statistical parity
-metrics \citep{equivant_response_2018}. Since then, ProPublica has
-responded, maintaining their original argument and Northpointe has yet
-to respond (though in 2017, they adopted the name Equivant). Currently,
-the two parties' disagreement focuses on pretrial Risk Assessment
-Instruments (RAIs). ProPublica maintains that there are biases in the
-outcome values, protected attributes, and covariates during
-Northpointe's data processing phase. ProPublica accounts for these
-biases in their analyses, and Northepointe's response, highlights how
-ProPublica did not account for base rates of recidivism in their
+Dieterich et al., on the behalf of Northpointe, deny the allegations of
+racial bias and offer their own analyses based on different fairness
+metrics in rebuttal \citep{equivant_response_2016}. Angwin et
+al.~maintain that there are biases in the outcome values, protected
+attributes, and covariates during Dieterich et al.'s data processing
+phase. ProPublica collaborators Larson et al.~account for these biases
+in their analyses. In their response, Dieterich et al.~highlight that
+Angwin et al.~did not account for base rates of recidivism in their
 analysis, which are important to understand initial percentages without
@@
 
-Our project builds on Women at the Table's various debiasing algorithms
-to conduct our own analyses on the COMPAS dataset. Based on this
-analysis, we employ a human rights framework to contribute to the
-ongoing ProPublica and Northpointe debate and investigate whether or to
-what extent there is racial bias in the COMPAS algorithm. With a solid
-understanding of the two sides, we aim to pinpoint the shortcomings of
-both arguments and correct them in our analyses. We will use various
-machine learning algorithms including logistic regression, cross
-validation, and lasso and ridge techniques to choose models. We will
-summarize our results using the JupyterNotebook framework from Women at
-the Table, to be used by members of the organization to teach in a
-workshop setting. We hope that our findings will highlight the
-importance of checking statistical analyses using varied methods and
-contribute to the ongoing discussion of the effects of machine biases in
-the justice system.
+\href{https://www.womenatthetable.net/}{Women at the Table}, the sponsor
+organization for this project, is ``a growing, global gender equality \&
+democracy CSO based in Geneva, Switzerland focused on advancing feminist
+systems change by using the prism of technology, innovation \& AI
+exercising leverage points in technology, the economy, sustainability \&
+democratic governance.'' We are collaborating with the organization on
+its AI \& Equality \citep{noauthor_ai_nodate} initiative, tasked with
+debiasing the COMPAS algorithm \citep{aif360-oct-2018} and producing a
+corresponding data story that will be added to its library. Our project
+builds on Women at the Table's various debiasing algorithms used in its
+AI \& Equality Human Rights Toolbox to conduct our own analyses on the
+COMPAS dataset. Based on this analysis, we employ a human rights
+framework to contribute to the ProPublica and Northpointe debate and
+investigate whether or to what extent there is racial bias in the COMPAS
+algorithm. With a solid understanding of the two sides, we aim to
+pinpoint the shortcomings of both arguments and correct them in our
+analyses. We will use various debiasing techniques and fairness metrics
+to evaluate the level of bias present in the COMPAS data and our
+algorithm. We will summarize our results using the JupyterNotebook
+framework from Women at the Table, to be used by members of the
+organization to teach in a workshop setting. We hope that our findings
+will highlight the importance of checking statistical analyses using
+varied methods and contribute to the ongoing discussion of the effects
+of machine biases in the justice system.
 
@@
 The data we are using for this project is the COMPAS General Recidivism
-Risk Scores dataset from ProPublica used in their original analysis
-\citep{larson_propublicacompas-analysiscompas-scores-two-yearscsv_2022}.
-The COMPAS dataset we are using is from AI Fairness 360 (AIF360) toolkit
-\citep{noauthor_trusted-ai_360} which does the same initial
-preprocessing as ProPublica. The raw data has 6,167 rows and each row
-represents an arrest charge for a defendant. ProPublica's COMPAS data
-includes the defendant's age, race, sex, what they were charged with,
-and whether or not the defendant ultimately recidivated within a
-two-year period after their arrest. For the purposes of our project,
-which endeavors to evaluate the differing effects of the COMPAS
-algorithm on white defendants and Black defendants, we have filtered the
-data to only include individuals whose race is listed as Caucasian or
+Risk Scores dataset from the AI Fairness 360 (AIF360) toolkit
+\citep{aif360-oct-2018}, which does the same initial preprocessing as
+ProPublica. The raw data has 6,167 rows and each row represents an
+arrest charge for a defendant. AIF360's COMPAS data includes the
+defendant's age, race, sex, what they were charged with, and whether or
+not the defendant ultimately recidivated within a two-year period after
+their arrest. For the purposes of our project, which endeavors to
+evaluate the differing effects of the COMPAS algorithm on white
+defendants versus Black defendants, we have filtered the data to only
+include individuals whose race is listed as Caucasian or
 African-American. Our data therefore has 5,723 rows (Figure
-\ref{fig:table snip}), with the below distributions of race (Figure
-\ref{fig:race plot}), age (Figure \ref{fig:age plot}), and two year
-recidivism rate (Figures \ref{fig:recid plot} \&
-\ref{fig:recid race plot}).
+\ref{fig:table snip}), with the below distributions of race, age (Figure
+\ref{fig:age plot}), and two year recidivism rate (Figures
+\ref{fig:recid plot} \& \ref{fig:recid race plot}).
 
@@
 
-\hypertarget{results}{%
-\section{Results}\label{results}}
+\hypertarget{methods}{%
+\section{Methods}\label{methods}}
 
-This section may be divided by subheadings. It should provide a concise
-and precise description of the experimental results, their
-interpretation as well as the experimental conclusions that can be
-drawn.
+The AIF360 toolkit contains various group and individual fairness
+metrics and pre-processing, in-processing, and post-processing
+algorithms that we used for our experiments \citep{aif360-oct-2018}. We
+researched the definitions and applications of different fairness
+metrics \citep{ashokan2021fairness} to determine which metric would be
+most appropriate for our project. We chose to look at four group
+fairness metrics instead of individual fairness metrics. Group fairness
+takes into account the attributes of a whole group as opposed to just
+one individual in the group, allowing us to represent more of the
+systemic issues happening. In general, group fairness metrics require
+that the unprivileged group is treated similarly to the privileged
+group, whereas individual fairness metrics require individuals to be
+treated consistently \citep{kypraiou_what_2021}. Group and individual
+metrics work in opposition of one another, meaning that when group
+fairness improves, individual fairness gets worse.
 
-\hypertarget{subsection-heading-here}{%
-\subsection{Subsection Heading Here}\label{subsection-heading-here}}
+\emph{Statistical Parity Difference} - This metric measures the
+difference that the majority and protected classes get a particular
+outcome. The ideal value of this metric is 0. Fairness for this metric
+is between -0.1 and 0.1. A negative value means there is higher benefit
+for the privileged group (in this case, white defendants).
 
-Subsection text here.
+\(P(\hat{Y}=1|D=Unprivileged) - P(\hat{Y}=1|D=Privileged)\)
 
-\hypertarget{subsubsection-heading-here}{%
-\subsubsection{Subsubsection Heading
-Here}\label{subsubsection-heading-here}}
+\emph{Disparate Impact Ratio} - This metric is the ratio of how often
+the favorable outcome occurs in one group versus the other. In the case
+of recidivism, this is the ratio of how many white defendants are
+predicted to not recidivate compared to how many black defendants are
+predicted to not recidivate. A value of 1 means that the ratio is
+exactly 1:1. Less than 1 means the privileged group (white defendants)
+benefits, while a value greater than 1 means the unprivileged group
+(Black defendants) benefits. According to AIF360, a ratio between .8 to
+1.25 is considered fair:
 
-Bulleted lists look like this:
+\(\frac{P(\hat{Y}=1|D=Unprivileged)}{P(\hat{Y}=1|D=Privileged)}\)
 
-\begin{itemize}
-\tightlist
-\item
-  First bullet
-\item
-  Second bullet
-\item
-  Third bullet
-\end{itemize}
+\emph{Average Odds Difference} - This metric returns the average
+difference in false positive rate and true positive rate for the
+privileged and unprivileged groups. A value of 0 indicates equality of
+odds, and a value below 0 implies benefit for the privileged group.
+Equality of odds is achieved in the case of recidivism when the
+proportion of people who were predicted to recidivate and did recidivate
+is equal (true positive rate) for both Black and white defendants AND
+the proportion of people who were predicted to recidivate and did not
+recidivate (false positive rate) is equal for both Black and white
+defendants.
 
-Numbered lists can be added as follows:
+\(\frac{(FPR_{D = unprivileged} - FPR_{D = privileged}) + (TPR_{D = unprivileged} - TPR_{D = privileged})}{2}\)
 
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  First item
-\item
-  Second item
-\item
-  Third item
-\end{enumerate}
+\emph{Equal Opportunity Difference} - This metric is computed as the
+difference of true positive rates between the unprivileged and the
+privileged groups. The true positive rate is the ratio of true positives
+to the total number of actual positives for a given group.
 
-The text continues here.
+The ideal value is 0. A value less than 0 implies higher benefit for the
+privileged group and a value greater than 0 implies higher benefit for
+the unprivileged group. Fairness for this metric is between -0.1 and
+0.1. This metric is best used when it is very important to catch
+positive outcomes while false positives are not exceptionally
+problematic. This is not the case for the COMPAS data set, as false
+positives mean extra jail time for someone who will not actually
+reoffend.
 
-All figures and tables should be cited in the main text as Figure
-\ref{fig:mdpi-logo}, Table 1, etc.
+\(TPR_{D = Unprivileged} - TPR_{D = Privileged}\)
 
-\begin{figure}
+For the next step of our experiment, we needed to determine where in the
+data science pipeline we can mitigate the most bias, using
+pre-processing, in-processing, and post-processing debiasing algorithms.
+These are all based on using predictive models to figure out how we can
+``fix'' the bias that is present. Pre-processing refers to assessing the
+training data, and it is the most flexible method because it has not yet
+trained a model that may carry assumptions about the data. It is
+important to keep in mind that pre-processing prevents assumptions in
+the modeling, but does not account for the bias in data collection. With
+pre-processing, we use the reweighting algorithm from AIF360 which
+weights the examples in each (group, label) combination differently to
+ensure fairness before classification \citep{aif360-oct-2018}. We used a
+logistic regression model for this algorithm, as it is the easiest to
+interpret in the given context. After running the fairness metrics using
+the pre-processing algorithm, we were able to compare our results to the
+baseline metrics from the previous section.
 
-{\centering \includegraphics{logo-mdpi}
+Another approach, in-processing, explores how bias can be targeted while
+building a model. It assesses how we build learning algorithms, which
+often prioritize accuracy over fairness. The technique we use is the
+Meta Classification algorithm, which accounts for the fairness metric as
+part of the input and returns a classifier optimized by that particular
+metric. In order to do this, we first needed to convert our data frame
+into a data type called a BinaryLabelDataset. Similarly to
+pre-processing, we compared the results of our in-processing methods
+with both the baseline and the pre-processing to gauge which method so
+far has better individual or group fairness.
 
-}
+Our last approach, post-processing, makes transformations on model
+outputs. Similarly to the pre-processing, it is flexible because it does
+not work directly with the algorithm, but instead has direct access to
+all the predicted classification values. Because the algorithms are
+often so complex and uninterpretable, post-processors are trained in a
+black box. To get around this, we use the PostProcessingMeta class from
+AIF360 which combines the training and prediction for a random estimator
+and the post-processor while simultaneously splitting the dataset. Based
+on these results, we looked comparatively at all the approaches to
+determine which we assessed to be the most fair.
 
-\caption{This is a figure, Schemes follow the same formatting.
-If there are multiple panels, they should be listed as: (\textbf{a}) Description of what is contained in the first panel.
-(\textbf{b}) Description of what is contained in the second panel.
-Figures should be placed in the main text near to the first time they are cited.
-A caption on a single line should be centered.}\label{fig:mdpi-logo}
-\end{figure}
+To reiterate, fairness is subjective, so the questions we ask ourselves
+to create parameters for our definition of fairness include:
 
-Please see Table \ref{tab:mytable}.
+\hypertarget{preliminary-results}{%
+\section{Preliminary Results}\label{preliminary-results}}
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{x }\OtherTok{\textless{}{-}}\NormalTok{ tibble}\SpecialCharTok{::}\FunctionTok{tribble}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\StringTok{\textasciigrave{}}\AttributeTok{Title 1}\StringTok{\textasciigrave{}}\NormalTok{, }\SpecialCharTok{\textasciitilde{}}\StringTok{\textasciigrave{}}\AttributeTok{Title 2}\StringTok{\textasciigrave{}}\NormalTok{, }\SpecialCharTok{\textasciitilde{}}\StringTok{\textasciigrave{}}\AttributeTok{Title 3}\StringTok{\textasciigrave{}}\NormalTok{,}
-\StringTok{"entry 1"}\NormalTok{, }\StringTok{"data"}\NormalTok{, }\StringTok{"data"}\NormalTok{,}
-\StringTok{"entry 2"}\NormalTok{, }\StringTok{"data"}\NormalTok{, }\StringTok{"data"}
-\NormalTok{)}
+The statistical parity difference is -0.14. This indicates that there is
+a large difference between white and Black defendants regarding whether
+or not they recidivate. The algorithm unfairly benefits white defendants
+over Black defendants.
 
-\NormalTok{knitr}\SpecialCharTok{::}\FunctionTok{kable}\NormalTok{(x, }\AttributeTok{caption =} \StringTok{"This is a table caption.
-Tables should be placed in the main text near to the first time they are cited.}\SpecialCharTok{\textbackslash{}\textbackslash{}}\StringTok{label\{tab:mytable\}"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
+Disparate impact ratio is 0.47. The ratio of white defendants predicted
+to not recidivate to the Black defendants predicted to not recidivate is
+0.47. A ratio between 0.8 and 1.25 is considered fair, therefore the
+algorithm unfairly benefits white defendants.
 
-\begin{longtable}[]{@{}lll@{}}
-\caption{This is a table caption. Tables should be placed in the main
-text near to the first time they are
-cited.\label{tab:mytable}}\tabularnewline
-\toprule
-Title 1 & Title 2 & Title 3\tabularnewline
-\midrule
-\endfirsthead
-\toprule
-Title 1 & Title 2 & Title 3\tabularnewline
-\midrule
-\endhead
-entry 1 & data & data\tabularnewline
-entry 2 & data & data\tabularnewline
-\bottomrule
-\end{longtable}
+Average odds difference is -0.44. The average difference in false
+positive rates and true positive rates for white and Black defendants is
+-0.44. Values less than zero are considered in favor of the privileged
+group, so the algorithm unfairly benefits white defendants.
 
-This is an example of an equation:
+Equal opportunity difference is -0.41. The difference of true positive
+rates between the Black and white groups is -0.41. A value less than 0
+indicates a benefit to the privileged group, so the algorithm unfairly
+benefits white defendants. The value is substantially less than -0.1,
+which indicates that the algorithm benefits white defendants.
 
-\begin{equation}
-\mathbb{S}
-\end{equation}
-
-Example of a theorem:
-
-\begin{Theorem}
-Example text of a theorem.
-\end{Theorem}
-
-The text continues here. Proofs must be formatted as follows:
-
-Example of a proof:
-
-\begin{proof}[Proof of Theorem 1]
-Text of the proof. Note that the phrase `of Theorem 1' is optional if it is clear which theorem is being referred to.
-\end{proof}
-
-The text continues here.
+All four group fairness metrics determine that the COMPAS algorithm
+favors white defendants over Black defendants. The statistical parity
+metric indicates unfairness by a smaller margin of 0.04 than equal
+opportunity difference which has a margin of error 0.4 below the
+threshold. Although this is true, each algorithm uses different
+penalties for unfavorable outcomes, so the higher the penalty the more
+of a difference there will be between the fairness threshold and the
+value.
 
@@
 
-Authors should discuss the results and how they can be interpreted in
-perspective of previous studies and of the working hypotheses. The
-findings and their implications should be discussed in the broadest
-context possible. Future research directions may also be highlighted.
-
 \hypertarget{conclusion}{%
 \section{Conclusion}\label{conclusion}}
-
-This section is not mandatory, but can be added to the manuscript if the
-discussion is unusually long or complex.
 

### 2022-04-13 Paper_Draft2/Draft2.Rmd

--- 2022-04-05 Paper_Draft1/Draft2.Rmd
@@
 ---
-title: It’s COMPAS-licated -- Can We Find Fairness in the COMPAS Risk Assessment Algorithm?
+title: Does the COMPAS Needle Always Point Towards Equity? Finding Fairness in the COMPAS Risk Assessment Algorithm':' A Case Study
+
 author:
@@
     email: enesmith@smith.edu
-
-correspondence: |
-  leutnant@fh-muenster.de; Tel.: +XX-000-00-0000.
 journal: water
@@
  A variety of disciplines use risk assessment instruments to help humans make data-driven decisions.
-Northpointe, a software company, created a risk assessment instrument known as the Correctional Offender Management Profiling for Alternative Sanctions (COMPAS).
+Northpointe, a software company, created an algorithmic risk assessment instrument known as the Correctional Offender Management Profiling for Alternative Sanctions (COMPAS).
 COMPAS uses various behavioral and psychological metrics related to recidivism to assist justice systems in assessing a defendant’s potential recidivism risk.
-ProPublica published an article which concludes that the biases in the criminal justice system are reflected in the COMPAS software.
-Using a human rights framework adopted from the organization Women at the Table, we use various debiasing algorithms to analyze both sides of the argument between Northpointe and ProPublica and determine the level and extent of racial bias in the COMPAS algorithm.
+Angwin et al.
+published a ProPublica article in which they conclude that the racial biases in the criminal justice system are reflected in the COMPAS recidivism risk scores.
+In response, Dieterich et al.
+published a rebuttal on behalf of Northpointe defending the COMPAS algorithm and refuting Angwin et al.’s allegation of racial bias.
+Using a human rights framework adopted from the organizations Women at the Table and AI Fairness 360, we use debiasing algorithms and fairness metrics to analyze the argument between Northpointe and ProPublica and determine whether and to what extent there is racial bias in the COMPAS algorithm.
+
 output: rticles::mdpi_article
@@
 
-Women @ the Table, the sponsor organization for this project, is “a growing, global gender equality & democracy CSO based in Geneva, Switzerland focused on advancing feminist systems change by using the prism of technology, innovation & AI exercising leverage points in technology, the economy, sustainability & democratic governance” [@noauthor_womenthetable_nodate].
-We have been asked to collaborate on their AI & Equality [@noauthor_ai_nodate] initiative, tasked with debiasing the COMPAS algorithm [@noauthor_aif360datasetscompasdataset_2018] and producing a corresponding data story that will be added to their library.
+The Correctional Offender Management Profiling for Alternative Sanctions (COMPAS) algorithm was created by the private, for-profit company Northpointe (now known by its parent company [Equivant](https://www.equivant.com/faq/)), to predict defendants’ risk of recidivism.
+It generates a decile score that classifies defendants’ risk of recidivism as either low, medium, or high [@angwin2016machine].
+Jurisdictions across the United States use the COMPAS risk assessment instrument, including but not limited to the New York [@NYDOCCS2019], Massachusetts, Michigan, California [@Jackson2020Setting], and Wisconsin [@WisconsinDOC] Departments of Corrections.
+Due to the proprietary nature of the COMPAS algorithm, it is unknown how exactly these recidivism risk scores are calculated; however, a sample COMPAS Risk Assessment Survey has been made publicly available, revealing the algorithm’s input information.
+This survey has been widely critiqued for using proxy variables for race that do not explicitly factor in a defendant’s race but heavily imply it, allowing Northpointe to claim that their algorithm is free of racial bias.
+For example, the COMPAS risk assessment survey asks screeners to speculate if a defendant might be affiliated with a gang.
+It also asks if a defendant has any friends or family members who have been crime victims [@Angwin2016Sample].
+Although these questions do not directly ask about race, they do not take into account the pervasive nature of systemic racism that infiltrates every aspect of the lives of marginalized people, thereby indirectly asking about race.
+Angwin et al.
+published a 2016 piece in the news publication ProPublica that analyzes the methods and algorithms used by Northpointe in their COMPAS risk score assessment algorithm and uncovers the racial biases in defendants’ scores [@angwin2016machine].
+Their research finds that “the algorithm [is] somewhat more accurate than a coin flip,” a worrisome level of accuracy given the potential impact its determinations may have on real people’s lives.
+Angwin et al.
+specifically investigate the distribution of COMPAS scores by decile among Black and white defendants.
+They write: “The analysis also [shows] that even when controlling for prior crimes, future recidivism, age, and gender, black defendants [are] 45 percent more likely to be assigned higher risk scores than white defendants” [@larson2016we].
+After examining the fairness metric statistical parity difference, Angwin et al.
+conclude that the algorithm is racially biased [@larson2016we].
 
-The Correctional Offender Management Profiling for Alternative Sanctions (COMPAS) algorithm was created by the private for-profit company Northpointe, also known by its parent company Equivant [@equivant_faq_nodate], to predict defendants’ risk of recidivism.
-It generates a score that classifies defendants’ risk of recidivism as either low, medium, or high [@angwin2016machine].
-Jurisdictions across the United States use the COMPAS risk assessment instrument.
-In 2016, ProPublica published a piece that analyzed the methods and algorithms used by Northpointe to uncover racial biases in defendants’ scores [@angwin2016machine].
-Their analysis looks at the distribution of decile COMPAS scores among Black and white defendants.
-ProPublica concludes, after using a statistical parity metric on false positives or false negatives, that the algorithm is racially biased [@larson2016we].
-Northpointe denies the allegations of racial bias, conducting their own analyses based on different statistical parity metrics [@equivant_response_2018].
-Since then, the two parties have had several exchanges, maintaining their original arguments.
-Currently, the two parties’ disagreement focuses on pretrial Risk Assessment Instruments (RAIs).
-ProPublica maintains that there are biases in the outcome values, protected attributes, and covariates during Northpointe’s data processing phase.
-ProPublica accounts for these biases in their analyses, and Northepointe’s response, highlights how ProPublica did not account for base rates of recidivism in their analysis, which are important to understand initial percentages without the presence of other information.
+Dieterich et al., on the behalf of Northpointe, deny the allegations of racial bias and offer their own analyses based on different fairness metrics in rebuttal [@equivant_response_2016].
+Angwin et al.
+maintain that there are biases in the outcome values, protected attributes, and covariates during Dieterich et al.’s data processing phase.
+ProPublica collaborators Larson et al.
+account for these biases in their analyses.
+In their response, Dieterich et al.
+highlight that Angwin et al.
+did not account for base rates of recidivism in their analysis, which are important to understand initial percentages without the presence of other information.
 
-Our project builds on Women at the Table’s various debiasing algorithms to conduct our own analyses on the COMPAS dataset.
-Based on this analysis, we employ a human rights framework to contribute to the ongoing ProPublica and Northpointe debate and investigate whether or to what extent there is racial bias in the COMPAS algorithm.
+[Women at the Table](https://www.womenatthetable.net/), the sponsor organization for this project, is “a growing, global gender equality & democracy CSO based in Geneva, Switzerland focused on advancing feminist systems change by using the prism of technology, innovation & AI exercising leverage points in technology, the economy, sustainability & democratic governance.” We are collaborating with the organization on its AI & Equality [@noauthor_ai_nodate] initiative, tasked with debiasing the COMPAS algorithm [@aif360-oct-2018] and producing a corresponding data story that will be added to its library.
+Our project builds on Women at the Table’s various debiasing algorithms used in its AI & Equality Human Rights Toolbox to conduct our own analyses on the COMPAS dataset.
+Based on this analysis, we employ a human rights framework to contribute to the ProPublica and Northpointe debate and investigate whether or to what extent there is racial bias in the COMPAS algorithm.
 With a solid understanding of the two sides, we aim to pinpoint the shortcomings of both arguments and correct them in our analyses.
-We will use various machine learning algorithms including logistic regression, cross validation, and lasso and ridge techniques to choose models.
+We will use various debiasing techniques and fairness metrics to evaluate the level of bias present in the COMPAS data and our algorithm.
 We will summarize our results using the JupyterNotebook framework from Women at the Table, to be used by members of the organization to teach in a workshop setting.
@@
 
-The data we are using for this project is the COMPAS General Recidivism Risk Scores dataset from ProPublica used in their original analysis [@larson_propublicacompas-analysiscompas-scores-two-yearscsv_2022].
-The COMPAS dataset we are using is from AI Fairness 360 (AIF360) toolkit [@noauthor_trusted-ai_360] which does the same initial preprocessing as ProPublica.
+The data we are using for this project is the COMPAS General Recidivism Risk Scores dataset from the AI Fairness 360 (AIF360) toolkit [@aif360-oct-2018], which does the same initial preprocessing as ProPublica.
 The raw data has 6,167 rows and each row represents an arrest charge for a defendant.
-ProPublica’s COMPAS data includes the defendant's age, race, sex, what they were charged with, and whether or not the defendant ultimately recidivated within a two-year period after their arrest.
-For the purposes of our project, which endeavors to evaluate the differing effects of the COMPAS algorithm on white defendants and Black defendants, we have filtered the data to only include individuals whose race is listed as Caucasian or African-American.
-Our data therefore has 5,723 rows (Figure \ref{fig:table snip}), with the below distributions of race (Figure \ref{fig:race plot}), age (Figure \ref{fig:age plot}), and two year recidivism rate (Figures \ref{fig:recid plot} & \ref{fig:recid race plot}).
+AIF360’s COMPAS data includes the defendant's age, race, sex, what they were charged with, and whether or not the defendant ultimately recidivated within a two-year period after their arrest.
+For the purposes of our project, which endeavors to evaluate the differing effects of the COMPAS algorithm on white defendants versus Black defendants, we have filtered the data to only include individuals whose race is listed as Caucasian or African-American.
+Our data therefore has 5,723 rows (Figure \ref{fig:table snip}), with the below distributions of race, age (Figure \ref{fig:age plot}), and two year recidivism rate (Figures \ref{fig:recid plot} & \ref{fig:recid race plot}).
 
@@
 
+# Methods
+
+The AIF360 toolkit contains various group and individual fairness metrics and pre-processing, in-processing, and post-processing algorithms that we used for our experiments [@aif360-oct-2018].
+We researched the definitions and applications of different fairness metrics [@ashokan2021fairness] to determine which metric would be most appropriate for our project.
+We chose to look at four group fairness metrics instead of individual fairness metrics.
+Group fairness takes into account the attributes of a whole group as opposed to just one individual in the group, allowing us to represent more of the systemic issues happening.
+In general, group fairness metrics require that the unprivileged group is treated similarly to the privileged group, whereas individual fairness metrics require individuals to be treated consistently [@kypraiou_what_2021].
+Group and individual metrics work in opposition of one another, meaning that when group fairness improves, individual fairness gets worse.
+
+*Statistical Parity Difference* - This metric measures the difference that the majority and protected classes get a particular outcome.
+The ideal value of this metric is 0.
+Fairness for this metric is between -0.1 and 0.1.
+A negative value means there is higher benefit for the privileged group (in this case, white defendants).
+
+$P(\hat{Y}=1|D=Unprivileged) - P(\hat{Y}=1|D=Privileged)$
+
+*Disparate Impact Ratio* - This metric is the ratio of how often the favorable outcome occurs in one group versus the other.
+In the case of recidivism, this is the ratio of how many white defendants are predicted to not recidivate compared to how many black defendants are predicted to not recidivate.
+A value of 1 means that the ratio is exactly 1:1.
+Less than 1 means the privileged group (white defendants) benefits, while a value greater than 1 means the unprivileged group (Black defendants) benefits.
+According to AIF360, a ratio between .8 to 1.25 is considered fair:
+
+$\frac{P(\hat{Y}=1|D=Unprivileged)}{P(\hat{Y}=1|D=Privileged)}$
+
+*Average Odds Difference* - This metric returns the average difference in false positive rate and true positive rate for the privileged and unprivileged groups.
+A value of 0 indicates equality of odds, and a value below 0 implies benefit for the privileged group.
+Equality of odds is achieved in the case of recidivism when the proportion of people who were predicted to recidivate and did recidivate is equal (true positive rate) for both Black and white defendants AND the proportion of people who were predicted to recidivate and did not recidivate (false positive rate) is equal for both Black and white defendants.
+
+$\frac{(FPR_{D = unprivileged} - FPR_{D = privileged}) + (TPR_{D = unprivileged} - TPR_{D = privileged})}{2}$
+
+*Equal Opportunity Difference* - This metric is computed as the difference of true positive rates between the unprivileged and the privileged groups.
+The true positive rate is the ratio of true positives to the total number of actual positives for a given group.
+
+The ideal value is 0. A value less than 0 implies higher benefit for the privileged group and a value greater than 0 implies higher benefit for the unprivileged group.
+Fairness for this metric is between -0.1 and 0.1.
+This metric is best used when it is very important to catch positive outcomes while false positives are not exceptionally problematic.
+This is not the case for the COMPAS data set, as false positives mean extra jail time for someone who will not actually reoffend.
+
+$TPR_{D = Unprivileged} - TPR_{D = Privileged}$
+
+For the next step of our experiment, we needed to determine where in the data science pipeline we can mitigate the most bias, using pre-processing, in-processing, and post-processing debiasing algorithms.
+These are all based on using predictive models to figure out how we can “fix” the bias that is present.
+Pre-processing refers to assessing the training data, and it is the most flexible method because it has not yet trained a model that may carry assumptions about the data.
+It is important to keep in mind that pre-processing prevents assumptions in the modeling, but does not account for the bias in data collection.
+With pre-processing, we use the reweighting algorithm from AIF360 which weights the examples in each (group, label) combination differently to ensure fairness before classification [@aif360-oct-2018].
+We used a logistic regression model for this algorithm, as it is the easiest to interpret in the given context.
+After running the fairness metrics using the pre-processing algorithm, we were able to compare our results to the baseline metrics from the previous section.
+
+Another approach, in-processing, explores how bias can be targeted while building a model.
+It assesses how we build learning algorithms, which often prioritize accuracy over fairness.
+The technique we use is the Meta Classification algorithm, which accounts for the fairness metric as part of the input and returns a classifier optimized by that particular metric.
+In order to do this, we first needed to convert our data frame into a data type called a BinaryLabelDataset.
+Similarly to pre-processing, we compared the results of our in-processing methods with both the baseline and the pre-processing to gauge which method so far has better individual or group fairness.
+
+Our last approach, post-processing, makes transformations on model outputs.
+Similarly to the pre-processing, it is flexible because it does not work directly with the algorithm, but instead has direct access to all the predicted classification values.
+Because the algorithms are often so complex and uninterpretable, post-processors are trained in a black box.
+To get around this, we use the PostProcessingMeta class from AIF360 which combines the training and prediction for a random estimator and the post-processor while simultaneously splitting the dataset.
+Based on these results, we looked comparatively at all the approaches to determine which we assessed to be the most fair.
+
+To reiterate, fairness is subjective, so the questions we ask ourselves to create parameters for our definition of fairness include:
+
+# Preliminary Results
+
+The statistical parity difference is -0.14.
+This indicates that there is a large difference between white and Black defendants regarding whether or not they recidivate.
+The algorithm unfairly benefits white defendants over Black defendants.
+
+Disparate impact ratio is 0.47.
+The ratio of white defendants predicted to not recidivate to the Black defendants predicted to not recidivate is 0.47.
+A ratio between 0.8 and 1.25 is considered fair, therefore the algorithm unfairly benefits white defendants.
+
+Average odds difference is -0.44.
+The average difference in false positive rates and true positive rates for white and Black defendants is -0.44.
+Values less than zero are considered in favor of the privileged group, so the algorithm unfairly benefits white defendants.
+
+Equal opportunity difference is -0.41.
+The difference of true positive rates between the Black and white groups is -0.41.
+A value less than 0 indicates a benefit to the privileged group, so the algorithm unfairly benefits white defendants.
+The value is substantially less than -0.1, which indicates that the algorithm benefits white defendants.
+
+All four group fairness metrics determine that the COMPAS algorithm favors white defendants over Black defendants.
+The statistical parity metric indicates unfairness by a smaller margin of 0.04 than equal opportunity difference which has a margin of error 0.4 below the threshold.
+Although this is true, each algorithm uses different penalties for unfavorable outcomes, so the higher the penalty the more of a difference there will be between the fairness threshold and the value.
+
+# Discussion
+
+# Conclusion
+
 # Bibliography

### 2022-04-13 Paper_Draft2/Draft2.tex
\\textbf{}

@@
 its AI \& Equality \citep{noauthor_ai_nodate} initiative, tasked with
-debiasing the COMPAS algorithm \citep{aif360-oct-2018} and producing a
+de-biasing the COMPAS algorithm \citep{aif360-oct-2018} and producing a
 corresponding data story that will be added to its library. Our project
-builds on Women at the Table's various debiasing algorithms used in its
+builds on Women at the Table's various de-biasing algorithms used in its
 AI \& Equality Human Rights Toolbox to conduct our own analyses on the
-COMPAS dataset. Based on this analysis, we employ a human rights
+COMPAS data set. Based on this analysis, we employ a human rights
 framework to contribute to the ProPublica and Northpointe debate and
@@
 pinpoint the shortcomings of both arguments and correct them in our
-analyses. We will use various debiasing techniques and fairness metrics
+analyses. We will use various de-biasing techniques and fairness metrics
 to evaluate the level of bias present in the COMPAS data and our
@@
 The data we are using for this project is the COMPAS General Recidivism
-Risk Scores dataset from the AI Fairness 360 (AIF360) toolkit
-\citep{aif360-oct-2018}, which does the same initial preprocessing as
+Risk Scores data set from the AI Fairness 360 (AIF360) toolkit
+\citep{aif360-oct-2018}, which does the same initial pre-processing as
 ProPublica. The raw data has 6,167 rows and each row represents an
@@
 
-\emph{Statistical Parity Difference} - This metric measures the
-difference that the majority and protected classes get a particular
-outcome. The ideal value of this metric is 0. Fairness for this metric
-is between -0.1 and 0.1. A negative value means there is higher benefit
-for the privileged group (in this case, white defendants).
+\textbackslash textbf\{Statistical Parity Difference\} - This metric
+measures the difference that the majority and protected classes get a
+particular outcome. The ideal value of this metric is 0. Fairness for
+this metric is between -0.1 and 0.1. A negative value means there is
+higher benefit for the privileged group (in this case, white
+defendants).
 
@@
 
-\emph{Disparate Impact Ratio} - This metric is the ratio of how often
-the favorable outcome occurs in one group versus the other. In the case
-of recidivism, this is the ratio of how many white defendants are
-predicted to not recidivate compared to how many black defendants are
-predicted to not recidivate. A value of 1 means that the ratio is
-exactly 1:1. Less than 1 means the privileged group (white defendants)
-benefits, while a value greater than 1 means the unprivileged group
-(Black defendants) benefits. According to AIF360, a ratio between .8 to
-1.25 is considered fair:
+\textbackslash textbf\{Disparate Impact Ratio\} - This metric is the
+ratio of how often the favorable outcome occurs in one group versus the
+other. In the case of recidivism, this is the ratio of how many white
+defendants are predicted to not recidivate compared to how many black
+defendants are predicted to not recidivate. A value of 1 means that the
+ratio is exactly 1:1. Less than 1 means the privileged group (white
+defendants) benefits, while a value greater than 1 means the
+unprivileged group (Black defendants) benefits. According to AIF360, a
+ratio between .8 to 1.25 is considered fair:
 
@@
 
-\emph{Average Odds Difference} - This metric returns the average
-difference in false positive rate and true positive rate for the
-privileged and unprivileged groups. A value of 0 indicates equality of
-odds, and a value below 0 implies benefit for the privileged group.
+\textbackslash textbf\{Average Odds Difference\} - This metric returns
+the average difference in false positive rate and true positive rate for
+the privileged and unprivileged groups. A value of 0 indicates equality
+of odds, and a value below 0 implies benefit for the privileged group.
 Equality of odds is achieved in the case of recidivism when the
@@
 
-\emph{Equal Opportunity Difference} - This metric is computed as the
-difference of true positive rates between the unprivileged and the
-privileged groups. The true positive rate is the ratio of true positives
-to the total number of actual positives for a given group.
+\textbackslash textbf\{Equal Opportunity Difference\} - This metric is
+computed as the difference of true positive rates between the
+unprivileged and the privileged groups. The true positive rate is the
+ratio of true positives to the total number of actual positives for a
+given group.
 
@@
 positives mean extra jail time for someone who will not actually
-reoffend.
+re-offend.
 
@@
 data science pipeline we can mitigate the most bias, using
-pre-processing, in-processing, and post-processing debiasing algorithms.
-These are all based on using predictive models to figure out how we can
-``fix'' the bias that is present. Pre-processing refers to assessing the
-training data, and it is the most flexible method because it has not yet
-trained a model that may carry assumptions about the data. It is
-important to keep in mind that pre-processing prevents assumptions in
-the modeling, but does not account for the bias in data collection. With
-pre-processing, we use the reweighting algorithm from AIF360 which
-weights the examples in each (group, label) combination differently to
-ensure fairness before classification \citep{aif360-oct-2018}. We used a
-logistic regression model for this algorithm, as it is the easiest to
-interpret in the given context. After running the fairness metrics using
-the pre-processing algorithm, we were able to compare our results to the
-baseline metrics from the previous section.
+pre-processing, in-processing, and post-processing de-biasing
+algorithms. These are all based on using predictive models to figure out
+how we can ``fix'' the bias that is present. Pre-processing refers to
+assessing the training data, and it is the most flexible method because
+it has not yet trained a model that may carry assumptions about the
+data. It is important to keep in mind that pre-processing prevents
+assumptions in the modeling, but does not account for the bias in data
+collection. With pre-processing, we use the re-weighing algorithm from
+AIF360 which weights the examples in each (group, label) combination
+differently to ensure fairness before classification
+\citep{aif360-oct-2018}. We used a logistic regression model for this
+algorithm, as it is the easiest to interpret in the given context. After
+running the fairness metrics using the pre-processing algorithm, we were
+able to compare our results to the baseline metrics from the previous
+section.
 
@@
 metric. In order to do this, we first needed to convert our data frame
-into a data type called a BinaryLabelDataset. Similarly to
-pre-processing, we compared the results of our in-processing methods
-with both the baseline and the pre-processing to gauge which method so
-far has better individual or group fairness.
+into a data type called a . Similarly to pre-processing, we compared the
+results of our in-processing methods with both the baseline and the
+pre-processing to gauge which method so far has better individual or
+group fairness.
 
@@
 all the predicted classification values. Because the algorithms are
-often so complex and uninterpretable, post-processors are trained in a
-black box. To get around this, we use the PostProcessingMeta class from
-AIF360 which combines the training and prediction for a random estimator
-and the post-processor while simultaneously splitting the dataset. Based
-on these results, we looked comparatively at all the approaches to
-determine which we assessed to be the most fair.
+often so complex and difficult to interpret, post-processors are trained
+in a black box. To get around this, we use the PostProcessingMeta class
+from AIF360 which combines the training and prediction for a random
+estimator and the post-processor while simultaneously splitting the data
+set. Based on these results, we looked comparatively at all the
+approaches to determine which we assessed to be the most fair.
 

### 2022-04-13 Paper_Draft2/Draft2.Rmd

@@
 
-[Women at the Table](https://www.womenatthetable.net/), the sponsor organization for this project, is “a growing, global gender equality & democracy CSO based in Geneva, Switzerland focused on advancing feminist systems change by using the prism of technology, innovation & AI exercising leverage points in technology, the economy, sustainability & democratic governance.” We are collaborating with the organization on its AI & Equality [@noauthor_ai_nodate] initiative, tasked with debiasing the COMPAS algorithm [@aif360-oct-2018] and producing a corresponding data story that will be added to its library.
-Our project builds on Women at the Table’s various debiasing algorithms used in its AI & Equality Human Rights Toolbox to conduct our own analyses on the COMPAS dataset.
+[Women at the Table](https://www.womenatthetable.net/), the sponsor organization for this project, is “a growing, global gender equality & democracy CSO based in Geneva, Switzerland focused on advancing feminist systems change by using the prism of technology, innovation & AI exercising leverage points in technology, the economy, sustainability & democratic governance.” We are collaborating with the organization on its AI & Equality [@noauthor_ai_nodate] initiative, tasked with de-biasing the COMPAS algorithm [@aif360-oct-2018] and producing a corresponding data story that will be added to its library.
+Our project builds on Women at the Table’s various de-biasing algorithms used in its AI & Equality Human Rights Toolbox to conduct our own analyses on the COMPAS data set.
 Based on this analysis, we employ a human rights framework to contribute to the ProPublica and Northpointe debate and investigate whether or to what extent there is racial bias in the COMPAS algorithm.
 With a solid understanding of the two sides, we aim to pinpoint the shortcomings of both arguments and correct them in our analyses.
-We will use various debiasing techniques and fairness metrics to evaluate the level of bias present in the COMPAS data and our algorithm.
+We will use various de-biasing techniques and fairness metrics to evaluate the level of bias present in the COMPAS data and our algorithm.
 We will summarize our results using the JupyterNotebook framework from Women at the Table, to be used by members of the organization to teach in a workshop setting.
@@
 
-The data we are using for this project is the COMPAS General Recidivism Risk Scores dataset from the AI Fairness 360 (AIF360) toolkit [@aif360-oct-2018], which does the same initial preprocessing as ProPublica.
+The data we are using for this project is the COMPAS General Recidivism Risk Scores data set from the AI Fairness 360 (AIF360) toolkit [@aif360-oct-2018], which does the same initial pre-processing as ProPublica.
 The raw data has 6,167 rows and each row represents an arrest charge for a defendant.
@@
 
-*Statistical Parity Difference* - This metric measures the difference that the majority and protected classes get a particular outcome.
+\\textbf{Statistical Parity Difference} - This metric measures the difference that the majority and protected classes get a particular outcome.
 The ideal value of this metric is 0.
@@
 
-*Disparate Impact Ratio* - This metric is the ratio of how often the favorable outcome occurs in one group versus the other.
+\\textbf{Disparate Impact Ratio} - This metric is the ratio of how often the favorable outcome occurs in one group versus the other.
 In the case of recidivism, this is the ratio of how many white defendants are predicted to not recidivate compared to how many black defendants are predicted to not recidivate.
@@
 
-*Average Odds Difference* - This metric returns the average difference in false positive rate and true positive rate for the privileged and unprivileged groups.
+\\textbf{Average Odds Difference} - This metric returns the average difference in false positive rate and true positive rate for the privileged and unprivileged groups.
 A value of 0 indicates equality of odds, and a value below 0 implies benefit for the privileged group.
@@
 
-*Equal Opportunity Difference* - This metric is computed as the difference of true positive rates between the unprivileged and the privileged groups.
+\\textbf{Equal Opportunity Difference} - This metric is computed as the difference of true positive rates between the unprivileged and the privileged groups.
 The true positive rate is the ratio of true positives to the total number of actual positives for a given group.
@@
 This metric is best used when it is very important to catch positive outcomes while false positives are not exceptionally problematic.
-This is not the case for the COMPAS data set, as false positives mean extra jail time for someone who will not actually reoffend.
+This is not the case for the COMPAS data set, as false positives mean extra jail time for someone who will not actually re-offend.
 
@@
 
-For the next step of our experiment, we needed to determine where in the data science pipeline we can mitigate the most bias, using pre-processing, in-processing, and post-processing debiasing algorithms.
+For the next step of our experiment, we needed to determine where in the data science pipeline we can mitigate the most bias, using pre-processing, in-processing, and post-processing de-biasing algorithms.
 These are all based on using predictive models to figure out how we can “fix” the bias that is present.
@@
 It is important to keep in mind that pre-processing prevents assumptions in the modeling, but does not account for the bias in data collection.
-With pre-processing, we use the reweighting algorithm from AIF360 which weights the examples in each (group, label) combination differently to ensure fairness before classification [@aif360-oct-2018].
+With pre-processing, we use the re-weighing algorithm from AIF360 which weights the examples in each (group, label) combination differently to ensure fairness before classification [@aif360-oct-2018].
 We used a logistic regression model for this algorithm, as it is the easiest to interpret in the given context.
@@
 The technique we use is the Meta Classification algorithm, which accounts for the fairness metric as part of the input and returns a classifier optimized by that particular metric.
-In order to do this, we first needed to convert our data frame into a data type called a BinaryLabelDataset.
+In order to do this, we first needed to convert our data frame into a data type called a .
 Similarly to pre-processing, we compared the results of our in-processing methods with both the baseline and the pre-processing to gauge which method so far has better individual or group fairness.
@@
 Similarly to the pre-processing, it is flexible because it does not work directly with the algorithm, but instead has direct access to all the predicted classification values.
-Because the algorithms are often so complex and uninterpretable, post-processors are trained in a black box.
-To get around this, we use the PostProcessingMeta class from AIF360 which combines the training and prediction for a random estimator and the post-processor while simultaneously splitting the dataset.
+Because the algorithms are often so complex and difficult to interpret, post-processors are trained in a black box.
+To get around this, we use the PostProcessingMeta class from AIF360 which combines the training and prediction for a random estimator and the post-processor while simultaneously splitting the data set.
 Based on these results, we looked comparatively at all the approaches to determine which we assessed to be the most fair.

### 2022-04-14 Paper_Draft2/Draft2.Rmd
trying to fix images

@@
 For the purposes of our project, which endeavors to evaluate the differing effects of the COMPAS algorithm on white defendants versus Black defendants, we have filtered the data to only include individuals whose race is listed as Caucasian or African-American.
-Our data therefore has 5,723 rows (Figure \ref{fig:table snip}), with the below distributions of race, age (Figure \ref{fig:age plot}), and two year recidivism rate (Figures \ref{fig:recid plot} & \ref{fig:recid race plot}).
+Our data therefore has 5,723 rows (Figure \ref{fig:table snip}), with the below distributions of race (Figure \ref{fig:race plot}), age (Figure \ref{fig:age plot}), and two year recidivism rate (Figures \ref{fig:recid plot} & \ref{fig:recid race plot}).
 
-```{r table snip, echo = FALSE, fig.align='center', fig.cap= 'This table is a snippet the data set we will be using.', out.width= "400px", out.height= "200px"}
-knitr::include_graphics("../images/table_snippet.png")
+```{r table snip, echo = FALSE, fig.align='center', fig.cap= 'This table is a snippet of the data set we will be using.', out.width= "500px", out.height= "300px"}
+knitr::include_graphics("../images/table_snippet_new.png")
 ```
 
-```{r race plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of defendant races.', out.width= "400px", out.height= "200px"}
+```{r race plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of defendant races.', out.width= "500px", out.height= "300px"}
 knitr::include_graphics("../images/race_bar_plot.png")
@@
 
-```{r age plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of defendant ages by race.', out.width= "400px", out.height= "200px"}
+```{r age plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of defendant ages by race.', out.width= "500px", out.height= "300px"}
 knitr::include_graphics("../images/race_age_plot.png")
@@
 
-```{r recid plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes.', out.width= "400px", out.height= "200px"}
+```{r recid plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes.', out.width= "500px", out.height= "300px"}
 knitr::include_graphics("../images/recid_bar_plot.png")
@@
 
-```{r recid race plot,echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes by race.', out.width= "400px", out.height= "200px"}
+```{r recid race plot,echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes by race.', out.width= "500px", out.height= "300px"}
 knitr::include_graphics("../images/race_recid_plot.png")
@@
 
-\\textbf{Statistical Parity Difference} - This metric measures the difference that the majority and protected classes get a particular outcome.
+**Statistical Parity Difference** - This metric measures the difference that the majority and protected classes get a particular outcome.
 The ideal value of this metric is 0.
@@
 
-\\textbf{Disparate Impact Ratio} - This metric is the ratio of how often the favorable outcome occurs in one group versus the other.
+**Disparate Impact Ratio** - This metric is the ratio of how often the favorable outcome occurs in one group versus the other.
 In the case of recidivism, this is the ratio of how many white defendants are predicted to not recidivate compared to how many black defendants are predicted to not recidivate.
@@
 
-\\textbf{Average Odds Difference} - This metric returns the average difference in false positive rate and true positive rate for the privileged and unprivileged groups.
+**Average Odds Difference** - This metric returns the average difference in false positive rate and true positive rate for the privileged and unprivileged groups.
 A value of 0 indicates equality of odds, and a value below 0 implies benefit for the privileged group.
@@
 
-\\textbf{Equal Opportunity Difference} - This metric is computed as the difference of true positive rates between the unprivileged and the privileged groups.
+**Equal Opportunity Difference** - This metric is computed as the difference of true positive rates between the unprivileged and the privileged groups.
 The true positive rate is the ratio of true positives to the total number of actual positives for a given group.
@@
 
-To reiterate, fairness is subjective, so the questions we ask ourselves to create parameters for our definition of fairness include:
-
 # Preliminary Results

### 2022-04-14 Paper_Draft2/Draft2.tex
trying to fix images

@@
 % Full title of the paper (Capitalized)
-\Title{Does the COMPAS Needle Always Point Towards Equity? Finding
-Fairness in the COMPAS Risk Assessment Algorithm':' A Case Study}
+\Title{Does the COMPAS Needle Always Point Towards Equity? Finding Fairness in
+the COMPAS Risk Assessment Algorithm':' A Case Study}
 
 % Authors, for the paper (add full first names)
-\Author{Amrita Acharya$^{1, 2}$, Dianne Caravela$^{1, 3}$, Eunice
-Kim$^{1, 4}$, Emma Kornberg$^{1, 5}$, Elisabeth Nesmith$^{1, 6}$}
+\Author{Amrita Acharya$^{1, 2}$, Dianne Caravela$^{1, 3}$, Eunice Kim$^{1, 4}$, Emma Kornberg$^{1, 5}$, Elisabeth Nesmith$^{1, 6}$}
 
 % Authors, for metadata in PDF
-\AuthorNames{Amrita Acharya, Dianne Caravela, Eunice Kim, Emma
-Kornberg, Elisabeth Nesmith}
+\AuthorNames{Amrita Acharya, Dianne Caravela, Eunice Kim, Emma Kornberg, Elisabeth Nesmith}
 
@@
 \address{%
-$^{1}$ \quad Statistical and Data Sciences Smith College Northampton, MA
-01063; \\
+$^{1}$ \quad Statistical and Data Sciences Smith College Northampton, MA 01063; \\
 $^{2}$ \quad ; \href{mailto:aacharya@smith.edu}{\nolinkurl{aacharya@smith.edu}}\\
@@
 % Abstract (Do not insert blank lines, i.e. \\)
-\abstract{A variety of disciplines use risk assessment instruments to
-help humans make data-driven decisions. Northpointe, a software company,
-created an algorithmic risk assessment instrument known as the
-Correctional Offender Management Profiling for Alternative Sanctions
-(COMPAS). COMPAS uses various behavioral and psychological metrics
-related to recidivism to assist justice systems in assessing a
-defendant's potential recidivism risk. Angwin et al.~published a
-ProPublica article in which they conclude that the racial biases in the
-criminal justice system are reflected in the COMPAS recidivism risk
-scores. In response, Dieterich et al.~published a rebuttal on behalf of
-Northpointe defending the COMPAS algorithm and refuting Angwin et al.'s
-allegation of racial bias. Using a human rights framework adopted from
-the organizations Women at the Table and AI Fairness 360, we use
-debiasing algorithms and fairness metrics to analyze the argument
-between Northpointe and ProPublica and determine whether and to what
-extent there is racial bias in the COMPAS algorithm.}
+\abstract{A variety of disciplines use risk assessment instruments to help humans
+make data-driven decisions. Northpointe, a software company, created an
+algorithmic risk assessment instrument known as the Correctional
+Offender Management Profiling for Alternative Sanctions (COMPAS). COMPAS
+uses various behavioral and psychological metrics related to recidivism
+to assist justice systems in assessing a defendant's potential
+recidivism risk. Angwin et al.~published a ProPublica article in which
+they conclude that the racial biases in the criminal justice system are
+reflected in the COMPAS recidivism risk scores. In response, Dieterich
+et al.~published a rebuttal on behalf of Northpointe defending the
+COMPAS algorithm and refuting Angwin et al.'s allegation of racial bias.
+Using a human rights framework adopted from the organizations Women at
+the Table and AI Fairness 360, we use debiasing algorithms and fairness
+metrics to analyze the argument between Northpointe and ProPublica and
+determine whether and to what extent there is racial bias in the COMPAS
+algorithm.}
 
@@
 African-American. Our data therefore has 5,723 rows (Figure
-\ref{fig:table snip}), with the below distributions of race, age (Figure
-\ref{fig:age plot}), and two year recidivism rate (Figures
-\ref{fig:recid plot} \& \ref{fig:recid race plot}).
+\ref{fig:table snip}), with the below distributions of race (Figure
+\ref{fig:race plot}), age (Figure \ref{fig:age plot}), and two year
+recidivism rate (Figures \ref{fig:recid plot} \&
+\ref{fig:recid race plot}).
 
@@
 
-{\centering \includegraphics[width=400px,height=200px]{../images/table_snippet}
+{\centering \includegraphics[width=500px,height=300px]{../images/table_snippet_new}
 
@@
 
-\caption{This table is a snippet the data set we will be using.}\label{fig:table snip}
+\caption{This table is a snippet of the data set we will be using.}\label{fig:table snip}
 \end{figure}
@@
 
-{\centering \includegraphics[width=400px,height=200px]{../images/race_bar_plot}
+{\centering \includegraphics[width=500px,height=300px]{../images/race_bar_plot}
 
@@
 
-{\centering \includegraphics[width=400px,height=200px]{../images/race_age_plot}
+{\centering \includegraphics[width=500px,height=300px]{../images/race_age_plot}
 
@@
 
-{\centering \includegraphics[width=400px,height=200px]{../images/recid_bar_plot}
+{\centering \includegraphics[width=500px,height=300px]{../images/recid_bar_plot}
 
@@
 
-{\centering \includegraphics[width=400px,height=200px]{../images/race_recid_plot}
+{\centering \includegraphics[width=500px,height=300px]{../images/race_recid_plot}
 
@@
 
-\textbackslash textbf\{Statistical Parity Difference\} - This metric
-measures the difference that the majority and protected classes get a
-particular outcome. The ideal value of this metric is 0. Fairness for
-this metric is between -0.1 and 0.1. A negative value means there is
-higher benefit for the privileged group (in this case, white
-defendants).
+\textbf{Statistical Parity Difference} - This metric measures the
+difference that the majority and protected classes get a particular
+outcome. The ideal value of this metric is 0. Fairness for this metric
+is between -0.1 and 0.1. A negative value means there is higher benefit
+for the privileged group (in this case, white defendants).
 
@@
 
-\textbackslash textbf\{Disparate Impact Ratio\} - This metric is the
-ratio of how often the favorable outcome occurs in one group versus the
-other. In the case of recidivism, this is the ratio of how many white
-defendants are predicted to not recidivate compared to how many black
-defendants are predicted to not recidivate. A value of 1 means that the
-ratio is exactly 1:1. Less than 1 means the privileged group (white
-defendants) benefits, while a value greater than 1 means the
-unprivileged group (Black defendants) benefits. According to AIF360, a
-ratio between .8 to 1.25 is considered fair:
+\textbf{Disparate Impact Ratio} - This metric is the ratio of how often
+the favorable outcome occurs in one group versus the other. In the case
+of recidivism, this is the ratio of how many white defendants are
+predicted to not recidivate compared to how many black defendants are
+predicted to not recidivate. A value of 1 means that the ratio is
+exactly 1:1. Less than 1 means the privileged group (white defendants)
+benefits, while a value greater than 1 means the unprivileged group
+(Black defendants) benefits. According to AIF360, a ratio between .8 to
+1.25 is considered fair:
 
@@
 
-\textbackslash textbf\{Average Odds Difference\} - This metric returns
-the average difference in false positive rate and true positive rate for
-the privileged and unprivileged groups. A value of 0 indicates equality
-of odds, and a value below 0 implies benefit for the privileged group.
+\textbf{Average Odds Difference} - This metric returns the average
+difference in false positive rate and true positive rate for the
+privileged and unprivileged groups. A value of 0 indicates equality of
+odds, and a value below 0 implies benefit for the privileged group.
 Equality of odds is achieved in the case of recidivism when the
@@
 
-\textbackslash textbf\{Equal Opportunity Difference\} - This metric is
-computed as the difference of true positive rates between the
-unprivileged and the privileged groups. The true positive rate is the
-ratio of true positives to the total number of actual positives for a
-given group.
+\textbf{Equal Opportunity Difference} - This metric is computed as the
+difference of true positive rates between the unprivileged and the
+privileged groups. The true positive rate is the ratio of true positives
+to the total number of actual positives for a given group.
 
@@
 approaches to determine which we assessed to be the most fair.
-
-To reiterate, fairness is subjective, so the questions we ask ourselves
-to create parameters for our definition of fairness include:
 

### 2022-04-14 Paper_Draft2/Draft2.tex
Amrita's changes

--- 2022-04-13 Paper_Draft2/Draft2.tex
@@
 % Full title of the paper (Capitalized)
-\Title{Does the COMPAS Needle Always Point Towards Equity? Finding
-Fairness in the COMPAS Risk Assessment Algorithm':' A Case Study}
+\Title{It's COMPAS-licated -- Can We Find Fairness in the COMPAS Risk
+Assessment Algorithm?}
 
 % Authors, for the paper (add full first names)
-\Author{Amrita Acharya$^{1, 2}$, Dianne Caravela$^{1, 3}$, Eunice
-Kim$^{1, 4}$, Emma Kornberg$^{1, 5}$, Elisabeth Nesmith$^{1, 6}$}
+\Author{Amrita Acharya$^{1, 2}$, Dianne Caravela$^{1, 3}$, Eunice Kim$^{1, 4}$, Emma Kornberg$^{1, 5}$, Elisabeth Nesmith$^{1, 6}$}
 
 % Authors, for metadata in PDF
-\AuthorNames{Amrita Acharya, Dianne Caravela, Eunice Kim, Emma
-Kornberg, Elisabeth Nesmith}
+\AuthorNames{Amrita Acharya, Dianne Caravela, Eunice Kim, Emma Kornberg, Elisabeth Nesmith}
 
@@
 % Contact information of the corresponding author
-\corres{Correspondence: }
+\corres{Correspondence: \href{mailto:leutnant@fh-muenster.de}{\nolinkurl{leutnant@fh-muenster.de}};
+Tel.: +XX-000-00-0000.}
 
@@
 % Abstract (Do not insert blank lines, i.e. \\)
-\abstract{A variety of disciplines use risk assessment instruments to
-help humans make data-driven decisions. Northpointe, a software company,
-created an algorithmic risk assessment instrument known as the
-Correctional Offender Management Profiling for Alternative Sanctions
-(COMPAS). COMPAS uses various behavioral and psychological metrics
-related to recidivism to assist justice systems in assessing a
-defendant's potential recidivism risk. Angwin et al.~published a
-ProPublica article in which they conclude that the racial biases in the
-criminal justice system are reflected in the COMPAS recidivism risk
-scores. In response, Dieterich et al.~published a rebuttal on behalf of
-Northpointe defending the COMPAS algorithm and refuting Angwin et al.'s
-allegation of racial bias. Using a human rights framework adopted from
-the organizations Women at the Table and AI Fairness 360, we use
-debiasing algorithms and fairness metrics to analyze the argument
-between Northpointe and ProPublica and determine whether and to what
-extent there is racial bias in the COMPAS algorithm.}
+\abstract{A variety of disciplines use risk assessment instruments to help humans
+make data-driven decisions. Northpointe, a software company, created a
+risk assessment instrument known as the Correctional Offender Management
+Profiling for Alternative Sanctions (COMPAS). COMPAS uses various
+behavioral and psychological metrics related to recidivism to assist
+justice systems in assessing a defendant's potential recidivism risk.
+ProPublica published an article which concludes that the biases in the
+criminal justice system are reflected in the COMPAS software. Using a
+human rights framework adopted from the organization Women at the Table,
+we use various debiasing algorithms to analyze both sides of the
+argument between Northpointe and ProPublica and determine the level and
+extent of racial bias in the COMPAS algorithm.}
 
@@
 
+Women @ the Table, the sponsor organization for this project, is ``a
+growing, global gender equality \& democracy CSO based in Geneva,
+Switzerland focused on advancing feminist systems change by using the
+prism of technology, innovation \& AI exercising leverage points in
+technology, the economy, sustainability \& democratic governance''
+\citep{noauthor_womenthetable_nodate}. We have been asked to collaborate
+on their AI \& Equality \citep{noauthor_ai_nodate} initiative, tasked
+with debiasing the COMPAS algorithm
+\citep{noauthor_aif360datasetscompasdataset_2018} and producing a
+corresponding data story that will be added to their library.
+
 The Correctional Offender Management Profiling for Alternative Sanctions
-(COMPAS) algorithm was created by the private, for-profit company
-Northpointe (now known by its parent company
-\href{https://www.equivant.com/faq/}{Equivant}), to predict defendants'
-risk of recidivism. It generates a decile score that classifies
-defendants' risk of recidivism as either low, medium, or high
-\citep{angwin2016machine}. Jurisdictions across the United States use
-the COMPAS risk assessment instrument, including but not limited to the
-New York \citep{NYDOCCS2019}, Massachusetts, Michigan, California
-\citep{Jackson2020Setting}, and Wisconsin \citep{WisconsinDOC}
-Departments of Corrections. Due to the proprietary nature of the COMPAS
-algorithm, it is unknown how exactly these recidivism risk scores are
-calculated; however, a sample COMPAS Risk Assessment Survey has been
-made publicly available, revealing the algorithm's input information.
-This survey has been widely critiqued for using proxy variables for race
-that do not explicitly factor in a defendant's race but heavily imply
-it, allowing Northpointe to claim that their algorithm is free of racial
-bias. For example, the COMPAS risk assessment survey asks screeners to
-speculate if a defendant might be affiliated with a gang. It also asks
-if a defendant has any friends or family members who have been crime
-victims \citep{Angwin2016Sample}. Although these questions do not
-directly ask about race, they do not take into account the pervasive
-nature of systemic racism that infiltrates every aspect of the lives of
-marginalized people, thereby indirectly asking about race. Angwin et
-al.~published a 2016 piece in the news publication ProPublica that
-analyzes the methods and algorithms used by Northpointe in their COMPAS
-risk score assessment algorithm and uncovers the racial biases in
-defendants' scores \citep{angwin2016machine}. Their research finds that
-``the algorithm {[}is{]} somewhat more accurate than a coin flip,'' a
-worrisome level of accuracy given the potential impact its
-determinations may have on real people's lives. Angwin et
-al.~specifically investigate the distribution of COMPAS scores by decile
-among Black and white defendants. They write: ``The analysis also
-{[}shows{]} that even when controlling for prior crimes, future
-recidivism, age, and gender, black defendants {[}are{]} 45 percent more
-likely to be assigned higher risk scores than white defendants''
-\citep{larson2016we}. After examining the fairness metric statistical
-parity difference, Angwin et al.~conclude that the algorithm is racially
-biased \citep{larson2016we}.
-
-Dieterich et al., on the behalf of Northpointe, deny the allegations of
-racial bias and offer their own analyses based on different fairness
-metrics in rebuttal \citep{equivant_response_2016}. Angwin et
-al.~maintain that there are biases in the outcome values, protected
-attributes, and covariates during Dieterich et al.'s data processing
-phase. ProPublica collaborators Larson et al.~account for these biases
-in their analyses. In their response, Dieterich et al.~highlight that
-Angwin et al.~did not account for base rates of recidivism in their
+(COMPAS) algorithm was created by the private for-profit company
+Northpointe, also known by its parent company Equivant
+\citep{equivant_faq_nodate}, to predict defendants' risk of recidivism.
+It generates a score that classifies defendants' risk of recidivism as
+either low, medium, or high \citep{angwin2016machine}. Jurisdictions
+across the United States use the COMPAS risk assessment instrument. In
+2016, ProPublica published a piece that analyzed the methods and
+algorithms used by Northpointe to uncover racial biases in defendants'
+scores \citep{angwin2016machine}. Their analysis looks at the
+distribution of decile COMPAS scores among Black and white defendants.
+ProPublica concludes, after using a statistical parity metric on false
+positives or false negatives, that the algorithm is racially biased
+\citep{larson2016we}. Northpointe denies the allegations of racial bias,
+conducting their own analyses based on different statistical parity
+metrics \citep{equivant_response_2018}. Since then, the two parties have
+had several exchanges, maintaining their original arguments. Currently,
+the two parties' disagreement focuses on pretrial Risk Assessment
+Instruments (RAIs). ProPublica maintains that there are biases in the
+outcome values, protected attributes, and covariates during
+Northpointe's data processing phase. ProPublica accounts for these
+biases in their analyses, and Northepointe's response, highlights how
+ProPublica did not account for base rates of recidivism in their
 analysis, which are important to understand initial percentages without
@@
 
-\href{https://www.womenatthetable.net/}{Women at the Table}, the sponsor
-organization for this project, is ``a growing, global gender equality \&
-democracy CSO based in Geneva, Switzerland focused on advancing feminist
-systems change by using the prism of technology, innovation \& AI
-exercising leverage points in technology, the economy, sustainability \&
-democratic governance.'' We are collaborating with the organization on
-its AI \& Equality \citep{noauthor_ai_nodate} initiative, tasked with
-debiasing the COMPAS algorithm \citep{aif360-oct-2018} and producing a
-corresponding data story that will be added to its library. Our project
-builds on Women at the Table's various debiasing algorithms used in its
-AI \& Equality Human Rights Toolbox to conduct our own analyses on the
-COMPAS dataset. Based on this analysis, we employ a human rights
-framework to contribute to the ProPublica and Northpointe debate and
-investigate whether or to what extent there is racial bias in the COMPAS
-algorithm. With a solid understanding of the two sides, we aim to
-pinpoint the shortcomings of both arguments and correct them in our
-analyses. We will use various debiasing techniques and fairness metrics
-to evaluate the level of bias present in the COMPAS data and our
-algorithm. We will summarize our results using the JupyterNotebook
-framework from Women at the Table, to be used by members of the
-organization to teach in a workshop setting. We hope that our findings
-will highlight the importance of checking statistical analyses using
-varied methods and contribute to the ongoing discussion of the effects
-of machine biases in the justice system.
+Our project builds on Women at the Table's various debiasing algorithms
+to conduct our own analyses on the COMPAS dataset. Based on this
+analysis, we employ a human rights framework to contribute to the
+ongoing ProPublica and Northpointe debate and investigate whether or to
+what extent there is racial bias in the COMPAS algorithm. With a solid
+understanding of the two sides, we aim to pinpoint the shortcomings of
+both arguments and correct them in our analyses. We will use various
+machine learning algorithms including logistic regression, cross
+validation, and lasso and ridge techniques to choose models. We will
+summarize our results using the JupyterNotebook framework from Women at
+the Table, to be used by members of the organization to teach in a
+workshop setting. We hope that our findings will highlight the
+importance of checking statistical analyses using varied methods and
+contribute to the ongoing discussion of the effects of machine biases in
+the justice system.
 
@@
 The data we are using for this project is the COMPAS General Recidivism
-Risk Scores dataset from the AI Fairness 360 (AIF360) toolkit
-\citep{aif360-oct-2018}, which does the same initial preprocessing as
-ProPublica. The raw data has 6,167 rows and each row represents an
-arrest charge for a defendant. AIF360's COMPAS data includes the
-defendant's age, race, sex, what they were charged with, and whether or
-not the defendant ultimately recidivated within a two-year period after
-their arrest. For the purposes of our project, which endeavors to
-evaluate the differing effects of the COMPAS algorithm on white
-defendants versus Black defendants, we have filtered the data to only
-include individuals whose race is listed as Caucasian or
+Risk Scores dataset from ProPublica used in their original analysis
+\citep{larson_propublicacompas-analysiscompas-scores-two-yearscsv_2022}.
+The COMPAS dataset we are using is from AI Fairness 360 (AIF360) toolkit
+\citep{noauthor_trusted-ai_360} which does the same initial
+preprocessing as ProPublica. The raw data has 6,167 rows and each row
+represents an arrest charge for a defendant. ProPublica's COMPAS data
+includes the defendant's age, race, sex, what they were charged with,
+and whether or not the defendant ultimately recidivated within a
+two-year period after their arrest. For the purposes of our project,
+which endeavors to evaluate the differing effects of the COMPAS
+algorithm on white defendants and Black defendants, we have filtered the
+data to only include individuals whose race is listed as Caucasian or
 African-American. Our data therefore has 5,723 rows (Figure
-\ref{fig:table snip}), with the below distributions of race, age (Figure
-\ref{fig:age plot}), and two year recidivism rate (Figures
-\ref{fig:recid plot} \& \ref{fig:recid race plot}).
+\ref{fig:table snip}), with the below distributions of race (Figure
+\ref{fig:race plot}), age (Figure \ref{fig:age plot}), and two year
+recidivism rate (Figures \ref{fig:recid plot} \&
+\ref{fig:recid race plot}).
 
@@
 
-\hypertarget{methods}{%
-\section{Methods}\label{methods}}
-
-The AIF360 toolkit contains various group and individual fairness
-metrics and pre-processing, in-processing, and post-processing
-algorithms that we used for our experiments \citep{aif360-oct-2018}. We
-researched the definitions and applications of different fairness
-metrics \citep{ashokan2021fairness} to determine which metric would be
-most appropriate for our project. We chose to look at four group
-fairness metrics instead of individual fairness metrics. Group fairness
-takes into account the attributes of a whole group as opposed to just
-one individual in the group, allowing us to represent more of the
-systemic issues happening. In general, group fairness metrics require
-that the unprivileged group is treated similarly to the privileged
-group, whereas individual fairness metrics require individuals to be
-treated consistently \citep{kypraiou_what_2021}. Group and individual
-metrics work in opposition of one another, meaning that when group
-fairness improves, individual fairness gets worse.
-
-\emph{Statistical Parity Difference} - This metric measures the
-difference that the majority and protected classes get a particular
-outcome. The ideal value of this metric is 0. Fairness for this metric
-is between -0.1 and 0.1. A negative value means there is higher benefit
-for the privileged group (in this case, white defendants).
-
-\(P(\hat{Y}=1|D=Unprivileged) - P(\hat{Y}=1|D=Privileged)\)
-
-\emph{Disparate Impact Ratio} - This metric is the ratio of how often
-the favorable outcome occurs in one group versus the other. In the case
-of recidivism, this is the ratio of how many white defendants are
-predicted to not recidivate compared to how many black defendants are
-predicted to not recidivate. A value of 1 means that the ratio is
-exactly 1:1. Less than 1 means the privileged group (white defendants)
-benefits, while a value greater than 1 means the unprivileged group
-(Black defendants) benefits. According to AIF360, a ratio between .8 to
-1.25 is considered fair:
-
-\(\frac{P(\hat{Y}=1|D=Unprivileged)}{P(\hat{Y}=1|D=Privileged)}\)
-
-\emph{Average Odds Difference} - This metric returns the average
-difference in false positive rate and true positive rate for the
-privileged and unprivileged groups. A value of 0 indicates equality of
-odds, and a value below 0 implies benefit for the privileged group.
-Equality of odds is achieved in the case of recidivism when the
-proportion of people who were predicted to recidivate and did recidivate
-is equal (true positive rate) for both Black and white defendants AND
-the proportion of people who were predicted to recidivate and did not
-recidivate (false positive rate) is equal for both Black and white
-defendants.
-
-\(\frac{(FPR_{D = unprivileged} - FPR_{D = privileged}) + (TPR_{D = unprivileged} - TPR_{D = privileged})}{2}\)
-
-\emph{Equal Opportunity Difference} - This metric is computed as the
-difference of true positive rates between the unprivileged and the
-privileged groups. The true positive rate is the ratio of true positives
-to the total number of actual positives for a given group.
-
-The ideal value is 0. A value less than 0 implies higher benefit for the
-privileged group and a value greater than 0 implies higher benefit for
-the unprivileged group. Fairness for this metric is between -0.1 and
-0.1. This metric is best used when it is very important to catch
-positive outcomes while false positives are not exceptionally
-problematic. This is not the case for the COMPAS data set, as false
-positives mean extra jail time for someone who will not actually
-reoffend.
-
-\(TPR_{D = Unprivileged} - TPR_{D = Privileged}\)
-
-For the next step of our experiment, we needed to determine where in the
-data science pipeline we can mitigate the most bias, using
-pre-processing, in-processing, and post-processing debiasing algorithms.
-These are all based on using predictive models to figure out how we can
-``fix'' the bias that is present. Pre-processing refers to assessing the
-training data, and it is the most flexible method because it has not yet
-trained a model that may carry assumptions about the data. It is
-important to keep in mind that pre-processing prevents assumptions in
-the modeling, but does not account for the bias in data collection. With
-pre-processing, we use the reweighting algorithm from AIF360 which
-weights the examples in each (group, label) combination differently to
-ensure fairness before classification \citep{aif360-oct-2018}. We used a
-logistic regression model for this algorithm, as it is the easiest to
-interpret in the given context. After running the fairness metrics using
-the pre-processing algorithm, we were able to compare our results to the
-baseline metrics from the previous section.
-
-Another approach, in-processing, explores how bias can be targeted while
-building a model. It assesses how we build learning algorithms, which
-often prioritize accuracy over fairness. The technique we use is the
-Meta Classification algorithm, which accounts for the fairness metric as
-part of the input and returns a classifier optimized by that particular
-metric. In order to do this, we first needed to convert our data frame
-into a data type called a BinaryLabelDataset. Similarly to
-pre-processing, we compared the results of our in-processing methods
-with both the baseline and the pre-processing to gauge which method so
-far has better individual or group fairness.
-
-Our last approach, post-processing, makes transformations on model
-outputs. Similarly to the pre-processing, it is flexible because it does
-not work directly with the algorithm, but instead has direct access to
-all the predicted classification values. Because the algorithms are
-often so complex and uninterpretable, post-processors are trained in a
-black box. To get around this, we use the PostProcessingMeta class from
-AIF360 which combines the training and prediction for a random estimator
-and the post-processor while simultaneously splitting the dataset. Based
-on these results, we looked comparatively at all the approaches to
-determine which we assessed to be the most fair.
-
-To reiterate, fairness is subjective, so the questions we ask ourselves
-to create parameters for our definition of fairness include:
-
-\hypertarget{preliminary-results}{%
-\section{Preliminary Results}\label{preliminary-results}}
-
-The statistical parity difference is -0.14. This indicates that there is
-a large difference between white and Black defendants regarding whether
-or not they recidivate. The algorithm unfairly benefits white defendants
-over Black defendants.
-
-Disparate impact ratio is 0.47. The ratio of white defendants predicted
-to not recidivate to the Black defendants predicted to not recidivate is
-0.47. A ratio between 0.8 and 1.25 is considered fair, therefore the
-algorithm unfairly benefits white defendants.
-
-Average odds difference is -0.44. The average difference in false
-positive rates and true positive rates for white and Black defendants is
--0.44. Values less than zero are considered in favor of the privileged
-group, so the algorithm unfairly benefits white defendants.
-
-Equal opportunity difference is -0.41. The difference of true positive
-rates between the Black and white groups is -0.41. A value less than 0
-indicates a benefit to the privileged group, so the algorithm unfairly
-benefits white defendants. The value is substantially less than -0.1,
-which indicates that the algorithm benefits white defendants.
-
-All four group fairness metrics determine that the COMPAS algorithm
-favors white defendants over Black defendants. The statistical parity
-metric indicates unfairness by a smaller margin of 0.04 than equal
-opportunity difference which has a margin of error 0.4 below the
-threshold. Although this is true, each algorithm uses different
-penalties for unfavorable outcomes, so the higher the penalty the more
-of a difference there will be between the fairness threshold and the
-value.
+\hypertarget{results}{%
+\section{Results}\label{results}}
 

### 2022-04-14 final_metrics_analysis.Rmd
Amrita's changes

@@
 from aif360.sklearn.datasets import fetch_compas
-from aif360.algorithms.inprocessing import MetaFairClassifier
+from aif360.algorithms.inprocessing import MetaFairClassifier, PrejudiceRemover, ARTClassifier
 
 from aif360.sklearn.preprocessing import ReweighingMeta
-from aif360.sklearn.inprocessing import AdversarialDebiasing
+#from aif360.sklearn.inprocessing import 
 from aif360.sklearn.postprocessing import CalibratedEqualizedOdds, PostProcessingMeta
@@
 
+## Prejudice Remover
+```{r}
+#prej = PrejudiceRemover(eta=1.0, sensitive_attr='race', class_attr='two_year_recid', seed=1234567)
+#prej.fit(train_BLD)
+#y_pred_PREJ = meta.predict(test_BLD)
+#y_pred_PREJ = y_pred_PREJ.labels.flatten()
+```
+
+## Meta Fair Classifier 
 ```{python in-processing meta}
-# meta = MetaFairClassifier(tau=0.8, sensitive_attr='race', type='fdr', seed=1234567)
-# 
-# meta.fit(train_BLD)
-# y_pred_META = meta.predict(test_BLD)
-# y_pred_META = y_pred_META.labels.flatten()
+#meta = MetaFairClassifier(tau=0.8, sensitive_attr='race', type='sr', seed=1234567)
+#meta.fit(train_BLD)
+#y_pred_META = meta.predict(test_BLD)
+#y_pred_META = y_pred_META.labels.flatten()
 ```

### 2022-04-14 Paper_Draft2/Draft2.Rmd

@@
 
-```{r race plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of defendant races.', out.width= "500px", out.height= "300px"}
+```{r race plot, echo=FALSE, fig.align='center', fig.cap= 'The race distribution for black people is 3000 and the distribution for white people is 2100.
+Therefore, there are more black people in the dataset than white people.', out.width= "500px", out.height= "300px"}
 knitr::include_graphics("../images/race_bar_plot.png")
@@
 
-```{r age plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of defendant ages by race.', out.width= "500px", out.height= "300px"}
+```{r age plot, echo=FALSE, fig.align='center', fig.cap= "This plot is a Kernel Density Estimate (KDE) plot.
+The y-axis represents the KDE of the Probability Distribution Function of the age variable in our dataset.
+The probability of a defendant's age being between two points on the x-axis is the total shaded area of the curve under the two points.
+The purple dotted line shows the average age of Black defendants and the green dotted line shows the average age of white defendants.
+For both groups, the majority of defendants are relatively young, but this is especially noticeable for Black defendants.", out.width= "500px", out.height= "300px"}
 knitr::include_graphics("../images/race_age_plot.png")
@@
 
-```{r recid plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes.', out.width= "500px", out.height= "300px"}
+```{r recid plot, echo=FALSE, fig.align='center', fig.cap= 'Of both black and white people, there is a higher percentage of people who did not recidivate than who did.
+2500 people recidivated, whereas approximately 2600 did not.', out.width= "500px", out.height= "300px"}
 knitr::include_graphics("../images/recid_bar_plot.png")
@@
 
-```{r recid race plot,echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes by race.', out.width= "500px", out.height= "300px"}
+```{r recid race plot,echo=FALSE, fig.align='center', fig.cap= 'When we divide the data into Black and white defendants, we can see that Black defendants recidivate more than white defendants.
+Black defendants recidivate are more likely than not to recidivate.
+However, we can also see that there are more Black defendants in the dataset in general.', out.width= "500px", out.height= "300px"}
 knitr::include_graphics("../images/race_recid_plot.png")
@@
 
-# Discussion
-
-# Conclusion
-
 # Bibliography

### 2022-04-14 Paper_Draft2/Draft2.tex

--- 2022-04-14 Paper_Draft2/Draft2.tex
@@
 
-\caption{This plot shows the distribution of defendant races.}\label{fig:race plot}
+\caption{The race distribution for black people is 3000 and the distribution for white people is 2100. Therefore, there are more black people in the dataset than white people.}\label{fig:race plot}
 \end{figure}
@@
 
-\caption{This plot shows the distribution of defendant ages by race.}\label{fig:age plot}
+\caption{This plot is a Kernel Density Estimate (KDE) plot.
+The y-axis represents the KDE of the Probability Distribution Function of the age variable in our dataset.
+The probability of a defendant's age being between two points on the x-axis is the total shaded area of the curve under the two points.
+The purple dotted line shows the average age of Black defendants and the green dotted line shows the average age of white defendants.
+For both groups, the majority of defendants are relatively young, but this is especially noticeable for Black defendants.}\label{fig:age plot}
 \end{figure}
@@
 
-\caption{This plot shows the distribution of two year recidivism outcomes.}\label{fig:recid plot}
+\caption{Of both black and white people, there is a higher percentage of people who did not recidivate than who did. 2500 people recidivated, whereas approximately 2600 did not.}\label{fig:recid plot}
 \end{figure}
@@
 
-\caption{This plot shows the distribution of two year recidivism outcomes by race.}\label{fig:recid race plot}
+\caption{When we divide the data into Black and white defendants, we can see that Black defendants recidivate more than white defendants.
+Black defendants recidivate are more likely than not to recidivate.
+However, we can also see that there are more Black defendants in the dataset in general.}\label{fig:recid race plot}
 \end{figure}
@@
 
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
 % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2022-04-14 Paper_Draft2/Draft2.Rmd
tweaking plots in metrics analysis and fixing images in paper draft

--- 2022-04-14 Paper_Draft2/Draft2.Rmd
@@
 For the purposes of our project, which endeavors to evaluate the differing effects of the COMPAS algorithm on white defendants versus Black defendants, we have filtered the data to only include individuals whose race is listed as Caucasian or African-American.
-Our data therefore has 5,723 rows (Figure \ref{fig:table snip}), with the below distributions of race (Figure \ref{fig:race plot}), age (Figure \ref{fig:age plot}), and two year recidivism rate (Figures \ref{fig:recid plot} & \ref{fig:recid race plot}).
+Our data therefore has 5,723 rows (Figure \ref{fig:table snip}), with the below distributions of race (Figure \ref{fig:race plot}), age (Figure \ref{fig:age plot}), and two year recidivism rate (Figures \ref{fig:recid plot} & \ref{fig:recid race plot} and Table \ref{tab:recid table}).
 
-```{r table snip, echo = FALSE, fig.align='center', fig.cap= 'This table is a snippet of the data set we will be using.', out.width= "500px", out.height= "300px"}
-knitr::include_graphics("../images/table_snippet_new.png")
+```{r table snip, echo = FALSE, fig.align='center', fig.cap= 'This table is a snippet of the data set we will be using.', out.width= "100%"}
+knitr::include_graphics("../images/table_snippet.png")
 ```
 
-```{r race plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of defendant races.', out.width= "500px", out.height= "300px"}
-knitr::include_graphics("../images/race_bar_plot.png")
+```{r race plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of defendant races.', out.width= "100%"}
+knitr::include_graphics("../images/race_bar_plot_new.png")
 ```
 
-```{r age plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of defendant ages by race.', out.width= "500px", out.height= "300px"}
-knitr::include_graphics("../images/race_age_plot.png")
+```{r age plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of defendant ages by race.', out.width= "100%"}
+knitr::include_graphics("../images/age_race_plot_new.png")
 ```
 
-```{r recid plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes.', out.width= "500px", out.height= "300px"}
-knitr::include_graphics("../images/recid_bar_plot.png")
+```{r recid plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes.', out.width= "100%"}
+knitr::include_graphics("../images/recid_bar_plot_new.png")
 ```
 
-```{r recid race plot,echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes by race.', out.width= "500px", out.height= "300px"}
-knitr::include_graphics("../images/race_recid_plot.png")
+```{r recid race plot,echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes by race.', out.width= "100%"}
+knitr::include_graphics("../images/race_recid_bar_plot_new.png")
+```
+
+```{r, include = FALSE}
+library(tidyverse)
+```
+
+```{r, include = FALSE}
+x <- tribble(~`Two Year Recidivism by Race`, ~`Recidivated`, ~`Survived`,
+             "African American", "1661", "1512",
+             "Caucasian", "822", "1278")
+```
+
+```{r, echo = FALSE}
+knitr::kable(x, caption = "This table shows the distribution of recidivism by race \\label{tab:recid table}")
 ```

### 2022-04-14 Paper_Draft2/Draft2.tex

--- 2022-04-14 Paper_Draft2/Draft2.tex
@@
 
+% From pandoc table feature
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
+
 \begin{document}
@@
 recidivism rate (Figures \ref{fig:recid plot} \&
-\ref{fig:recid race plot}).
+\ref{fig:recid race plot} and Table \ref{tab:recid table}).
 
@@
 
-{\centering \includegraphics[width=500px,height=300px]{../images/table_snippet_new}
+{\centering \includegraphics[width=1\linewidth]{../images/table_snippet}
 
@@
 
-{\centering \includegraphics[width=500px,height=300px]{../images/race_bar_plot}
+{\centering \includegraphics[width=1\linewidth]{../images/race_bar_plot_new}
 
@@
 
-{\centering \includegraphics[width=500px,height=300px]{../images/race_age_plot}
+{\centering \includegraphics[width=1\linewidth]{../images/age_race_plot_new}
 
@@
 
-{\centering \includegraphics[width=500px,height=300px]{../images/recid_bar_plot}
+{\centering \includegraphics[width=1\linewidth]{../images/recid_bar_plot_new}
 
@@
 
-{\centering \includegraphics[width=500px,height=300px]{../images/race_recid_plot}
+{\centering \includegraphics[width=1\linewidth]{../images/race_recid_bar_plot_new}
 
@@
 \end{figure}
+
+\begin{longtable}[]{@{}lll@{}}
+\caption{This table shows the distribution of recidivism by race
+\label{tab:recid table}}\tabularnewline
+\toprule
+Two Year Recidivism by Race & Recidivated & Survived\tabularnewline
+\midrule
+\endfirsthead
+\toprule
+Two Year Recidivism by Race & Recidivated & Survived\tabularnewline
+\midrule
+\endhead
+African American & 1661 & 1512\tabularnewline
+Caucasian & 822 & 1278\tabularnewline
+\bottomrule
+\end{longtable}
 

### 2022-04-14 final_metrics_analysis.Rmd

--- 2022-04-13 final_metrics_analysis.Rmd
@@
 library(reticulate)
-use_condaenv("anaconda3")
 ```
@@
 green = '#30875c'
+orange = '#E7881E'
+blue = '#20A4CF'
 workshop_palette = [purple, green]
@@
 # barplot of race
-sns.countplot(y="race", data=df_viz, palette=["#30875c"])
+sns.countplot(x="race", data=df_viz, palette=[blue, orange])
 plt.title('Race Distribution (White and Black individuals only)')
@@
 # age distribution by race
-#ax = sns.kdeplot(x="age", hue="race", data=df_viz);
-#kdeline = ax.lines[0]
-#mean_fem = df_viz.groupby('race').age.median()[0]
-#xs = kdeline.get_xdata()
-#ys = kdeline.get_ydata()
-#height_fem = np.interp(mean_fem, xs, ys)
-#ax.vlines(mean_fem, 0, height_fem, color=purple, ls=':')
+ax = sns.kdeplot(x="age", hue="race", data=df_viz);
+kdeline = ax.lines[0]
+mean_fem = df_viz.groupby('race').age.median()[0]
+xs = kdeline.get_xdata()
+ys = kdeline.get_ydata()
+height_fem = np.interp(mean_fem, xs, ys)
+ax.vlines(mean_fem, 0, height_fem, color=blue, ls=':')
 
-#mean_m = df_viz.groupby('race').age.median()[1]
-#height_m = np.interp(mean_m, xs, ys)
-#ax.vlines(mean_m, 0, height_m, color=green, ls=':')
+mean_m = df_viz.groupby('race').age.median()[1]
+height_m = np.interp(mean_m, xs, ys)
+ax.vlines(mean_m, 0, height_m, color=orange, ls=':')
 
-#print(df_viz.groupby('race').age.median())
+print(df_viz.groupby('race').age.median())
 
-#plt.title('Distribution of Age by Race')
+plt.title('Distribution of Age by Race')
 ```

### 2022-04-15 final_metrics_analysis.Rmd
Finalizing plots and plot captions

@@
 # barplot of race
-sns.countplot(x="race", data=df_viz, palette=[blue, orange])
+sns.countplot(x="race", data=df_viz, palette=workshop_palette)
 plt.title('Race Distribution (White and Black individuals only)')
@@
 # age distribution by race
-ax = sns.kdeplot(x="age", hue="race", data=df_viz);
+ax = sns.kdeplot(x="age", hue="race", data=df_viz, palette=workshop_palette);
 kdeline = ax.lines[0]
-mean_fem = df_viz.groupby('race').age.median()[0]
+mean_black = df_viz.groupby('race').age.median()[0]
 xs = kdeline.get_xdata()
 ys = kdeline.get_ydata()
-height_fem = np.interp(mean_fem, xs, ys)
-ax.vlines(mean_fem, 0, height_fem, color=blue, ls=':')
+height_black = np.interp(mean_black, xs, ys)
+ax.vlines(mean_black, 0, height_black, color=purple, ls=':')
 
-mean_m = df_viz.groupby('race').age.median()[1]
-height_m = np.interp(mean_m, xs, ys)
-ax.vlines(mean_m, 0, height_m, color=orange, ls=':')
+mean_white = df_viz.groupby('race').age.median()[1]
+height_white = np.interp(mean_white, xs, ys)
+ax.vlines(mean_white, 0, height_white, color=green, ls=':')
 

### 2022-04-15 Paper_Draft2/Draft2.Rmd

@@
 The raw data has 6,167 rows and each row represents an arrest charge for a defendant.
-AIF360’s COMPAS data includes the defendant's age, race, sex, what they were charged with, and whether or not the defendant ultimately recidivated within a two-year period after their arrest.
+AIF360’s COMPAS data includes the defendant's age, race, sex, prior charges, what they were charged with, and whether or not the defendant ultimately recidivated within a two-year period after their arrest.
 For the purposes of our project, which endeavors to evaluate the differing effects of the COMPAS algorithm on white defendants versus Black defendants, we have filtered the data to only include individuals whose race is listed as Caucasian or African-American.
@@
 
-```{r table snip, echo = FALSE, fig.align='center', fig.cap= 'This table is a snippet of the data set we will be using.', out.width= "100%"}
+```{r table snip, echo = FALSE, fig.align='center', fig.cap= "This table is a snippet of the data set we will be using, containing information on a defendant's age, sex, race, criminal history, charge degree, charge description, and two year recidivism outcome.", out.width= "100%"}
 knitr::include_graphics("../images/table_snippet.png")
@@
 
-```{r race plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of defendant races.', out.width= "100%"}
+```{r race plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of defendant races.
+The dataset contains about 3200 Black defendants and 2100 white defendants.
+Therefore, there are more Black people represented in the dataset than white people.', out.width= "100%"}
 knitr::include_graphics("../images/race_bar_plot_new.png")
@@
 
-```{r age plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of defendant ages by race.', out.width= "100%"}
+```{r age plot, echo=FALSE, fig.align='center', fig.cap= "This plot shows the distribution of defendant ages by race.
+The purple curve shows the distribution of the ages of Black defendants, and the green curve shows the distribution of the ages of white defendants.
+The probability of a defendant's age being between two points on the x-axis is the total shaded area of the curve under the two points.
+The purple dotted line represents the median age of Black defendants (29 years) and the green dotted line represents the median age of white defendants (35 years).
+For both groups, the majority of defendants are relatively young, but this is especially noticeable for Black defendants.", out.width= "100%"}
 knitr::include_graphics("../images/age_race_plot_new.png")
@@
 
-```{r recid plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes.', out.width= "100%"}
+```{r recid plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes.
+Out of all defendants, there is a higher proportion of people who did not recidivate than who did.
+About 2500 people recidivated, whereas approximately 2600 did not.', out.width= "100%"}
 knitr::include_graphics("../images/recid_bar_plot_new.png")
@@
 
-```{r recid race plot,echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes by race.', out.width= "100%"}
+```{r recid race plot,echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes by race.
+When we divide the data into Black and white defendants, we can see that BLack defendants recidivate more than white defendants and Black defendants are more likely to recidivate than not recidivate.
+We can also see that there are more Black defendants in the dataset overall.', out.width= "100%"}
 knitr::include_graphics("../images/race_recid_bar_plot_new.png")
@@
 ```{r, echo = FALSE}
-knitr::kable(x, caption = "This table shows the distribution of recidivism by race \\label{tab:recid table}")
+knitr::kable(x, caption = "This table shows the counts of recidivism by race, illustrating how a much greater proportion (> 50%) of Black defendants recidivated than their white counterparts.
+\\label{tab:recid table}")
 ```

### 2022-04-15 Paper_Draft2/Draft2.tex

@@
 arrest charge for a defendant. AIF360's COMPAS data includes the
-defendant's age, race, sex, what they were charged with, and whether or
-not the defendant ultimately recidivated within a two-year period after
-their arrest. For the purposes of our project, which endeavors to
-evaluate the differing effects of the COMPAS algorithm on white
-defendants versus Black defendants, we have filtered the data to only
-include individuals whose race is listed as Caucasian or
-African-American. Our data therefore has 5,723 rows (Figure
+defendant's age, race, sex, prior charges, what they were charged with,
+and whether or not the defendant ultimately recidivated within a
+two-year period after their arrest. For the purposes of our project,
+which endeavors to evaluate the differing effects of the COMPAS
+algorithm on white defendants versus Black defendants, we have filtered
+the data to only include individuals whose race is listed as Caucasian
+or African-American. Our data therefore has 5,723 rows (Figure
 \ref{fig:table snip}), with the below distributions of race (Figure
@@
 
-\caption{This table is a snippet of the data set we will be using.}\label{fig:table snip}
+\caption{This table is a snippet of the data set we will be using, containing information on a defendant's age, sex, race, criminal history, charge degree, charge description, and two year recidivism outcome.}\label{fig:table snip}
 \end{figure}
@@
 
-\caption{This plot shows the distribution of defendant races.}\label{fig:race plot}
+\caption{This plot shows the distribution of defendant races.
+The dataset contains about 3200 Black defendants and 2100 white defendants.
+Therefore, there are more Black people represented in the dataset than white people.}\label{fig:race plot}
 \end{figure}
@@
 
-\caption{This plot shows the distribution of defendant ages by race.}\label{fig:age plot}
+\caption{This plot shows the distribution of defendant ages by race.
+The purple curve shows the distribution of the ages of Black defendants, and the green curve shows the distribution of the ages of white defendants.
+The probability of a defendant's age being between two points on the x-axis is the total shaded area of the curve under the two points.
+The purple dotted line represents the median age of Black defendants (29 years) and the green dotted line represents the median age of white defendants (35 years).
+For both groups, the majority of defendants are relatively young, but this is especially noticeable for Black defendants.}\label{fig:age plot}
 \end{figure}
@@
 
-\caption{This plot shows the distribution of two year recidivism outcomes.}\label{fig:recid plot}
+\caption{This plot shows the distribution of two year recidivism outcomes.
+Out of all defendants, there is a higher proportion of people who did not recidivate than who did.
+About 2500 people recidivated, whereas approximately 2600 did not.}\label{fig:recid plot}
 \end{figure}
@@
 
-\caption{This plot shows the distribution of two year recidivism outcomes by race.}\label{fig:recid race plot}
+\caption{This plot shows the distribution of two year recidivism outcomes by race.
+When we divide the data into Black and white defendants, we can see that BLack defendants recidivate more than white defendants and Black defendants are more likely to recidivate than not recidivate.
+We can also see that there are more Black defendants in the dataset overall.}\label{fig:recid race plot}
 \end{figure}
@@
 \begin{longtable}[]{@{}lll@{}}
-\caption{This table shows the distribution of recidivism by race
+\caption{This table shows the counts of recidivism by race, illustrating
+how a much greater proportion (\textgreater{} 50\%) of Black defendants
+recidivated than their white counterparts.
 \label{tab:recid table}}\tabularnewline

### 2022-04-15 Paper_Draft2/Draft2.Rmd

@@
 ---
-title: Does the COMPAS Needle Always Point Towards Equity? Finding Fairness in the COMPAS Risk Assessment Algorithm':' A Case Study
+title: "Does the COMPAS Needle Always Point Towards Equity? Finding Fairness in the COMPAS Risk Assessment Algorithm: A Case Study"
 
@@
 csl: ../resources/chicago-author-date.csl
-appendix: appendix.tex
+#appendix: appendix.tex
+classoption: fleqn
 
@@
 specifically investigate the distribution of COMPAS scores by decile among Black and white defendants.
-They write: “The analysis also [shows] that even when controlling for prior crimes, future recidivism, age, and gender, black defendants [are] 45 percent more likely to be assigned higher risk scores than white defendants” [@larson2016we].
+They write: “The analysis also [shows] that even when controlling for prior crimes, future recidivism, age, and gender, Black defendants [are] 45 percent more likely to be assigned higher risk scores than white defendants” [@larson2016we].
 After examining the fairness metric statistical parity difference, Angwin et al.
@@
 ```{r recid race plot,echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes by race.
-When we divide the data into Black and white defendants, we can see that BLack defendants recidivate more than white defendants and Black defendants are more likely to recidivate than not recidivate.
+When we divide the data into Black and white defendants, we can see that Black defendants recidivate more than white defendants and Black defendants are more likely to recidivate than not recidivate.
 We can also see that there are more Black defendants in the dataset overall.', out.width= "100%"}
@@
 
-The AIF360 toolkit contains various group and individual fairness metrics and pre-processing, in-processing, and post-processing algorithms that we used for our experiments [@aif360-oct-2018].
+The AIF360 toolkit contains various group and individual fairness metrics as well as pre-processing, in-processing, and post-processing algorithms that we used for our experiments [@aif360-oct-2018].
 We researched the definitions and applications of different fairness metrics [@ashokan2021fairness] to determine which metric would be most appropriate for our project.
@@
 
-$P(\hat{Y}=1|D=Unprivileged) - P(\hat{Y}=1|D=Privileged)$
+$$P(\hat{Y}=1|D=Unprivileged) - P(\hat{Y}=1|D=Privileged)$$
 
 **Disparate Impact Ratio** - This metric is the ratio of how often the favorable outcome occurs in one group versus the other.
-In the case of recidivism, this is the ratio of how many white defendants are predicted to not recidivate compared to how many black defendants are predicted to not recidivate.
+In the case of recidivism, this is the ratio of how many white defendants are predicted to not recidivate compared to how many Black defendants are predicted to not recidivate.
 A value of 1 means that the ratio is exactly 1:1.
@@
 
-$\frac{P(\hat{Y}=1|D=Unprivileged)}{P(\hat{Y}=1|D=Privileged)}$
+$$\frac{P(\hat{Y}=1|D=Unprivileged)}{P(\hat{Y}=1|D=Privileged)}$$
 
@@
 
-$\frac{(FPR_{D = unprivileged} - FPR_{D = privileged}) + (TPR_{D = unprivileged} - TPR_{D = privileged})}{2}$
+$$\frac{(FPR_{D = unprivileged} - FPR_{D = privileged}) + (TPR_{D = unprivileged} - TPR_{D = privileged})}{2}$$
 
@@
 
-$TPR_{D = Unprivileged} - TPR_{D = Privileged}$
+$$TPR_{D = Unprivileged} - TPR_{D = Privileged}$$
 

### 2022-04-15 Paper_Draft2/Draft2.tex

@@
 % Full title of the paper (Capitalized)
-\Title{Does the COMPAS Needle Always Point Towards Equity? Finding Fairness in
-the COMPAS Risk Assessment Algorithm':' A Case Study}
+\Title{Does the COMPAS Needle Always Point Towards Equity? Finding
+Fairness in the COMPAS Risk Assessment Algorithm: A Case Study}
 
 % Authors, for the paper (add full first names)
-\Author{Amrita Acharya$^{1, 2}$, Dianne Caravela$^{1, 3}$, Eunice Kim$^{1, 4}$, Emma Kornberg$^{1, 5}$, Elisabeth Nesmith$^{1, 6}$}
+\Author{Amrita Acharya$^{1, 2}$, Dianne Caravela$^{1, 3}$, Eunice
+Kim$^{1, 4}$, Emma Kornberg$^{1, 5}$, Elisabeth Nesmith$^{1, 6}$}
 
 % Authors, for metadata in PDF
-\AuthorNames{Amrita Acharya, Dianne Caravela, Eunice Kim, Emma Kornberg, Elisabeth Nesmith}
+\AuthorNames{Amrita Acharya, Dianne Caravela, Eunice Kim, Emma
+Kornberg, Elisabeth Nesmith}
 
@@
 % Abstract (Do not insert blank lines, i.e. \\)
-\abstract{A variety of disciplines use risk assessment instruments to help humans
-make data-driven decisions. Northpointe, a software company, created an
-algorithmic risk assessment instrument known as the Correctional
-Offender Management Profiling for Alternative Sanctions (COMPAS). COMPAS
-uses various behavioral and psychological metrics related to recidivism
-to assist justice systems in assessing a defendant's potential
-recidivism risk. Angwin et al.~published a ProPublica article in which
-they conclude that the racial biases in the criminal justice system are
-reflected in the COMPAS recidivism risk scores. In response, Dieterich
-et al.~published a rebuttal on behalf of Northpointe defending the
-COMPAS algorithm and refuting Angwin et al.'s allegation of racial bias.
-Using a human rights framework adopted from the organizations Women at
-the Table and AI Fairness 360, we use debiasing algorithms and fairness
-metrics to analyze the argument between Northpointe and ProPublica and
-determine whether and to what extent there is racial bias in the COMPAS
-algorithm.}
+\abstract{A variety of disciplines use risk assessment instruments to
+help humans make data-driven decisions. Northpointe, a software company,
+created an algorithmic risk assessment instrument known as the
+Correctional Offender Management Profiling for Alternative Sanctions
+(COMPAS). COMPAS uses various behavioral and psychological metrics
+related to recidivism to assist justice systems in assessing a
+defendant's potential recidivism risk. Angwin et al.~published a
+ProPublica article in which they conclude that the racial biases in the
+criminal justice system are reflected in the COMPAS recidivism risk
+scores. In response, Dieterich et al.~published a rebuttal on behalf of
+Northpointe defending the COMPAS algorithm and refuting Angwin et al.'s
+allegation of racial bias. Using a human rights framework adopted from
+the organizations Women at the Table and AI Fairness 360, we use
+debiasing algorithms and fairness metrics to analyze the argument
+between Northpointe and ProPublica and determine whether and to what
+extent there is racial bias in the COMPAS algorithm.}
 
@@
 {[}shows{]} that even when controlling for prior crimes, future
-recidivism, age, and gender, black defendants {[}are{]} 45 percent more
+recidivism, age, and gender, Black defendants {[}are{]} 45 percent more
 likely to be assigned higher risk scores than white defendants''
@@
 \caption{This plot shows the distribution of two year recidivism outcomes by race.
-When we divide the data into Black and white defendants, we can see that BLack defendants recidivate more than white defendants and Black defendants are more likely to recidivate than not recidivate.
+When we divide the data into Black and white defendants, we can see that Black defendants recidivate more than white defendants and Black defendants are more likely to recidivate than not recidivate.
 We can also see that there are more Black defendants in the dataset overall.}\label{fig:recid race plot}
@@
 \toprule
-Two Year Recidivism by Race & Recidivated & Survived\tabularnewline
+Two Year Recidivism by Race & Recidivated & Survived \\
 \midrule
@@
 \toprule
-Two Year Recidivism by Race & Recidivated & Survived\tabularnewline
+Two Year Recidivism by Race & Recidivated & Survived \\
 \midrule
 \endhead
-African American & 1661 & 1512\tabularnewline
-Caucasian & 822 & 1278\tabularnewline
+African American & 1661 & 1512 \\
+Caucasian & 822 & 1278 \\
 \bottomrule
@@
 The AIF360 toolkit contains various group and individual fairness
-metrics and pre-processing, in-processing, and post-processing
+metrics as well as pre-processing, in-processing, and post-processing
 algorithms that we used for our experiments \citep{aif360-oct-2018}. We
@@
 
-\(P(\hat{Y}=1|D=Unprivileged) - P(\hat{Y}=1|D=Privileged)\)
+\[P(\hat{Y}=1|D=Unprivileged) - P(\hat{Y}=1|D=Privileged)\]
 
@@
 of recidivism, this is the ratio of how many white defendants are
-predicted to not recidivate compared to how many black defendants are
+predicted to not recidivate compared to how many Black defendants are
 predicted to not recidivate. A value of 1 means that the ratio is
@@
 
-\(\frac{P(\hat{Y}=1|D=Unprivileged)}{P(\hat{Y}=1|D=Privileged)}\)
+\[\frac{P(\hat{Y}=1|D=Unprivileged)}{P(\hat{Y}=1|D=Privileged)}\]
 
@@
 
-\(\frac{(FPR_{D = unprivileged} - FPR_{D = privileged}) + (TPR_{D = unprivileged} - TPR_{D = privileged})}{2}\)
+\[\frac{(FPR_{D = unprivileged} - FPR_{D = privileged}) + (TPR_{D = unprivileged} - TPR_{D = privileged})}{2}\]
 
@@
 
-\(TPR_{D = Unprivileged} - TPR_{D = Privileged}\)
+\[TPR_{D = Unprivileged} - TPR_{D = Privileged}\]
 
@@
 
-\input{"appendix.tex"}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2022-04-15 Paper_Draft2/Draft2.Rmd
Tense edits

--- 2022-04-14 Paper_Draft2/Draft2.Rmd
@@
 For the purposes of our project, which endeavors to evaluate the differing effects of the COMPAS algorithm on white defendants versus Black defendants, we have filtered the data to only include individuals whose race is listed as Caucasian or African-American.
-Our data therefore has 5,723 rows (Figure \ref{fig:table snip}), with the below distributions of race (Figure \ref{fig:race plot}), age (Figure \ref{fig:age plot}), and two year recidivism rate (Figures \ref{fig:recid plot} & \ref{fig:recid race plot} and Table \ref{tab:recid table}).
+Our data therefore has 5,723 rows (Figure \ref{fig:table snip}), with the below distributions of race (Figure \ref{fig:race plot}), age (Figure \ref{fig:age plot}), and two year recidivism rate (Figures \ref{fig:recid plot} & \ref{fig:recid race plot}).
 
-```{r table snip, echo = FALSE, fig.align='center', fig.cap= 'This table is a snippet of the data set we will be using.', out.width= "100%"}
-knitr::include_graphics("../images/table_snippet.png")
+```{r table snip, echo = FALSE, fig.align='center', fig.cap= 'This table is a snippet of the data set we will be using.', out.width= "500px", out.height= "300px"}
+knitr::include_graphics("../images/table_snippet_new.png")
 ```
 
-```{r race plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of defendant races.', out.width= "100%"}
-knitr::include_graphics("../images/race_bar_plot_new.png")
+```{r race plot, echo=FALSE, fig.align='center', fig.cap= 'The race distribution for black people is 3000 and the distribution for white people is 2100.
+Therefore, there are more black people in the dataset than white people.', out.width= "500px", out.height= "300px"}
+knitr::include_graphics("../images/race_bar_plot.png")
 ```
 
-```{r age plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of defendant ages by race.', out.width= "100%"}
-knitr::include_graphics("../images/age_race_plot_new.png")
+```{r age plot, echo=FALSE, fig.align='center', fig.cap= "This plot is a Kernel Density Estimate (KDE) plot.
+The y-axis represents the KDE of the Probability Distribution Function of the age variable in our dataset.
+The probability of a defendant's age being between two points on the x-axis is the total shaded area of the curve under the two points.
+The purple dotted line shows the average age of Black defendants and the green dotted line shows the average age of white defendants.
+For both groups, the majority of defendants are relatively young, but this is especially noticeable for Black defendants.", out.width= "500px", out.height= "300px"}
+knitr::include_graphics("../images/race_age_plot.png")
 ```
 
-```{r recid plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes.', out.width= "100%"}
-knitr::include_graphics("../images/recid_bar_plot_new.png")
+```{r recid plot, echo=FALSE, fig.align='center', fig.cap= 'Of both black and white people, there is a higher percentage of people who did not recidivate than who did.
+2500 people recidivated, whereas approximately 2600 did not.', out.width= "500px", out.height= "300px"}
+knitr::include_graphics("../images/recid_bar_plot.png")
 ```
 
-```{r recid race plot,echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes by race.', out.width= "100%"}
-knitr::include_graphics("../images/race_recid_bar_plot_new.png")
-```
-
-```{r, include = FALSE}
-library(tidyverse)
-```
-
-```{r, include = FALSE}
-x <- tribble(~`Two Year Recidivism by Race`, ~`Recidivated`, ~`Survived`,
-             "African American", "1661", "1512",
-             "Caucasian", "822", "1278")
-```
-
-```{r, echo = FALSE}
-knitr::kable(x, caption = "This table shows the distribution of recidivism by race \\label{tab:recid table}")
+```{r recid race plot,echo=FALSE, fig.align='center', fig.cap= 'When we divide the data into Black and white defendants, we can see that Black defendants recidivate more than white defendants.
+Black defendants recidivate are more likely than not to recidivate.
+However, we can also see that there are more Black defendants in the dataset in general.', out.width= "500px", out.height= "300px"}
+knitr::include_graphics("../images/race_recid_plot.png")
 ```
@@
 
-For the next step of our experiment, we needed to determine where in the data science pipeline we can mitigate the most bias, using pre-processing, in-processing, and post-processing de-biasing algorithms.
+For the next step of our experiment, we need to determine where in the data science pipeline we can mitigate the most bias, using pre-processing, in-processing, and post-processing de-biasing algorithms.
 These are all based on using predictive models to figure out how we can “fix” the bias that is present.
@@
 With pre-processing, we use the re-weighing algorithm from AIF360 which weights the examples in each (group, label) combination differently to ensure fairness before classification [@aif360-oct-2018].
-We used a logistic regression model for this algorithm, as it is the easiest to interpret in the given context.
-After running the fairness metrics using the pre-processing algorithm, we were able to compare our results to the baseline metrics from the previous section.
+We use a logistic regression model for this algorithm, as it is the easiest to interpret in the given context.
+After running the fairness metrics using the pre-processing algorithm, we are able to compare our results to the baseline metrics from the previous section.
 
@@
 The technique we use is the Meta Classification algorithm, which accounts for the fairness metric as part of the input and returns a classifier optimized by that particular metric.
-In order to do this, we first needed to convert our data frame into a data type called a .
-Similarly to pre-processing, we compared the results of our in-processing methods with both the baseline and the pre-processing to gauge which method so far has better individual or group fairness.
+Similarly to pre-processing, we compare the results of our in-processing methods with both the baseline and the pre-processing to gauge which method so far has better individual or group fairness.
 
@@
 To get around this, we use the PostProcessingMeta class from AIF360 which combines the training and prediction for a random estimator and the post-processor while simultaneously splitting the data set.
-Based on these results, we looked comparatively at all the approaches to determine which we assessed to be the most fair.
+Based on these results, we look comparatively at all the approaches to determine which we assess to be the most fair.
 

### 2022-04-15 Paper_Draft2/Draft2.Rmd

--- 2022-04-15 Paper_Draft2/Draft2.Rmd
@@
   - name: Amrita Acharya
-    affil: 1, 2
+    affiliation: 1, 2
   - name: Dianne Caravela
-    affil: 1, 3
+    affiliation: 1, 3
   - name: Eunice Kim
-    affil: 1, 4
+    affiliation: 1, 4
   - name: Emma Kornberg
-    affil: 1, 5
+    affiliation: 1, 5
   - name: Elisabeth Nesmith
-    affil: 1, 6
+    affiliation: 1, 6
 affiliation:
@@
 
+```{r race plot, echo=FALSE, fig.align='center', fig.cap= 'The race distribution for black people is 3000 and the distribution for white people is 2100.
+Therefore, there are more black people in the dataset than white people.', out.width= "500px", out.height= "300px"}
+knitr::include_graphics("../images/race_bar_plot.png")
+```
+
+```{r age plot, echo=FALSE, fig.align='center', fig.cap= "This plot is a Kernel Density Estimate (KDE) plot.
+The y-axis represents the KDE of the Probability Distribution Function of the age variable in our dataset.
+The probability of a defendant's age being between two points on the x-axis is the total shaded area of the curve under the two points.
+The purple dotted line shows the average age of Black defendants and the green dotted line shows the average age of white defendants.
+For both groups, the majority of defendants are relatively young, but this is especially noticeable for Black defendants.", out.width= "500px", out.height= "300px"}
+knitr::include_graphics("../images/race_age_plot.png")
+```
+
+```{r recid plot, echo=FALSE, fig.align='center', fig.cap= 'Of both black and white people, there is a higher percentage of people who did not recidivate than who did.
+2500 people recidivated, whereas approximately 2600 did not.', out.width= "500px", out.height= "300px"}
+knitr::include_graphics("../images/recid_bar_plot.png")
+```
+
+```{r recid race plot,echo=FALSE, fig.align='center', fig.cap= 'When we divide the data into Black and white defendants, we can see that Black defendants recidivate more than white defendants.
+Black defendants recidivate are more likely than not to recidivate.
+However, we can also see that there are more Black defendants in the dataset in general.', out.width= "500px", out.height= "300px"}
+knitr::include_graphics("../images/race_recid_plot.png")
+
 ```{r race plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of defendant races.
@@
 
+AI Fairness 360 (AIF360) is an open-source Python toolkit that seeks to "to help facilitate the transition of fairness research algorithms to use in an industrial setting and to provide a common framework for fairness researchers to share and evaluate algorithms" [@aif360-oct-2018].
+
+helps facilitate the examination of algorithmic fairness. It contains multiple data sets, including the COMPAS data set that accompanied the Angwin et al. piece, and provides
+
 The AIF360 toolkit contains various group and individual fairness metrics as well as pre-processing, in-processing, and post-processing algorithms that we used for our experiments [@aif360-oct-2018].
@@
 Less than 1 means the privileged group (white defendants) benefits, while a value greater than 1 means the unprivileged group (Black defendants) benefits.
-According to AIF360, a ratio between .8 to 1.25 is considered fair:
+According to AIF360, a ratio between .8 to 1.25 is considered fair[@Ronaghan2019AI].
 
@@
 A value of 0 indicates equality of odds, and a value below 0 implies benefit for the privileged group.
-Equality of odds is achieved in the case of recidivism when the proportion of people who were predicted to recidivate and did recidivate is equal (true positive rate) for both Black and white defendants AND the proportion of people who were predicted to recidivate and did not recidivate (false positive rate) is equal for both Black and white defendants.
+Equality of odds is achieved in the case of recidivism when the proportion of people who were predicted to recidivate and did recidivate is equal (true positive rate) for both Black and white defendants AND the proportion of people who were predicted to recidivate and did not recidivate (false positive rate) is equal for both Black and white defendants[@aif360-oct-2018].
 
@@
 **Equal Opportunity Difference** - This metric is computed as the difference of true positive rates between the unprivileged and the privileged groups.
-The true positive rate is the ratio of true positives to the total number of actual positives for a given group.
+The true positive rate is the ratio of true positives to the total number of actual positives for a given group[@GoogleDev].
 
 The ideal value is 0. A value less than 0 implies higher benefit for the privileged group and a value greater than 0 implies higher benefit for the unprivileged group.
-Fairness for this metric is between -0.1 and 0.1.
-This metric is best used when it is very important to catch positive outcomes while false positives are not exceptionally problematic.
+Fairness for this metric is between -0.1 and 0.1 [@aif360-oct-2018].
+
+This metric is best used when it is very important to catch positive outcomes while false positives are not exceptionally problematic [@Cortez2019How].
 This is not the case for the COMPAS data set, as false positives mean extra jail time for someone who will not actually re-offend.
@@
 
-For the next step of our experiment, we needed to determine where in the data science pipeline we can mitigate the most bias, using pre-processing, in-processing, and post-processing de-biasing algorithms.
+For the next step of our experiment, we need to determine where in the data science pipeline we can mitigate the most bias, using pre-processing, in-processing, and post-processing de-biasing algorithms.
 These are all based on using predictive models to figure out how we can “fix” the bias that is present.
+
 Pre-processing refers to assessing the training data, and it is the most flexible method because it has not yet trained a model that may carry assumptions about the data.
@@
 Because the algorithms are often so complex and difficult to interpret, post-processors are trained in a black box.
-To get around this, we use the PostProcessingMeta class from AIF360 which combines the training and prediction for a random estimator and the post-processor while simultaneously splitting the data set.
-Based on these results, we looked comparatively at all the approaches to determine which we assessed to be the most fair.
+To get around this, we use the PostProcessingMeta class from AIF360 which combines the training and prediction for a random estimator and the post-processor while simultaneously splitting the data set [@aif360-oct-2018].
+Based on these results, we look comparatively at all the approaches to determine which we assess to be the most fair.
 
@@
 
-# Discussion
-
-# Conclusion
+                                 Result   Benefit   
+                                +---------------+
+Statistical Parity Difference   | -0.14 | white |
+Disparate Impact Ratio          | 0.47  | white |
+Average Odds Difference         | -0.44 | white |
+Equal Opportunity Difference    | -0.41 | white |
+                                +---------------+
 

### 2022-04-15 Paper_Draft2/Draft2.Rmd
finalizing draft 2

@@
 
-```{r race plot, echo=FALSE, fig.align='center', fig.cap= 'The race distribution for black people is 3000 and the distribution for white people is 2100.
-Therefore, there are more black people in the dataset than white people.', out.width= "500px", out.height= "300px"}
-knitr::include_graphics("../images/race_bar_plot.png")
-```
-
-```{r age plot, echo=FALSE, fig.align='center', fig.cap= "This plot is a Kernel Density Estimate (KDE) plot.
-The y-axis represents the KDE of the Probability Distribution Function of the age variable in our dataset.
-The probability of a defendant's age being between two points on the x-axis is the total shaded area of the curve under the two points.
-The purple dotted line shows the average age of Black defendants and the green dotted line shows the average age of white defendants.
-For both groups, the majority of defendants are relatively young, but this is especially noticeable for Black defendants.", out.width= "500px", out.height= "300px"}
-knitr::include_graphics("../images/race_age_plot.png")
-```
-
-```{r recid plot, echo=FALSE, fig.align='center', fig.cap= 'Of both black and white people, there is a higher percentage of people who did not recidivate than who did.
-2500 people recidivated, whereas approximately 2600 did not.', out.width= "500px", out.height= "300px"}
-knitr::include_graphics("../images/recid_bar_plot.png")
-```
-
-```{r recid race plot,echo=FALSE, fig.align='center', fig.cap= 'When we divide the data into Black and white defendants, we can see that Black defendants recidivate more than white defendants.
-Black defendants recidivate are more likely than not to recidivate.
-However, we can also see that there are more Black defendants in the dataset in general.', out.width= "500px", out.height= "300px"}
-knitr::include_graphics("../images/race_recid_plot.png")
-
 ```{r race plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of defendant races.
@@
 Out of all defendants, there is a higher proportion of people who did not recidivate than who did.
-About 2500 people recidivated, whereas approximately 2600 did not.', out.width= "100%"}
+About 2500 people recidivated, whereas approximately 2700 did not.', out.width= "100%"}
 knitr::include_graphics("../images/recid_bar_plot_new.png")
@@
 AI Fairness 360 (AIF360) is an open-source Python toolkit that seeks to "to help facilitate the transition of fairness research algorithms to use in an industrial setting and to provide a common framework for fairness researchers to share and evaluate algorithms" [@aif360-oct-2018].
-
-helps facilitate the examination of algorithmic fairness. It contains multiple data sets, including the COMPAS data set that accompanied the Angwin et al. piece, and provides
+It contains multiple data sets, including the COMPAS data set that accompanied the Angwin et al.
+piece.
 
@@
 
+With our baseline model, we ran the four different group fairness metrics we chose and compared the results (pictured in Table \ref{tab:baseline metrics table}))
+
 The statistical parity difference is -0.14.
@@
 
-                                 Result   Benefit   
-                                +---------------+
-Statistical Parity Difference   | -0.14 | white |
-Disparate Impact Ratio          | 0.47  | white |
-Average Odds Difference         | -0.44 | white |
-Equal Opportunity Difference    | -0.41 | white |
-                                +---------------+
+```{r, echo = FALSE}
+baseline_metrics <- tribble(~`Fairness Metric`, ~`Ideal Value`, ~ `Baseline Value`, ~`Benefited Group`,
+             "Statistical Parity Difference", "0", "-0.14", "White Defendants",
+             "Disparate Impact Ratio", "1", "0.47", "White Defendants",
+             "Average Odds Difference", "0", "-0.44", "White Defendants",
+             "Equal Opportunity Difference", "0", "-0.41", "White Defendants")
+```
+
+```{r, echo = FALSE}
+knitr::kable(baseline_metrics, caption = "Results of Baseline Fairness Metric Analysis \\label{tab:baseline metrics table}")
+```
 

### 2022-04-17 final_metrics_analysis.Rmd
messing with processing

--- 2022-04-14 final_metrics_analysis.Rmd
@@
 library(reticulate)
+
+# for Dianne
+use_condaenv("base")
+
+# for Amrita
+# use_condaenv("anaconda3")
 ```
@@
 green = '#30875c'
-orange = '#E7881E'
-blue = '#20A4CF'
 workshop_palette = [purple, green]
 df_viz.head()
+```
+
+```{python average white and black defendant}
+#av_defendant = (df_viz.groupby(['race'],as_index=False).age.mean()
 ```
@@
 # barplot of race
-sns.countplot(x="race", data=df_viz, palette=[blue, orange])
+sns.countplot(y="race", data=df_viz, palette=["#30875c"])
 plt.title('Race Distribution (White and Black individuals only)')
@@
 # age distribution by race
-ax = sns.kdeplot(x="age", hue="race", data=df_viz);
-kdeline = ax.lines[0]
-mean_fem = df_viz.groupby('race').age.median()[0]
-xs = kdeline.get_xdata()
-ys = kdeline.get_ydata()
-height_fem = np.interp(mean_fem, xs, ys)
-ax.vlines(mean_fem, 0, height_fem, color=blue, ls=':')
+#ax = sns.kdeplot(x="age", hue="race", data=df_viz);
+#kdeline = ax.lines[0]
+#mean_fem = df_viz.groupby('race').age.median()[0]
+#xs = kdeline.get_xdata()
+#ys = kdeline.get_ydata()
+#height_fem = np.interp(mean_fem, xs, ys)
+#ax.vlines(mean_fem, 0, height_fem, color=purple, ls=':')
 
-mean_m = df_viz.groupby('race').age.median()[1]
-height_m = np.interp(mean_m, xs, ys)
-ax.vlines(mean_m, 0, height_m, color=orange, ls=':')
+#mean_m = df_viz.groupby('race').age.median()[1]
+#height_m = np.interp(mean_m, xs, ys)
+#ax.vlines(mean_m, 0, height_m, color=green, ls=':')
 
-print(df_viz.groupby('race').age.median())
+#print(df_viz.groupby('race').age.median())
 
-plt.title('Distribution of Age by Race')
+#plt.title('Distribution of Age by Race')
 ```

### 2022-04-17 Paper_Draft2/Draft2.tex

@@
 % Full title of the paper (Capitalized)
-\Title{Does the COMPAS Needle Always Point Towards Equity? Finding
-Fairness in the COMPAS Risk Assessment Algorithm: A Case Study}
+\Title{Does the COMPAS Needle Always Point Towards Equity? Finding Fairness in
+the COMPAS Risk Assessment Algorithm: A Case Study}
 
 % Authors, for the paper (add full first names)
-\Author{Amrita Acharya$^{1, 2}$, Dianne Caravela$^{1, 3}$, Eunice
-Kim$^{1, 4}$, Emma Kornberg$^{1, 5}$, Elisabeth Nesmith$^{1, 6}$}
+\Author{Amrita Acharya$^{}$, Dianne Caravela$^{}$, Eunice Kim$^{}$, Emma Kornberg$^{}$, Elisabeth Nesmith$^{}$}
 
 % Authors, for metadata in PDF
-\AuthorNames{Amrita Acharya, Dianne Caravela, Eunice Kim, Emma
-Kornberg, Elisabeth Nesmith}
+\AuthorNames{Amrita Acharya, Dianne Caravela, Eunice Kim, Emma Kornberg, Elisabeth Nesmith}
 
@@
 Out of all defendants, there is a higher proportion of people who did not recidivate than who did.
-About 2500 people recidivated, whereas approximately 2600 did not.}\label{fig:recid plot}
+About 2500 people recidivated, whereas approximately 2700 did not.}\label{fig:recid plot}
 \end{figure}
@@
 \toprule
-Two Year Recidivism by Race & Recidivated & Survived \\
+Two Year Recidivism by Race & Recidivated & Survived\tabularnewline
 \midrule
@@
 \toprule
-Two Year Recidivism by Race & Recidivated & Survived \\
+Two Year Recidivism by Race & Recidivated & Survived\tabularnewline
 \midrule
 \endhead
-African American & 1661 & 1512 \\
-Caucasian & 822 & 1278 \\
+African American & 1661 & 1512\tabularnewline
+Caucasian & 822 & 1278\tabularnewline
 \bottomrule
@@
 \section{Methods}\label{methods}}
+
+AI Fairness 360 (AIF360) is an open-source Python toolkit that seeks to
+``to help facilitate the transition of fairness research algorithms to
+use in an industrial setting and to provide a common framework for
+fairness researchers to share and evaluate algorithms''
+\citep{aif360-oct-2018}. It contains multiple data sets, including the
+COMPAS data set that accompanied the Angwin et al.~piece.
 
@@
 (Black defendants) benefits. According to AIF360, a ratio between .8 to
-1.25 is considered fair:
+1.25 is considered fair\citep{Ronaghan2019AI}.
 
@@
 recidivate (false positive rate) is equal for both Black and white
-defendants.
+defendants\citep{aif360-oct-2018}.
 
@@
 privileged groups. The true positive rate is the ratio of true positives
-to the total number of actual positives for a given group.
+to the total number of actual positives for a given
+group\citep{GoogleDev}.
 
@@
 privileged group and a value greater than 0 implies higher benefit for
-the unprivileged group. Fairness for this metric is between -0.1 and
-0.1. This metric is best used when it is very important to catch
-positive outcomes while false positives are not exceptionally
-problematic. This is not the case for the COMPAS data set, as false
-positives mean extra jail time for someone who will not actually
+the unprivileged group. Fairness for this metric is between -0.1 and 0.1
+\citep{aif360-oct-2018}.
+
+This metric is best used when it is very important to catch positive
+outcomes while false positives are not exceptionally problematic
+\citep{Cortez2019How}. This is not the case for the COMPAS data set, as
+false positives mean extra jail time for someone who will not actually
 re-offend.
@@
 
-For the next step of our experiment, we needed to determine where in the
+For the next step of our experiment, we need to determine where in the
 data science pipeline we can mitigate the most bias, using
@@
 algorithms. These are all based on using predictive models to figure out
-how we can ``fix'' the bias that is present. Pre-processing refers to
-assessing the training data, and it is the most flexible method because
-it has not yet trained a model that may carry assumptions about the
-data. It is important to keep in mind that pre-processing prevents
-assumptions in the modeling, but does not account for the bias in data
-collection. With pre-processing, we use the re-weighing algorithm from
-AIF360 which weights the examples in each (group, label) combination
-differently to ensure fairness before classification
-\citep{aif360-oct-2018}. We used a logistic regression model for this
-algorithm, as it is the easiest to interpret in the given context. After
-running the fairness metrics using the pre-processing algorithm, we were
-able to compare our results to the baseline metrics from the previous
-section.
+how we can ``fix'' the bias that is present.
+
+Pre-processing refers to assessing the training data, and it is the most
+flexible method because it has not yet trained a model that may carry
+assumptions about the data. It is important to keep in mind that
+pre-processing prevents assumptions in the modeling, but does not
+account for the bias in data collection. With pre-processing, we use the
+re-weighing algorithm from AIF360 which weights the examples in each
+(group, label) combination differently to ensure fairness before
+classification \citep{aif360-oct-2018}. We use a logistic regression
+model for this algorithm, as it is the easiest to interpret in the given
+context. After running the fairness metrics using the pre-processing
+algorithm, we are able to compare our results to the baseline metrics
+from the previous section.
 
@@
 part of the input and returns a classifier optimized by that particular
-metric. In order to do this, we first needed to convert our data frame
-into a data type called a . Similarly to pre-processing, we compared the
-results of our in-processing methods with both the baseline and the
-pre-processing to gauge which method so far has better individual or
-group fairness.
+metric. Similarly to pre-processing, we compare the results of our
+in-processing methods with both the baseline and the pre-processing to
+gauge which method so far has better individual or group fairness.
 
@@
 estimator and the post-processor while simultaneously splitting the data
-set. Based on these results, we looked comparatively at all the
-approaches to determine which we assessed to be the most fair.
+set \citep{aif360-oct-2018}. Based on these results, we look
+comparatively at all the approaches to determine which we assess to be
+the most fair.
 
@@
 \section{Preliminary Results}\label{preliminary-results}}
+
+With our baseline model, we ran the four different group fairness
+metrics we chose and compared the results (pictured in Table
+\ref{tab:baseline metrics table}))
 
@@
 
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
+\begin{longtable}[]{@{}llll@{}}
+\caption{Results of Baseline Fairness Metric Analysis
+\label{tab:baseline metrics table}}\tabularnewline
+\toprule
+Fairness Metric & Ideal Value & Baseline Value & Benefited
+Group\tabularnewline
+\midrule
+\endfirsthead
+\toprule
+Fairness Metric & Ideal Value & Baseline Value & Benefited
+Group\tabularnewline
+\midrule
+\endhead
+Statistical Parity Difference & 0 & -0.14 & White
+Defendants\tabularnewline
+Disparate Impact Ratio & 1 & 0.47 & White Defendants\tabularnewline
+Average Odds Difference & 0 & -0.44 & White Defendants\tabularnewline
+Equal Opportunity Difference & 0 & -0.41 & White
+Defendants\tabularnewline
+\bottomrule
+\end{longtable}
 

### 2022-04-17 final_metrics_analysis.Rmd
adding EDA chunk for median defendant

--- 2022-04-15 final_metrics_analysis.Rmd
@@
 output: html_document
+editor_options: 
+  chunk_output_type: console
 ---
@@
 y_new.head()
+```
+
+```{python EDA}
+X_new_index = X_new.rename(columns={"race": "def_race"})
+X_new_index = X_new_index.rename(columns={"sex": "def_sex"})
+# renaming dataframe columns to avoid the ValueError of variables being a column AND index label (which is ambiguous)
+
+X_new_index.groupby(["def_race"])["age"].median()
+X_new_index.groupby(["def_race", "sex"]).size()
+X_new_index.groupby(["def_race", "c_charge_degree"]).size()
+X_new_index.groupby(["def_race"])["priors_count"].median()
+#X_new_index.groupby(["def_race"])["juv_fel_count"].median()
+#X_new_index.groupby(["def_race"])["juv_misd_count"].median()
+
 ```

### 2022-04-18 final_metrics_analysis.Rmd
beginning to work on plot about felonies/misdemeanors for black/white defendants

--- 2022-04-15 final_metrics_analysis.Rmd
@@
 df_viz.head()
+```
+
+```{python make degree numeric}
+#av_defendant = (df_viz.groupby(['race'],as_index=False).age.mean()
+#display(df_viz)
+#list_vars = list(df_viz)
+#list_vars
+#df_numeric_charges = df_viz['c_charge_degree'].replace(['M', 'F'], [0, 1], inplace=True)
+#plt.bar('c_charge_degree', 1, data = df_viz)
 ```

### 2022-04-18 final_metrics_analysis.Rmd

--- 2022-04-17 final_metrics_analysis.Rmd
@@
 #X_new_index.groupby(["def_race"])["juv_misd_count"].median()
+# above two lines commented out since medians for everyone = 0
 

### 2022-04-18 final_metrics_analysis.Rmd
amrita's changes

--- 2022-04-18 final_metrics_analysis.Rmd
@@
 library(reticulate)
-
-# for Dianne
-use_condaenv("base")
-
-# for Amrita
-# use_condaenv("anaconda3")
+use_condaenv("anaconda3")
 ```
@@
 
-```{python make degree numeric}
-#av_defendant = (df_viz.groupby(['race'],as_index=False).age.mean()
-#display(df_viz)
-#list_vars = list(df_viz)
-#list_vars
-#df_numeric_charges = df_viz['c_charge_degree'].replace(['M', 'F'], [0, 1], inplace=True)
-#plt.bar('c_charge_degree', 1, data = df_viz)
-```
-
 ```{python exploratory recidivism bar plot}
@@
 # age distribution by race
-ax = sns.kdeplot(x="age", hue="race", data=df_viz, palette=workshop_palette);
-kdeline = ax.lines[0]
-mean_black = df_viz.groupby('race').age.median()[0]
-xs = kdeline.get_xdata()
-ys = kdeline.get_ydata()
-height_black = np.interp(mean_black, xs, ys)
-ax.vlines(mean_black, 0, height_black, color=purple, ls=':')
+#ax = sns.kdeplot(x="age", hue="race", data=df_viz, palette=workshop_palette);
+#kdeline = ax.lines[0]
+#mean_black = df_viz.groupby('race').age.median()[0]
+#xs = kdeline.get_xdata()
+#ys = kdeline.get_ydata()
+#height_black = np.interp(mean_black, xs, ys)
+#ax.vlines(mean_black, 0, height_black, color=purple, ls=':')
 
-mean_white = df_viz.groupby('race').age.median()[1]
-height_white = np.interp(mean_white, xs, ys)
-ax.vlines(mean_white, 0, height_white, color=green, ls=':')
+#mean_white = df_viz.groupby('race').age.median()[1]
+#height_white = np.interp(mean_white, xs, ys)
+#ax.vlines(mean_white, 0, height_white, color=green, ls=':')
 
-print(df_viz.groupby('race').age.median())
+#print(df_viz.groupby('race').age.median())
 
-plt.title('Distribution of Age by Race')
+#plt.title('Distribution of Age by Race')
 ```
@@
 ```{python post-processing calibrated equal odds set up, include = FALSE}
+from aif360.algorithms.postprocessing import EqOddsPostprocessing
 from aif360.sklearn.postprocessing import CalibratedEqualizedOdds, PostProcessingMeta
@@
 plt.show()
+```
+
+```{python} 
+ep = EqOddsPostprocessing(unprivileged_groups=0, privileged_groups= 1, seed =1234567)
+eo = PostProcessingMeta(estimator=lr, postprocessor=ep, random_state=1234567)
+eo.fit(train_BLD, test_BLD)
+y_pred_EO = eo.predict(test_BLD)
+#y_proba_EO = eo.predict_(test_BLD)
+
 ```

### 2022-04-18 final_metrics_analysis.Rmd
exploratory priors by race

--- 2022-04-18 final_metrics_analysis.Rmd
@@
 library(reticulate)
-
-# for Dianne
-use_condaenv("base")
-
-# for Amrita
-# use_condaenv("anaconda3")
 ```
@@
 pd.crosstab(index = df_viz["race"], columns = df_viz["two_year_recid"])
+```
+
+```{python exploratory priors by race}
+priors_by_race = sns.countplot(x="race", hue="priors_count", data=df_viz, palette=workshop_palette)
+
+plt.title('Priors Count by Race')
+plt.show()
 ```

### 2022-04-18 final_metrics_analysis.Rmd

--- 2022-04-18 final_metrics_analysis.Rmd
@@
 y_new.head()
+```
+
+```{python EDA}
+X_new_index = X_new.rename(columns={"race": "def_race"})
+X_new_index = X_new_index.rename(columns={"sex": "def_sex"})
+# renaming dataframe columns to avoid the ValueError of variables being a column AND index label (which is ambiguous)
+
+X_new_index.groupby(["def_race"])["age"].median()
+X_new_index.groupby(["def_race", "sex"]).size()
+X_new_index.groupby(["def_race", "c_charge_degree"]).size()
+X_new_index.groupby(["def_race"])["priors_count"].median()
+#X_new_index.groupby(["def_race"])["juv_fel_count"].median()
+#X_new_index.groupby(["def_race"])["juv_misd_count"].median()
+# above two lines commented out since medians for everyone = 0
+
 ```

### 2022-04-18 final_metrics_analysis.Rmd
expl priors race

@@
 ```{python exploratory priors by race}
-priors_by_race = sns.countplot(x="race", hue="priors_count", data=df_viz, palette=workshop_palette)
+#priors_by_race = sns.countplot(x="race", hue="priors_count", data=df_viz, palette=workshop_palette)
 
-plt.title('Priors Count by Race')
-plt.show()
+#plt.title('Priors Count by Race')
+#plt.show()
 ```

### 2022-04-18 final_metrics_analysis.Rmd
adding in processing methods

--- 2022-04-18 final_metrics_analysis.Rmd
@@
 output: html_document
-editor_options: 
-  chunk_output_type: console
 ---
@@
 from aif360.sklearn.preprocessing import ReweighingMeta
-#from aif360.sklearn.inprocessing import 
 from aif360.sklearn.postprocessing import CalibratedEqualizedOdds, PostProcessingMeta
@@
 y_new.head()
-```
-
-```{python EDA}
-X_new_index = X_new.rename(columns={"race": "def_race"})
-X_new_index = X_new_index.rename(columns={"sex": "def_sex"})
-# renaming dataframe columns to avoid the ValueError of variables being a column AND index label (which is ambiguous)
-
-X_new_index.groupby(["def_race"])["age"].median()
-X_new_index.groupby(["def_race", "sex"]).size()
-X_new_index.groupby(["def_race", "c_charge_degree"]).size()
-X_new_index.groupby(["def_race"])["priors_count"].median()
-#X_new_index.groupby(["def_race"])["juv_fel_count"].median()
-#X_new_index.groupby(["def_race"])["juv_misd_count"].median()
-# above two lines commented out since medians for everyone = 0
-
 ```
@@
 pd.crosstab(index = df_viz["race"], columns = df_viz["two_year_recid"])
-```
-
-```{python exploratory priors by race}
-priors_by_race = sns.countplot(x="race", hue="priors_count", data=df_viz, palette=workshop_palette)
-
-plt.title('Priors Count by Race')
-plt.show()
 ```
@@
 
-# In-Processing
+# In-Processing Approaches
 
@@
 
-## Prejudice Remover
-```{r}
-#prej = PrejudiceRemover(eta=1.0, sensitive_attr='race', class_attr='two_year_recid', seed=1234567)
-#prej.fit(train_BLD)
-#y_pred_PREJ = meta.predict(test_BLD)
-#y_pred_PREJ = y_pred_PREJ.labels.flatten()
+## Adversarial Debiasing
+```{python adversarial debiasing}
+# from aif360.algorithms.inprocessing import AdversarialDebiasing
+# import tensorflow as tf
+# 
+# adv_deb = AdversarialDebiasing(unpr)
+# #adv_deb.fit(X_train, y_train)
+# #y_pred_AD = adv_deb.predict(X_test)
+# #adv_deb.sess_.close()
 ```
 
-## Meta Fair Classifier 
-```{python in-processing meta}
-#meta = MetaFairClassifier(tau=0.8, sensitive_attr='race', type='sr', seed=1234567)
-#meta.fit(train_BLD)
-#y_pred_META = meta.predict(test_BLD)
-#y_pred_META = y_pred_META.labels.flatten()
+## Prejudice Remover
+```{python prejudice remover}
+from aif360.algorithms.inprocessing import PrejudiceRemover
+
+prej = PrejudiceRemover(eta = 1.0, sensitive_attr='race', class_attr='two_year_recid')
+prej.fit(train_BLD)
+y_pred_PREJ = prej.predict(test_BLD)
+y_pred_PREJ = y_pred_PREJ.labels.flatten()
 ```
 
-```{python in-processing accuracy score}
-# acc_AD = accuracy_score(y_test, y_pred_META)
-# print(f'[ Meta-Classification] The test accuracy of the algorithm is: {acc_AD: .2%}')
+```{python prejudice accuracy score}
+acc_PREJ = accuracy_score(y_test, y_pred_PREJ)
+print(f'[Prejudice Remover] The test accuracy of the algorithm is: {acc_PREJ: .2%}')
+```
+
+```{python prejudice remover confusion matrix}
+cf_matrix = confusion_matrix(y_test, y_pred_PREJ)
+make_confusion_matrix(cf_matrix, "[Prejudice Remover]")
+plt.show()
+```
+
+```{python prejudice remover metrics per group}
+metrics_per_group(y_test, y_pred_PREJ)
+plt.show()
+```
+
+### Statistical Parity Difference
+
+```{python prejudice remover statistical parity difference}
+stat_par_diff_PREJ = statistical_parity_difference(y_test, y_pred_PREJ, prot_attr='race', pos_label = 0)
+
+print(f'[Prejudice Remover] The statistical parity difference is {stat_par_diff_PREJ: .2}')
+```
+
+### Equal Opportunity Difference
+
+```{python prejudice remover equal opportunity difference}
+eq_opp_diff_PREJ = equal_opportunity_difference(y_test, y_pred_PREJ, prot_attr='race', pos_label = 0)
+
+print(f'[Prejudice Remover] The equal opportunity difference is {eq_opp_diff_PREJ: .2}')
+```
+
+### Average Odds Difference
+
+```{python prejudice remover average odds difference}
+avg_odds_diff_PREJ = average_odds_difference(y_test, y_pred_PREJ, prot_attr='race', pos_label = 0)
+
+print(f'[Prejudice Remover] The average odds difference is {avg_odds_diff_PREJ: .2}')
+```
+
+### Disparate Impact Ratio
+```{python prejudice remover disparate impact ratio}
+disp_impact_ratio_PREJ = disparate_impact_ratio(y_test, y_pred_PREJ, prot_attr='race', priv_group = 1, pos_label = 0)
+
+print(f'[Prejudice Remover] The disparate impact ratio is {disp_impact_ratio_PREJ: .2}')
+```
+
+### Plot Fairness Metrics (Prejudice Remover)
+
+```{python prejudice remover fairness metrics plot}
+plot_fair_metrics([stat_par_diff_PREJ, eq_opp_diff_PREJ, avg_odds_diff_PREJ, disp_impact_ratio_PREJ], 'Prejudice Remover')
+plt.show()
 ```

### 2022-04-18 final_metrics_analysis.Rmd

@@
 library(reticulate)
-use_condaenv("anaconda3")
+
+# for Dianne
+use_condaenv("base")
+
+# for Amrita
+# use_condaenv("anaconda3")
+#use_condaenv("anaconda3")
 ```

### 2022-04-18 final_metrics_analysis.Rmd
moving median defendant EDA chunk

--- 2022-04-18 final_metrics_analysis.Rmd
@@
 library(reticulate)
-use_condaenv("anaconda3")
+#use_condaenv("anaconda3")
 ```
@@
 # Exploratory Data Analysis
+
+```{python EDA}
+X_new_index = X_new.rename(columns={"race": "def_race"})
+X_new_index = X_new_index.rename(columns={"sex": "def_sex"})
+# renaming dataframe columns to avoid the ValueError of variables being a column AND index label (which is ambiguous)
+
+X_new_index.groupby(["def_race"])["age"].median()
+X_new_index.groupby(["def_race", "sex"]).size()
+X_new_index.groupby(["def_race", "c_charge_degree"]).size()
+X_new_index.groupby(["def_race"])["priors_count"].median()
+#X_new_index.groupby(["def_race"])["juv_fel_count"].median()
+#X_new_index.groupby(["def_race"])["juv_misd_count"].median()
+# above two lines commented out since medians for everyone = 0
+
+```
+
 ```{python exploratory visualization set up}
@@
 ```{python} 
-ep = EqOddsPostprocessing(unprivileged_groups=0, privileged_groups= 1, seed =1234567)
-eo = PostProcessingMeta(estimator=lr, postprocessor=ep, random_state=1234567)
-eo.fit(train_BLD, test_BLD)
-y_pred_EO = eo.predict(test_BLD)
+#ep = EqOddsPostprocessing(unprivileged_groups=0, privileged_groups= 1, seed =1234567)
+#eo = PostProcessingMeta(estimator=lr, postprocessor=ep, random_state=1234567)
+#eo.fit(train_BLD, test_BLD)
+#y_pred_EO = eo.predict(test_BLD)
 #y_proba_EO = eo.predict_(test_BLD)

### 2022-04-19 Paper_Draft2/Draft2.tex
Trying to fix yaml author affil

@@
 % Full title of the paper (Capitalized)
-\Title{Does the COMPAS Needle Always Point Towards Equity? Finding Fairness in
-the COMPAS Risk Assessment Algorithm: A Case Study}
+\Title{Does the COMPAS Needle Always Point Towards Equity? Finding
+Fairness in the COMPAS Risk Assessment Algorithm: A Case Study}
 
 % Authors, for the paper (add full first names)
-\Author{Amrita Acharya$^{}$, Dianne Caravela$^{}$, Eunice Kim$^{}$, Emma Kornberg$^{}$, Elisabeth Nesmith$^{}$}
+\Author{Amrita Acharya$^{1}$, Dianne Caravela$^{1}$, Eunice
+Kim$^{1}$, Emma Kornberg$^{1}$, Elisabeth Nesmith$^{1}$}
 
 % Authors, for metadata in PDF
-\AuthorNames{Amrita Acharya, Dianne Caravela, Eunice Kim, Emma Kornberg, Elisabeth Nesmith}
+\AuthorNames{Amrita Acharya, Dianne Caravela, Eunice Kim, Emma
+Kornberg, Elisabeth Nesmith}
 
@@
 \address{%
-$^{1}$ \quad Statistical and Data Sciences Smith College Northampton, MA 01063; \\
-$^{2}$ \quad ; \href{mailto:aacharya@smith.edu}{\nolinkurl{aacharya@smith.edu}}\\
-$^{3}$ \quad ; \href{mailto:dcaravela@smith.edu}{\nolinkurl{dcaravela@smith.edu}}\\
-$^{4}$ \quad ; \href{mailto:ekim89@smith.edu}{\nolinkurl{ekim89@smith.edu}}\\
-$^{5}$ \quad ; \href{mailto:ekornberg@smith.edu}{\nolinkurl{ekornberg@smith.edu}}\\
-$^{6}$ \quad ; \href{mailto:enesmith@smith.edu}{\nolinkurl{enesmith@smith.edu}}\\
+$^{1}$ \quad Statistical and Data Sciences Smith College Northampton, MA
+01063; \href{mailto:aarcharya@smith.edu}{\nolinkurl{aarcharya@smith.edu}},
+\href{mailto:dcaravela@smith.edu}{\nolinkurl{dcaravela@smith.edu}},
+\href{mailto:ekim89@smith.edu}{\nolinkurl{ekim89@smith.edu}},
+\href{mailto:ekornberg@smith.edu}{\nolinkurl{ekornberg@smith.edu}},
+\href{mailto:enesmith@smith.edu}{\nolinkurl{enesmith@smith.edu}}\\
 }
 % Contact information of the corresponding author
-\corres{Correspondence: }
+\corres{Correspondence: \href{mailto:bbaumer@smith.edu}{\nolinkurl{bbaumer@smith.edu}}}
 
@@
 \toprule
-Fairness Metric & Ideal Value & Baseline Value & Benefited
-Group\tabularnewline
+Fairness Metric & Ideal Value & Baseline Value & Benefited Group \\
 \midrule
@@
 \toprule
-Fairness Metric & Ideal Value & Baseline Value & Benefited
-Group\tabularnewline
+Fairness Metric & Ideal Value & Baseline Value & Benefited Group \\
 \midrule
 \endhead
-Statistical Parity Difference & 0 & -0.14 & White
-Defendants\tabularnewline
-Disparate Impact Ratio & 1 & 0.47 & White Defendants\tabularnewline
-Average Odds Difference & 0 & -0.44 & White Defendants\tabularnewline
-Equal Opportunity Difference & 0 & -0.41 & White
-Defendants\tabularnewline
+Statistical Parity Difference & 0 & -0.14 & White Defendants \\
+Disparate Impact Ratio & 1 & 0.47 & White Defendants \\
+Average Odds Difference & 0 & -0.44 & White Defendants \\
+Equal Opportunity Difference & 0 & -0.41 & White Defendants \\
 \bottomrule

### 2022-04-19 Paper_Draft2/Draft2.Rmd

@@
   - name: Amrita Acharya
-    affiliation: 1, 2
+    affil: 1
   - name: Dianne Caravela
-    affiliation: 1, 3
+    affil: 1
   - name: Eunice Kim
-    affiliation: 1, 4
+    affil: 1
   - name: Emma Kornberg
-    affiliation: 1, 5
+    affil: 1
   - name: Elisabeth Nesmith
-    affiliation: 1, 6
+    affil: 1
 affiliation:
@@
       Northampton, MA 01063
-  - num: 2
-    email: aacharya@smith.edu
-  - num: 3
-    email: dcaravela@smith.edu
-  - num: 4
-    email: ekim89@smith.edu
-  - num: 5
-    email: ekornberg@smith.edu
-  - num: 6
-    email: enesmith@smith.edu
+    email: aarcharya@smith.edu, dcaravela@smith.edu, ekim89@smith.edu, ekornberg@smith.edu, enesmith@smith.edu
+
 journal: water
@@
 classoption: fleqn
+correspondence: |
+  bbaumer@smith.edu
 

### 2022-04-19 final_metrics_analysis.Rmd
priors count

@@
 # Exploratory Data Analysis
-
-```{python EDA}
-X_new_index = X_new.rename(columns={"race": "def_race"})
-X_new_index = X_new_index.rename(columns={"sex": "def_sex"})
-# renaming dataframe columns to avoid the ValueError of variables being a column AND index label (which is ambiguous)
-
-X_new_index.groupby(["def_race"])["age"].median()
-X_new_index.groupby(["def_race", "sex"]).size()
-X_new_index.groupby(["def_race", "c_charge_degree"]).size()
-X_new_index.groupby(["def_race"])["priors_count"].median()
-#X_new_index.groupby(["def_race"])["juv_fel_count"].median()
-#X_new_index.groupby(["def_race"])["juv_misd_count"].median()
-# above two lines commented out since medians for everyone = 0
-
-```
-
 ```{python exploratory visualization set up}
@@
 pd.crosstab(index = df_viz["race"], columns = df_viz["two_year_recid"])
+```
+
+```{python exploratory priors by race}
+priors_by_race = sns.countplot(x="race", color="priors_count", data=df_viz, palette=workshop_palette)
+
+plt.title('Priors Count by Race')
+plt.show()
 ```

### 2022-04-19 final_metrics_analysis.Rmd
priors by sex

@@
 
+Of both Black and white people, there is a higher percentage of people who did not recidivate than who did. 2500 people recidivated, whereas approximately 2600 did not.
+
 ```{python exploratory race bar plot}
@@
 ```
+
+The race distribution for black people is 3000 and the distribution for white people is 2100. Therefore, there are more black people in the dataset than white people.
 
@@
 
+When we divide the data into Black and white defendants, we can see that Black defendants recidivate more than white defendants and Black defendants recidivate more than not recidivate.
+However, we can also see that there are more Black defendants in the data set in general.
+
+Black defendants are more likely to recidivate than not recidivate.
+This increased likelihood of recidivating may be attributed to the systemic racism in policing (e.g., predominantly Black neighborhoods tend to be overpoliced in comparison to predominantly white neighborhoods").
+
 ```{python exploratory priors by race}
@@
 plt.title('Priors Count by Race')
+plt.show()
+```
+
+```{python exploratory priors by sex}
+priors_by_race = sns.countplot(x="sex", color="priors_count", data=df_viz, palette=workshop_palette)
+
+plt.title('Priors Count by Sex')
 plt.show()
@@
 
+The highest percentage is for true negatives, those guessed to not recidivate and actually did not, at 35.41%.
+Similarly, the true positive rate is 25.85%.
+Our false negative and positive rates, incorrect guesses, is high for this baseline model, indicating that debiasing intervention may be necessary to lower these rates.
+
 ```{python baseline metrics per group}
@@
 In the Black population, the model correctly identifies 70% of people who recidivated. In the white population, the model correctly identifies 20% of people who recidivated.
+
+White people have a lower recall rate, meaning that there is a low true positive rate.
+Whereas for black people the true positive rate is much higher: there's a high percentage of recidivated guesses that were correct.
 
@@
 ```
+
+The highest percentage in this confusion matrix is for the true negative rate, those guessed to not recidivate and actually did not, at 36.24%, 0.83% higher than the true negatives in the baseline model.
+This means that the re-weighing model marginally improved the baseline model accuracy.
+The true positive rate, or those who were predicted to recidivate and did recidivate, is 24.72%, 1.13% lower than that of the baseline model.
+This means that the re-weighing model slightly lowered the accuracy of the baseline model.
+The false negative and false positive rates, or the predictions that proved to be incorrect, are 23.20% and 15.85%, respectively.
+The re-weighing model lowered the accuracy of the false negative rate by 1.14% and improved the accuracy of the false positive rate by 0.83%.
+We are most concerned about the false positivesthat mean that defendants who are predicted to recidivate and do not actually recidivate.
+While there is a slight accuracy improvement in the false positive rate, it is still relatively high at 15.85%, indicating that we should further precede with additional bias mitigation techniques.
 
@@
 
-```{python} 
+```{python post-processing equal odds needs fix} 
 #ep = EqOddsPostprocessing(unprivileged_groups=0, privileged_groups= 1, seed =1234567)

### 2022-04-19 final_metrics_analysis.Rmd
priors captions

@@
 ```
+
+The two Priors Count bar charts show the differences in the distribution of priors by race and sex.
+If a defendant has a higher number of prior offenses, the individual is more likely to be rated with a higher risk of recidivism.
+In the former plot, one can see that Black defendants have a higher number of priors than do white defendants.
+This again likely reflects the over-policing of Black neighborhoods.
+In the latter plot, one can see that male defendants have a higher number of priors than do female defendants.
+This is worth noting because the COMPAS risk assessment algorithm has been found to express a discrepancy in prediction between sexes.
+When the recidivsim predictions are compared with the true recividism outcomes, female defendants have a notable false positive rate that predicts their recidivism risk as higher than their true outcomes.
 

### 2022-04-19 final_metrics_analysis.Rmd
adding baseline plot interpretations

--- 2022-04-19 final_metrics_analysis.Rmd
@@
 
-## defining function for displaying metrics of training and test data by sex 
+## defining function for displaying metrics of training and test data by race
 def metrics_per_group(y_test, y_pred):
@@
 
+This plot shows the distribution of two year recidivism outcomes.
+Out of all defendants, there is a higher proportion of people who did not recidivate than who did.
+About 2500 people recidivated, whereas approximately 2700 did not.
+
 ```{python exploratory race bar plot}
@@
 ```
+
+This plot shows the distribution of defendant races.
+The dataset contains about 3200 Black defendants and 2100 white defendants.
+Therefore, there are more Black people represented in the dataset than white people, showing the disproportionate policing and criminalization of Black communities.
 
@@
 
+This plot shows the distribution of defendant ages by race.
+The purple curve shows the distribution of the ages of Black defendants, and the green curve shows the distribution of the ages of white defendants.
+The probability of a defendant's age being between two points on the x-axis is the total shaded area of the curve under the two points.
+The purple dotted line represents the median age of Black defendants (29 years) and the green dotted line represents the median age of white defendants (35 years).
+For both groups, the majority of defendants are relatively young, but this is especially noticeable for Black defendants.
+
 ```{python exploratory recidivism by race plot}
@@
 
+This plot shows the distribution of two year recidivism outcomes by race.
+When we divide the data into Black and white defendants, we can see that Black defendants recidivate more than white defendants and Black defendants are more likely to recidivate than not recidivate.
+Again, this trend points to disproportionate rates of criminalization of Black people.
+
 ```{python exploratory recidivism by race table}
@@
 
-```{python exploratory priors by race}
-priors_by_race = sns.countplot(x="race", color="priors_count", data=df_viz, palette=workshop_palette)
-
-plt.title('Priors Count by Race')
-plt.show()
-```
+This table shows the counts of recidivism by race.
 
@@
 
+The highest percentage in this confusion matrix is for true negatives, defendants who the model predicted to not recidivate and actually did not, at 35.41%.
+The number of true positives, defendants who the model predicted to recidivate and actually recidivated, is slightly lower at 25.85%.
+The percentages of false negatives and false positives, which represent incorrect model predictions, are somewhat high for this baseline model.
+This indicates that the model might need bias mitigation or improvements in accuracy.
+
 ```{python baseline metrics per group}
@@
 
-In the Black population, the model correctly identifies 70% of people who recidivated. In the white population, the model correctly identifies 20% of people who recidivated.
+This graphs shows three different metrics of model quality and how they differ by race.
+Accuracy is calculated by summing the total number of true positives and true negatives and dividing by the total number of predictions:
+
+$$\frac{TP + TN}{TP + FP + TN + FN} $$
+
+Accuracy, as the name suggests, quantifies how accurately the model makes predictions.
+Looking at the graph, we can see that accuracy is fairly similar for Black, white, and all defendants, but the model is slightly less accurate in the case of Black defendants.
+The model makes accurate predictions for all defendants about 61% of the time, for Black defendants about 59% of the time, and for white defendants about 63% of the time.
+
+The value of recall represents the proportion of actual positives which were predicted correctly.
+In other words, recall is the number of people who were predicted to recidivate and did recidivate out of the total number of people actually recidivated:
+
+$$\frac{TP}{TP + FN}$$
+
+In the case of our baseline mode, recall is highest for Black defendants at nearly 0.7, while recall for white defendants is much lower at just above 0.2.
+The model correctly identifies defendants who recidivate about 54% of the time, Black defendants who recidivate about 70% of the time, and white defendants who recidivate about 20% of the time.
+This means that the model frequently predicts white defendants to not recidivate (the more favorable outcome) when they actually do.
+This points to potential bias in the model, as the recall score for white defendants is so much lower than that for Black defendants.
+
+Slightly different from recall, precision is the proportion of predicted positives which were actually correct.
+Precision is therefore the number of people who were predicted to recidivate and did recidivate out of the total number of people who were predicted to recidivate:
+
+$$\frac{TP}{TP + FP} $$ 
+We can see that precision is very similar across groups, all approximately 0.6.
+This means that the model's predictions of a defendant, either Black or white, who recidivates are accurate about 60% of the time.
 
@@
 ```{python group fairnesss statistical parity difference}
-stat_par_diff = statistical_parity_difference(y_test, prot_attr='race', pos_label = 0)
+stat_par_diff = statistical_parity_difference(y_test, prot_attr='race', priv_group = 1, pos_label = 0)
 
@@
 
-$\frac{P(\hat{Y}=0|D=0)}{P(\hat{Y}=0|D=1)}$
+$$\frac{P(\hat{Y}=0|D=0)}{P(\hat{Y}=0|D=1)}$$
 
@@
 
-$\frac{0.44}{0.66} = 0.667$
+$$\frac{0.44}{0.66} = 0.667$$
 
@@
 ```{python} 
-#ep = EqOddsPostprocessing(unprivileged_groups=0, privileged_groups= 1, seed =1234567)
-#eo = PostProcessingMeta(estimator=lr, postprocessor=ep, random_state=1234567)
-#eo.fit(train_BLD, test_BLD)
-#y_pred_EO = eo.predict(test_BLD)
+# ep = EqOddsPostprocessing(unprivileged_groups=0, privileged_groups= 1, seed =1234567)
+# eo = PostProcessingMeta(estimator=lr, postprocessor=ep, random_state=1234567)
+# eo.fit(train_BLD, test_BLD)
+# y_pred_EO = eo.predict(test_BLD)
 #y_proba_EO = eo.predict_(test_BLD)

### 2022-04-19 final_metrics_analysis.Rmd

@@
 # Exploratory Data Analysis
+```{python EDA} 
+# needs interpretation
+X_new_index = X_new.rename(columns={"race": "def_race"})
+X_new_index = X_new_index.rename(columns={"sex": "def_sex"})
+# renaming dataframe columns to avoid the ValueError of variables being a column AND index label (which is ambiguous)
+
+X_new_index.groupby(["def_race"])["age"].median()
+X_new_index.groupby(["def_race", "sex"]).size()
+X_new_index.groupby(["def_race", "c_charge_degree"]).size()
+X_new_index.groupby(["def_race"])["priors_count"].median()
+#X_new_index.groupby(["def_race"])["juv_fel_count"].median()
+#X_new_index.groupby(["def_race"])["juv_misd_count"].median()
+# above two lines commented out since medians for everyone = 0
+
+```
+
 ```{python exploratory visualization set up}
@@
 Again, this trend points to disproportionate rates of criminalization of Black people.
+This increased likelihood of recidivating may be attributed to the systemic racism in policing (e.g., predominantly Black neighborhoods tend to be overpoliced in comparison to predominantly white neighborhoods).
 
@@
 This table shows the counts of recidivism by race.
+
+```{python exploratory priors by race}
+priors_by_race = sns.countplot(x="race", color="priors_count", data=df_viz, palette=workshop_palette)
+
+plt.title('Priors Count by Race')
+plt.show()
+```
+
+```{python exploratory priors by sex}
+priors_by_race = sns.countplot(x="sex", color="priors_count", data=df_viz, palette=workshop_palette)
+
+plt.title('Priors Count by Sex')
+plt.show()
+```
+
+The two Priors Count bar charts show the differences in the distribution of priors by race and sex.
+If a defendant has a higher number of prior offenses, the individual is more likely to be rated with a higher risk of recidivism.
+In the former plot, one can see that Black defendants have a higher number of priors than do white defendants.
+This again likely reflects the over-policing of Black neighborhoods.
+In the latter plot, one can see that male defendants have a higher number of priors than do female defendants.
+This is worth noting because the COMPAS risk assessment algorithm has been found to express a discrepancy in prediction between sexes.
+When the recidivsim predictions are compared with the true recividism outcomes, female defendants have a notable false positive rate that predicts their recidivism risk as higher than their true outcomes.
 
@@
 This means that the model's predictions of a defendant, either Black or white, who recidivates are accurate about 60% of the time.
+
+White people have a lower recall rate, meaning that there is a low true positive rate.
+Whereas for black people the true positive rate is much higher: there's a high percentage of recidivated guesses that were correct.
 
@@
 ```
+
+The highest percentage in this confusion matrix is for the true negative rate, those who were predicted to not recidivate and actually did not, at 36.24%, 0.83% higher than the true negatives in the baseline model.
+This means that the re-weighing model marginally improved the baseline model accuracy.
+The true positive rate, or those who were predicted to recidivate and did recidivate, is 24.72%, 1.13% lower than that of the baseline model.
+This means that the re-weighing model slightly lowered the accuracy of the baseline model.
+The false negative and false positive rates, or the predictions that proved to be incorrect, are 23.20% and 15.85%, respectively.
+The re-weighing model lowered the accuracy of the false negative rate by 1.14% and improved the accuracy of the false positive rate by 0.83%.
+We are most concerned about the false positivesthat mean that defendants who are predicted to recidivate and do not actually recidivate.
+While there is a slight accuracy improvement in the false positive rate, it is still relatively high at 15.85%, indicating that we should further precede with additional bias mitigation techniques.
 
@@
 
-```{python} 
-# ep = EqOddsPostprocessing(unprivileged_groups=0, privileged_groups= 1, seed =1234567)
-# eo = PostProcessingMeta(estimator=lr, postprocessor=ep, random_state=1234567)
-# eo.fit(train_BLD, test_BLD)
-# y_pred_EO = eo.predict(test_BLD)
+```{python post-processing equal odds needs fix} 
+#ep = EqOddsPostprocessing(unprivileged_groups=0, privileged_groups= 1, seed =1234567)
+#eo = PostProcessingMeta(estimator=lr, postprocessor=ep, random_state=1234567)
+#eo.fit(train_BLD, test_BLD)
+#y_pred_EO = eo.predict(test_BLD)
 #y_proba_EO = eo.predict_(test_BLD)
-
 ```

### 2022-04-19 final_metrics_analysis.Rmd
commenting out code in order to knit

--- 2022-04-19 final_metrics_analysis.Rmd
@@
 library(reticulate)
+
+# for Dianne
+use_condaenv("base")
+
+# for Amrita
+# use_condaenv("anaconda3")
 #use_condaenv("anaconda3")
@@
 
+```{python make degree numeric}
+#av_defendant = (df_viz.groupby(['race'],as_index=False).age.mean()
+#display(df_viz)
+#list_vars = list(df_viz)
+#list_vars
+#df_numeric_charges = df_viz['c_charge_degree'].replace(['M', 'F'], [0, 1], inplace=True)
+#plt.bar('c_charge_degree', 1, data = df_viz)
+```
+
 ```{python exploratory recidivism bar plot}
@@
 
-Of both Black and white people, there is a higher percentage of people who did not recidivate than who did. 2500 people recidivated, whereas approximately 2600 did not.
-
 ```{python exploratory race bar plot}
@@
 ```
-
-The race distribution for black people is 3000 and the distribution for white people is 2100. Therefore, there are more black people in the dataset than white people.
 
@@
 ```
-
-When we divide the data into Black and white defendants, we can see that Black defendants recidivate more than white defendants and Black defendants recidivate more than not recidivate.
-However, we can also see that there are more Black defendants in the data set in general.
-
-Black defendants are more likely to recidivate than not recidivate.
-This increased likelihood of recidivating may be attributed to the systemic racism in policing (e.g., predominantly Black neighborhoods tend to be overpoliced in comparison to predominantly white neighborhoods").
-
-```{python exploratory priors by race}
-priors_by_race = sns.countplot(x="race", color="priors_count", data=df_viz, palette=workshop_palette)
-
-plt.title('Priors Count by Race')
-plt.show()
-```
-
-```{python exploratory priors by sex}
-priors_by_race = sns.countplot(x="sex", color="priors_count", data=df_viz, palette=workshop_palette)
-
-plt.title('Priors Count by Sex')
-plt.show()
-```
-
-The two Priors Count bar charts show the differences in the distribution of priors by race and sex.
-If a defendant has a higher number of prior offenses, the individual is more likely to be rated with a higher risk of recidivism.
-In the former plot, one can see that Black defendants have a higher number of priors than do white defendants.
-This again likely reflects the over-policing of Black neighborhoods.
-In the latter plot, one can see that male defendants have a higher number of priors than do female defendants.
-This is worth noting because the COMPAS risk assessment algorithm has been found to express a discrepancy in prediction between sexes.
-When the recidivsim predictions are compared with the true recividism outcomes, female defendants have a notable false positive rate that predicts their recidivism risk as higher than their true outcomes.
 
@@
 
-The highest percentage is for true negatives, those guessed to not recidivate and actually did not, at 35.41%.
-Similarly, the true positive rate is 25.85%.
-Our false negative and positive rates, incorrect guesses, is high for this baseline model, indicating that debiasing intervention may be necessary to lower these rates.
-
 ```{python baseline metrics per group}
@@
 In the Black population, the model correctly identifies 70% of people who recidivated. In the white population, the model correctly identifies 20% of people who recidivated.
-
-White people have a lower recall rate, meaning that there is a low true positive rate.
-Whereas for black people the true positive rate is much higher: there's a high percentage of recidivated guesses that were correct.
 
@@
 ```
-
-The highest percentage in this confusion matrix is for the true negative rate, those guessed to not recidivate and actually did not, at 36.24%, 0.83% higher than the true negatives in the baseline model.
-This means that the re-weighing model marginally improved the baseline model accuracy.
-The true positive rate, or those who were predicted to recidivate and did recidivate, is 24.72%, 1.13% lower than that of the baseline model.
-This means that the re-weighing model slightly lowered the accuracy of the baseline model.
-The false negative and false positive rates, or the predictions that proved to be incorrect, are 23.20% and 15.85%, respectively.
-The re-weighing model lowered the accuracy of the false negative rate by 1.14% and improved the accuracy of the false positive rate by 0.83%.
-We are most concerned about the false positivesthat mean that defendants who are predicted to recidivate and do not actually recidivate.
-While there is a slight accuracy improvement in the false positive rate, it is still relatively high at 15.85%, indicating that we should further precede with additional bias mitigation techniques.
 
@@
 
-```{python post-processing equal odds needs fix} 
+```{python} 
 #ep = EqOddsPostprocessing(unprivileged_groups=0, privileged_groups= 1, seed =1234567)

### 2022-04-19 Paper_Draft2/Draft2.Rmd

@@
 COMPAS uses various behavioral and psychological metrics related to recidivism to assist justice systems in assessing a defendant’s potential recidivism risk.
-Angwin et al.
-published a ProPublica article in which they conclude that the racial biases in the criminal justice system are reflected in the COMPAS recidivism risk scores.
-In response, Dieterich et al.
-published a rebuttal on behalf of Northpointe defending the COMPAS algorithm and refuting Angwin et al.’s allegation of racial bias.
+@angwin2016machine published a ProPublica article in which they conclude that the racial biases in the criminal justice system are reflected in the COMPAS recidivism risk scores.
+In response, @equivant_response_2016 published a rebuttal on behalf of Northpointe defending the COMPAS algorithm and refuting @angwin2016machine’s allegation of racial bias.
 Using a human rights framework adopted from the organizations Women at the Table and AI Fairness 360, we use debiasing algorithms and fairness metrics to analyze the argument between Northpointe and ProPublica and determine whether and to what extent there is racial bias in the COMPAS algorithm.
@@
 Although these questions do not directly ask about race, they do not take into account the pervasive nature of systemic racism that infiltrates every aspect of the lives of marginalized people, thereby indirectly asking about race.
-Angwin et al.
+@angwin2016machine.
 published a 2016 piece in the news publication ProPublica that analyzes the methods and algorithms used by Northpointe in their COMPAS risk score assessment algorithm and uncovers the racial biases in defendants’ scores [@angwin2016machine].
 Their research finds that “the algorithm [is] somewhat more accurate than a coin flip,” a worrisome level of accuracy given the potential impact its determinations may have on real people’s lives.
-Angwin et al.
-specifically investigate the distribution of COMPAS scores by decile among Black and white defendants.
+@angwin2016machine specifically investigate the distribution of COMPAS scores by decile among Black and white defendants.
 They write: “The analysis also [shows] that even when controlling for prior crimes, future recidivism, age, and gender, Black defendants [are] 45 percent more likely to be assigned higher risk scores than white defendants” [@larson2016we].
-After examining the fairness metric statistical parity difference, Angwin et al.
-conclude that the algorithm is racially biased [@larson2016we].
+After examining the fairness metric statistical parity difference, @angwin2016machine conclude that the algorithm is racially biased [@larson2016we].
 
-Dieterich et al., on the behalf of Northpointe, deny the allegations of racial bias and offer their own analyses based on different fairness metrics in rebuttal [@equivant_response_2016].
-Angwin et al.
-maintain that there are biases in the outcome values, protected attributes, and covariates during Dieterich et al.’s data processing phase.
-ProPublica collaborators Larson et al.
-account for these biases in their analyses.
-In their response, Dieterich et al.
-highlight that Angwin et al.
-did not account for base rates of recidivism in their analysis, which are important to understand initial percentages without the presence of other information.
+@equivant_response_2016, on the behalf of Northpointe, deny the allegations of racial bias and offer their own analyses based on different fairness metrics in rebuttal [@equivant_response_2016].
+@angwin2016machine maintain that there are biases in the outcome values, protected attributes, and covariates during @equivant_response_2016’s data processing phase.
+ProPublica collaborators @larson2016we account for these biases in their analyses.
+In their response, @equivant_response_2016 highlight that @angwin2016machine did not account for base rates of recidivism in their analysis, which are important to understand initial percentages without the presence of other information.
 
@@
 AI Fairness 360 (AIF360) is an open-source Python toolkit that seeks to "to help facilitate the transition of fairness research algorithms to use in an industrial setting and to provide a common framework for fairness researchers to share and evaluate algorithms" [@aif360-oct-2018].
-It contains multiple data sets, including the COMPAS data set that accompanied the Angwin et al.
-piece.
+It contains multiple data sets, including the COMPAS data set that accompanied the @angwin2016machine piece.
 

### 2022-04-19 final_metrics_analysis.Rmd
Mosaic of charge degree by race

--- 2022-04-19 final_metrics_analysis.Rmd
@@
 library(reticulate)
+
+# for Dianne
+use_condaenv("base")
+
+# for Amrita
 #use_condaenv("anaconda3")
@@
 import matplotlib.patches as mpatches
+from statsmodels.graphics.mosaicplot import mosaic
+from itertools import product
 
@@
 ```
+
+```{python crosstable of charge degrees by race}
+crosstable=pd.crosstab(df_viz['race'],df_viz['c_charge_degree'])
+crosstable
+```
+
+```{python mosiac of charge degrees by race}
+props={}
+props[('Caucasian','F')]={'facecolor':'red', 'edgecolor':'white'}
+props[('Caucasian','M')]={'facecolor':'red', 'edgecolor':'white'}
+props[('African-American','F')]={'facecolor':'xkcd:aqua','edgecolor':'white'}
+props[('African-American','M')]=        {'facecolor':'xkcd:aqua','edgecolor':'white'}
+labelizer=lambda k:{('Caucasian','F'):1242,('African-American','F'):2194,('Caucasian','M'):858,('African-American','M'):979}[k]
+mosaic(df_viz,['race','c_charge_degree'],labelizer=labelizer,properties=props)
+plt.show()
+```
+
+59.14% of white defendants were charged with felonies versus 69.15% of black defendants.
+Likewise, a higher percentage of white defendants (40.86%) were charged with a misdemeanor than black defendants (30.85%).
 

### 2022-04-19 Paper_Draft2/Draft2.Rmd
hyperlink places that use COMPAS instead of reference

@@
       Northampton, MA 01063
-    email: aarcharya@smith.edu, dcaravela@smith.edu, ekim89@smith.edu, ekornberg@smith.edu, enesmith@smith.edu
+    email: aacharya@smith.edu, dcaravela@smith.edu, ekim89@smith.edu, ekornberg@smith.edu, enesmith@smith.edu
 
@@
 
-The Correctional Offender Management Profiling for Alternative Sanctions (COMPAS) algorithm was created by the private, for-profit company Northpointe (now known by its parent company [Equivant](https://www.equivant.com/faq/)), to predict defendants’ risk of recidivism.
+The Correctional Offender Management Profiling for Alternative Sanctions (COMPAS) algorithm was created by the private, for-profit company Northpointe (now known by its parent company [equivant](https://www.equivant.com/faq/)), to predict defendants’ risk of recidivism.
 It generates a decile score that classifies defendants’ risk of recidivism as either low, medium, or high [@angwin2016machine].
-Jurisdictions across the United States use the COMPAS risk assessment instrument, including but not limited to the New York [@NYDOCCS2019], Massachusetts, Michigan, California [@Jackson2020Setting], and Wisconsin [@WisconsinDOC] Departments of Corrections.
+Jurisdictions across the United States use the COMPAS risk assessment instrument, including but not limited to the [New York](https://doccs.ny.gov/system/files/documents/2020/11/8500.pdf), [Massachusetts](https://hdsr.mitpress.mit.edu/pub/hzwo7ax4/release/4), [Michigan](https://hdsr.mitpress.mit.edu/pub/hzwo7ax4/release/4), [California](https://hdsr.mitpress.mit.edu/pub/hzwo7ax4/release/4), and [Wisconsin](https://doc.wi.gov/Pages/AboutDOC/COMPAS.aspx) Departments of Corrections.
+
 Due to the proprietary nature of the COMPAS algorithm, it is unknown how exactly these recidivism risk scores are calculated; however, a sample COMPAS Risk Assessment Survey has been made publicly available, revealing the algorithm’s input information.
@@
 The algorithm unfairly benefits white defendants over Black defendants.
-
 Disparate impact ratio is 0.47.

### 2022-04-19 Paper_Draft2/Draft2.tex
hyperlink places that use COMPAS instead of reference

--- 2022-04-17 Paper_Draft2/Draft2.tex
@@
 \address{%
-$^{1}$ \quad Statistical and Data Sciences Smith College Northampton, MA 01063; \\
-$^{2}$ \quad ; \href{mailto:aacharya@smith.edu}{\nolinkurl{aacharya@smith.edu}}\\
-$^{3}$ \quad ; \href{mailto:dcaravela@smith.edu}{\nolinkurl{dcaravela@smith.edu}}\\
-$^{4}$ \quad ; \href{mailto:ekim89@smith.edu}{\nolinkurl{ekim89@smith.edu}}\\
-$^{5}$ \quad ; \href{mailto:ekornberg@smith.edu}{\nolinkurl{ekornberg@smith.edu}}\\
-$^{6}$ \quad ; \href{mailto:enesmith@smith.edu}{\nolinkurl{enesmith@smith.edu}}\\
+$^{1}$ \quad Statistical and Data Sciences Smith College Northampton, MA
+01063; \href{mailto:aacharya@smith.edu}{\nolinkurl{aacharya@smith.edu}},
+\href{mailto:dcaravela@smith.edu}{\nolinkurl{dcaravela@smith.edu}},
+\href{mailto:ekim89@smith.edu}{\nolinkurl{ekim89@smith.edu}},
+\href{mailto:ekornberg@smith.edu}{\nolinkurl{ekornberg@smith.edu}},
+\href{mailto:enesmith@smith.edu}{\nolinkurl{enesmith@smith.edu}}\\
 }
 % Contact information of the corresponding author
-\corres{Correspondence: }
+\corres{Correspondence: \href{mailto:bbaumer@smith.edu}{\nolinkurl{bbaumer@smith.edu}}}
 
@@
 % Abstract (Do not insert blank lines, i.e. \\)
-\abstract{A variety of disciplines use risk assessment instruments to help humans
-make data-driven decisions. Northpointe, a software company, created an
-algorithmic risk assessment instrument known as the Correctional
-Offender Management Profiling for Alternative Sanctions (COMPAS). COMPAS
-uses various behavioral and psychological metrics related to recidivism
-to assist justice systems in assessing a defendant's potential
-recidivism risk. Angwin et al.~published a ProPublica article in which
-they conclude that the racial biases in the criminal justice system are
-reflected in the COMPAS recidivism risk scores. In response, Dieterich
-et al.~published a rebuttal on behalf of Northpointe defending the
-COMPAS algorithm and refuting Angwin et al.'s allegation of racial bias.
-Using a human rights framework adopted from the organizations Women at
-the Table and AI Fairness 360, we use debiasing algorithms and fairness
-metrics to analyze the argument between Northpointe and ProPublica and
-determine whether and to what extent there is racial bias in the COMPAS
-algorithm.}
+\abstract{A variety of disciplines use risk assessment instruments to
+help humans make data-driven decisions. Northpointe, a software company,
+created an algorithmic risk assessment instrument known as the
+Correctional Offender Management Profiling for Alternative Sanctions
+(COMPAS). COMPAS uses various behavioral and psychological metrics
+related to recidivism to assist justice systems in assessing a
+defendant's potential recidivism risk. \citet{angwin2016machine}
+published a ProPublica article in which they conclude that the racial
+biases in the criminal justice system are reflected in the COMPAS
+recidivism risk scores. In response, \citet{equivant_response_2016}
+published a rebuttal on behalf of Northpointe defending the COMPAS
+algorithm and refuting \citet{angwin2016machine}'s allegation of racial
+bias. Using a human rights framework adopted from the organizations
+Women at the Table and AI Fairness 360, we use debiasing algorithms and
+fairness metrics to analyze the argument between Northpointe and
+ProPublica and determine whether and to what extent there is racial bias
+in the COMPAS algorithm.}
 
@@
 \citep{Jackson2020Setting}, and Wisconsin \citep{WisconsinDOC}
-Departments of Corrections. Due to the proprietary nature of the COMPAS
-algorithm, it is unknown how exactly these recidivism risk scores are
-calculated; however, a sample COMPAS Risk Assessment Survey has been
-made publicly available, revealing the algorithm's input information.
-This survey has been widely critiqued for using proxy variables for race
-that do not explicitly factor in a defendant's race but heavily imply
-it, allowing Northpointe to claim that their algorithm is free of racial
-bias. For example, the COMPAS risk assessment survey asks screeners to
-speculate if a defendant might be affiliated with a gang. It also asks
-if a defendant has any friends or family members who have been crime
-victims \citep{Angwin2016Sample}. Although these questions do not
-directly ask about race, they do not take into account the pervasive
-nature of systemic racism that infiltrates every aspect of the lives of
-marginalized people, thereby indirectly asking about race. Angwin et
-al.~published a 2016 piece in the news publication ProPublica that
-analyzes the methods and algorithms used by Northpointe in their COMPAS
-risk score assessment algorithm and uncovers the racial biases in
-defendants' scores \citep{angwin2016machine}. Their research finds that
-``the algorithm {[}is{]} somewhat more accurate than a coin flip,'' a
-worrisome level of accuracy given the potential impact its
-determinations may have on real people's lives. Angwin et
-al.~specifically investigate the distribution of COMPAS scores by decile
-among Black and white defendants. They write: ``The analysis also
-{[}shows{]} that even when controlling for prior crimes, future
-recidivism, age, and gender, Black defendants {[}are{]} 45 percent more
-likely to be assigned higher risk scores than white defendants''
-\citep{larson2016we}. After examining the fairness metric statistical
-parity difference, Angwin et al.~conclude that the algorithm is racially
-biased \citep{larson2016we}.
+Departments of Corrections.
 
-Dieterich et al., on the behalf of Northpointe, deny the allegations of
-racial bias and offer their own analyses based on different fairness
-metrics in rebuttal \citep{equivant_response_2016}. Angwin et
-al.~maintain that there are biases in the outcome values, protected
-attributes, and covariates during Dieterich et al.'s data processing
-phase. ProPublica collaborators Larson et al.~account for these biases
-in their analyses. In their response, Dieterich et al.~highlight that
-Angwin et al.~did not account for base rates of recidivism in their
-analysis, which are important to understand initial percentages without
-the presence of other information.
+Due to the proprietary nature of the COMPAS algorithm, it is unknown how
+exactly these recidivism risk scores are calculated; however, a sample
+COMPAS Risk Assessment Survey has been made publicly available,
+revealing the algorithm's input information. This survey has been widely
+critiqued for using proxy variables for race that do not explicitly
+factor in a defendant's race but heavily imply it, allowing Northpointe
+to claim that their algorithm is free of racial bias. For example, the
+COMPAS risk assessment survey asks screeners to speculate if a defendant
+might be affiliated with a gang. It also asks if a defendant has any
+friends or family members who have been crime victims
+\citep{Angwin2016Sample}. Although these questions do not directly ask
+about race, they do not take into account the pervasive nature of
+systemic racism that infiltrates every aspect of the lives of
+marginalized people, thereby indirectly asking about race.
+\citet{angwin2016machine}. published a 2016 piece in the news
+publication ProPublica that analyzes the methods and algorithms used by
+Northpointe in their COMPAS risk score assessment algorithm and uncovers
+the racial biases in defendants' scores \citep{angwin2016machine}. Their
+research finds that ``the algorithm {[}is{]} somewhat more accurate than
+a coin flip,'' a worrisome level of accuracy given the potential impact
+its determinations may have on real people's lives.
+\citet{angwin2016machine} specifically investigate the distribution of
+COMPAS scores by decile among Black and white defendants. They write:
+``The analysis also {[}shows{]} that even when controlling for prior
+crimes, future recidivism, age, and gender, Black defendants {[}are{]}
+45 percent more likely to be assigned higher risk scores than white
+defendants'' \citep{larson2016we}. After examining the fairness metric
+statistical parity difference, \citet{angwin2016machine} conclude that
+the algorithm is racially biased \citep{larson2016we}.
+
+\citet{equivant_response_2016}, on the behalf of Northpointe, deny the
+allegations of racial bias and offer their own analyses based on
+different fairness metrics in rebuttal \citep{equivant_response_2016}.
+\citet{angwin2016machine} maintain that there are biases in the outcome
+values, protected attributes, and covariates during
+\citet{equivant_response_2016}'s data processing phase. ProPublica
+collaborators \citet{larson2016we} account for these biases in their
+analyses. In their response, \citet{equivant_response_2016} highlight
+that \citet{angwin2016machine} did not account for base rates of
+recidivism in their analysis, which are important to understand initial
+percentages without the presence of other information.
 
@@
 \citep{aif360-oct-2018}. It contains multiple data sets, including the
-COMPAS data set that accompanied the Angwin et al.~piece.
+COMPAS data set that accompanied the \citet{angwin2016machine} piece.
 
@@
 or not they recidivate. The algorithm unfairly benefits white defendants
-over Black defendants.
-
-Disparate impact ratio is 0.47. The ratio of white defendants predicted
-to not recidivate to the Black defendants predicted to not recidivate is
-0.47. A ratio between 0.8 and 1.25 is considered fair, therefore the
-algorithm unfairly benefits white defendants.
+over Black defendants. Disparate impact ratio is 0.47. The ratio of
+white defendants predicted to not recidivate to the Black defendants
+predicted to not recidivate is 0.47. A ratio between 0.8 and 1.25 is
+considered fair, therefore the algorithm unfairly benefits white
+defendants.
 

### 2022-04-19 final_metrics_analysis.Rmd

--- 2022-04-19 final_metrics_analysis.Rmd
@@
 
+```{python}
+def get_attributes(data, selected_attr=None):
+    unprivileged_groups = []
+    privileged_groups = []
+    if selected_attr == None:
+        selected_attr = data.protected_attribute_names
+
+    for attr in selected_attr:
+            idx = data.protected_attribute_names.index(attr)
+            privileged_groups.append({attr:data.privileged_protected_attributes[idx]}) 
+            unprivileged_groups.append({attr:data.unprivileged_protected_attributes[idx]})
+
+    return privileged_groups, unprivileged_groups
+```
+
 ```{python post-processing equal odds needs fix} 
-#ep = EqOddsPostprocessing(unprivileged_groups=0, privileged_groups= 1, seed =1234567)
+privileged_groups, unprivileged_groups = get_attributes(train_BLD, selected_attr= 'race')
+ep = EqOddsPostprocessing(unprivileged_groups= unprivileged_groups, privileged_groups= privileged_groups, seed = 42)
 #eo = PostProcessingMeta(estimator=lr, postprocessor=ep, random_state=1234567)
-#eo.fit(train_BLD, test_BLD)
-#y_pred_EO = eo.predict(test_BLD)
-#y_proba_EO = eo.predict_(test_BLD)
+EOPP = ep.fit(train_BLD, test_BLD)
+y_pred_EO = ep.predict(test_BLD)
+#y_proba_EO = ep.predict_(test_BLD)
+
+acc_EO = accuracy_score(y_test, y_pred_EO)
 ```

### 2022-04-19 final_metrics_analysis.Rmd
Attempted mosaic plot of recidivism by race - caucasian not showing up

--- 2022-04-19 final_metrics_analysis.Rmd
@@
 
+```{python crosstable of recitivism by race}
+crosstable=pd.crosstab(df_viz['race'],df_viz['two_year_recid'])
+crosstable
+```
+
+```{python exploratory recidivism mosaic plot}
+props={}
+props[('Caucasian','Recidivated')]={'facecolor':'red', 'edgecolor':'white'}
+props[('Caucasian','Survived')]={'facecolor':'red', 'edgecolor':'white'}
+props[('African-American','Recidivated')]={'facecolor':'xkcd:aqua','edgecolor':'white'}
+props[('African-American','Survived')]=        {'facecolor':'xkcd:aqua','edgecolor':'white'}
+labelizer=lambda k:{('Caucasian','Recidivated'):1242,('African-American','Recidivated'):1661,('Caucasian','Survived'):858,('African-American','Survived'):1512}[k]
+mosaic(df_viz,['race','two_year_recid'],labelizer=labelizer,properties=props)
+plt.show()
+```
+
 This plot shows the distribution of two year recidivism outcomes.
@@
 plt.title('Race Distribution (White and Black individuals only)')
+plt.show()
 ```

### 2022-04-19 Paper_Draft2/Draft2.tex
not sure

--- 2022-04-19 Paper_Draft2/Draft2.tex
@@
 % Full title of the paper (Capitalized)
-\Title{Does the COMPAS Needle Always Point Towards Equity? Finding
-Fairness in the COMPAS Risk Assessment Algorithm: A Case Study}
+\Title{Does the COMPAS Needle Always Point Towards Equity? Finding Fairness in
+the COMPAS Risk Assessment Algorithm: A Case Study}
 
 % Authors, for the paper (add full first names)
-\Author{Amrita Acharya$^{1}$, Dianne Caravela$^{1}$, Eunice
-Kim$^{1}$, Emma Kornberg$^{1}$, Elisabeth Nesmith$^{1}$}
+\Author{Amrita Acharya$^{1}$, Dianne Caravela$^{1}$, Eunice Kim$^{1}$, Emma Kornberg$^{1}$, Elisabeth Nesmith$^{1}$}
 
 % Authors, for metadata in PDF
-\AuthorNames{Amrita Acharya, Dianne Caravela, Eunice Kim, Emma
-Kornberg, Elisabeth Nesmith}
+\AuthorNames{Amrita Acharya, Dianne Caravela, Eunice Kim, Emma Kornberg, Elisabeth Nesmith}
 
@@
 \address{%
-$^{1}$ \quad Statistical and Data Sciences Smith College Northampton, MA
-01063; \href{mailto:aarcharya@smith.edu}{\nolinkurl{aarcharya@smith.edu}},
+$^{1}$ \quad Statistical and Data Sciences Smith College Northampton, MA 01063; \href{mailto:aarcharya@smith.edu}{\nolinkurl{aarcharya@smith.edu}},
 \href{mailto:dcaravela@smith.edu}{\nolinkurl{dcaravela@smith.edu}},
@@
 \toprule
-Fairness Metric & Ideal Value & Baseline Value & Benefited Group \\
+Fairness Metric & Ideal Value & Baseline Value & Benefited
+Group\tabularnewline
 \midrule
@@
 \toprule
-Fairness Metric & Ideal Value & Baseline Value & Benefited Group \\
+Fairness Metric & Ideal Value & Baseline Value & Benefited
+Group\tabularnewline
 \midrule
 \endhead
-Statistical Parity Difference & 0 & -0.14 & White Defendants \\
-Disparate Impact Ratio & 1 & 0.47 & White Defendants \\
-Average Odds Difference & 0 & -0.44 & White Defendants \\
-Equal Opportunity Difference & 0 & -0.41 & White Defendants \\
+Statistical Parity Difference & 0 & -0.14 & White
+Defendants\tabularnewline
+Disparate Impact Ratio & 1 & 0.47 & White Defendants\tabularnewline
+Average Odds Difference & 0 & -0.44 & White Defendants\tabularnewline
+Equal Opportunity Difference & 0 & -0.41 & White
+Defendants\tabularnewline
 \bottomrule

### 2022-04-20 final_metrics_analysis.Rmd
more graph interpretations, fixing a few typos

--- 2022-04-19 final_metrics_analysis.Rmd
@@
 editor_options: 
-  chunk_output_type: console
+  chunk_output_type: inline
 ---
@@
   obj_fairness = [[0,0,0,1]]
-
   # row for objectives    
@@
   fair_metrics.loc['Baseline Model'] = [stat_par_diff, eq_opp_diff, avg_odds_diff, disp_impact_ratio]
-
   # row for mitigated bias
   fair_metrics.loc[model] = fair_metrics_mitigated
-
   metrics_len = len(cols)
-
   fig, ax = plt.subplots(figsize=(20,4), ncols=metrics_len, nrows=1)
-
   plt.subplots_adjust(
@@
   )
-
   y_title_margin = 1.2
-
   plt.suptitle("Fairness metrics", y = 1.09, fontsize=20)
   sns.set(style="dark")
-
   cols = fair_metrics.columns.values
@@
   bound = [[-0.1,0.1],[-0.1,0.1],[-0.1,0.1],[0.8,1.25]]
-
   for i in range(0,metrics_len):
@@
           ax.text(a.get_x()+a.get_width()/5, a.get_y()+a.get_height()+marg, round(val, 3), fontsize=15,color='black')
-
       plt.ylim(bottom[i], top[i])
@@
 X_new_index.groupby(["def_race"])["priors_count"].median()
-#X_new_index.groupby(["def_race"])["juv_fel_count"].median()
-#X_new_index.groupby(["def_race"])["juv_misd_count"].median()
-# above two lines commented out since medians for everyone = 0
+X_new_index.groupby(["def_race"])["juv_fel_count"].median()
+X_new_index.groupby(["def_race"])["juv_misd_count"].median()
+```
 
-```
+The typical Black defendant is 29 years old, male, and has committed two prior crimes.
+The typical white defendant is 35 years old, male, and has committed one prior crime.
+The typical defendant for both races does not have any juvenile convictions.
 
@@
 
-This plot shows the distribution of two year recidivism outcomes.
-Out of all defendants, there is a higher proportion of people who did not recidivate than who did.
-About 2500 people recidivated, whereas approximately 2700 did not.
+Out of all defendants, there is a higher proportion of people who did not recidivate than who did. About 2500 people recidivated, whereas approximately 2700 did not.
 
@@
 
-This plot shows the distribution of defendant races.
-The dataset contains about 3200 Black defendants and 2100 white defendants.
+The dataset we are using contains about 3200 Black defendants and 2100 white defendants.
 Therefore, there are more Black people represented in the dataset than white people, showing the disproportionate policing and criminalization of Black communities.
@@
 ```{python exploratory age by race plot}
-# age distribution by race
-#ax = sns.kdeplot(x="age", hue="race", data=df_viz, palette=workshop_palette);
-#kdeline = ax.lines[0]
-#mean_black = df_viz.groupby('race').age.median()[0]
-#xs = kdeline.get_xdata()
-#ys = kdeline.get_ydata()
-#height_black = np.interp(mean_black, xs, ys)
-#ax.vlines(mean_black, 0, height_black, color=purple, ls=':')
+# age distribution by race - hacking solution for purple dotted line for now
+ax = sns.kdeplot(x="age", hue="race", data=df_viz, palette=workshop_palette);
+kdeline = ax.lines[0]
+mean_black = df_viz.groupby('race').age.median()[0]
+xs = kdeline.get_xdata()
+ys = kdeline.get_ydata()
+height_black = 0.027 # np.interp(mean_black, xs, ys)
+ax.vlines(mean_black, 0, height_black, color=purple, ls=':')
 
-#mean_white = df_viz.groupby('race').age.median()[1]
-#height_white = np.interp(mean_white, xs, ys)
-#ax.vlines(mean_white, 0, height_white, color=green, ls=':')
+mean_white = df_viz.groupby('race').age.median()[1]
+height_white = np.interp(mean_white, xs, ys)
+ax.vlines(mean_white, 0, height_white, color=green, ls=':')
 
-#print(df_viz.groupby('race').age.median())
-
-#plt.title('Distribution of Age by Race')
+plt.title('Distribution of Age by Race')
 ```
 
-This plot shows the distribution of defendant ages by race.
-The purple curve shows the distribution of the ages of Black defendants, and the green curve shows the distribution of the ages of white defendants.
+The purple curve in the plot above shows the distribution of the ages of Black defendants, and the green curve shows the distribution of the ages of white defendants.
 The probability of a defendant's age being between two points on the x-axis is the total shaded area of the curve under the two points.
@@
 
-This plot shows the distribution of two year recidivism outcomes by race.
 When we divide the data into Black and white defendants, we can see that Black defendants recidivate more than white defendants and Black defendants are more likely to recidivate than not recidivate.
@@
 
-This table shows the counts of recidivism by race.
-
 ```{python exploratory priors by race}
@@
 
-plt.title('Priors Count by Race')
+plt.title('Prior Charges Count by Race')
 plt.show()
@@
 
-plt.title('Priors Count by Sex')
+plt.title('Prior Charges Count by Sex')
 plt.show()
@@
 
-The two Priors Count bar charts show the differences in the distribution of priors by race and sex.
-If a defendant has a higher number of prior offenses, the individual is more likely to be rated with a higher risk of recidivism.
-In the former plot, one can see that Black defendants have a higher number of priors than do white defendants.
-This again likely reflects the over-policing of Black neighborhoods.
-In the latter plot, one can see that male defendants have a higher number of priors than do female defendants.
-This is worth noting because the COMPAS risk assessment algorithm has been found to express a discrepancy in prediction between sexes.
-When the recidivsim predictions are compared with the true recividism outcomes, female defendants have a notable false positive rate that predicts their recidivism risk as higher than their true outcomes.
+If a defendant has a higher number of prior offenses, the individual is more likely to be rated with a higher risk of recidivism **do we know this for sure?**.
+In the former plot, we can see that Black defendants have a higher number of prior charges than do white defendants.
+This again likely reflects the over-policing of Black people and communities.
+In the latter plot, we can see that male defendants have a higher number of prior charges than do female defendants.
+This is worth noting because the COMPAS risk assessment algorithm has been found to express a discrepancy in prediction between sexes **who found this?**.
+When the recidivsim predictions are compared with the true recividism outcomes, female defendants have a notable false positive rate that predicts their recidivism risk as higher than their true outcomes**where does this come from?**.
 
@@
 
-$$\frac{TP}{TP + FP} $$ 
+$$\frac{TP}{TP + FP}$$
+
 We can see that precision is very similar across groups, all approximately 0.6.
 This means that the model's predictions of a defendant, either Black or white, who recidivates are accurate about 60% of the time.
-
-White people have a lower recall rate, meaning that there is a low true positive rate.
-Whereas for black people the true positive rate is much higher: there's a high percentage of recidivated guesses that were correct.
 
@@
 
+### Equal Opportunity Difference
+
+This metric is computed as the difference of true positive rates between the unprivileged and the privileged groups.
+The true positive rate is the ratio of true positives to the total number of actual positives for a given group.
+The ideal value is 0. A value of < 0 implies higher benefit for the privileged group and a value > 0 implies higher benefit for the unprivileged group.
+Fairness for this metric is between -0.1 and 0.1.
+
+$$TPR_{D = Unprivileged} - TPR_{D = Privileged}$$
+```{python group fairnesss equal opportunity difference}
+eq_opp_diff = equal_opportunity_difference(y_test, y_pred, prot_attr='race', pos_label = 0)
+
+print(f'[Baseline] The equal opportunity difference is {eq_opp_diff: .2}')
+```
+
+Since the equal opportunity difference is below 0 and beyond -0.1, the algorithm unfairly benefits white defendants.
+
 ### Average Odds Difference
 
-This metric returns the average difference in false positive rate and true positive rate for the privileged and unprivileged groups.
+This metric, an expansion of equal opportunity difference, returns the average difference in false positive rate and true positive rate for the privileged and unprivileged groups.
 A value of 0 is considered fair, and a value below 0 implies benefit for the privileged group.
@@
 
-$$\frac{(FPR_{D = unprivileged} - FPR_{D = privileged}) + (TPR_{D = unprivileged} - TPR_{D = privileged})}{2}$$
+$$\frac{1}{2}\left[(FPR_{D = Unprivileged} - FPR_{D = Privileged}) + \underbrace{(TPR_{D = Unprivileged} - TPR_{D = Privileged})}_\textrm{Equal Opportunity Difference}\right]$$
 
@@
 Since the average odds difference is below 0 by a substantial amount, white defendants are unfairly benefitted by the algorithm.
-
-### Equal Opportunity Difference
-
-This metric is computed as the difference of true positive rates between the unprivileged and the privileged groups.
-The true positive rate is the ratio of true positives to the total number of actual positives for a given group.
-The ideal value is 0. A value of < 0 implies higher benefit for the privileged group and a value > 0 implies higher benefit for the unprivileged group.
-Fairness for this metric is between -0.1 and 0.1.
-
-$$TPR_{D = Unprivileged} - TPR_{D = Privileged}$$
-
-```{python group fairnesss equal opportunity difference}
-eq_opp_diff = equal_opportunity_difference(y_test, y_pred, prot_attr='race', pos_label = 0)
-
-print(f'[Baseline] The equal opportunity difference is {eq_opp_diff: .2}')
-```
-
-Since the equal opportunity difference is below 0 and beyond -0.1, the algorithm unfairly benefits white defendants.
 
@@
 
-## Reweighing
+## Reweighting
 
@@
 
-The highest percentage in this confusion matrix is for the true negative rate, those who were predicted to not recidivate and actually did not, at 36.24%, 0.83% higher than the true negatives in the baseline model.
-This means that the re-weighing model marginally improved the baseline model accuracy.
-The true positive rate, or those who were predicted to recidivate and did recidivate, is 24.72%, 1.13% lower than that of the baseline model.
-This means that the re-weighing model slightly lowered the accuracy of the baseline model.
+The highest percentage in this confusion matrix is for the true negative rate, defendants who were predicted to not recidivate and actually did not, at 36.24%, 0.83% higher than the true negatives in the baseline model.
+This means that the re-weighing model marginally improved the baseline model accuracy in predicting people who did not recidivate.
+However, the true positive rate, or those who were predicted to recidivate and did recidivate, is 24.72%, 1.13% lower than that of the baseline model.
+This means that the re-weighing model slightly lowered the accuracy of the baseline model in predicting people who did recidivate.
+As a result, the overall reweighting model accuracy remains about the same as the baseline model accuracy.
 The false negative and false positive rates, or the predictions that proved to be incorrect, are 23.20% and 15.85%, respectively.
-The re-weighing model lowered the accuracy of the false negative rate by 1.14% and improved the accuracy of the false positive rate by 0.83%.
-We are most concerned about the false positivesthat mean that defendants who are predicted to recidivate and do not actually recidivate.
-While there is a slight accuracy improvement in the false positive rate, it is still relatively high at 15.85%, indicating that we should further precede with additional bias mitigation techniques.
+The re-weighing model increased the number of false negatives by 1.14% and decreased the number of false positives by 0.83%.
+We are most concerned about the false positives, which indicate defendants who are predicted to recidivate and do not actually recidivate.
+While there is a slight improvement in the percent of false positives, it is still relatively high at 15.85%, indicating that we should further precede with additional bias mitigation techniques.
 
@@
 ```
+
+Like we saw in the metrics plot for the baseline model, the model accuracy for defendants of all races hovers around 61%, being slightly lower for Black defendants and slightly higher for white defendants.
+However, we can note that the gap between accuracy for Black and white defendants is smaller that it was with the baseline model (about a 3% gap rather than 4%).
+
+We can see clear effects of the reweighting approach in the recall scores, which are now all very similar around 0.5.
+This is a marked improvement from the baseline model, where the model only correctly identified white defendants who recidivated 20% of the time.
+The reweighting model now correctly identifies defendants of any race who recidivate about half of the time.
+Notably, the reweighting also lowered the recall score for Black defendants.
+
+Whereas with the baseline model all three precision scores were relatively similar, with the reweighting model we see the precision score for Black defendants is the highest.
+The models predictions for Black defendants who recidivate are correct about 66% of the time.
+In contrast, these predictions are correct for white defendents only 50% of the time, and correct for all defendants 60% of the time.
+The increase in precision for Black defendants shows the effects of the bias mitigation, as fewer Black people are being incorrectly predicted to recidivate.
 
@@
 
-```{python}
-def get_attributes(data, selected_attr=None):
-    unprivileged_groups = []
-    privileged_groups = []
-    if selected_attr == None:
-        selected_attr = data.protected_attribute_names
-
-    for attr in selected_attr:
-            idx = data.protected_attribute_names.index(attr)
-            privileged_groups.append({attr:data.privileged_protected_attributes[idx]}) 
-            unprivileged_groups.append({attr:data.unprivileged_protected_attributes[idx]})
-
-    return privileged_groups, unprivileged_groups
-```
-
 ```{python post-processing equal odds needs fix} 
-privileged_groups, unprivileged_groups = get_attributes(train_BLD, selected_attr= 'race')
-ep = EqOddsPostprocessing(unprivileged_groups= unprivileged_groups, privileged_groups= privileged_groups, seed = 42)
+#ep = EqOddsPostprocessing(unprivileged_groups=0, privileged_groups= 1, seed =1234567)
 #eo = PostProcessingMeta(estimator=lr, postprocessor=ep, random_state=1234567)
-EOPP = ep.fit(train_BLD, test_BLD)
-y_pred_EO = ep.predict(test_BLD)
-#y_proba_EO = ep.predict_(test_BLD)
-
-acc_EO = accuracy_score(y_test, y_pred_EO)
+#eo.fit(train_BLD, test_BLD)
+#y_pred_EO = eo.predict(test_BLD)
+#y_proba_EO = eo.predict_(test_BLD)
 ```

### 2022-04-21 final_metrics_analysis.Rmd
Updating numbers for recitivated/survived mosaic chart

--- 2022-04-19 final_metrics_analysis.Rmd
@@
 props[('African-American','Survived')]=        {'facecolor':'xkcd:aqua','edgecolor':'white'}
-labelizer=lambda k:{('Caucasian','Recidivated'):1242,('African-American','Recidivated'):1661,('Caucasian','Survived'):858,('African-American','Survived'):1512}[k]
+labelizer=lambda k:{('Caucasian','Recidivated'):1278,('African-American','Recidivated'):1661,('Caucasian','Survived'):1278,('African-American','Survived'):1512}[k]
 mosaic(df_viz,['race','two_year_recid'],labelizer=labelizer,properties=props)

### 2022-04-21 final_metrics_analysis.Rmd
Adding description of recidivism mosaic chart

@@
 
+This plot shows the distribution of two year recidivism outcomes.
+Out of all defendants, there is a higher proportion of people who did not recidivate than who did.
+About 2500 people recidivated, whereas approximately 2700 did not.
+
 ```{python crosstable of recitivism by race}
@@
 props[('African-American','Survived')]=        {'facecolor':'xkcd:aqua','edgecolor':'white'}
-labelizer=lambda k:{('Caucasian','Recidivated'):1278,('African-American','Recidivated'):1661,('Caucasian','Survived'):1278,('African-American','Survived'):1512}[k]
+labelizer=lambda k:{('Caucasian','Recidivated'):822,('African-American','Recidivated'):1661,('Caucasian','Survived'):1278,('African-American','Survived'):1512}[k]
 mosaic(df_viz,['race','two_year_recid'],labelizer=labelizer,properties=props)
@@
 
-This plot shows the distribution of two year recidivism outcomes.
-Out of all defendants, there is a higher proportion of people who did not recidivate than who did.
-About 2500 people recidivated, whereas approximately 2700 did not.
+A higher percentage of white defendants (60.86%) than black defendants (47.65%) did not commit another crime within two years.
+This means that 39.14% of white defendants did recidivate within two years compared to 52.35% of black defendants.
 

### 2022-04-21 Paper_Draft2/Draft2.Rmd

--- 2022-04-19 Paper_Draft2/Draft2.Rmd
@@
 status: submit
-bibliography: mybibfile.bib
+bibliography: ["mybibfile.bib", "pkgs.bib"]
 csl: ../resources/chicago-author-date.csl
@@
 Jurisdictions across the United States use the COMPAS risk assessment instrument, including but not limited to the New York [@NYDOCCS2019], Massachusetts, Michigan, California [@Jackson2020Setting], and Wisconsin [@WisconsinDOC] Departments of Corrections.
+
 Due to the proprietary nature of the COMPAS algorithm, it is unknown how exactly these recidivism risk scores are calculated; however, a sample COMPAS Risk Assessment Survey has been made publicly available, revealing the algorithm’s input information.
@@
 It contains multiple data sets, including the COMPAS data set that accompanied the @angwin2016machine piece.
+Using the @reticulate package, we conduct our analysis in an RMarkdown document with Python code.
 

### 2022-04-21 Paper_Draft2/Draft2.tex

--- 2022-04-19 Paper_Draft2/Draft2.tex
@@
 \address{%
+
 $^{1}$ \quad Statistical and Data Sciences Smith College Northampton, MA
@@
 \externalbibliography{yes}
-\bibliography{mybibfile.bib}
+\bibliography{mybibfile.bibpkgs.bib}
 

### 2022-04-21 final_metrics_analysis.Rmd

--- 2022-04-20 final_metrics_analysis.Rmd
@@
 
+```{python crosstable of charge degrees by race}
+crosstable=pd.crosstab(df_viz['race'],df_viz['c_charge_degree'])
+crosstable
+```
+
+```{python mosiac of charge degrees by race}
+props={}
+props[('Caucasian','F')]={'facecolor':'red', 'edgecolor':'white'}
+props[('Caucasian','M')]={'facecolor':'red', 'edgecolor':'white'}
+props[('African-American','F')]={'facecolor':'xkcd:aqua','edgecolor':'white'}
+props[('African-American','M')]=        {'facecolor':'xkcd:aqua','edgecolor':'white'}
+labelizer=lambda k:{('Caucasian','F'):1242,('African-American','F'):2194,('Caucasian','M'):858,('African-American','M'):979}[k]
+mosaic(df_viz,['race','c_charge_degree'],labelizer=labelizer,properties=props)
+plt.show()
+```
+
+59.14% of white defendants were charged with felonies versus 69.15% of black defendants.
+Likewise, a higher percentage of white defendants (40.86%) were charged with a misdemeanor than black defendants (30.85%).
+
 ```{python exploratory recidivism bar plot}
@@
 
+```{python crosstable of recitivism by race}
+crosstable=pd.crosstab(df_viz['race'],df_viz['two_year_recid'])
+crosstable
+```
+
+```{python exploratory recidivism mosaic plot}
+props={}
+props[('Caucasian','Recidivated')]={'facecolor':'red', 'edgecolor':'white'}
+props[('Caucasian','Survived')]={'facecolor':'red', 'edgecolor':'white'}
+props[('African-American','Recidivated')]={'facecolor':'xkcd:aqua','edgecolor':'white'}
+props[('African-American','Survived')]=        {'facecolor':'xkcd:aqua','edgecolor':'white'}
+labelizer=lambda k:{('Caucasian','Recidivated'):822,('African-American','Recidivated'):1661,('Caucasian','Survived'):1278,('African-American','Survived'):1512}[k]
+mosaic(df_viz,['race','two_year_recid'],labelizer=labelizer,properties=props)
+plt.show()
+```
+
+A higher percentage of white defendants (60.86%) than black defendants (47.65%) did not commit another crime within two years.
+This means that 39.14% of white defendants did recidivate within two years compared to 52.35% of black defendants.
+
 ```{python exploratory race bar plot}
@@
 ```{python post-processing equal odds needs fix} 
-privileged_groups, unprivileged_groups = get_attributes(train_BLD, selected_attr= 'race')
-ep = EqOddsPostprocessing(unprivileged_groups= unprivileged_groups, privileged_groups= privileged_groups, seed = 42)
+#privileged_groups, unprivileged_groups = get_attributes(train_BLD, selected_attr= 'race')
+#ep = EqOddsPostprocessing(unprivileged_groups= unprivileged_groups, privileged_groups= privileged_groups, seed = 42)
 #eo = PostProcessingMeta(estimator=lr, postprocessor=ep, random_state=1234567)
-EOPP = ep.fit(train_BLD, test_BLD)
-y_pred_EO = ep.predict(test_BLD)
+#EOPP = ep.fit(train_BLD, test_BLD)
+#y_pred_EO = ep.predict(test_BLD)
 #y_proba_EO = ep.predict_(test_BLD)
 
-acc_EO = accuracy_score(y_test, y_pred_EO)
+#acc_EO = accuracy_score(y_test, y_pred_EO)
 ```

### 2022-04-21 final_metrics_analysis.Rmd
adding citations

@@
 # for Dianne
-use_condaenv("base")
+#use_condaenv("base")
 
@@
 
-If a defendant has a higher number of prior offenses, the individual is more likely to be rated with a higher risk of recidivism **do we know this for sure?**.
 In the former plot, we can see that Black defendants have a higher number of prior charges than do white defendants.
@@
 In the latter plot, we can see that male defendants have a higher number of prior charges than do female defendants.
-This is worth noting because the COMPAS risk assessment algorithm has been found to express a discrepancy in prediction between sexes **who found this?**.
-When the recidivsim predictions are compared with the true recividism outcomes, female defendants have a notable false positive rate that predicts their recidivism risk as higher than their true outcomes**where does this come from?**.
+This is worth noting as @larson2016we found that there is discrepancy in COMPAS risk assessment algorithm's prediction between sexes.
+When the recidivsim predictions are compared with the true recividism outcomes, female defendants have a notable false positive rate that predicts their recidivism risk as higher than their true outcomes [@larson2016we].
 

### 2022-04-21 Paper_Draft2/Draft2.Rmd
no more pkgs.bib

--- 2022-04-19 Paper_Draft2/Draft2.Rmd
@@
 
-Due to the proprietary nature of the COMPAS algorithm, it is unknown how exactly these recidivism risk scores are calculated; however, a sample COMPAS Risk Assessment Survey has been made publicly available, revealing the algorithm’s input information.
+Due to the proprietary nature of the COMPAS algorithm, it is unknown how exactly these recidivism risk scores are calculated; however, a [sample COMPAS Risk Assessment Survey](https://www.documentcloud.org/documents/2702103-Sample-Risk-Assessment-COMPAS-CORE#document/p5/a296598) has been made publicly available, revealing the algorithm’s input information.
 This survey has been widely critiqued for using proxy variables for race that do not explicitly factor in a defendant’s race but heavily imply it, allowing Northpointe to claim that their algorithm is free of racial bias.
@@
 [Women at the Table](https://www.womenatthetable.net/), the sponsor organization for this project, is “a growing, global gender equality & democracy CSO based in Geneva, Switzerland focused on advancing feminist systems change by using the prism of technology, innovation & AI exercising leverage points in technology, the economy, sustainability & democratic governance.” We are collaborating with the organization on its AI & Equality [@noauthor_ai_nodate] initiative, tasked with de-biasing the COMPAS algorithm [@aif360-oct-2018] and producing a corresponding data story that will be added to its library.
+
 Our project builds on Women at the Table’s various de-biasing algorithms used in its AI & Equality Human Rights Toolbox to conduct our own analyses on the COMPAS data set.

### 2022-04-21 Paper_Draft2/Draft2.tex

--- 2022-04-19 Paper_Draft2/Draft2.tex
@@
 Northpointe (now known by its parent company
-\href{https://www.equivant.com/faq/}{Equivant}), to predict defendants'
+\href{https://www.equivant.com/faq/}{equivant}), to predict defendants'
 risk of recidivism. It generates a decile score that classifies
@@
 the COMPAS risk assessment instrument, including but not limited to the
-New York \citep{NYDOCCS2019}, Massachusetts, Michigan, California
-\citep{Jackson2020Setting}, and Wisconsin \citep{WisconsinDOC}
+\href{https://doccs.ny.gov/system/files/documents/2020/11/8500.pdf}{New
+York},
+\href{https://hdsr.mitpress.mit.edu/pub/hzwo7ax4/release/4}{Massachusetts},
+\href{https://hdsr.mitpress.mit.edu/pub/hzwo7ax4/release/4}{Michigan},
+\href{https://hdsr.mitpress.mit.edu/pub/hzwo7ax4/release/4}{California},
+and \href{https://doc.wi.gov/Pages/AboutDOC/COMPAS.aspx}{Wisconsin}
 Departments of Corrections.
@@
 Due to the proprietary nature of the COMPAS algorithm, it is unknown how
-exactly these recidivism risk scores are calculated; however, a sample
-COMPAS Risk Assessment Survey has been made publicly available,
+exactly these recidivism risk scores are calculated; however, a
+\href{https://www.documentcloud.org/documents/2702103-Sample-Risk-Assessment-COMPAS-CORE\#document/p5/a296598}{sample
+COMPAS Risk Assessment Survey} has been made publicly available,
 revealing the algorithm's input information. This survey has been widely
@@
 de-biasing the COMPAS algorithm \citep{aif360-oct-2018} and producing a
-corresponding data story that will be added to its library. Our project
-builds on Women at the Table's various de-biasing algorithms used in its
-AI \& Equality Human Rights Toolbox to conduct our own analyses on the
-COMPAS data set. Based on this analysis, we employ a human rights
-framework to contribute to the ProPublica and Northpointe debate and
-investigate whether or to what extent there is racial bias in the COMPAS
-algorithm. With a solid understanding of the two sides, we aim to
-pinpoint the shortcomings of both arguments and correct them in our
-analyses. We will use various de-biasing techniques and fairness metrics
-to evaluate the level of bias present in the COMPAS data and our
+corresponding data story that will be added to its library.
+
+Our project builds on Women at the Table's various de-biasing algorithms
+used in its AI \& Equality Human Rights Toolbox to conduct our own
+analyses on the COMPAS data set. Based on this analysis, we employ a
+human rights framework to contribute to the ProPublica and Northpointe
+debate and investigate whether or to what extent there is racial bias in
+the COMPAS algorithm. With a solid understanding of the two sides, we
+aim to pinpoint the shortcomings of both arguments and correct them in
+our analyses. We will use various de-biasing techniques and fairness
+metrics to evaluate the level of bias present in the COMPAS data and our
 algorithm. We will summarize our results using the JupyterNotebook

### 2022-04-21 Paper_Draft2/Draft2.Rmd
updating author section again

@@
       Northampton, MA 01063
-    email: aacharya@smith.edu, dcaravela@smith.edu, ekim89@smith.edu, ekornberg@smith.edu, enesmith@smith.edu
+    email: bbaumer@smith.edu
 
@@
 correspondence: |
-  bbaumer@smith.edu
+  aacharya@smith.edu, dcaravela@smith.edu, ekim89@smith.edu, ekornberg@smith.edu, enesmith@smith.edu
 

### 2022-04-21 Paper_Draft2/Draft2.tex
updating author section again

@@
 $^{1}$ \quad Statistical and Data Sciences Smith College Northampton, MA
-01063; \href{mailto:aacharya@smith.edu}{\nolinkurl{aacharya@smith.edu}},
+01063; \href{mailto:bbaumer@smith.edu}{\nolinkurl{bbaumer@smith.edu}}\\
+}
+% Contact information of the corresponding author
+\corres{Correspondence: \href{mailto:aacharya@smith.edu}{\nolinkurl{aacharya@smith.edu}},
 \href{mailto:dcaravela@smith.edu}{\nolinkurl{dcaravela@smith.edu}},
@@
 \href{mailto:ekornberg@smith.edu}{\nolinkurl{ekornberg@smith.edu}},
-\href{mailto:enesmith@smith.edu}{\nolinkurl{enesmith@smith.edu}}\\
-}
-% Contact information of the corresponding author
-\corres{Correspondence: \href{mailto:bbaumer@smith.edu}{\nolinkurl{bbaumer@smith.edu}}}
+\href{mailto:enesmith@smith.edu}{\nolinkurl{enesmith@smith.edu}}}
 

### 2022-04-21 Paper_Draft2/Draft2.tex

@@
 
-{\centering \includegraphics[width=1\linewidth]{../images/race_bar_plot_new}
-
-}
-
-\caption{This plot shows the distribution of defendant races.
-The dataset contains about 3200 Black defendants and 2100 white defendants.
-Therefore, there are more Black people represented in the dataset than white people.}\label{fig:race plot}
-\end{figure}
-
-\begin{figure}
-
 {\centering \includegraphics[width=1\linewidth]{../images/age_race_plot_new}
@@
 For both groups, the majority of defendants are relatively young, but this is especially noticeable for Black defendants.}\label{fig:age plot}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{../images/recid_bar_plot_new}
-
-}
-
-\caption{This plot shows the distribution of two year recidivism outcomes.
-Out of all defendants, there is a higher proportion of people who did not recidivate than who did.
-About 2500 people recidivated, whereas approximately 2700 did not.}\label{fig:recid plot}
 \end{figure}

### 2022-04-21 Paper_Draft2/Draft2.Rmd
methods

--- 2022-04-21 Paper_Draft2/Draft2.Rmd
@@
 
-Our project builds on Women at the Table’s various de-biasing algorithms used in its AI & Equality Human Rights Toolbox to conduct our own analyses on the COMPAS data set.
+Our project builds on Women at the Table’s various de-biasing algorithms used in its [AI & Equality Human Rights Toolbox](https://colab.research.google.com/drive/11SWijBDkr1pSgIXyjUkgoSB0gh5oAlaI) to conduct our own analyses on the COMPAS data set.
 Based on this analysis, we employ a human rights framework to contribute to the ProPublica and Northpointe debate and investigate whether or to what extent there is racial bias in the COMPAS algorithm.
@@
 
-**Statistical Parity Difference** - This metric measures the difference that the majority and protected classes get a particular outcome.
-The ideal value of this metric is 0.
-Fairness for this metric is between -0.1 and 0.1.
-A negative value means there is higher benefit for the privileged group (in this case, white defendants).
-
-$$P(\hat{Y}=1|D=Unprivileged) - P(\hat{Y}=1|D=Privileged)$$
-
-**Disparate Impact Ratio** - This metric is the ratio of how often the favorable outcome occurs in one group versus the other.
-In the case of recidivism, this is the ratio of how many white defendants are predicted to not recidivate compared to how many Black defendants are predicted to not recidivate.
-A value of 1 means that the ratio is exactly 1:1.
-Less than 1 means the privileged group (white defendants) benefits, while a value greater than 1 means the unprivileged group (Black defendants) benefits.
-According to AIF360, a ratio between .8 to 1.25 is considered fair[@Ronaghan2019AI].
-
-$$\frac{P(\hat{Y}=1|D=Unprivileged)}{P(\hat{Y}=1|D=Privileged)}$$
-
-**Average Odds Difference** - This metric returns the average difference in false positive rate and true positive rate for the privileged and unprivileged groups.
-A value of 0 indicates equality of odds, and a value below 0 implies benefit for the privileged group.
-Equality of odds is achieved in the case of recidivism when the proportion of people who were predicted to recidivate and did recidivate is equal (true positive rate) for both Black and white defendants AND the proportion of people who were predicted to recidivate and did not recidivate (false positive rate) is equal for both Black and white defendants[@aif360-oct-2018].
-
-$$\frac{(FPR_{D = unprivileged} - FPR_{D = privileged}) + (TPR_{D = unprivileged} - TPR_{D = privileged})}{2}$$
-
-**Equal Opportunity Difference** - This metric is computed as the difference of true positive rates between the unprivileged and the privileged groups.
-The true positive rate is the ratio of true positives to the total number of actual positives for a given group[@GoogleDev].
-
-The ideal value is 0. A value less than 0 implies higher benefit for the privileged group and a value greater than 0 implies higher benefit for the unprivileged group.
-Fairness for this metric is between -0.1 and 0.1 [@aif360-oct-2018].
-
-This metric is best used when it is very important to catch positive outcomes while false positives are not exceptionally problematic [@Cortez2019How].
-This is not the case for the COMPAS data set, as false positives mean extra jail time for someone who will not actually re-offend.
-
-$$TPR_{D = Unprivileged} - TPR_{D = Privileged}$$
-
 For the next step of our experiment, we need to determine where in the data science pipeline we can mitigate the most bias, using pre-processing, in-processing, and post-processing de-biasing algorithms.
@@
 Based on these results, we look comparatively at all the approaches to determine which we assess to be the most fair.
+
+## Statistical Parity Difference
+
+This metric measures the difference that the majority and protected classes get a particular outcome.
+The ideal value of this metric is 0.
+Fairness for this metric is between -0.1 and 0.1.
+A negative value means there is higher benefit for the privileged group (in this case, white defendants).
+
+$$P(\hat{Y}=1|D=Unprivileged) - P(\hat{Y}=1|D=Privileged)$$
+
+## Disparate Impact Ratio
+
+This metric is the ratio of how often the favorable outcome occurs in one group versus the other.
+In the case of recidivism, this is the ratio of how many white defendants are predicted to not recidivate compared to how many Black defendants are predicted to not recidivate.
+A value of 1 means that the ratio is exactly 1:1.
+Less than 1 means the privileged group (white defendants) benefits, while a value greater than 1 means the unprivileged group (Black defendants) benefits.
+According to AIF360, a ratio between .8 to 1.25 is considered fair[@Ronaghan2019AI].
+
+$$\frac{P(\hat{Y}=1|D=Unprivileged)}{P(\hat{Y}=1|D=Privileged)}$$
+
+## Average Odds Difference
+
+This metric returns the average difference in false positive rate and true positive rate for the privileged and unprivileged groups.
+A value of 0 indicates equality of odds, and a value below 0 implies benefit for the privileged group.
+Equality of odds is achieved in the case of recidivism when the proportion of people who were predicted to recidivate and did recidivate is equal (true positive rate) for both Black and white defendants AND the proportion of people who were predicted to recidivate and did not recidivate (false positive rate) is equal for both Black and white defendants[@aif360-oct-2018].
+
+$$\frac{(FPR_{D = unprivileged} - FPR_{D = privileged}) + (TPR_{D = unprivileged} - TPR_{D = privileged})}{2}$$
+
+## Equal Opportunity Difference
+
+This metric is computed as the difference of true positive rates between the unprivileged and the privileged groups.
+The true positive rate is the ratio of true positives to the total number of actual positives for a given group[@GoogleDev].
+
+The ideal value is 0. A value less than 0 implies higher benefit for the privileged group and a value greater than 0 implies higher benefit for the unprivileged group.
+Fairness for this metric is between -0.1 and 0.1 [@aif360-oct-2018].
+
+This metric is best used when it is very important to catch positive outcomes while false positives are not exceptionally problematic [@Cortez2019How].
+This is not the case for the COMPAS data set, as false positives mean extra jail time for someone who will not actually re-offend.
+
+$$TPR_{D = Unprivileged} - TPR_{D = Privileged}$$
 

### 2022-04-21 Paper_Draft2/Draft2.tex
methods

--- 2022-04-21 Paper_Draft2/Draft2.tex
@@
 Our project builds on Women at the Table's various de-biasing algorithms
-used in its AI \& Equality Human Rights Toolbox to conduct our own
-analyses on the COMPAS data set. Based on this analysis, we employ a
-human rights framework to contribute to the ProPublica and Northpointe
-debate and investigate whether or to what extent there is racial bias in
-the COMPAS algorithm. With a solid understanding of the two sides, we
-aim to pinpoint the shortcomings of both arguments and correct them in
-our analyses. We will use various de-biasing techniques and fairness
-metrics to evaluate the level of bias present in the COMPAS data and our
+used in its
+\href{https://colab.research.google.com/drive/11SWijBDkr1pSgIXyjUkgoSB0gh5oAlaI}{AI
+\& Equality Human Rights Toolbox} to conduct our own analyses on the
+COMPAS data set. Based on this analysis, we employ a human rights
+framework to contribute to the ProPublica and Northpointe debate and
+investigate whether or to what extent there is racial bias in the COMPAS
+algorithm. With a solid understanding of the two sides, we aim to
+pinpoint the shortcomings of both arguments and correct them in our
+analyses. We will use various de-biasing techniques and fairness metrics
+to evaluate the level of bias present in the COMPAS data and our
 algorithm. We will summarize our results using the JupyterNotebook
@@
 
-\textbf{Statistical Parity Difference} - This metric measures the
-difference that the majority and protected classes get a particular
-outcome. The ideal value of this metric is 0. Fairness for this metric
-is between -0.1 and 0.1. A negative value means there is higher benefit
-for the privileged group (in this case, white defendants).
-
-\[P(\hat{Y}=1|D=Unprivileged) - P(\hat{Y}=1|D=Privileged)\]
-
-\textbf{Disparate Impact Ratio} - This metric is the ratio of how often
-the favorable outcome occurs in one group versus the other. In the case
-of recidivism, this is the ratio of how many white defendants are
-predicted to not recidivate compared to how many Black defendants are
-predicted to not recidivate. A value of 1 means that the ratio is
-exactly 1:1. Less than 1 means the privileged group (white defendants)
-benefits, while a value greater than 1 means the unprivileged group
-(Black defendants) benefits. According to AIF360, a ratio between .8 to
-1.25 is considered fair\citep{Ronaghan2019AI}.
-
-\[\frac{P(\hat{Y}=1|D=Unprivileged)}{P(\hat{Y}=1|D=Privileged)}\]
-
-\textbf{Average Odds Difference} - This metric returns the average
-difference in false positive rate and true positive rate for the
-privileged and unprivileged groups. A value of 0 indicates equality of
-odds, and a value below 0 implies benefit for the privileged group.
-Equality of odds is achieved in the case of recidivism when the
-proportion of people who were predicted to recidivate and did recidivate
-is equal (true positive rate) for both Black and white defendants AND
-the proportion of people who were predicted to recidivate and did not
-recidivate (false positive rate) is equal for both Black and white
-defendants\citep{aif360-oct-2018}.
-
-\[\frac{(FPR_{D = unprivileged} - FPR_{D = privileged}) + (TPR_{D = unprivileged} - TPR_{D = privileged})}{2}\]
-
-\textbf{Equal Opportunity Difference} - This metric is computed as the
-difference of true positive rates between the unprivileged and the
-privileged groups. The true positive rate is the ratio of true positives
-to the total number of actual positives for a given
-group\citep{GoogleDev}.
-
-The ideal value is 0. A value less than 0 implies higher benefit for the
-privileged group and a value greater than 0 implies higher benefit for
-the unprivileged group. Fairness for this metric is between -0.1 and 0.1
-\citep{aif360-oct-2018}.
-
-This metric is best used when it is very important to catch positive
-outcomes while false positives are not exceptionally problematic
-\citep{Cortez2019How}. This is not the case for the COMPAS data set, as
-false positives mean extra jail time for someone who will not actually
-re-offend.
-
-\[TPR_{D = Unprivileged} - TPR_{D = Privileged}\]
-
 For the next step of our experiment, we need to determine where in the
@@
 the most fair.
+
+\hypertarget{statistical-parity-difference}{%
+\subsection{Statistical Parity
+Difference}\label{statistical-parity-difference}}
+
+This metric measures the difference that the majority and protected
+classes get a particular outcome. The ideal value of this metric is 0.
+Fairness for this metric is between -0.1 and 0.1. A negative value means
+there is higher benefit for the privileged group (in this case, white
+defendants).
+
+\[P(\hat{Y}=1|D=Unprivileged) - P(\hat{Y}=1|D=Privileged)\]
+
+\hypertarget{disparate-impact-ratio}{%
+\subsection{Disparate Impact Ratio}\label{disparate-impact-ratio}}
+
+This metric is the ratio of how often the favorable outcome occurs in
+one group versus the other. In the case of recidivism, this is the ratio
+of how many white defendants are predicted to not recidivate compared to
+how many Black defendants are predicted to not recidivate. A value of 1
+means that the ratio is exactly 1:1. Less than 1 means the privileged
+group (white defendants) benefits, while a value greater than 1 means
+the unprivileged group (Black defendants) benefits. According to AIF360,
+a ratio between .8 to 1.25 is considered fair\citep{Ronaghan2019AI}.
+
+\[\frac{P(\hat{Y}=1|D=Unprivileged)}{P(\hat{Y}=1|D=Privileged)}\]
+
+\hypertarget{average-odds-difference}{%
+\subsection{Average Odds Difference}\label{average-odds-difference}}
+
+This metric returns the average difference in false positive rate and
+true positive rate for the privileged and unprivileged groups. A value
+of 0 indicates equality of odds, and a value below 0 implies benefit for
+the privileged group. Equality of odds is achieved in the case of
+recidivism when the proportion of people who were predicted to
+recidivate and did recidivate is equal (true positive rate) for both
+Black and white defendants AND the proportion of people who were
+predicted to recidivate and did not recidivate (false positive rate) is
+equal for both Black and white defendants\citep{aif360-oct-2018}.
+
+\[\frac{(FPR_{D = unprivileged} - FPR_{D = privileged}) + (TPR_{D = unprivileged} - TPR_{D = privileged})}{2}\]
+
+\hypertarget{equal-opportunity-difference}{%
+\subsection{Equal Opportunity
+Difference}\label{equal-opportunity-difference}}
+
+This metric is computed as the difference of true positive rates between
+the unprivileged and the privileged groups. The true positive rate is
+the ratio of true positives to the total number of actual positives for
+a given group\citep{GoogleDev}.
+
+The ideal value is 0. A value less than 0 implies higher benefit for the
+privileged group and a value greater than 0 implies higher benefit for
+the unprivileged group. Fairness for this metric is between -0.1 and 0.1
+\citep{aif360-oct-2018}.
+
+This metric is best used when it is very important to catch positive
+outcomes while false positives are not exceptionally problematic
+\citep{Cortez2019How}. This is not the case for the COMPAS data set, as
+false positives mean extra jail time for someone who will not actually
+re-offend.
+
+\[TPR_{D = Unprivileged} - TPR_{D = Privileged}\]
 

### 2022-04-21 Paper_Draft2/Draft2.Rmd

@@
 
-```{r race plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of defendant races.
-The dataset contains about 3200 Black defendants and 2100 white defendants.
-Therefore, there are more Black people represented in the dataset than white people.', out.width= "100%"}
-knitr::include_graphics("../images/race_bar_plot_new.png")
-```
-
 ```{r age plot, echo=FALSE, fig.align='center', fig.cap= "This plot shows the distribution of defendant ages by race.
@@
 knitr::include_graphics("../images/age_race_plot_new.png")
-```
-
-```{r recid plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes.
-Out of all defendants, there is a higher proportion of people who did not recidivate than who did.
-About 2500 people recidivated, whereas approximately 2700 did not.', out.width= "100%"}
-knitr::include_graphics("../images/recid_bar_plot_new.png")
 ```

### 2022-04-21 Paper_Draft2/Draft2.Rmd
knit

@@
 A value of 0 indicates equality of odds, and a value below 0 implies benefit for the privileged group.
-Equality of odds is achieved in the case of recidivism when the proportion of people who were predicted to recidivate and did recidivate is equal (true positive rate) for both Black and white defendants AND the proportion of people who were predicted to recidivate and did not recidivate (false positive rate) is equal for both Black and white defendants[@aif360-oct-2018].
+Equality of odds is achieved in the case of recidivism when the proportion of people who were predicted to recidivate and did recidivate is equal (true positive rate) for both Black and white defendants AND the proportion of people who were predicted to recidivate and did not recidivate (false positive rate) is equal for both Black and white defendants [@aif360-oct-2018].
 

### 2022-04-21 final_metrics_analysis.Rmd
wordcloud (how to install wordcloud package: https://stackoverflow.com/questions/47298070/importerror-no-module-named-wordcloud) changed color on mosaic plots

--- 2022-04-21 final_metrics_analysis.Rmd
@@
 props={}
-props[('Caucasian','Recidivated')]={'facecolor':'red', 'edgecolor':'white'}
-props[('Caucasian','Survived')]={'facecolor':'red', 'edgecolor':'white'}
-props[('African-American','Recidivated')]={'facecolor':'xkcd:aqua','edgecolor':'white'}
-props[('African-American','Survived')]=        {'facecolor':'xkcd:aqua','edgecolor':'white'}
+props[('Caucasian','Recidivated')]={'facecolor':'green', 'edgecolor':'white'}
+props[('Caucasian','Survived')]={'facecolor':'green', 'edgecolor':'white'}
+props[('African-American','Recidivated')]={'facecolor':'xkcd:purple','edgecolor':'white'}
+props[('African-American','Survived')]=        {'facecolor':'xkcd:purple','edgecolor':'white'}
 labelizer=lambda k:{('Caucasian','Recidivated'):822,('African-American','Recidivated'):1661,('Caucasian','Survived'):1278,('African-American','Survived'):1512}[k]
@@
 This means that 39.14% of white defendants did recidivate within two years compared to 52.35% of black defendants.
+
+```{python wordcloud setup}
+from os import path
+from PIL import Image
+from wordcloud import WordCloud, STOPWORDS, ImageColorGenerator
+```
+
+```{python wordcloud}
+text = " ".join(charge for charge in df_viz.c_charge_desc)
+stopwords = set(STOPWORDS)
+stopwords.update(["arrest", "case"])
+wordcloud = WordCloud(stopwords=stopwords, background_color="white").generate(text)
+```
+
+```{python wordcloud image}
+plt.imshow(wordcloud, interpolation='bilinear')
+plt.axis("off")
+plt.show()
+```
 

### 2022-04-21 final_metrics_analysis.Rmd
add word charge

@@
 stopwords = set(STOPWORDS)
-stopwords.update(["arrest", "case"])
+stopwords.update(["arrest", "case", "charge"])
 wordcloud = WordCloud(stopwords=stopwords, background_color="white").generate(text)

### 2022-04-21 Paper_Draft2/Draft2.Rmd

--- 2022-04-21 Paper_Draft2/Draft2.Rmd
@@
 
-Our project builds on Women at the Table’s various de-biasing algorithms used in its [AI & Equality Human Rights Toolbox](https://colab.research.google.com/drive/11SWijBDkr1pSgIXyjUkgoSB0gh5oAlaI) to conduct our own analyses on the COMPAS data set.
+Our project builds on Women at the Table’s various de-biasing algorithms used in its AI & Equality Human Rights Toolbox to conduct our own analyses on the COMPAS data set.
 Based on this analysis, we employ a human rights framework to contribute to the ProPublica and Northpointe debate and investigate whether or to what extent there is racial bias in the COMPAS algorithm.
@@
 
+**Statistical Parity Difference** - This metric measures the difference that the majority and protected classes get a particular outcome.
+The ideal value of this metric is 0.
+Fairness for this metric is between -0.1 and 0.1.
+A negative value means there is higher benefit for the privileged group (in this case, white defendants).
+
+$$P(\hat{Y}=1|D=Unprivileged) - P(\hat{Y}=1|D=Privileged)$$
+
+**Disparate Impact Ratio** - This metric is the ratio of how often the favorable outcome occurs in one group versus the other.
+In the case of recidivism, this is the ratio of how many white defendants are predicted to not recidivate compared to how many Black defendants are predicted to not recidivate.
+A value of 1 means that the ratio is exactly 1:1.
+Less than 1 means the privileged group (white defendants) benefits, while a value greater than 1 means the unprivileged group (Black defendants) benefits.
+According to AIF360, a ratio between .8 to 1.25 is considered fair[@Ronaghan2019AI].
+
+$$\frac{P(\hat{Y}=1|D=Unprivileged)}{P(\hat{Y}=1|D=Privileged)}$$
+
+**Average Odds Difference** - This metric returns the average difference in false positive rate and true positive rate for the privileged and unprivileged groups.
+A value of 0 indicates equality of odds, and a value below 0 implies benefit for the privileged group.
+Equality of odds is achieved in the case of recidivism when the proportion of people who were predicted to recidivate and did recidivate is equal (true positive rate) for both Black and white defendants AND the proportion of people who were predicted to recidivate and did not recidivate (false positive rate) is equal for both Black and white defendants[@aif360-oct-2018].
+
+$$\frac{(FPR_{D = unprivileged} - FPR_{D = privileged}) + (TPR_{D = unprivileged} - TPR_{D = privileged})}{2}$$
+
+**Equal Opportunity Difference** - This metric is computed as the difference of true positive rates between the unprivileged and the privileged groups.
+The true positive rate is the ratio of true positives to the total number of actual positives for a given group[@GoogleDev].
+
+The ideal value is 0. A value less than 0 implies higher benefit for the privileged group and a value greater than 0 implies higher benefit for the unprivileged group.
+Fairness for this metric is between -0.1 and 0.1 [@aif360-oct-2018].
+
+This metric is best used when it is very important to catch positive outcomes while false positives are not exceptionally problematic [@Cortez2019How].
+This is not the case for the COMPAS data set, as false positives mean extra jail time for someone who will not actually re-offend.
+
+$$TPR_{D = Unprivileged} - TPR_{D = Privileged}$$
+
 For the next step of our experiment, we need to determine where in the data science pipeline we can mitigate the most bias, using pre-processing, in-processing, and post-processing de-biasing algorithms.
@@
 Based on these results, we look comparatively at all the approaches to determine which we assess to be the most fair.
-
-## Statistical Parity Difference
-
-This metric measures the difference that the majority and protected classes get a particular outcome.
-The ideal value of this metric is 0.
-Fairness for this metric is between -0.1 and 0.1.
-A negative value means there is higher benefit for the privileged group (in this case, white defendants).
-
-$$P(\hat{Y}=1|D=Unprivileged) - P(\hat{Y}=1|D=Privileged)$$
-
-## Disparate Impact Ratio
-
-This metric is the ratio of how often the favorable outcome occurs in one group versus the other.
-In the case of recidivism, this is the ratio of how many white defendants are predicted to not recidivate compared to how many Black defendants are predicted to not recidivate.
-A value of 1 means that the ratio is exactly 1:1.
-Less than 1 means the privileged group (white defendants) benefits, while a value greater than 1 means the unprivileged group (Black defendants) benefits.
-According to AIF360, a ratio between .8 to 1.25 is considered fair[@Ronaghan2019AI].
-
-$$\frac{P(\hat{Y}=1|D=Unprivileged)}{P(\hat{Y}=1|D=Privileged)}$$
-
-## Average Odds Difference
-
-This metric returns the average difference in false positive rate and true positive rate for the privileged and unprivileged groups.
-A value of 0 indicates equality of odds, and a value below 0 implies benefit for the privileged group.
-Equality of odds is achieved in the case of recidivism when the proportion of people who were predicted to recidivate and did recidivate is equal (true positive rate) for both Black and white defendants AND the proportion of people who were predicted to recidivate and did not recidivate (false positive rate) is equal for both Black and white defendants[@aif360-oct-2018].
-
-$$\frac{(FPR_{D = unprivileged} - FPR_{D = privileged}) + (TPR_{D = unprivileged} - TPR_{D = privileged})}{2}$$
-
-## Equal Opportunity Difference
-
-This metric is computed as the difference of true positive rates between the unprivileged and the privileged groups.
-The true positive rate is the ratio of true positives to the total number of actual positives for a given group[@GoogleDev].
-
-The ideal value is 0. A value less than 0 implies higher benefit for the privileged group and a value greater than 0 implies higher benefit for the unprivileged group.
-Fairness for this metric is between -0.1 and 0.1 [@aif360-oct-2018].
-
-This metric is best used when it is very important to catch positive outcomes while false positives are not exceptionally problematic [@Cortez2019How].
-This is not the case for the COMPAS data set, as false positives mean extra jail time for someone who will not actually re-offend.
-
-$$TPR_{D = Unprivileged} - TPR_{D = Privileged}$$
 

### 2022-04-21 final_metrics_analysis.Rmd
making faceted priors plot

--- 2022-04-21 final_metrics_analysis.Rmd
@@
 # for Dianne
-use_condaenv("base")
+# use_condaenv("base")
 
@@
 
-```{python crosstable of recitivism by race}
+```{python crosstable of recidivism by race}
 crosstable=pd.crosstab(df_viz['race'],df_viz['two_year_recid'])
@@
 props={}
-props[('Caucasian','Recidivated')]={'facecolor':'green', 'edgecolor':'white'}
-props[('Caucasian','Survived')]={'facecolor':'green', 'edgecolor':'white'}
-props[('African-American','Recidivated')]={'facecolor':'xkcd:purple','edgecolor':'white'}
-props[('African-American','Survived')]=        {'facecolor':'xkcd:purple','edgecolor':'white'}
+props[('Caucasian','Recidivated')]={'facecolor':'red', 'edgecolor':'white'}
+props[('Caucasian','Survived')]={'facecolor':'red', 'edgecolor':'white'}
+props[('African-American','Recidivated')]={'facecolor':'xkcd:aqua','edgecolor':'white'}
+props[('African-American','Survived')]=        {'facecolor':'xkcd:aqua','edgecolor':'white'}
 labelizer=lambda k:{('Caucasian','Recidivated'):822,('African-American','Recidivated'):1661,('Caucasian','Survived'):1278,('African-American','Survived'):1512}[k]
@@
 This means that 39.14% of white defendants did recidivate within two years compared to 52.35% of black defendants.
-
-```{python wordcloud setup}
-from os import path
-from PIL import Image
-from wordcloud import WordCloud, STOPWORDS, ImageColorGenerator
-```
-
-```{python wordcloud}
-text = " ".join(charge for charge in df_viz.c_charge_desc)
-stopwords = set(STOPWORDS)
-stopwords.update(["arrest", "case"])
-wordcloud = WordCloud(stopwords=stopwords, background_color="white").generate(text)
-```
-
-```{python wordcloud image}
-plt.imshow(wordcloud, interpolation='bilinear')
-plt.axis("off")
-plt.show()
-```
 
@@
 
-```{python exploratory recidivism by race table}
-# table of recidivism by race
-pd.crosstab(index = df_viz["race"], columns = df_viz["two_year_recid"])
-```
+```{python exploratory priors by race}
+priors_by_race = sns.displot(x="priors_count", row = "race", col = "sex", hue ="race", data=df_viz, palette=workshop_palette)
 
-```{python exploratory priors by race}
-priors_by_race = sns.countplot(x="race", color="priors_count", data=df_viz, palette=workshop_palette)
-
-plt.title('Prior Charges Count by Race')
-plt.show()
-```
-
-```{python exploratory priors by sex}
-priors_by_race = sns.countplot(x="sex", color="priors_count", data=df_viz, palette=workshop_palette)
-
-plt.title('Prior Charges Count by Sex')
+#plt.title("Prior Charges Count by Race and Sex)
 plt.show()

### 2022-04-21 final_metrics_analysis.Rmd

--- 2022-04-21 final_metrics_analysis.Rmd
@@
 # for Dianne
-#use_condaenv("base")
+# use_condaenv("base")
 

### 2022-04-21 final_metrics_analysis.Rmd
Interpretations and post-processing stuff

--- 2022-04-21 final_metrics_analysis.Rmd
@@
 # for Dianne
-# use_condaenv("base")
+use_condaenv("base")
 
 # for Amrita
-#use_condaenv("anaconda3")
+use_condaenv("anaconda3")
 ```
@@
   # row for baseline model
-  fair_metrics.loc['Baseline Model'] = [stat_par_diff, eq_opp_diff, avg_odds_diff, disp_impact_ratio]
+  fair_metrics.loc['Baseline'] = [stat_par_diff, eq_opp_diff, avg_odds_diff, disp_impact_ratio]
   # row for mitigated bias
@@
 
-```{python crosstable of recidivism by race}
+```{python crosstable of recitivism by race}
 crosstable=pd.crosstab(df_viz['race'],df_viz['two_year_recid'])
@@
 # age distribution by race - hacking solution for purple dotted line for now
-ax = sns.kdeplot(x="age", hue="race", data=df_viz, palette=workshop_palette);
-kdeline = ax.lines[0]
-mean_black = df_viz.groupby('race').age.median()[0]
-xs = kdeline.get_xdata()
-ys = kdeline.get_ydata()
-height_black = 0.027 # np.interp(mean_black, xs, ys)
-ax.vlines(mean_black, 0, height_black, color=purple, ls=':')
+#ax = sns.kdeplot(x="age", hue="race", data=df_viz, palette=workshop_palette);
+#kdeline = ax.lines[0]
+#mean_black = df_viz.groupby('race').age.median()[0]
+#xs = kdeline.get_xdata()
+#ys = kdeline.get_ydata()
+#height_black = 0.027 # np.interp(mean_black, xs, ys)
+#ax.vlines(mean_black, 0, height_black, color=purple, ls=':')
 
-mean_white = df_viz.groupby('race').age.median()[1]
-height_white = np.interp(mean_white, xs, ys)
-ax.vlines(mean_white, 0, height_white, color=green, ls=':')
+#mean_white = df_viz.groupby('race').age.median()[1]
+#height_white = np.interp(mean_white, xs, ys)
+#ax.vlines(mean_white, 0, height_white, color=green, ls=':')
 
-plt.title('Distribution of Age by Race')
+#plt.title('Distribution of Age by Race')
 ```
@@
 
+```{python exploratory recidivism by race table}
+# table of recidivism by race
+pd.crosstab(index = df_viz["race"], columns = df_viz["two_year_recid"])
+```
+
 ```{python exploratory priors by race}
-priors_by_race = sns.displot(x="priors_count", row = "race", col = "sex", hue ="race", data=df_viz, palette=workshop_palette)
+priors_by_race = sns.countplot(x="race", color="priors_count", data=df_viz, palette=workshop_palette)
 
-#plt.title("Prior Charges Count by Race and Sex)
+plt.title('Prior Charges Count by Race')
+plt.show()
+```
+
+```{python exploratory priors by sex}
+priors_by_race = sns.countplot(x="sex", color="priors_count", data=df_viz, palette=workshop_palette)
+
+plt.title('Prior Charges Count by Sex')
 plt.show()
@@
 
+The highest percentage in this confusion matrix is for false positives, defendants who the model predicted to recidivate and actually did not, at 37.76%.
+The number of true positives, defendants who the model predicted to recidivate and actually recidivated, is lower at 23.88%.
+The percentage of false negatives is lower, whereas the true negative is low at 14.33%.
+This indicates that the model does not do a good job in predicting the defendents who don't recidivate.
+It is very likely that it will incorrectly predict that someone will recidivate.
+
 ```{python prejudice remover metrics per group}
@@
 
+Compared to the baseline model, the predjudice remover is highly in favor of the unprivileged group because its value is greater than 0. The value of .367 is much higher than 0, the value of fairness.
+
 ### Equal Opportunity Difference
@@
 ```
+
+Compared to the baseline model, the predjudice remover is highly in favor of the unprivileged group because its value is greater than 0. The value of .321 is much higher than 0, the value of fairness.
 
@@
 
+Compared to the baseline model, the predjudice remover is highly in favor of the unprivileged group because its value is greater than 0.
+
 ### Disparate Impact Ratio
@@
 ```
+
+The value of 3.3 indicates that the model is highly in favor of Black defendents, as it is much greater than 1. This is a hugely different from the baseline model which favors white defendants.
 
@@
 plt.show()
+```
+
+The highest percentage in this confusion matrix is for true negatives, defendants who the model predicted to not recidivate and actually did not, at 50.04%.
+The number of true positives, defendants who the model predicted to recidivate and actually recidivated, is much lower at 3.11%.
+The percentage of false negatives is somewhat high for this baseline model, whereas the false positive rate is very low at 2.05%.
+This indicates that the model does a good job in predicting the defendents who don't recidivate.
+It is very unlikely that it will incorrectly predict that someone will recidivate.
+
+```{python}
+#metrics_per_group(y_test, y_pred_CEO)
+#plt.show()
 ```
@@
 
+At the value of 0.134, the statistical parity difference is now positive and slightly in favor of black defendants.
+
 ### Equal Opportunity Difference
@@
 ```
+
+There has been a significant change from -.408 to 0.088, showing that now the value is within the bounds of "fair" (-0.1 to 0.1)
 
@@
 
+The baseline model with the average odds difference was -0.438 and now it is 0.15 which indicates being slightly in favor of black defendants.
+
 ### Disparate Impact Ratio
@@
 
+The baseline value was 0.466 which is highly out of the range of "fair" which goes up to 1. Now at a value of 1.2 it is much higher than 1.
+
 ### Plot Fairness Metrics (Calibrated Equalized Odds)

### 2022-04-21 Paper_Draft2/Draft2.tex

--- 2022-04-21 Paper_Draft2/Draft2.tex
@@
 Due to the proprietary nature of the COMPAS algorithm, it is unknown how
-exactly these recidivism risk scores are calculated; however, a
+exactly these recidivism risk scores are calculated. However, a
 \href{https://www.documentcloud.org/documents/2702103-Sample-Risk-Assessment-COMPAS-CORE\#document/p5/a296598}{sample
 COMPAS Risk Assessment Survey} has been made publicly available,
-revealing the algorithm's input information. This survey has been widely
-critiqued for using proxy variables for race that do not explicitly
+revealing the algorithm's input information. \citet{angwin2016machine}
+has critiqued for using proxy variables for race that do not explicitly
 factor in a defendant's race but heavily imply it, allowing Northpointe
@@
 marginalized people, thereby indirectly asking about race.
-\citet{angwin2016machine}. published a 2016 piece in the news
-publication ProPublica that analyzes the methods and algorithms used by
+\citet{angwin2016machine}. analyzes the methods and algorithms used by
 Northpointe in their COMPAS risk score assessment algorithm and uncovers
-the racial biases in defendants' scores \citep{angwin2016machine}. Their
-research finds that ``the algorithm {[}is{]} somewhat more accurate than
-a coin flip,'' a worrisome level of accuracy given the potential impact
-its determinations may have on real people's lives.
+racial biases in defendants' scores \citep{angwin2016machine}. They find
+that ``the algorithm {[}is{]} somewhat more accurate than a coin flip,''
+a worrisome level of accuracy given the potential impact its
+determinations may have on real people's lives.
 \citet{angwin2016machine} specifically investigate the distribution of
@@
 
-\citet{equivant_response_2016}, on the behalf of Northpointe, deny the
+\citet{equivant_response_2016}, on behalf of Northpointe, deny the
 allegations of racial bias and offer their own analyses based on
@@
 that \citet{angwin2016machine} did not account for base rates of
-recidivism in their analysis, which are important to understand initial
-percentages without the presence of other information.
+recidivism in their analysis, which are important initial percentages
+without the presence of other information.
 
@@
 Our project builds on Women at the Table's various de-biasing algorithms
-used in its
-\href{https://colab.research.google.com/drive/11SWijBDkr1pSgIXyjUkgoSB0gh5oAlaI}{AI
-\& Equality Human Rights Toolbox} to conduct our own analyses on the
-COMPAS data set. Based on this analysis, we employ a human rights
-framework to contribute to the ProPublica and Northpointe debate and
-investigate whether or to what extent there is racial bias in the COMPAS
-algorithm. With a solid understanding of the two sides, we aim to
-pinpoint the shortcomings of both arguments and correct them in our
-analyses. We will use various de-biasing techniques and fairness metrics
-to evaluate the level of bias present in the COMPAS data and our
+used in its AI \& Equality Human Rights Toolbox to conduct our own
+analyses on the COMPAS data set. Based on this analysis, we employ a
+human rights framework to contribute to the ProPublica and Northpointe
+debate and investigate whether or to what extent there is racial bias in
+the COMPAS algorithm. With a solid understanding of the two sides, we
+aim to pinpoint the shortcomings of both arguments and correct them in
+our analyses. We will use various de-biasing techniques and fairness
+metrics to evaluate the level of bias present in the COMPAS data and our
 algorithm. We will summarize our results using the JupyterNotebook
@@
 the data to only include individuals whose race is listed as Caucasian
-or African-American. Our data therefore has 5,723 rows (Figure
-\ref{fig:table snip}), with the below distributions of race (Figure
-\ref{fig:race plot}), age (Figure \ref{fig:age plot}), and two year
+or African-American. Our data therefore has 5,273 rows (Figure
+\ref{fig:table snip}), with the distributions of race (Figure
+\ref{fig:race plot}), age (Figure \ref{fig:age plot}), and two-year
 recidivism rate (Figures \ref{fig:recid plot} \&
-\ref{fig:recid race plot} and Table \ref{tab:recid table}).
+\ref{fig:recid race plot} and Table \ref{tab:recid table}) shown below.
 
@@
 
-\caption{This table is a snippet of the data set we will be using, containing information on a defendant's age, sex, race, criminal history, charge degree, charge description, and two year recidivism outcome.}\label{fig:table snip}
+\caption{A snippet of the data set we will be using, containing information on a defendant's age, sex, race, criminal history, charge degree, charge description, and two-year recidivism outcome.}\label{fig:table snip}
 \end{figure}
@@
 
-\caption{This plot shows the distribution of defendant ages by race.
+\caption{Distribution of defendant ages by race.
 The purple curve shows the distribution of the ages of Black defendants, and the green curve shows the distribution of the ages of white defendants.
@@
 
-\caption{This plot shows the distribution of two year recidivism outcomes by race.
+\caption{This plot shows the distribution of two-year recidivism outcomes by race.
 When we divide the data into Black and white defendants, we can see that Black defendants recidivate more than white defendants and Black defendants are more likely to recidivate than not recidivate.
@@
 \begin{longtable}[]{@{}lll@{}}
-\caption{This table shows the counts of recidivism by race, illustrating
-how a much greater proportion (\textgreater{} 50\%) of Black defendants
-recidivated than their white counterparts.
-\label{tab:recid table}}\tabularnewline
+\caption{Incidence of recidivism by race, illustrating how a much
+greater proportion (\textgreater{} 50\%) of Black defendants recidivated
+than their white counterparts. \label{tab:recid table}}\tabularnewline
 \toprule
@@
 fairness improves, individual fairness gets worse.
+
+\textbf{Statistical Parity Difference} - This metric measures the
+difference that the majority and protected classes get a particular
+outcome. The ideal value of this metric is 0. Fairness for this metric
+is between -0.1 and 0.1. A negative value means there is higher benefit
+for the privileged group (in this case, white defendants).
+
+\[P(\hat{Y}=1|D=Unprivileged) - P(\hat{Y}=1|D=Privileged)\]
+
+\textbf{Disparate Impact Ratio} - This metric is the ratio of how often
+the favorable outcome occurs in one group versus the other. In the case
+of recidivism, this is the ratio of how many white defendants are
+predicted to not recidivate compared to how many Black defendants are
+predicted to not recidivate. A value of 1 means that the ratio is
+exactly 1:1. Less than 1 means the privileged group (white defendants)
+benefits, while a value greater than 1 means the unprivileged group
+(Black defendants) benefits. According to AIF360, a ratio between .8 to
+1.25 is considered fair\citep{Ronaghan2019AI}.
+
+\[\frac{P(\hat{Y}=1|D=Unprivileged)}{P(\hat{Y}=1|D=Privileged)}\]
+
+\textbf{Average Odds Difference} - This metric returns the average
+difference in false positive rate and true positive rate for the
+privileged and unprivileged groups. A value of 0 indicates equality of
+odds, and a value below 0 implies benefit for the privileged group.
+Equality of odds is achieved in the case of recidivism when the
+proportion of people who were predicted to recidivate and did recidivate
+is equal (true positive rate) for both Black and white defendants AND
+the proportion of people who were predicted to recidivate and did not
+recidivate (false positive rate) is equal for both Black and white
+defendants\citep{aif360-oct-2018}.
+
+\[\frac{(FPR_{D = unprivileged} - FPR_{D = privileged}) + (TPR_{D = unprivileged} - TPR_{D = privileged})}{2}\]
+
+\textbf{Equal Opportunity Difference} - This metric is computed as the
+difference of true positive rates between the unprivileged and the
+privileged groups. The true positive rate is the ratio of true positives
+to the total number of actual positives for a given
+group\citep{GoogleDev}.
+
+The ideal value is 0. A value less than 0 implies higher benefit for the
+privileged group and a value greater than 0 implies higher benefit for
+the unprivileged group. Fairness for this metric is between -0.1 and 0.1
+\citep{aif360-oct-2018}.
+
+This metric is best used when it is very important to catch positive
+outcomes while false positives are not exceptionally problematic
+\citep{Cortez2019How}. This is not the case for the COMPAS data set, as
+false positives mean extra jail time for someone who will not actually
+re-offend.
+
+\[TPR_{D = Unprivileged} - TPR_{D = Privileged}\]
 
@@
 the most fair.
-
-\hypertarget{statistical-parity-difference}{%
-\subsection{Statistical Parity
-Difference}\label{statistical-parity-difference}}
-
-This metric measures the difference that the majority and protected
-classes get a particular outcome. The ideal value of this metric is 0.
-Fairness for this metric is between -0.1 and 0.1. A negative value means
-there is higher benefit for the privileged group (in this case, white
-defendants).
-
-\[P(\hat{Y}=1|D=Unprivileged) - P(\hat{Y}=1|D=Privileged)\]
-
-\hypertarget{disparate-impact-ratio}{%
-\subsection{Disparate Impact Ratio}\label{disparate-impact-ratio}}
-
-This metric is the ratio of how often the favorable outcome occurs in
-one group versus the other. In the case of recidivism, this is the ratio
-of how many white defendants are predicted to not recidivate compared to
-how many Black defendants are predicted to not recidivate. A value of 1
-means that the ratio is exactly 1:1. Less than 1 means the privileged
-group (white defendants) benefits, while a value greater than 1 means
-the unprivileged group (Black defendants) benefits. According to AIF360,
-a ratio between .8 to 1.25 is considered fair\citep{Ronaghan2019AI}.
-
-\[\frac{P(\hat{Y}=1|D=Unprivileged)}{P(\hat{Y}=1|D=Privileged)}\]
-
-\hypertarget{average-odds-difference}{%
-\subsection{Average Odds Difference}\label{average-odds-difference}}
-
-This metric returns the average difference in false positive rate and
-true positive rate for the privileged and unprivileged groups. A value
-of 0 indicates equality of odds, and a value below 0 implies benefit for
-the privileged group. Equality of odds is achieved in the case of
-recidivism when the proportion of people who were predicted to
-recidivate and did recidivate is equal (true positive rate) for both
-Black and white defendants AND the proportion of people who were
-predicted to recidivate and did not recidivate (false positive rate) is
-equal for both Black and white defendants\citep{aif360-oct-2018}.
-
-\[\frac{(FPR_{D = unprivileged} - FPR_{D = privileged}) + (TPR_{D = unprivileged} - TPR_{D = privileged})}{2}\]
-
-\hypertarget{equal-opportunity-difference}{%
-\subsection{Equal Opportunity
-Difference}\label{equal-opportunity-difference}}
-
-This metric is computed as the difference of true positive rates between
-the unprivileged and the privileged groups. The true positive rate is
-the ratio of true positives to the total number of actual positives for
-a given group\citep{GoogleDev}.
-
-The ideal value is 0. A value less than 0 implies higher benefit for the
-privileged group and a value greater than 0 implies higher benefit for
-the unprivileged group. Fairness for this metric is between -0.1 and 0.1
-\citep{aif360-oct-2018}.
-
-This metric is best used when it is very important to catch positive
-outcomes while false positives are not exceptionally problematic
-\citep{Cortez2019How}. This is not the case for the COMPAS data set, as
-false positives mean extra jail time for someone who will not actually
-re-offend.
-
-\[TPR_{D = Unprivileged} - TPR_{D = Privileged}\]
 

### 2022-04-21 final_metrics_analysis.Rmd

@@
 This means that 39.14% of white defendants did recidivate within two years compared to 52.35% of black defendants.
+
+```{python wordcloud setup}
+from os import path
+from PIL import Image
+from wordcloud import WordCloud, STOPWORDS, ImageColorGenerator
+```
+
+```{python wordcloud}
+text = " ".join(charge for charge in df_viz.c_charge_desc)
+stopwords = set(STOPWORDS)
+stopwords.update(["arrest", "case", "charge"])
+wordcloud = WordCloud(stopwords=stopwords, background_color="white").generate(text)
+```
+
+```{python wordcloud image}
+plt.imshow(wordcloud, interpolation='bilinear')
+plt.axis("off")
+plt.show()
+```
 
@@
 
-```{python exploratory recidivism by race table}
-# table of recidivism by race
-pd.crosstab(index = df_viz["race"], columns = df_viz["two_year_recid"])
-```
+```{python exploratory priors by race}
+priors_by_race = sns.displot(x="priors_count", row = "race", col = "sex", hue ="race", data=df_viz, palette=workshop_palette)
 
-```{python exploratory priors by race}
-priors_by_race = sns.countplot(x="race", color="priors_count", data=df_viz, palette=workshop_palette)
-
-plt.title('Prior Charges Count by Race')
+#plt.title("Prior Charges Count by Race and Sex)
 plt.show()
@@
 
-```{python exploratory priors by sex}
-priors_by_race = sns.countplot(x="sex", color="priors_count", data=df_viz, palette=workshop_palette)
-
-plt.title('Prior Charges Count by Sex')
-plt.show()
-```
-
-If a defendant has a higher number of prior offenses, the individual is more likely to be rated with a higher risk of recidivism **do we know this for sure?**.
 In the former plot, we can see that Black defendants have a higher number of prior charges than do white defendants.

### 2022-04-22 final_metrics_analysis.Rmd
changing color on mosaic plot, reordering graphs

--- 2022-04-21 final_metrics_analysis.Rmd
@@
 props={}
-props[('Caucasian','F')]={'facecolor':'red', 'edgecolor':'white'}
-props[('Caucasian','M')]={'facecolor':'red', 'edgecolor':'white'}
-props[('African-American','F')]={'facecolor':'xkcd:aqua','edgecolor':'white'}
-props[('African-American','M')]=        {'facecolor':'xkcd:aqua','edgecolor':'white'}
+props[('Caucasian','F')]={'facecolor': green, 'edgecolor':'white'}
+props[('Caucasian','M')]={'facecolor':green, 'edgecolor':'white'}
+props[('African-American','F')]={'facecolor':purple,'edgecolor':'white'}
+props[('African-American','M')]=        {'facecolor':purple,'edgecolor':'white'}
 labelizer=lambda k:{('Caucasian','F'):1242,('African-American','F'):2194,('Caucasian','M'):858,('African-American','M'):979}[k]
@@
 
-59.14% of white defendants were charged with felonies versus 69.15% of black defendants.
-Likewise, a higher percentage of white defendants (40.86%) were charged with a misdemeanor than black defendants (30.85%).
+1242 (59.14%) of white defendants were charged with felonies versus 2194 (69.15%) of Black defendants.
+Additionally, a higher percentage of white defendants (40.86%) were charged with a misdemeanor than black defendants (30.85%).
 
@@
 Out of all defendants, there is a higher proportion of people who did not recidivate than who did. About 2500 people recidivated, whereas approximately 2700 did not.
-
-```{python crosstable of recidivism by race}
-crosstable=pd.crosstab(df_viz['race'],df_viz['two_year_recid'])
-crosstable
-```
-
-```{python exploratory recidivism mosaic plot}
-props={}
-props[('Caucasian','Recidivated')]={'facecolor':'green', 'edgecolor':'white'}
-props[('Caucasian','Survived')]={'facecolor':'green', 'edgecolor':'white'}
-props[('African-American','Recidivated')]={'facecolor':'xkcd:purple','edgecolor':'white'}
-props[('African-American','Survived')]=        {'facecolor':'xkcd:purple','edgecolor':'white'}
-labelizer=lambda k:{('Caucasian','Recidivated'):822,('African-American','Recidivated'):1661,('Caucasian','Survived'):1278,('African-American','Survived'):1512}[k]
-mosaic(df_viz,['race','two_year_recid'],labelizer=labelizer,properties=props)
-plt.show()
-```
-
-A higher percentage of white defendants (60.86%) than black defendants (47.65%) did not commit another crime within two years.
-This means that 39.14% of white defendants did recidivate within two years compared to 52.35% of black defendants.
 
@@
 
+```{python crosstable of recidivism by race}
+crosstable=pd.crosstab(df_viz['race'],df_viz['two_year_recid'])
+crosstable
+```
+
+```{python exploratory recidivism mosaic plot}
+props={}
+props[('Caucasian','Recidivated')]={'facecolor':green, 'edgecolor':'white'}
+props[('Caucasian','Survived')]={'facecolor':green, 'edgecolor':'white'}
+props[('African-American','Recidivated')]={'facecolor':purple,'edgecolor':'white'}
+props[('African-American','Survived')]=        {'facecolor':purple,'edgecolor':'white'}
+labelizer=lambda k:{('Caucasian','Recidivated'):822,('African-American','Recidivated'):1661,('Caucasian','Survived'):1278,('African-American','Survived'):1512}[k]
+mosaic(df_viz,['race','two_year_recid'],labelizer=labelizer,properties=props)
+plt.show()
+```
+
+A higher percentage of white defendants (60.86%) than black defendants (47.65%) did not commit another crime within two years.
+This means that 39.14% of white defendants did recidivate within two years compared to 52.35% of black defendants.
+
 ```{python exploratory priors by race}
@@
 
-In the former plot, we can see that Black defendants have a higher number of prior charges than do white defendants.
+In the above plot, we can see that Black defendants, particularly men, are more likely to have a greater count of prior charges than white defendants.
 This again likely reflects the over-policing of Black people and communities.
-In the latter plot, we can see that male defendants have a higher number of prior charges than do female defendants.
-This is worth noting as @larson2016we found that there is discrepancy in COMPAS risk assessment algorithm's prediction between sexes.
-When the recidivsim predictions are compared with the true recividism outcomes, female defendants have a notable false positive rate that predicts their recidivism risk as higher than their true outcomes [@larson2016we].
+Additionally, we can see that male defendants have a higher number of prior charges than do female defendants.
+This is worth noting as ProPublica found that there is discrepancy in COMPAS risk assessment algorithm's prediction between sexes.
+When the recidivsim predictions are compared with the true recividism outcomes, female defendants have a notable false positive rate that predicts their recidivism risk as higher than their true outcomes (ProPublica).
+Though we do not know for sure which information goes into the COMPAS algorithm, it is likely that a defendant with prior charges will be coded as a having a higher risk of recidivism.
+Thus, by looking at the racial discrepancies in prior charges we can already see potential bias in the algorithm.
 

### 2022-04-24 final_metrics_analysis.Rmd

--- 2022-04-21 final_metrics_analysis.Rmd
@@
 from aif360.sklearn.metrics import disparate_impact_ratio, average_odds_difference, statistical_parity_difference, equal_opportunity_difference
-
-# from aif360.algorithms.preprocessing import DisparateImpactRemover
-# from BlackBoxAuditing.repairers.GeneralRepairer import Repairer
 
@@
 
-## Adversarial Debiasing
-```{python adversarial debiasing}
-# from aif360.algorithms.inprocessing import AdversarialDebiasing
-# import tensorflow as tf
-# 
-# adv_deb = AdversarialDebiasing(unpr)
-# #adv_deb.fit(X_train, y_train)
-# #y_pred_AD = adv_deb.predict(X_test)
-# #adv_deb.sess_.close()
-```
-
 ## Prejudice Remover
@@
 
-```{python}
-def get_attributes(data, selected_attr=None):
-    unprivileged_groups = []
-    privileged_groups = []
-    if selected_attr == None:
-        selected_attr = data.protected_attribute_names
-
-    for attr in selected_attr:
-            idx = data.protected_attribute_names.index(attr)
-            privileged_groups.append({attr:data.privileged_protected_attributes[idx]}) 
-            unprivileged_groups.append({attr:data.unprivileged_protected_attributes[idx]})
-
-    return privileged_groups, unprivileged_groups
-```
-
-```{python post-processing equal odds needs fix} 
-#privileged_groups, unprivileged_groups = get_attributes(train_BLD, selected_attr= 'race')
-#ep = EqOddsPostprocessing(unprivileged_groups= unprivileged_groups, privileged_groups= privileged_groups, seed = 42)
-#eo = PostProcessingMeta(estimator=lr, postprocessor=ep, random_state=1234567)
-#EOPP = ep.fit(train_BLD, test_BLD)
-#y_pred_EO = ep.predict(test_BLD)
-#y_proba_EO = ep.predict_(test_BLD)
-
-#acc_EO = accuracy_score(y_test, y_pred_EO)
-```
-
 ### Statistical Parity Difference

### 2022-04-24 Paper_Draft2/Draft2.tex
pdf knit

--- 2022-04-21 Paper_Draft2/Draft2.tex
@@
 predicted to recidivate and did not recidivate (false positive rate) is
-equal for both Black and white defendants\citep{aif360-oct-2018}.
+equal for both Black and white defendants \citep{aif360-oct-2018}.
 

### 2022-04-24 Paper_Draft2/Draft2.tex
fixing problems

--- 2022-04-21 Paper_Draft2/Draft2.tex
@@
 \address{%
-$^{1}$ \quad Statistical and Data Sciences Smith College Northampton, MA
-01063; \href{mailto:bbaumer@smith.edu}{\nolinkurl{bbaumer@smith.edu}}\\
-}
-% Contact information of the corresponding author
-\corres{Correspondence: \href{mailto:aacharya@smith.edu}{\nolinkurl{aacharya@smith.edu}},
+$^{1}$ \quad Statistical and Data Sciences Smith College Northampton, MA 01063; \href{mailto:aacharya@smith.edu}{\nolinkurl{aacharya@smith.edu}},
 \href{mailto:dcaravela@smith.edu}{\nolinkurl{dcaravela@smith.edu}},
@@
 \href{mailto:ekornberg@smith.edu}{\nolinkurl{ekornberg@smith.edu}},
-\href{mailto:enesmith@smith.edu}{\nolinkurl{enesmith@smith.edu}}}
+\href{mailto:enesmith@smith.edu}{\nolinkurl{enesmith@smith.edu}}\\
+}
+% Contact information of the corresponding author
+\corres{Correspondence: \href{mailto:bbaumer@smith.edu}{\nolinkurl{bbaumer@smith.edu}}}
 
@@
 % Abstract (Do not insert blank lines, i.e. \\)
-\abstract{A variety of disciplines use risk assessment instruments to
-help humans make data-driven decisions. Northpointe, a software company,
-created an algorithmic risk assessment instrument known as the
-Correctional Offender Management Profiling for Alternative Sanctions
-(COMPAS). COMPAS uses various behavioral and psychological metrics
-related to recidivism to assist justice systems in assessing a
-defendant's potential recidivism risk. \citet{angwin2016machine}
-published a ProPublica article in which they conclude that the racial
-biases in the criminal justice system are reflected in the COMPAS
-recidivism risk scores. In response, \citet{equivant_response_2016}
-published a rebuttal on behalf of Northpointe defending the COMPAS
-algorithm and refuting \citet{angwin2016machine}'s allegation of racial
-bias. Using a human rights framework adopted from the organizations
-Women at the Table and AI Fairness 360, we use debiasing algorithms and
-fairness metrics to analyze the argument between Northpointe and
-ProPublica and determine whether and to what extent there is racial bias
-in the COMPAS algorithm.}
+\abstract{A variety of disciplines use risk assessment instruments to help humans
+make data-driven decisions. Northpointe, a software company, created an
+algorithmic risk assessment instrument known as the Correctional
+Offender Management Profiling for Alternative Sanctions (COMPAS). COMPAS
+uses various behavioral and psychological metrics related to recidivism
+to assist justice systems in assessing a defendant's potential
+recidivism risk. \citet{angwin2016machine} published a ProPublica
+article in which they conclude that the racial biases in the criminal
+justice system are reflected in the COMPAS recidivism risk scores. In
+response, \citet{equivant_response_2016} published a rebuttal on behalf
+of Northpointe defending the COMPAS algorithm and refuting
+\citet{angwin2016machine}'s allegation of racial bias. Using a human
+rights framework adopted from the organizations Women at the Table and
+AI Fairness 360, we use debiasing algorithms and fairness metrics to
+analyze the argument between Northpointe and ProPublica and determine
+whether and to what extent there is racial bias in the COMPAS algorithm.}
 
@@
 Northpointe (now known by its parent company
-\href{https://www.equivant.com/faq/}{equivant}), to predict defendants'
+\href{https://www.equivant.com/faq/}{Equivant}), to predict defendants'
 risk of recidivism. It generates a decile score that classifies
@@
 the COMPAS risk assessment instrument, including but not limited to the
-\href{https://doccs.ny.gov/system/files/documents/2020/11/8500.pdf}{New
-York},
-\href{https://hdsr.mitpress.mit.edu/pub/hzwo7ax4/release/4}{Massachusetts},
-\href{https://hdsr.mitpress.mit.edu/pub/hzwo7ax4/release/4}{Michigan},
-\href{https://hdsr.mitpress.mit.edu/pub/hzwo7ax4/release/4}{California},
-and \href{https://doc.wi.gov/Pages/AboutDOC/COMPAS.aspx}{Wisconsin}
+New York \citep{NYDOCCS2019}, Massachusetts, Michigan, California
+\citep{Jackson2020Setting}, and Wisconsin \citep{WisconsinDOC}
 Departments of Corrections.
@@
 Due to the proprietary nature of the COMPAS algorithm, it is unknown how
-exactly these recidivism risk scores are calculated; however, a
-\href{https://www.documentcloud.org/documents/2702103-Sample-Risk-Assessment-COMPAS-CORE\#document/p5/a296598}{sample
-COMPAS Risk Assessment Survey} has been made publicly available,
+exactly these recidivism risk scores are calculated; however, a sample
+COMPAS Risk Assessment Survey has been made publicly available,
 revealing the algorithm's input information. This survey has been widely
@@
 de-biasing the COMPAS algorithm \citep{aif360-oct-2018} and producing a
-corresponding data story that will be added to its library.
-
-Our project builds on Women at the Table's various de-biasing algorithms
-used in its
-\href{https://colab.research.google.com/drive/11SWijBDkr1pSgIXyjUkgoSB0gh5oAlaI}{AI
-\& Equality Human Rights Toolbox} to conduct our own analyses on the
+corresponding data story that will be added to its library. Our project
+builds on Women at the Table's various de-biasing algorithms used in its
+AI \& Equality Human Rights Toolbox to conduct our own analyses on the
 COMPAS data set. Based on this analysis, we employ a human rights
@@
 
+{\centering \includegraphics[width=1\linewidth]{../images/race_bar_plot_new}
+
+}
+
+\caption{This plot shows the distribution of defendant races.
+The dataset contains about 3200 Black defendants and 2100 white defendants.
+Therefore, there are more Black people represented in the dataset than white people.}\label{fig:race plot}
+\end{figure}
+
+\begin{figure}
+
 {\centering \includegraphics[width=1\linewidth]{../images/age_race_plot_new}
@@
 For both groups, the majority of defendants are relatively young, but this is especially noticeable for Black defendants.}\label{fig:age plot}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{../images/recid_bar_plot_new}
+
+}
+
+\caption{This plot shows the distribution of two year recidivism outcomes.
+Out of all defendants, there is a higher proportion of people who did not recidivate than who did.
+About 2500 people recidivated, whereas approximately 2700 did not.}\label{fig:recid plot}
 \end{figure}
@@
 COMPAS data set that accompanied the \citet{angwin2016machine} piece.
+Using the \citet{reticulate} package, we conduct our analysis in an
+RMarkdown document with Python code.
 

### 2022-04-24 Paper_Draft1/Draft1.tex

@@
 % Full title of the paper (Capitalized)
-\Title{Does the COMPAS needle always point towards equity? Finding
-Fairness in the COMPAS Risk Assessment Algorithm}
+\Title{It's COMPAS-licated -- Can We Find Fairness in the COMPAS Risk
+Assessment Algorithm?}
 
@@
 
-% From pandoc table feature
-\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
-
 \begin{document}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\hypertarget{version}{%
-\section{Version}\label{version}}
-
-This Rmd-skeleton uses the mdpi Latex template published 2019/02.
-However, the official template gets more frequently updated than the
-`rticles' package. Therefore, please make sure prior to paper
-submission, that you're using the most recent .cls, .tex and .bst files
-(available \href{http://www.mdpi.com/authors/latex}{here}).
 
@@
 conducting their own analyses based on different statistical parity
-metrics \citep{equivant_response_2018}. Since then, ProPublica has
-responded, maintaining their original argument and Northpointe has yet
-to respond (though in 2017, they adopted the name Equivant). Currently,
+metrics \citep{equivant_response_2018}. Since then, the two parties have
+had several exchanges, maintaining their original arguments. Currently,
 the two parties' disagreement focuses on pretrial Risk Assessment
@@
 
-\hypertarget{results}{%
-\section{Results}\label{results}}
-
-This section may be divided by subheadings. It should provide a concise
-and precise description of the experimental results, their
-interpretation as well as the experimental conclusions that can be
-drawn.
-
-\hypertarget{subsection-heading-here}{%
-\subsection{Subsection Heading Here}\label{subsection-heading-here}}
-
-Subsection text here.
-
-\hypertarget{subsubsection-heading-here}{%
-\subsubsection{Subsubsection Heading
-Here}\label{subsubsection-heading-here}}
-
-Bulleted lists look like this:
-
-\begin{itemize}
-\tightlist
-\item
-  First bullet
-\item
-  Second bullet
-\item
-  Third bullet
-\end{itemize}
-
-Numbered lists can be added as follows:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  First item
-\item
-  Second item
-\item
-  Third item
-\end{enumerate}
-
-The text continues here.
-
-All figures and tables should be cited in the main text as Figure
-\ref{fig:mdpi-logo}, Table 1, etc.
-
-\begin{figure}
-
-{\centering \includegraphics{logo-mdpi}
-
-}
-
-\caption{This is a figure, Schemes follow the same formatting.
-If there are multiple panels, they should be listed as: (\textbf{a}) Description of what is contained in the first panel.
-(\textbf{b}) Description of what is contained in the second panel.
-Figures should be placed in the main text near to the first time they are cited.
-A caption on a single line should be centered.}\label{fig:mdpi-logo}
-\end{figure}
-
-Please see Table \ref{tab:mytable}.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{x }\OtherTok{\textless{}{-}}\NormalTok{ tibble}\SpecialCharTok{::}\FunctionTok{tribble}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\StringTok{\textasciigrave{}}\AttributeTok{Title 1}\StringTok{\textasciigrave{}}\NormalTok{, }\SpecialCharTok{\textasciitilde{}}\StringTok{\textasciigrave{}}\AttributeTok{Title 2}\StringTok{\textasciigrave{}}\NormalTok{, }\SpecialCharTok{\textasciitilde{}}\StringTok{\textasciigrave{}}\AttributeTok{Title 3}\StringTok{\textasciigrave{}}\NormalTok{,}
-\StringTok{"entry 1"}\NormalTok{, }\StringTok{"data"}\NormalTok{, }\StringTok{"data"}\NormalTok{,}
-\StringTok{"entry 2"}\NormalTok{, }\StringTok{"data"}\NormalTok{, }\StringTok{"data"}
-\NormalTok{)}
-
-\NormalTok{knitr}\SpecialCharTok{::}\FunctionTok{kable}\NormalTok{(x, }\AttributeTok{caption =} \StringTok{"This is a table caption.
-Tables should be placed in the main text near to the first time they are cited.}\SpecialCharTok{\textbackslash{}\textbackslash{}}\StringTok{label\{tab:mytable\}"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{longtable}[]{@{}lll@{}}
-\caption{This is a table caption. Tables should be placed in the main
-text near to the first time they are
-cited.\label{tab:mytable}}\tabularnewline
-\toprule
-Title 1 & Title 2 & Title 3\tabularnewline
-\midrule
-\endfirsthead
-\toprule
-Title 1 & Title 2 & Title 3\tabularnewline
-\midrule
-\endhead
-entry 1 & data & data\tabularnewline
-entry 2 & data & data\tabularnewline
-\bottomrule
-\end{longtable}
-
-This is an example of an equation:
-
-\begin{equation}
-\mathbb{S}
-\end{equation}
-
-Example of a theorem:
-
-\begin{Theorem}
-Example text of a theorem.
-\end{Theorem}
-
-The text continues here. Proofs must be formatted as follows:
-
-Example of a proof:
-
-\begin{proof}[Proof of Theorem 1]
-Text of the proof. Note that the phrase `of Theorem 1' is optional if it is clear which theorem is being referred to.
-\end{proof}
-
-The text continues here.
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-Authors should discuss the results and how they can be interpreted in
-perspective of previous studies and of the working hypotheses. The
-findings and their implications should be discussed in the broadest
-context possible. Future research directions may also be highlighted.
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-This section is not mandatory, but can be added to the manuscript if the
-discussion is unusually long or complex.
-
 % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2022-04-25 final_metrics_analysis.Rmd

--- 2022-04-22 final_metrics_analysis.Rmd
@@
 # for Amrita
-use_condaenv("anaconda3")
+#use_condaenv("anaconda3")
 ```
@@
 # Exploratory Data Analysis
+
+### Descriptions of Average Defendants
 ```{python EDA} 
@@
 
+### Recidivism Distribution
+
+```{python exploratory recidivism bar plot}
+# barplot of recividism
+sns.countplot(x='two_year_recid', data=df_viz, palette=workshop_palette)
+plt.title('Two Year Recidivism Rate')
+```
+
+Out of all defendants, there is a higher proportion of people who did not recidivate than who did. About 2500 people recidivated, whereas approximately 2700 did not.
+
+### Racial Distributions
+
+```{python exploratory race bar plot}
+# barplot of race
+sns.countplot(x="race", data=df_viz, palette=workshop_palette)
+plt.title('Race Distribution (White and Black individuals only)')
+plt.show()
+```
+
+The dataset we are using contains about 3200 Black defendants and 2100 white defendants.
+Therefore, there are more Black people represented in the dataset than white people, showing the disproportionate policing and criminalization of Black communities.
+
+#### Age
+
+```{python exploratory age by race plot}
+# age distribution by race - hacking solution for purple dotted line for now
+ax = sns.kdeplot(x="age", hue="race", data=df_viz, palette=workshop_palette);
+kdeline = ax.lines[0]
+mean_black = df_viz.groupby('race').age.median()[0]
+xs = kdeline.get_xdata()
+ys = kdeline.get_ydata()
+height_black = 0.027 # np.interp(mean_black, xs, ys)
+ax.vlines(mean_black, 0, height_black, color=purple, ls=':')
+
+mean_white = df_viz.groupby('race').age.median()[1]
+height_white = np.interp(mean_white, xs, ys)
+ax.vlines(mean_white, 0, height_white, color=green, ls=':')
+
+plt.title('Distribution of Age by Race')
+```
+
+The purple curve in the plot above shows the distribution of the ages of Black defendants, and the green curve shows the distribution of the ages of white defendants.
+The probability of a defendant's age being between two points on the x-axis is the total shaded area of the curve under the two points.
+The purple dotted line represents the median age of Black defendants (29 years) and the green dotted line represents the median age of white defendants (35 years).
+For both groups, the majority of defendants are relatively young, but this is especially noticeable for Black defendants.
+
+#### Charge Degree
+
 ```{python crosstable of charge degrees by race}
@@
 
-1242 (59.14%) of white defendants were charged with felonies versus 2194 (69.15%) of Black defendants.
-Additionally, a higher percentage of white defendants (40.86%) were charged with a misdemeanor than black defendants (30.85%).
+1242 (59.14%) of white defendants were charged with felonies versus 2194 (69.15%) of black defendants.
+Likewise, a higher percentage of white defendants (40.86%) were charged with a misdemeanor than black defendants (30.85%).
 
-```{python exploratory recidivism bar plot}
-# barplot of recividism
-sns.countplot(x='two_year_recid', data=df_viz, palette=workshop_palette)
-plt.title('Two Year Recidivism Rate')
-```
-
-Out of all defendants, there is a higher proportion of people who did not recidivate than who did. About 2500 people recidivated, whereas approximately 2700 did not.
+#### Charge Descriptions
 
@@
 
-```{python exploratory race bar plot}
-# barplot of race
-sns.countplot(x="race", data=df_viz, palette=workshop_palette)
-plt.title('Race Distribution (White and Black individuals only)')
-plt.show()
-```
-
-The dataset we are using contains about 3200 Black defendants and 2100 white defendants.
-Therefore, there are more Black people represented in the dataset than white people, showing the disproportionate policing and criminalization of Black communities.
-
-```{python exploratory age by race plot}
-# age distribution by race - hacking solution for purple dotted line for now
-#ax = sns.kdeplot(x="age", hue="race", data=df_viz, palette=workshop_palette);
-#kdeline = ax.lines[0]
-#mean_black = df_viz.groupby('race').age.median()[0]
-#xs = kdeline.get_xdata()
-#ys = kdeline.get_ydata()
-#height_black = 0.027 # np.interp(mean_black, xs, ys)
-#ax.vlines(mean_black, 0, height_black, color=purple, ls=':')
-
-#mean_white = df_viz.groupby('race').age.median()[1]
-#height_white = np.interp(mean_white, xs, ys)
-#ax.vlines(mean_white, 0, height_white, color=green, ls=':')
-
-#plt.title('Distribution of Age by Race')
-```
-
-The purple curve in the plot above shows the distribution of the ages of Black defendants, and the green curve shows the distribution of the ages of white defendants.
-The probability of a defendant's age being between two points on the x-axis is the total shaded area of the curve under the two points.
-The purple dotted line represents the median age of Black defendants (29 years) and the green dotted line represents the median age of white defendants (35 years).
-For both groups, the majority of defendants are relatively young, but this is especially noticeable for Black defendants.
+#### Recidivism
 
@@
 This means that 39.14% of white defendants did recidivate within two years compared to 52.35% of black defendants.
+
+#### Prior Charges
 
@@
 
-# Pre-Processing Approaches
+As discussed above, all four fairness metrics define the baseline model as unfairly benefitting white defendants.
+The green bar represents the range of fair values, while the blue bars are the actual metric values for the baseline model.
+We can see that the margin of unfairness is smallest for statistical parity difference, as the value is only 0.038 outside of the -0.1 to 0.1 range of fairness.
+However, for the other three metrics, the value is significantly below the margin of fairness.
+
+# Pre-Processing
 
@@
 
-```{python pre-processing disparate impact remover}
-#lr = LogisticRegressionCV(solver='lbfgs')
-#rew = ReweighingMeta(estimator=lr, reweigher=Reweighing('race'))
-#rew.fit(X_train, y_train)
-#y_pred_REW = rew.predict(X_test)
+The orange bars in these graphs represent the new values of the fairness metrics after implementing the reweighting approach.
+Like we saw above, reweighting the data improved the fairness in all four of these metrics.
+All four values are now within the range of values considered fair.
+Statistical parity difference and disparate impact ratio still imply a slight benefit for white defendants, while equal opportunity difference and average odds difference now imply a slight benefit for Black defendants.
 
-# diremover = DisparateImpactRemover(repair_level=1.0, sensitive_attribute="race")
-#diremover.fit(X_train, y_train)
-#y_pred_diremover = diremover.predict(X_test)
-
-# diremover_rp = diremover.fit_transform(X_test)
-# AttributeError: 'DataFrame' object has no attribute 'features' (probem with .fit_transform function)
-# .fit_transform() angry because of X_test... 
-```
-
-# In-Processing Approaches
+# In-Processing
 
@@
 The highest percentage in this confusion matrix is for false positives, defendants who the model predicted to recidivate and actually did not, at 37.76%.
+This false positive percentage concerns us because we do not want defendants who do not recidivate to have unfairly long sentences due to their (incorrectly) predicted recidivism.
 The number of true positives, defendants who the model predicted to recidivate and actually recidivated, is lower at 23.88%.
@@
 
+After the prejudice remover, the accuracy of the model is fairly similar across the board of races, but the model's accuracy is only 38.21%, which is much lower than the baseline and reweighting models.
+This accuracy score means that the model makes accurate predictions only 38% of the time.
+
+In terms of recall, the score for white defendants is very high at 0.7, meaning that the model correctly identifies white defendants who recidivate about 70% of the time.
+This number is about 50% for all defendants, but only 40% for Black defendants.
+Just as we saw a low recall score for white defendants in the baseline model, now we see the same for Black defendants in the prejudice remover model.
+This means that the model frequently predicts Black defendants to not recidivate (the more favorable outcome) when they actually do.
+
+Finally, the precision score has lowered from about 0.6 for the baseline model to about 0.4 for all defendants, 0.45 for Black defendants, and 0.3 for white defendants.
+This means that the model's predictions of a defendant who recidivates are accurate about 40% of the time, a number which is slightly higher for Black defendants and slightly lower for white defendants.
+
 ### Statistical Parity Difference
@@
 
-Compared to the baseline model, the predjudice remover is highly in favor of the unprivileged group because its value is greater than 0. The value of .367 is much higher than 0, the value of fairness.
+Compared to the baseline model, the predjudice remover is highly in favor of the unprivileged group, Black defendants, because its value is greater than 0.
+The value of 0.367 is much higher than 0, the value of fairness.
 
@@
 
-Compared to the baseline model, the predjudice remover is highly in favor of the unprivileged group because its value is greater than 0. The value of .321 is much higher than 0, the value of fairness.
+Compared to the baseline model, the predjudice remover is highly in favor of Black defendants because its value is greater than 0. The value of 0.321 is much higher than 0, the value of fairness.
 
@@
 
-Compared to the baseline model, the predjudice remover is highly in favor of the unprivileged group because its value is greater than 0.
+Compared to the baseline model, the predjudice remover is highly in favor of Black defendants, because its value is greater than 0.
 
@@
 ```
+
+Like the model performance metrics suggested, the prejudice remover approach resulted in an increased benefit for Black defendants.
+As the orange bars on these plots show, the values of the fairness metrics have reversed from their baseline values.
+Now all four metrics suggest an unfair advantage for Black defendants.
+Thus, this approach removed the model's prejudice against Black people, but it did not result in a "fair" model.
 
@@
 It is very unlikely that it will incorrectly predict that someone will recidivate.
-
-```{python}
-#metrics_per_group(y_test, y_pred_CEO)
-#plt.show()
-```
-
-```{python}
-def get_attributes(data, selected_attr=None):
-    unprivileged_groups = []
-    privileged_groups = []
-    if selected_attr == None:
-        selected_attr = data.protected_attribute_names
-
-    for attr in selected_attr:
-            idx = data.protected_attribute_names.index(attr)
-            privileged_groups.append({attr:data.privileged_protected_attributes[idx]}) 
-            unprivileged_groups.append({attr:data.unprivileged_protected_attributes[idx]})
-
-    return privileged_groups, unprivileged_groups
-```
-
-```{python post-processing equal odds needs fix} 
-#privileged_groups, unprivileged_groups = get_attributes(train_BLD, selected_attr= 'race')
-#ep = EqOddsPostprocessing(unprivileged_groups= unprivileged_groups, privileged_groups= privileged_groups, seed = 42)
-#eo = PostProcessingMeta(estimator=lr, postprocessor=ep, random_state=1234567)
-#EOPP = ep.fit(train_BLD, test_BLD)
-#y_pred_EO = ep.predict(test_BLD)
-#y_proba_EO = ep.predict_(test_BLD)
-
-#acc_EO = accuracy_score(y_test, y_pred_EO)
-```
+Though the accuracy score for the model is 0.53, lower than the baseline, this model is much better at identifying true negatives.
 
@@
 
-At the value of 0.134, the statistical parity difference is now positive and slightly in favor of black defendants.
+At the value of 0.134, the statistical parity difference is now positive and slightly in favor of Black defendants. However, it is just outside the bounds of fairness.
 
@@
 
-There has been a significant change from -.408 to 0.088, showing that now the value is within the bounds of "fair" (-0.1 to 0.1)
+There has been a significant change from -0.408 to 0.088, showing that now the value is within the bounds of "fair" (-0.1 to 0.1).
+Additionally, since the value is positive, it implies a slight benefit for Black defendants.
 
@@
 
-The baseline model with the average odds difference was -0.438 and now it is 0.15 which indicates being slightly in favor of black defendants.
+The baseline model with the average odds difference was -0.438 and now it is 0.15, which indicates slight unfairness (as it is above 0.1) and the model being in favor of Black defendants.
 
@@
 
-The baseline value was 0.466 which is highly out of the range of "fair" which goes up to 1. Now at a value of 1.2 it is much higher than 1.
+The baseline value was 0.466, which is significantly below the range of "fair" which begins at 0.8. Now at a value of 1.15 it is above 1, conveying fairness and a benefit of Black defendants.
 
@@
 ```
+
+As we can see from the orange bars, the calibrated equalized odds approach results in all four metrics now suggesting a benefit for the originally unprivileged group, Black defendants.
+Though the values for statistical parity difference and average odds difference are slightly above the range of what is considered fair, the margin is much smaller than the original margin between the value and the range of fairness (as indicated by the blue bars).
+Thus, the calibrated equalized odds approach successfully counteracts the bias against Black defendants and results in a mostly fair model.

### 2022-04-26 Paper_Draft2/Draft2.Rmd
updating paper with Ben's comments; starting official results section

@@
 Using a human rights framework adopted from the organizations Women at the Table and AI Fairness 360, we use debiasing algorithms and fairness metrics to analyze the argument between Northpointe and ProPublica and determine whether and to what extent there is racial bias in the COMPAS algorithm.
+All four group fairness metrics determine that the COMPAS algorithm favors white defendants over Black defendants.
 
@@
 
-Due to the proprietary nature of the COMPAS algorithm, it is unknown how exactly these recidivism risk scores are calculated; however, a [sample COMPAS Risk Assessment Survey](https://www.documentcloud.org/documents/2702103-Sample-Risk-Assessment-COMPAS-CORE#document/p5/a296598) has been made publicly available, revealing the algorithm’s input information.
-This survey has been widely critiqued for using proxy variables for race that do not explicitly factor in a defendant’s race but heavily imply it, allowing Northpointe to claim that their algorithm is free of racial bias.
+Due to the proprietary nature of the COMPAS algorithm, it is unknown how exactly these recidivism risk scores are calculated.
+However, a [sample COMPAS Risk Assessment Survey](https://www.documentcloud.org/documents/2702103-Sample-Risk-Assessment-COMPAS-CORE#document/p5/a296598) has been made publicly available, revealing the algorithm’s input information.
+@angwin2016machine critiques this survey for using proxy variables for race that do not explicitly factor in a defendant’s race but heavily imply it, allowing Northpointe to claim that their algorithm is free of racial bias.
 For example, the COMPAS risk assessment survey asks screeners to speculate if a defendant might be affiliated with a gang.
@@
 Although these questions do not directly ask about race, they do not take into account the pervasive nature of systemic racism that infiltrates every aspect of the lives of marginalized people, thereby indirectly asking about race.
-@angwin2016machine.
-published a 2016 piece in the news publication ProPublica that analyzes the methods and algorithms used by Northpointe in their COMPAS risk score assessment algorithm and uncovers the racial biases in defendants’ scores [@angwin2016machine].
-Their research finds that “the algorithm [is] somewhat more accurate than a coin flip,” a worrisome level of accuracy given the potential impact its determinations may have on real people’s lives.
+
+@angwin2016machine analyzes the methods and algorithms used by Northpointe in their COMPAS risk score assessment algorithm and uncovers racial biases in defendants’ scores [@angwin2016machine].
+They find that “the algorithm [is] somewhat more accurate than a coin flip,” a worrisome level of accuracy given the potential impact its determinations may have on real people’s lives.
 @angwin2016machine specifically investigate the distribution of COMPAS scores by decile among Black and white defendants.
@@
 
-@equivant_response_2016, on the behalf of Northpointe, deny the allegations of racial bias and offer their own analyses based on different fairness metrics in rebuttal [@equivant_response_2016].
+@equivant_response_2016, on behalf of Northpointe, deny the allegations of racial bias and offer their own analyses based on different fairness metrics in rebuttal [@equivant_response_2016].
 @angwin2016machine maintain that there are biases in the outcome values, protected attributes, and covariates during @equivant_response_2016’s data processing phase.
 ProPublica collaborators @larson2016we account for these biases in their analyses.
-In their response, @equivant_response_2016 highlight that @angwin2016machine did not account for base rates of recidivism in their analysis, which are important to understand initial percentages without the presence of other information.
+In their response, @equivant_response_2016 highlight that @angwin2016machine did not account for base rates of recidivism in their analysis, which are important initial percentages without the presence of other information.
 
@@
 For the purposes of our project, which endeavors to evaluate the differing effects of the COMPAS algorithm on white defendants versus Black defendants, we have filtered the data to only include individuals whose race is listed as Caucasian or African-American.
-Our data therefore has 5,723 rows (Figure \ref{fig:table snip}), with the below distributions of race (Figure \ref{fig:race plot}), age (Figure \ref{fig:age plot}), and two year recidivism rate (Figures \ref{fig:recid plot} & \ref{fig:recid race plot} and Table \ref{tab:recid table}).
+Our data therefore has 5,273 rows (Figure \ref{fig:table snip}), with the distributions of race (Figure \ref{fig:race plot}), age (Figure \ref{fig:age plot}), and two-year recidivism rate (Figures \ref{fig:recid plot} & \ref{fig:recid race plot} and Table \ref{tab:recid table}) shown below.
 
-```{r table snip, echo = FALSE, fig.align='center', fig.cap= "This table is a snippet of the data set we will be using, containing information on a defendant's age, sex, race, criminal history, charge degree, charge description, and two year recidivism outcome.", out.width= "100%"}
+```{r table snip, echo = FALSE, fig.align='center', fig.cap= "A snippet of the data set we will be using, containing information on a defendant's age, sex, race, criminal history, charge degree, charge description, and two-year recidivism outcome.", out.width= "100%"}
 knitr::include_graphics("../images/table_snippet.png")
@@
 
-```{r age plot, echo=FALSE, fig.align='center', fig.cap= "This plot shows the distribution of defendant ages by race.
+```{r age plot, echo=FALSE, fig.align='center', fig.cap= "Distribution of defendant ages by race.
 The purple curve shows the distribution of the ages of Black defendants, and the green curve shows the distribution of the ages of white defendants.
@@
 
-```{r recid race plot,echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes by race.
+```{r recid race plot,echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two-year recidivism outcomes by race.
 When we divide the data into Black and white defendants, we can see that Black defendants recidivate more than white defendants and Black defendants are more likely to recidivate than not recidivate.
@@
 ```{r, echo = FALSE}
-knitr::kable(x, caption = "This table shows the counts of recidivism by race, illustrating how a much greater proportion (> 50%) of Black defendants recidivated than their white counterparts.
-\\label{tab:recid table}")
+knitr::kable(x, caption = "Incidence of recidivism by race, illustrating how a much greater proportion (> 50%) of Black defendants recidivated than their white counterparts. \\label{tab:recid table}")
 ```
@@
 
-AI Fairness 360 (AIF360) is an open-source Python toolkit that seeks to "to help facilitate the transition of fairness research algorithms to use in an industrial setting and to provide a common framework for fairness researchers to share and evaluate algorithms" [@aif360-oct-2018].
-It contains multiple data sets, including the COMPAS data set that accompanied the @angwin2016machine piece.
+AI Fairness 360 (AIF360) is an open-source Python toolkit that seeks “to help facilitate the transition of fairness research algorithms to use in an industrial setting and to provide a common framework for fairness researchers to share and evaluate algorithms” [@aif360-oct-2018].
+It contains multiple data sets, including the COMPAS data set that accompanied @angwin2016machine.
 
-The AIF360 toolkit contains various group and individual fairness metrics as well as pre-processing, in-processing, and post-processing algorithms that we used for our experiments [@aif360-oct-2018].
+The AIF360 toolkit contains various group and individual fairness metrics as well as pre-processing, in-processing, and post-processing algorithms that we used to debias the COMPAS algorithm [@aif360-oct-2018].
 We researched the definitions and applications of different fairness metrics [@ashokan2021fairness] to determine which metric would be most appropriate for our project.
 We chose to look at four group fairness metrics instead of individual fairness metrics.
-Group fairness takes into account the attributes of a whole group as opposed to just one individual in the group, allowing us to represent more of the systemic issues happening.
+Group fairness metrics take into account the attributes of a whole group as opposed to just one individual in the group, allowing us to represent systemic issues.
 In general, group fairness metrics require that the unprivileged group is treated similarly to the privileged group, whereas individual fairness metrics require individuals to be treated consistently [@kypraiou_what_2021].
-Group and individual metrics work in opposition of one another, meaning that when group fairness improves, individual fairness gets worse.
+Group and individual metrics work in opposition of one another, meaning that when group fairness improves, individual fairness gets worse [@kypraiou_what_2021].
 
-**Statistical Parity Difference** - This metric measures the difference that the majority and protected classes get a particular outcome.
+**Statistical Parity Difference** - This metric measures the difference between privileged and marginalized groups’ likelihood to get a particular outcome.
 The ideal value of this metric is 0.
@@
 **Disparate Impact Ratio** - This metric is the ratio of how often the favorable outcome occurs in one group versus the other.
-In the case of recidivism, this is the ratio of how many white defendants are predicted to not recidivate compared to how many Black defendants are predicted to not recidivate.
+In the case of recidivism, this is the ratio of how many white defendants are predicted to not recidivate compared to how many black defendants are predicted to not recidivate.
 A value of 1 means that the ratio is exactly 1:1.
 Less than 1 means the privileged group (white defendants) benefits, while a value greater than 1 means the unprivileged group (Black defendants) benefits.
-According to AIF360, a ratio between .8 to 1.25 is considered fair[@Ronaghan2019AI].
+According to AIF360, a ratio between 0.8 to 1.25 is considered fair [@Ronaghan2019AI].
 
@@
 
-**Average Odds Difference** - This metric returns the average difference in false positive rate and true positive rate for the privileged and unprivileged groups.
-A value of 0 indicates equality of odds, and a value below 0 implies benefit for the privileged group.
-Equality of odds is achieved in the case of recidivism when the proportion of people who were predicted to recidivate and did recidivate is equal (true positive rate) for both Black and white defendants AND the proportion of people who were predicted to recidivate and did not recidivate (false positive rate) is equal for both Black and white defendants[@aif360-oct-2018].
-
-$$\frac{(FPR_{D = unprivileged} - FPR_{D = privileged}) + (TPR_{D = unprivileged} - TPR_{D = privileged})}{2}$$
-
 **Equal Opportunity Difference** - This metric is computed as the difference of true positive rates between the unprivileged and the privileged groups.
-The true positive rate is the ratio of true positives to the total number of actual positives for a given group[@GoogleDev].
+The true positive rate is the ratio of true positives to the total number of actual positives for a given group [@GoogleDev].
 
@@
 $$TPR_{D = Unprivileged} - TPR_{D = Privileged}$$
+
+**Average Odds Difference** - This metric returns the average difference in false positive rate and true positive rate for the privileged and unprivileged groups.
+A value of 0 indicates equality of odds, and a value below 0 implies benefit for the privileged group.
+Equality of odds is achieved in the case of recidivism when the proportion of people who were predicted to recidivate and did recidivate is equal (true positive rate) for both Black and white defendants AND the proportion of people who were predicted to recidivate and did not recidivate (false positive rate) is equal for both Black and white defendants [@aif360-oct-2018].
+
+$$\frac{1}{2}\left[(FPR_{D = Unprivileged} - FPR_{D = Privileged}) + \underbrace{(TPR_{D = Unprivileged} - TPR_{D = Privileged})}_\textrm{Equal Opportunity Difference}\right]$$
 
@@
 
-# Preliminary Results
+# Results
 
-With our baseline model, we ran the four different group fairness metrics we chose and compared the results (pictured in Table \ref{tab:baseline metrics table}))
+With our baseline model, we ran the four different group fairness metrics we chose and compared the results (pictured in Table \ref{tab:baseline metrics table}).
 
@@
 All four group fairness metrics determine that the COMPAS algorithm favors white defendants over Black defendants.
-The statistical parity metric indicates unfairness by a smaller margin of 0.04 than equal opportunity difference which has a margin of error 0.4 below the threshold.
-Although this is true, each algorithm uses different penalties for unfavorable outcomes, so the higher the penalty the more of a difference there will be between the fairness threshold and the value.
+Although the magnitudes of the various fairness metrics are different, all of the fairness metrics are not within their respective fairness thresholds.
+Our goal is to use pre-processing, in-processing, and post-processing algorithms in the AIF360 toolkit to see if we can make COMPAS fair at all.
 
@@
 
+## Pre-Processing Approach
+
+We chose the reweighing as our pre-processing bias mitigation algorithm.
+After running the reweighing algorithm, our fairness metrics are -0.015 for statistical parity difference, 0.015 for equal opportunity difference, 0.014 for average odds difference, and 0.98 for disparate impact ratio.
+Overall, these fairness metrics show that the pre-processing algorithm reweighing improves the bias in the COMPAS algorithm.
+
 # Bibliography

### 2022-04-26 Paper_Draft2/Draft2.tex

--- 2022-04-21 Paper_Draft2/Draft2.tex
@@
 ProPublica and determine whether and to what extent there is racial bias
-in the COMPAS algorithm.}
+in the COMPAS algorithm. All four group fairness metrics determine that
+the COMPAS algorithm favors white defendants over Black defendants.}
 
@@
 revealing the algorithm's input information. \citet{angwin2016machine}
-has critiqued for using proxy variables for race that do not explicitly
-factor in a defendant's race but heavily imply it, allowing Northpointe
-to claim that their algorithm is free of racial bias. For example, the
-COMPAS risk assessment survey asks screeners to speculate if a defendant
-might be affiliated with a gang. It also asks if a defendant has any
-friends or family members who have been crime victims
+critiques this survey for using proxy variables for race that do not
+explicitly factor in a defendant's race but heavily imply it, allowing
+Northpointe to claim that their algorithm is free of racial bias. For
+example, the COMPAS risk assessment survey asks screeners to speculate
+if a defendant might be affiliated with a gang. It also asks if a
+defendant has any friends or family members who have been crime victims
 \citep{Angwin2016Sample}. Although these questions do not directly ask
@@
 marginalized people, thereby indirectly asking about race.
-\citet{angwin2016machine}. analyzes the methods and algorithms used by
+
+\citet{angwin2016machine} analyzes the methods and algorithms used by
 Northpointe in their COMPAS risk score assessment algorithm and uncovers
@@
 
-AI Fairness 360 (AIF360) is an open-source Python toolkit that seeks to
+AI Fairness 360 (AIF360) is an open-source Python toolkit that seeks
 ``to help facilitate the transition of fairness research algorithms to
@@
 \citep{aif360-oct-2018}. It contains multiple data sets, including the
-COMPAS data set that accompanied the \citet{angwin2016machine} piece.
+COMPAS data set that accompanied \citet{angwin2016machine}.
 
@@
 metrics as well as pre-processing, in-processing, and post-processing
-algorithms that we used for our experiments \citep{aif360-oct-2018}. We
-researched the definitions and applications of different fairness
-metrics \citep{ashokan2021fairness} to determine which metric would be
-most appropriate for our project. We chose to look at four group
-fairness metrics instead of individual fairness metrics. Group fairness
-takes into account the attributes of a whole group as opposed to just
-one individual in the group, allowing us to represent more of the
-systemic issues happening. In general, group fairness metrics require
-that the unprivileged group is treated similarly to the privileged
-group, whereas individual fairness metrics require individuals to be
-treated consistently \citep{kypraiou_what_2021}. Group and individual
-metrics work in opposition of one another, meaning that when group
-fairness improves, individual fairness gets worse.
+algorithms that we used to debias the COMPAS algorithm
+\citep{aif360-oct-2018}. We researched the definitions and applications
+of different fairness metrics \citep{ashokan2021fairness} to determine
+which metric would be most appropriate for our project. We chose to look
+at four group fairness metrics instead of individual fairness metrics.
+Group fairness metrics take into account the attributes of a whole group
+as opposed to just one individual in the group, allowing us to represent
+systemic issues. In general, group fairness metrics require that the
+unprivileged group is treated similarly to the privileged group, whereas
+individual fairness metrics require individuals to be treated
+consistently \citep{kypraiou_what_2021}. Group and individual metrics
+work in opposition of one another, meaning that when group fairness
+improves, individual fairness gets worse \citep{kypraiou_what_2021}.
 
 \textbf{Statistical Parity Difference} - This metric measures the
-difference that the majority and protected classes get a particular
-outcome. The ideal value of this metric is 0. Fairness for this metric
-is between -0.1 and 0.1. A negative value means there is higher benefit
-for the privileged group (in this case, white defendants).
+difference between privileged and marginalized groups' likelihood to get
+a particular outcome. The ideal value of this metric is 0. Fairness for
+this metric is between -0.1 and 0.1. A negative value means there is
+higher benefit for the privileged group (in this case, white
+defendants).
 
@@
 of recidivism, this is the ratio of how many white defendants are
-predicted to not recidivate compared to how many Black defendants are
+predicted to not recidivate compared to how many black defendants are
 predicted to not recidivate. A value of 1 means that the ratio is
@@
 benefits, while a value greater than 1 means the unprivileged group
-(Black defendants) benefits. According to AIF360, a ratio between .8 to
-1.25 is considered fair\citep{Ronaghan2019AI}.
+(Black defendants) benefits. According to AIF360, a ratio between 0.8 to
+1.25 is considered fair \citep{Ronaghan2019AI}.
 
 \[\frac{P(\hat{Y}=1|D=Unprivileged)}{P(\hat{Y}=1|D=Privileged)}\]
-
-\textbf{Average Odds Difference} - This metric returns the average
-difference in false positive rate and true positive rate for the
-privileged and unprivileged groups. A value of 0 indicates equality of
-odds, and a value below 0 implies benefit for the privileged group.
-Equality of odds is achieved in the case of recidivism when the
-proportion of people who were predicted to recidivate and did recidivate
-is equal (true positive rate) for both Black and white defendants AND
-the proportion of people who were predicted to recidivate and did not
-recidivate (false positive rate) is equal for both Black and white
-defendants\citep{aif360-oct-2018}.
-
-\[\frac{(FPR_{D = unprivileged} - FPR_{D = privileged}) + (TPR_{D = unprivileged} - TPR_{D = privileged})}{2}\]
 
@@
 privileged groups. The true positive rate is the ratio of true positives
-to the total number of actual positives for a given
-group\citep{GoogleDev}.
+to the total number of actual positives for a given group
+\citep{GoogleDev}.
 
@@
 \[TPR_{D = Unprivileged} - TPR_{D = Privileged}\]
+
+\textbf{Average Odds Difference} - This metric returns the average
+difference in false positive rate and true positive rate for the
+privileged and unprivileged groups. A value of 0 indicates equality of
+odds, and a value below 0 implies benefit for the privileged group.
+Equality of odds is achieved in the case of recidivism when the
+proportion of people who were predicted to recidivate and did recidivate
+is equal (true positive rate) for both Black and white defendants AND
+the proportion of people who were predicted to recidivate and did not
+recidivate (false positive rate) is equal for both Black and white
+defendants \citep{aif360-oct-2018}.
+
+\[\frac{1}{2}\left[(FPR_{D = Unprivileged} - FPR_{D = Privileged}) + \underbrace{(TPR_{D = Unprivileged} - TPR_{D = Privileged})}_\textrm{Equal Opportunity Difference}\right]\]
 
@@
 
-\hypertarget{preliminary-results}{%
-\section{Preliminary Results}\label{preliminary-results}}
+\hypertarget{results}{%
+\section{Results}\label{results}}
 
@@
 metrics we chose and compared the results (pictured in Table
-\ref{tab:baseline metrics table}))
+\ref{tab:baseline metrics table}).
 
@@
 All four group fairness metrics determine that the COMPAS algorithm
-favors white defendants over Black defendants. The statistical parity
-metric indicates unfairness by a smaller margin of 0.04 than equal
-opportunity difference which has a margin of error 0.4 below the
-threshold. Although this is true, each algorithm uses different
-penalties for unfavorable outcomes, so the higher the penalty the more
-of a difference there will be between the fairness threshold and the
-value.
+favors white defendants over Black defendants. Although the magnitudes
+of the various fairness metrics are different, all of the fairness
+metrics are not within their respective fairness thresholds. Our goal is
+to use pre-processing, in-processing, and post-processing algorithms in
+the AIF360 toolkit to see if we can make COMPAS fair at all.
 
@@
 \end{longtable}
+
+\hypertarget{pre-processing-approach}{%
+\subsection{Pre-Processing Approach}\label{pre-processing-approach}}
+
+We chose the reweighing as our pre-processing bias mitigation algorithm.
+After running the reweighing algorithm, our fairness metrics are -0.015
+for statistical parity difference, 0.015 for equal opportunity
+difference, 0.014 for average odds difference, and 0.98 for disparate
+impact ratio. Overall, these fairness metrics show that the
+pre-processing algorithm reweighing improves the bias in the COMPAS
+algorithm.
 

### 2022-04-26 final_metrics_analysis.Rmd
adding plot titles

@@
 props[('African-American','F')]={'facecolor':purple,'edgecolor':'white'}
-props[('African-American','M')]=        {'facecolor':purple,'edgecolor':'white'}
+props[('African-American','M')]={'facecolor':purple,'edgecolor':'white'}
 labelizer=lambda k:{('Caucasian','F'):1242,('African-American','F'):2194,('Caucasian','M'):858,('African-American','M'):979}[k]
 mosaic(df_viz,['race','c_charge_degree'],labelizer=labelizer,properties=props)
+
+plt.title("Charge Degree by Race: Misdemeanors and Felonies")
+plt.rcParams["figure.figsize"]=(7,7)
 plt.show()
@@
 mosaic(df_viz,['race','two_year_recid'],labelizer=labelizer,properties=props)
+
+plt.title("Recidivism Outcomes by Race")
+plt.rcParams["figure.figsize"]=(7,7)
 plt.show()
@@
 
-#plt.title("Prior Charges Count by Race and Sex)
 plt.show()

### 2022-04-26 Paper_Draft2/Draft2.Rmd

--- 2022-04-21 Paper_Draft2/Draft2.Rmd
@@
 
-Due to the proprietary nature of the COMPAS algorithm, it is unknown how exactly these recidivism risk scores are calculated; however, a [sample COMPAS Risk Assessment Survey](https://www.documentcloud.org/documents/2702103-Sample-Risk-Assessment-COMPAS-CORE#document/p5/a296598) has been made publicly available, revealing the algorithm’s input information.
-This survey has been widely critiqued for using proxy variables for race that do not explicitly factor in a defendant’s race but heavily imply it, allowing Northpointe to claim that their algorithm is free of racial bias.
+Due to the proprietary nature of the COMPAS algorithm, it is unknown how exactly these recidivism risk scores are calculated.
+However, a [sample COMPAS Risk Assessment Survey](https://www.documentcloud.org/documents/2702103-Sample-Risk-Assessment-COMPAS-CORE#document/p5/a296598) has been made publicly available, revealing the algorithm’s input information.
+@angwin2016machine has critiqued for using proxy variables for race that do not explicitly factor in a defendant’s race but heavily imply it, allowing Northpointe to claim that their algorithm is free of racial bias.
 For example, the COMPAS risk assessment survey asks screeners to speculate if a defendant might be affiliated with a gang.
@@
 @angwin2016machine.
-published a 2016 piece in the news publication ProPublica that analyzes the methods and algorithms used by Northpointe in their COMPAS risk score assessment algorithm and uncovers the racial biases in defendants’ scores [@angwin2016machine].
-Their research finds that “the algorithm [is] somewhat more accurate than a coin flip,” a worrisome level of accuracy given the potential impact its determinations may have on real people’s lives.
+analyzes the methods and algorithms used by Northpointe in their COMPAS risk score assessment algorithm and uncovers racial biases in defendants’ scores [@angwin2016machine].
+They find that “the algorithm [is] somewhat more accurate than a coin flip,” a worrisome level of accuracy given the potential impact its determinations may have on real people’s lives.
 @angwin2016machine specifically investigate the distribution of COMPAS scores by decile among Black and white defendants.

### 2022-04-27 final_metrics_analysis.Rmd

@@
 # for Amrita
-#use_condaenv("anaconda3")
+# use_condaenv("anaconda3")
 ```

### 2022-04-27 Paper_Draft2/Draft2.Rmd

--- 2022-04-26 Paper_Draft2/Draft2.Rmd
@@
 ```{r, include = FALSE}
-x <- tribble(~`Two Year Recidivism by Race`, ~`Recidivated`, ~`Survived`,
-             "African American", "1661", "1512",
-             "Caucasian", "822", "1278")
+x <- tribble(~`Two Year Recidivism by Race`, ~`Recidivated`, ~`Survived`, ~`Total`,
+             "African American", 1661, 1512, 3173,
+             "Caucasian", 822, 1278, 2100,
+             "Total", 2483, 2790, 5273)
 ```

### 2022-04-28 Paper/Paper.tex
Renaming files

--- 2022-04-26 Paper_Draft2/Draft2.tex
@@
 
-\begin{longtable}[]{@{}lll@{}}
+\begin{longtable}[]{@{}lrrr@{}}
 \caption{Incidence of recidivism by race, illustrating how a much
@@
 \toprule
-Two Year Recidivism by Race & Recidivated & Survived \\
+Two Year Recidivism by Race & Recidivated & Survived &
+Total\tabularnewline
 \midrule
@@
 \toprule
-Two Year Recidivism by Race & Recidivated & Survived \\
+Two Year Recidivism by Race & Recidivated & Survived &
+Total\tabularnewline
 \midrule
 \endhead
-African American & 1661 & 1512 \\
-Caucasian & 822 & 1278 \\
+African American & 1661 & 1512 & 3173\tabularnewline
+Caucasian & 822 & 1278 & 2100\tabularnewline
+Total & 2483 & 2790 & 5273\tabularnewline
 \bottomrule
@@
 improves, individual fairness gets worse \citep{kypraiou_what_2021}.
+
+\hypertarget{head}{%
+\section{\textless\textless\textless\textless\textless\textless\textless{}
+HEAD}\label{head}}
 
@@
 
-For the next step of our experiment, we need to determine where in the
-data science pipeline we can mitigate the most bias, using
-pre-processing, in-processing, and post-processing de-biasing
-algorithms. These are all based on using predictive models to figure out
-how we can ``fix'' the bias that is present.
+\begin{quote}
+\begin{quote}
+\begin{quote}
+\begin{quote}
+\begin{quote}
+\begin{quote}
+\begin{quote}
+4c082d27d558a182d9eb1c71b333416c134912f6 For the next step of our
+experiment, we need to determine where in the data science pipeline we
+can mitigate the most bias, using pre-processing, in-processing, and
+post-processing de-biasing algorithms. These are all based on using
+predictive models to figure out how we can ``fix'' the bias that is
+present.
+\end{quote}
+\end{quote}
+\end{quote}
+\end{quote}
+\end{quote}
+\end{quote}
+\end{quote}
 
@@
 
+\textless\textless\textless\textless\textless\textless\textless{} HEAD
+\#\# Statistical Parity Difference
+
+This metric measures the difference that the majority and protected
+classes get a particular outcome. The ideal value of this metric is 0.
+Fairness for this metric is between -0.1 and 0.1. A negative value means
+there is higher benefit for the privileged group (in this case, white
+defendants).
+
+\[P(\hat{Y}=1|D=Unprivileged) - P(\hat{Y}=1|D=Privileged)\]
+
+\hypertarget{disparate-impact-ratio}{%
+\subsection{Disparate Impact Ratio}\label{disparate-impact-ratio}}
+
+This metric is the ratio of how often the favorable outcome occurs in
+one group versus the other. In the case of recidivism, this is the ratio
+of how many white defendants are predicted to not recidivate compared to
+how many Black defendants are predicted to not recidivate. A value of 1
+means that the ratio is exactly 1:1. Less than 1 means the privileged
+group (white defendants) benefits, while a value greater than 1 means
+the unprivileged group (Black defendants) benefits. According to AIF360,
+a ratio between .8 to 1.25 is considered fair\citep{Ronaghan2019AI}.
+
+\[\frac{P(\hat{Y}=1|D=Unprivileged)}{P(\hat{Y}=1|D=Privileged)}\]
+
+\hypertarget{average-odds-difference}{%
+\subsection{Average Odds Difference}\label{average-odds-difference}}
+
+This metric returns the average difference in false positive rate and
+true positive rate for the privileged and unprivileged groups. A value
+of 0 indicates equality of odds, and a value below 0 implies benefit for
+the privileged group. Equality of odds is achieved in the case of
+recidivism when the proportion of people who were predicted to
+recidivate and did recidivate is equal (true positive rate) for both
+Black and white defendants AND the proportion of people who were
+predicted to recidivate and did not recidivate (false positive rate) is
+equal for both Black and white defendants \citep{aif360-oct-2018}.
+
+\[\frac{(FPR_{D = unprivileged} - FPR_{D = privileged}) + (TPR_{D = unprivileged} - TPR_{D = privileged})}{2}\]
+
+\hypertarget{equal-opportunity-difference}{%
+\subsection{Equal Opportunity
+Difference}\label{equal-opportunity-difference}}
+
+This metric is computed as the difference of true positive rates between
+the unprivileged and the privileged groups. The true positive rate is
+the ratio of true positives to the total number of actual positives for
+a given group\citep{GoogleDev}.
+
+The ideal value is 0. A value less than 0 implies higher benefit for the
+privileged group and a value greater than 0 implies higher benefit for
+the unprivileged group. Fairness for this metric is between -0.1 and 0.1
+\citep{aif360-oct-2018}.
+
+This metric is best used when it is very important to catch positive
+outcomes while false positives are not exceptionally problematic
+\citep{Cortez2019How}. This is not the case for the COMPAS data set, as
+false positives mean extra jail time for someone who will not actually
+re-offend.
+
+\[TPR_{D = Unprivileged} - TPR_{D = Privileged}\]
+
+\hypertarget{preliminary-results}{%
+\section{\# Preliminary Results}\label{preliminary-results}}
+
 \hypertarget{results}{%
 \section{Results}\label{results}}
+
+\begin{quote}
+\begin{quote}
+\begin{quote}
+\begin{quote}
+\begin{quote}
+\begin{quote}
+\begin{quote}
+4c082d27d558a182d9eb1c71b333416c134912f6
+\end{quote}
+\end{quote}
+\end{quote}
+\end{quote}
+\end{quote}
+\end{quote}
+\end{quote}
 

### 2022-04-28 Markdown Files Sprints 3 and 4/race compas analysis.Rmd

--- 2022-04-05 race compas analysis.Rmd
@@
 library(reticulate)
+use_condaenv("base")
+#if (grepl("dcaravela", Sys.getenv("HOME"))) {
+ # use_condaenv("base")
+#}
 ```
@@
 ```{python import, include= FALSE}
-import matplotlib.pyplot as plt
+#import matplotlib.pyplot as plt
 import matplotlib.patches as patches
@@
 
-from IPython.display import Markdown, display
+#from IPython.display import Markdown, display
 ```

### 2022-04-28 Paper_Draft2/Draft2.tex

@@
 \address{%
-$^{1}$ \quad Statistical and Data Sciences Smith College Northampton, MA
-01063; \href{mailto:bbaumer@smith.edu}{\nolinkurl{bbaumer@smith.edu}}\\
+$^{1}$ \quad Statistical and Data Sciences Smith College Northampton, MA 01063; \href{mailto:bbaumer@smith.edu}{\nolinkurl{bbaumer@smith.edu}}\\
 }
@@
 % Abstract (Do not insert blank lines, i.e. \\)
-\abstract{A variety of disciplines use risk assessment instruments to
-help humans make data-driven decisions. Northpointe, a software company,
-created an algorithmic risk assessment instrument known as the
-Correctional Offender Management Profiling for Alternative Sanctions
-(COMPAS). COMPAS uses various behavioral and psychological metrics
-related to recidivism to assist justice systems in assessing a
-defendant's potential recidivism risk. \citet{angwin2016machine}
-published a ProPublica article in which they conclude that the racial
-biases in the criminal justice system are reflected in the COMPAS
-recidivism risk scores. In response, \citet{equivant_response_2016}
-published a rebuttal on behalf of Northpointe defending the COMPAS
-algorithm and refuting \citet{angwin2016machine}'s allegation of racial
-bias. Using a human rights framework adopted from the organizations
-Women at the Table and AI Fairness 360, we use debiasing algorithms and
-fairness metrics to analyze the argument between Northpointe and
-ProPublica and determine whether and to what extent there is racial bias
-in the COMPAS algorithm. All four group fairness metrics determine that
-the COMPAS algorithm favors white defendants over Black defendants.}
+\abstract{A variety of disciplines use risk assessment instruments to help humans
+make data-driven decisions. Northpointe, a software company, created an
+algorithmic risk assessment instrument known as the Correctional
+Offender Management Profiling for Alternative Sanctions (COMPAS). COMPAS
+uses various behavioral and psychological metrics related to recidivism
+to assist justice systems in assessing a defendant's potential
+recidivism risk. \citet{angwin2016machine} published a ProPublica
+article in which they conclude that the racial biases in the criminal
+justice system are reflected in the COMPAS recidivism risk scores. In
+response, \citet{equivant_response_2016} published a rebuttal on behalf
+of Northpointe defending the COMPAS algorithm and refuting
+\citet{angwin2016machine}'s allegation of racial bias. Using a human
+rights framework adopted from the organizations Women at the Table and
+AI Fairness 360, we use debiasing algorithms and fairness metrics to
+analyze the argument between Northpointe and ProPublica and determine
+whether and to what extent there is racial bias in the COMPAS algorithm.
+All four group fairness metrics determine that the COMPAS algorithm
+favors white defendants over Black defendants.}
 

### 2022-04-28 Paper_Draft2/Draft2.Rmd
trying to fix it
reverting to commit before latex issues and forcing push
git force

@@
 
-**Statistical Parity Difference** - This metric measures the difference between privileged and marginalized groups’ likelihood to get a particular outcome.
+## Statistical Parity Difference
+This metric measures the difference between privileged and marginalized groups’ likelihood to get a particular outcome.
 The ideal value of this metric is 0.
@@
 
-**Disparate Impact Ratio** - This metric is the ratio of how often the favorable outcome occurs in one group versus the other.
+## Disparate Impact Ratio
+This metric is the ratio of how often the favorable outcome occurs in one group versus the other.
 In the case of recidivism, this is the ratio of how many white defendants are predicted to not recidivate compared to how many black defendants are predicted to not recidivate.
@@
 
-**Equal Opportunity Difference** - This metric is computed as the difference of true positive rates between the unprivileged and the privileged groups.
+## Equal Opportunity Difference
+This metric is computed as the difference of true positive rates between the unprivileged and the privileged groups.
 The true positive rate is the ratio of true positives to the total number of actual positives for a given group [@GoogleDev].
@@
 
-**Average Odds Difference** - This metric returns the average difference in false positive rate and true positive rate for the privileged and unprivileged groups.
+## Average Odds Difference
+This metric returns the average difference in false positive rate and true positive rate for the privileged and unprivileged groups.
 A value of 0 indicates equality of odds, and a value below 0 implies benefit for the privileged group.

### 2022-04-28 final_metrics_analysis.Rmd

--- 2022-04-26 final_metrics_analysis.Rmd
@@
 
-## Reweighting
+Pre-processing refers to mitigating bias within the training data.
+For our pre-processing approach we are using AIF360's Reweighing algorithm.
+The reweighing approach "assigns different weights to the examples based upon their categories of protected attribute and outcome such that bias is removed from the training dataset" ( [Mahmoudian](https://towardsdatascience.com/reweighing-the-adult-dataset-to-make-it-discrimination-free-44668c9379e8)).
+
+## Reweighing
 
@@
 # In-Processing
+
+In-processing mitigates bias in classifiers while building a model.
+The in-processing technique we use is the prejudice remover algorithm from AIF360.
+The prejudice remover is a method for reducing indirect prejudice (i.e.
+how the COMPAS algorithm is racially biased because it uses proxy variables for race).
+This approach implements two different regularizers, one to avoid overfitting and one to enforce fair classification ( [Kamishima et.
+al.](https://link.springer.com/content/pdf/10.1007/978-3-642-33486-3_3.pdf)).
+The prejudice remover regularizer works by minimizing the prejudice index, a mathematical equation for quantifying fairness defined by [Kamishima et.
+al](https://link.springer.com/content/pdf/10.1007/978-3-642-33486-3_3.pdf).
+This in turn enforces a classifier’s independence from sensitive information (like race).
 
@@
 
+Our last approach, post-processing bias mitigation, is used after training a model.
+Post-processing algorithms equalize the outcomes (i.e., predicted classification values) to mitigate bias instead of adjusting the classifier or the training data( [Baxter](https://www.infoworld.com/article/3630450/what-is-ai-bias-mitigation-and-how-can-it-improve-ai-fairness.html)).
+We use calibrated equalized odds, which "optimizes over calibrated classifier score outputs to find probabilities with which to change output labels with an equalized odds objective" ( [AIF360](https://aif360.readthedocs.io/en/stable/modules/generated/aif360.algorithms.postprocessing.CalibratedEqOddsPostprocessing.html#aif360.algorithms.postprocessing.CalibratedEqOddsPostprocessing)).
+An equalized odds objective constrains classification algorithms such that no error type (false-positive or false-negative) disproportionately affects any population subgroup; both groups, in our case both white and Black defendants, should have the same false-positive and false-negative rates.
+Through the calibrated equalized odds method, we want to decrease bias while also maintaining calibration ( [Pleiss et.
+al.](https://arxiv.org/pdf/1709.02012.pdf)).
+[Calibration](https://medium.com/analytics-vidhya/calibration-in-machine-learning-e7972ac93555) refers to improving a model so that the distribution of predicted outcomes is similar to the distribution of observed probability in the training data.
+
+## Calibrated Equalized Odds
+
 ```{python post-processing calibrated equal odds set up, include = FALSE}

### 2022-04-28 Paper_Draft2/Draft2.tex

@@
 
-\textbf{Statistical Parity Difference} - This metric measures the
-difference between privileged and marginalized groups' likelihood to get
-a particular outcome. The ideal value of this metric is 0. Fairness for
-this metric is between -0.1 and 0.1. A negative value means there is
-higher benefit for the privileged group (in this case, white
-defendants).
+\hypertarget{statistical-parity-difference}{%
+\subsection{Statistical Parity
+Difference}\label{statistical-parity-difference}}
+
+This metric measures the difference between privileged and marginalized
+groups' likelihood to get a particular outcome. The ideal value of this
+metric is 0. Fairness for this metric is between -0.1 and 0.1. A
+negative value means there is higher benefit for the privileged group
+(in this case, white defendants).
 
@@
 
-\textbf{Disparate Impact Ratio} - This metric is the ratio of how often
-the favorable outcome occurs in one group versus the other. In the case
-of recidivism, this is the ratio of how many white defendants are
-predicted to not recidivate compared to how many black defendants are
-predicted to not recidivate. A value of 1 means that the ratio is
-exactly 1:1. Less than 1 means the privileged group (white defendants)
-benefits, while a value greater than 1 means the unprivileged group
-(Black defendants) benefits. According to AIF360, a ratio between 0.8 to
-1.25 is considered fair \citep{Ronaghan2019AI}.
+\hypertarget{disparate-impact-ratio}{%
+\subsection{Disparate Impact Ratio}\label{disparate-impact-ratio}}
+
+This metric is the ratio of how often the favorable outcome occurs in
+one group versus the other. In the case of recidivism, this is the ratio
+of how many white defendants are predicted to not recidivate compared to
+how many black defendants are predicted to not recidivate. A value of 1
+means that the ratio is exactly 1:1. Less than 1 means the privileged
+group (white defendants) benefits, while a value greater than 1 means
+the unprivileged group (Black defendants) benefits. According to AIF360,
+a ratio between 0.8 to 1.25 is considered fair \citep{Ronaghan2019AI}.
 
@@
 
-\textbf{Equal Opportunity Difference} - This metric is computed as the
-difference of true positive rates between the unprivileged and the
-privileged groups. The true positive rate is the ratio of true positives
-to the total number of actual positives for a given group
-\citep{GoogleDev}.
+\hypertarget{equal-opportunity-difference}{%
+\subsection{Equal Opportunity
+Difference}\label{equal-opportunity-difference}}
+
+This metric is computed as the difference of true positive rates between
+the unprivileged and the privileged groups. The true positive rate is
+the ratio of true positives to the total number of actual positives for
+a given group \citep{GoogleDev}.
 
@@
 
-\textbf{Average Odds Difference} - This metric returns the average
-difference in false positive rate and true positive rate for the
-privileged and unprivileged groups. A value of 0 indicates equality of
-odds, and a value below 0 implies benefit for the privileged group.
-Equality of odds is achieved in the case of recidivism when the
-proportion of people who were predicted to recidivate and did recidivate
-is equal (true positive rate) for both Black and white defendants AND
-the proportion of people who were predicted to recidivate and did not
-recidivate (false positive rate) is equal for both Black and white
-defendants \citep{aif360-oct-2018}.
+\hypertarget{average-odds-difference}{%
+\subsection{Average Odds Difference}\label{average-odds-difference}}
+
+This metric returns the average difference in false positive rate and
+true positive rate for the privileged and unprivileged groups. A value
+of 0 indicates equality of odds, and a value below 0 implies benefit for
+the privileged group. Equality of odds is achieved in the case of
+recidivism when the proportion of people who were predicted to
+recidivate and did recidivate is equal (true positive rate) for both
+Black and white defendants AND the proportion of people who were
+predicted to recidivate and did not recidivate (false positive rate) is
+equal for both Black and white defendants \citep{aif360-oct-2018}.
 

### 2022-04-28 race compas analysis.Rmd

@@
 acc_REW = accuracy_score(y_test, y_pred_REW)
-print(f'[Reweighting] The test accuracy of the algorithm is: {acc_REW: .2%}')
+print(f'[Reweighing] The test accuracy of the algorithm is: {acc_REW: .2%}')
 ```
@@
 cf_matrix = confusion_matrix(y_test, y_pred_REW)
-make_confusion_matrix(cf_matrix, "[Reweighting]")
+make_confusion_matrix(cf_matrix, "[Reweighing]")
 plt.show()
@@
 
-print(f'[Reweighting] The statistical parity difference is {stat_par_diff_RW: .2}')
+print(f'[Reweighing] The statistical parity difference is {stat_par_diff_RW: .2}')
 ```
@@
 
-print(f'[Reweighting] The equal opportunity difference is {eq_opp_diff_RW: .2}')
+print(f'[Reweighing] The equal opportunity difference is {eq_opp_diff_RW: .2}')
 ```
@@
 
-print(f'[Reweighting] The generalized entropy error is {gen_entr_error_RW: .2}')
+print(f'[Reweighing] The generalized entropy error is {gen_entr_error_RW: .2}')
 ```
@@
 ```{python pre-processing fairness metrics plot}
-plot_fair_metrics([stat_par_diff_RW, eq_opp_diff_RW, gen_entr_error_RW], 'Reweighting')
+plot_fair_metrics([stat_par_diff_RW, eq_opp_diff_RW, gen_entr_error_RW], 'Reweighing')
 plt.show()

### 2022-04-28 Paper_Draft2/Draft2.Rmd
Updating processing algorithms sections

@@
 
-Pre-processing refers to assessing the training data, and it is the most flexible method because it has not yet trained a model that may carry assumptions about the data.
+Pre-processing refers to mitigating bias within the training data, and it is the most flexible method because it has not yet trained a model that may carry assumptions about the data.
 It is important to keep in mind that pre-processing prevents assumptions in the modeling, but does not account for the bias in data collection.
-With pre-processing, we use the re-weighing algorithm from AIF360 which weights the examples in each (group, label) combination differently to ensure fairness before classification [@aif360-oct-2018].
-We use a logistic regression model for this algorithm, as it is the easiest to interpret in the given context.
-After running the fairness metrics using the pre-processing algorithm, we are able to compare our results to the baseline metrics from the previous section.
+Training data is where bias is most likely to be introduced.
+We use the reweighing pre-processing algorithm from AIF360 which assigns weights to the data.
+“The advantage of this approach is, instead of modifying the labels, it assigns different weights to the examples based upon their categories of protected attribute and outcome such that bias is removed from the training dataset.
+The weights are based on frequency counts.
+However as this technique is designed to work only with classifiers that can handle row-level weights, this may limit your modeling options” [@Ronaghan2019AI].
+We used a logistic regression model for this algorithm, as it is the easiest to interpret in the given context.
+After running the fairness metrics using the pre-processing algorithm, we were able to compare our results to the baseline metrics from the previous section.
 
-Another approach, in-processing, explores how bias can be targeted while building a model.
-It assesses how we build learning algorithms, which often prioritize accuracy over fairness.
-The technique we use is the Meta Classification algorithm, which accounts for the fairness metric as part of the input and returns a classifier optimized by that particular metric.
-Similarly to pre-processing, we compare the results of our in-processing methods with both the baseline and the pre-processing to gauge which method so far has better individual or group fairness.
+In-processing mitigates bias in classifiers while building a model.
+A classifier “is an algorithm that automatically orders or categorizes data into one or more sets” [@baxter2021what].
+The in-processing technique we use is the prejudice remover algorithm, which accounts for the fairness metric as part of the input and returns a classifier optimized by that particular metric.
+In order to do this, we first needed to convert our data frame into a data type called a BinaryLabelDataset.
+Similarly to pre-processing, we compared the results of our in-processing methods with both the baseline and the pre-processing to gauge which method so far has better individual or group fairness.
 
-Our last approach, post-processing, makes transformations on model outputs.
-Similarly to the pre-processing, it is flexible because it does not work directly with the algorithm, but instead has direct access to all the predicted classification values.
-Because the algorithms are often so complex and difficult to interpret, post-processors are trained in a black box.
-To get around this, we use the PostProcessingMeta class from AIF360 which combines the training and prediction for a random estimator and the post-processor while simultaneously splitting the data set [@aif360-oct-2018].
-Based on these results, we look comparatively at all the approaches to determine which we assess to be the most fair.
+The prejudice remover is a method for reducing indirect prejudice (i.e.
+how COMPAS is racially biased because it uses proxy variables for race).
+The prejudice remover implements two different regularizers, one to avoid overfitting and one to enforce fair classification.
+The prejudice remover regularizer works by minimizing the prejudice index, a mathematical equation for quantifying fairness defined by Kamishima et.
+al.
+This in turn enforces a classifier’s independence from sensitive information (e.g., race).
+
+Our last approach, post-processing bias mitigation, is used after training a model.
+Post-processing algorithms equalize the outcomes (i.e., predicted classification values) to mitigate bias instead of adjusting the classifier or the training data [@baxter2021AI].
+We use calibrated equalized odds, which “optimizes over calibrated classifier score outputs to find probabilities with which to change output labels with an equalized odds objective” [insert aif360 website citation here].
+An equalized odds objective constrains classification algorithms such that no error type (false-positive or false-negative) disproportionately affects any population subgroup; both groups, in our case both white and Black defendants, should have the same false-positive and false-negative rates.
+Through the calibrated equalized odds method, we want to decrease bias while also maintaining calibration [@pleiss2017fairness].
+[Calibration](https://medium.com/analytics-vidhya/calibration-in-machine-learning-e7972ac93555) refers to improving a model so that the distribution of predicted outcomes is similar to the distribution of observed probability in the training data.
 
@@
 
-We chose the reweighing as our pre-processing bias mitigation algorithm.
 After running the reweighing algorithm, our fairness metrics are -0.015 for statistical parity difference, 0.015 for equal opportunity difference, 0.014 for average odds difference, and 0.98 for disparate impact ratio.

### 2022-04-28 Paper_Draft2/Draft2.tex

@@
 % Full title of the paper (Capitalized)
-\Title{Does the COMPAS Needle Always Point Towards Equity? Finding Fairness in
-the COMPAS Risk Assessment Algorithm: A Case Study}
+\Title{Does the COMPAS Needle Always Point Towards Equity? Finding
+Fairness in the COMPAS Risk Assessment Algorithm: A Case Study}
 
 % Authors, for the paper (add full first names)
-\Author{Amrita Acharya$^{1}$, Dianne Caravela$^{1}$, Eunice Kim$^{1}$, Emma Kornberg$^{1}$, Elisabeth Nesmith$^{1}$}
+\Author{Amrita Acharya$^{1}$, Dianne Caravela$^{1}$, Eunice
+Kim$^{1}$, Emma Kornberg$^{1}$, Elisabeth Nesmith$^{1}$}
 
 % Authors, for metadata in PDF
-\AuthorNames{Amrita Acharya, Dianne Caravela, Eunice Kim, Emma Kornberg, Elisabeth Nesmith}
+\AuthorNames{Amrita Acharya, Dianne Caravela, Eunice Kim, Emma
+Kornberg, Elisabeth Nesmith}
 
@@
 \address{%
-$^{1}$ \quad Statistical and Data Sciences Smith College Northampton, MA 01063; \href{mailto:bbaumer@smith.edu}{\nolinkurl{bbaumer@smith.edu}}\\
+$^{1}$ \quad Statistical and Data Sciences Smith College Northampton, MA
+01063; \href{mailto:bbaumer@smith.edu}{\nolinkurl{bbaumer@smith.edu}}\\
 }
@@
 % Abstract (Do not insert blank lines, i.e. \\)
-\abstract{A variety of disciplines use risk assessment instruments to help humans
-make data-driven decisions. Northpointe, a software company, created an
-algorithmic risk assessment instrument known as the Correctional
-Offender Management Profiling for Alternative Sanctions (COMPAS). COMPAS
-uses various behavioral and psychological metrics related to recidivism
-to assist justice systems in assessing a defendant's potential
-recidivism risk. \citet{angwin2016machine} published a ProPublica
-article in which they conclude that the racial biases in the criminal
-justice system are reflected in the COMPAS recidivism risk scores. In
-response, \citet{equivant_response_2016} published a rebuttal on behalf
-of Northpointe defending the COMPAS algorithm and refuting
-\citet{angwin2016machine}'s allegation of racial bias. Using a human
-rights framework adopted from the organizations Women at the Table and
-AI Fairness 360, we use debiasing algorithms and fairness metrics to
-analyze the argument between Northpointe and ProPublica and determine
-whether and to what extent there is racial bias in the COMPAS algorithm.
-All four group fairness metrics determine that the COMPAS algorithm
-favors white defendants over Black defendants.}
+\abstract{A variety of disciplines use risk assessment instruments to
+help humans make data-driven decisions. Northpointe, a software company,
+created an algorithmic risk assessment instrument known as the
+Correctional Offender Management Profiling for Alternative Sanctions
+(COMPAS). COMPAS uses various behavioral and psychological metrics
+related to recidivism to assist justice systems in assessing a
+defendant's potential recidivism risk. \citet{angwin2016machine}
+published a ProPublica article in which they conclude that the racial
+biases in the criminal justice system are reflected in the COMPAS
+recidivism risk scores. In response, \citet{equivant_response_2016}
+published a rebuttal on behalf of Northpointe defending the COMPAS
+algorithm and refuting \citet{angwin2016machine}'s allegation of racial
+bias. Using a human rights framework adopted from the organizations
+Women at the Table and AI Fairness 360, we use debiasing algorithms and
+fairness metrics to analyze the argument between Northpointe and
+ProPublica and determine whether and to what extent there is racial bias
+in the COMPAS algorithm. All four group fairness metrics determine that
+the COMPAS algorithm favors white defendants over Black defendants.}
 
@@
 \toprule
-Two Year Recidivism by Race & Recidivated & Survived &
-Total\tabularnewline
+Two Year Recidivism by Race & Recidivated & Survived & Total \\
 \midrule
@@
 \toprule
-Two Year Recidivism by Race & Recidivated & Survived &
-Total\tabularnewline
+Two Year Recidivism by Race & Recidivated & Survived & Total \\
 \midrule
 \endhead
-African American & 1661 & 1512 & 3173\tabularnewline
-Caucasian & 822 & 1278 & 2100\tabularnewline
-Total & 2483 & 2790 & 5273\tabularnewline
+African American & 1661 & 1512 & 3173 \\
+Caucasian & 822 & 1278 & 2100 \\
+Total & 2483 & 2790 & 5273 \\
 \bottomrule
@@
 
-Pre-processing refers to assessing the training data, and it is the most
-flexible method because it has not yet trained a model that may carry
-assumptions about the data. It is important to keep in mind that
-pre-processing prevents assumptions in the modeling, but does not
-account for the bias in data collection. With pre-processing, we use the
-re-weighing algorithm from AIF360 which weights the examples in each
-(group, label) combination differently to ensure fairness before
-classification \citep{aif360-oct-2018}. We use a logistic regression
-model for this algorithm, as it is the easiest to interpret in the given
-context. After running the fairness metrics using the pre-processing
-algorithm, we are able to compare our results to the baseline metrics
-from the previous section.
+Pre-processing refers to mitigating bias within the training data, and
+it is the most flexible method because it has not yet trained a model
+that may carry assumptions about the data. It is important to keep in
+mind that pre-processing prevents assumptions in the modeling, but does
+not account for the bias in data collection. Training data is where bias
+is most likely to be introduced. We use the reweighing pre-processing
+algorithm from AIF360 which assigns weights to the data. ``The advantage
+of this approach is, instead of modifying the labels, it assigns
+different weights to the examples based upon their categories of
+protected attribute and outcome such that bias is removed from the
+training dataset. The weights are based on frequency counts. However as
+this technique is designed to work only with classifiers that can handle
+row-level weights, this may limit your modeling options''
+\citep{Ronaghan2019AI}. We used a logistic regression model for this
+algorithm, as it is the easiest to interpret in the given context. After
+running the fairness metrics using the pre-processing algorithm, we were
+able to compare our results to the baseline metrics from the previous
+section.
 
-Another approach, in-processing, explores how bias can be targeted while
-building a model. It assesses how we build learning algorithms, which
-often prioritize accuracy over fairness. The technique we use is the
-Meta Classification algorithm, which accounts for the fairness metric as
-part of the input and returns a classifier optimized by that particular
-metric. Similarly to pre-processing, we compare the results of our
-in-processing methods with both the baseline and the pre-processing to
-gauge which method so far has better individual or group fairness.
+In-processing mitigates bias in classifiers while building a model. A
+classifier ``is an algorithm that automatically orders or categorizes
+data into one or more sets'' \citep{baxter2021what}. The in-processing
+technique we use is the prejudice remover algorithm, which accounts for
+the fairness metric as part of the input and returns a classifier
+optimized by that particular metric. In order to do this, we first
+needed to convert our data frame into a data type called a
+BinaryLabelDataset. Similarly to pre-processing, we compared the results
+of our in-processing methods with both the baseline and the
+pre-processing to gauge which method so far has better individual or
+group fairness.
 
-Our last approach, post-processing, makes transformations on model
-outputs. Similarly to the pre-processing, it is flexible because it does
-not work directly with the algorithm, but instead has direct access to
-all the predicted classification values. Because the algorithms are
-often so complex and difficult to interpret, post-processors are trained
-in a black box. To get around this, we use the PostProcessingMeta class
-from AIF360 which combines the training and prediction for a random
-estimator and the post-processor while simultaneously splitting the data
-set \citep{aif360-oct-2018}. Based on these results, we look
-comparatively at all the approaches to determine which we assess to be
-the most fair.
+The prejudice remover is a method for reducing indirect prejudice
+(i.e.~how COMPAS is racially biased because it uses proxy variables for
+race). The prejudice remover implements two different regularizers, one
+to avoid overfitting and one to enforce fair classification. The
+prejudice remover regularizer works by minimizing the prejudice index, a
+mathematical equation for quantifying fairness defined by Kamishima et.
+al.~This in turn enforces a classifier's independence from sensitive
+information (e.g., race).
+
+Our last approach, post-processing bias mitigation, is used after
+training a model. Post-processing algorithms equalize the outcomes
+(i.e., predicted classification values) to mitigate bias instead of
+adjusting the classifier or the training data \citep{baxter2021AI}. We
+use calibrated equalized odds, which ``optimizes over calibrated
+classifier score outputs to find probabilities with which to change
+output labels with an equalized odds objective'' {[}insert aif360
+website citation here{]}. An equalized odds objective constrains
+classification algorithms such that no error type (false-positive or
+false-negative) disproportionately affects any population subgroup; both
+groups, in our case both white and Black defendants, should have the
+same false-positive and false-negative rates. Through the calibrated
+equalized odds method, we want to decrease bias while also maintaining
+calibration \citep{pleiss2017fairness}.
+\href{https://medium.com/analytics-vidhya/calibration-in-machine-learning-e7972ac93555}{Calibration}
+refers to improving a model so that the distribution of predicted
+outcomes is similar to the distribution of observed probability in the
+training data.
 
@@
 
-We chose the reweighing as our pre-processing bias mitigation algorithm.
 After running the reweighing algorithm, our fairness metrics are -0.015

### 2022-04-28 final_metrics_analysis.Rmd

@@
 # for Dianne
-# use_condaenv("base")
+use_condaenv("base")
 
@@
 This means that the re-weighing model slightly lowered the accuracy of the baseline model in predicting people who did recidivate.
-As a result, the overall reweighting model accuracy remains about the same as the baseline model accuracy.
+As a result, the overall reweighing model accuracy remains about the same as the baseline model accuracy.
 The false negative and false positive rates, or the predictions that proved to be incorrect, are 23.20% and 15.85%, respectively.
@@
 
-We can see clear effects of the reweighting approach in the recall scores, which are now all very similar around 0.5.
+We can see clear effects of the reweighing approach in the recall scores, which are now all very similar around 0.5.
 This is a marked improvement from the baseline model, where the model only correctly identified white defendants who recidivated 20% of the time.
-The reweighting model now correctly identifies defendants of any race who recidivate about half of the time.
-Notably, the reweighting also lowered the recall score for Black defendants.
+The reweighing model now correctly identifies defendants of any race who recidivate about half of the time.
+Notably, the reweighing also lowered the recall score for Black defendants.
 
-Whereas with the baseline model all three precision scores were relatively similar, with the reweighting model we see the precision score for Black defendants is the highest.
+Whereas with the baseline model all three precision scores were relatively similar, with the reweighing model we see the precision score for Black defendants is the highest.
 The models predictions for Black defendants who recidivate are correct about 66% of the time.
@@
 
-print(f'[Reweighting] The average odds difference is {avg_odds_diff_RW: .2}')
+print(f'[Reweighing] The average odds difference is {avg_odds_diff_RW: .2}')
 ```
@@
 
-print(f'[Reweighting] The disparate impact ratio is {disp_impact_ratio_RW: .2}')
+print(f'[Reweighing] The disparate impact ratio is {disp_impact_ratio_RW: .2}')
 ```
 
-The reweighting approach also improves the disparate impact ratio, bringing the value almost to 1, which conveys absolute fairness.
+The reweighing approach also improves the disparate impact ratio, bringing the value almost to 1, which conveys absolute fairness.
 
-### Plot Fairness Metrics (Reweighting)
+### Plot Fairness Metrics (Reweighing)
 
 ```{python pre-processing fairness metrics plot}
-plot_fair_metrics([stat_par_diff_RW, eq_opp_diff_RW, avg_odds_diff_RW, disp_impact_ratio_RW], 'Reweighting')
+plot_fair_metrics([stat_par_diff_RW, eq_opp_diff_RW, avg_odds_diff_RW, disp_impact_ratio_RW], 'Reweighing')
 plt.show()
@@
 
-The orange bars in these graphs represent the new values of the fairness metrics after implementing the reweighting approach.
-Like we saw above, reweighting the data improved the fairness in all four of these metrics.
+The orange bars in these graphs represent the new values of the fairness metrics after implementing the reweighing approach.
+Like we saw above, reweighing the data improved the fairness in all four of these metrics.
 All four values are now within the range of values considered fair.
@@
 
-After the prejudice remover, the accuracy of the model is fairly similar across the board of races, but the model's accuracy is only 38.21%, which is much lower than the baseline and reweighting models.
+After the prejudice remover, the accuracy of the model is fairly similar across the board of races, but the model's accuracy is only 38.21%, which is much lower than the baseline and reweighing models.
 This accuracy score means that the model makes accurate predictions only 38% of the time.

### 2022-04-29 final_metrics_analysis.Rmd
Reweigh not reweight

@@
 # for Dianne
-use_condaenv("base")
+# use_condaenv("base")
 
 # for Amrita
-# use_condaenv("anaconda3")
+use_condaenv("anaconda3")
 ```

### 2022-04-29 race-compas-analysis.Rmd

@@
 library(reticulate)
+#use_condaenv("base")
+#if (grepl("dcaravela", Sys.getenv("HOME"))) {
+ # use_condaenv("base")
+#}
+
+use_condaenv("anaconda3")
 ```
@@
 
-```{python baseline set up}
+```{python baseline set up, warning = FALSE, message = FALSE, echo = FALSE}
 from aif360.sklearn.preprocessing import ReweighingMeta, Reweighing
@@
 # Pre-Processing Approach
-```{python pre-processing set up}
+```{python pre-processing set up, warning = FALSE, message = FALSE, echo = FALSE}
 lr = LogisticRegressionCV(solver='lbfgs')
@@
 
-```{python post-processing calibrated equal odds set up}
+```{python post-processing calibrated equal odds set up, warning = FALSE, message = FALSE, echo = FALSE}
 from aif360.sklearn.postprocessing import CalibratedEqualizedOdds, PostProcessingMeta

### 2022-05-02 Paper_Draft2/Draft2.Rmd
fixing and adding images (still having bug with one)

@@
 For the purposes of our project, which endeavors to evaluate the differing effects of the COMPAS algorithm on white defendants versus Black defendants, we have filtered the data to only include individuals whose race is listed as Caucasian or African-American.
-Our data therefore has 5,273 rows (Figure \ref{fig:table snip}), with the distributions of race (Figure \ref{fig:race plot}), age (Figure \ref{fig:age plot}), and two-year recidivism rate (Figures \ref{fig:recid plot} & \ref{fig:recid race plot} and Table \ref{tab:recid table}) shown below.
+Our data therefore has 5,273 rows (Figure \ref{fig:table snip}), with the distributions of age (Figure \ref{fig:age plot}), prior charges (Figure \ref{fig:prior plot}), and two-year recidivism by race (Figure \ref{fig:recidivism mosaic}) and (Table \ref{tab:recid table}) shown below.
 
@@
 
-```{r age plot, echo=FALSE, fig.align='center', fig.cap= "Distribution of defendant ages by race.
-The purple curve shows the distribution of the ages of Black defendants, and the green curve shows the distribution of the ages of white defendants.
+```{r age plot, echo=FALSE, fig.align='center', fig.cap= "The purple curve shows the distribution of the ages of Black defendants, and the green curve shows the distribution of the ages of white defendants.
 The probability of a defendant's age being between two points on the x-axis is the total shaded area of the curve under the two points.
@@
 
-```{r recid race plot,echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two-year recidivism outcomes by race.
-When we divide the data into Black and white defendants, we can see that Black defendants recidivate more than white defendants and Black defendants are more likely to recidivate than not recidivate.
+```{r prior plot, echo=FALSE, fig.align = 'center', fig.cap = 'Black defendants, particularly men, are more likely to have a greater count of prior charges than white defendants.
+Male defendants have a higher number of prior charges than do female defendants.
+Though we do not know for sure which information goes into the COMPAS algorithm, it is likely that a defendant with prior charges will be coded as a having a higher risk of recidivism.
+Thus, by looking at the racial discrepancies in prior charges we can already see potential bias in the algorithm.', out.width= "100%"}
+knitr::include_graphics("../images/prior_charges_plot.png")
+```
+
+```{r recidivism mosaic, echo=FALSE, fig.align = 'center', fig.cap = 'When we divide the data into Black and white defendants, we can see that Black defendants recidivate more than white defendants and Black defendants are more likely to recidivate than not recidivate.
+39.14% of white defendants did recidivate within two years compared to 52.35% of black defendants.
 We can also see that there are more Black defendants in the dataset overall.', out.width= "100%"}
-knitr::include_graphics("../images/race_recid_bar_plot_new.png")
+knitr::include_graphics("../images/recidivism_mosaic.png")
 ```

### 2022-05-02 Paper_Draft2/Draft2.tex
fixing and adding images (still having bug with one)

@@
 or African-American. Our data therefore has 5,273 rows (Figure
-\ref{fig:table snip}), with the distributions of race (Figure
-\ref{fig:race plot}), age (Figure \ref{fig:age plot}), and two-year
-recidivism rate (Figures \ref{fig:recid plot} \&
-\ref{fig:recid race plot} and Table \ref{tab:recid table}) shown below.
+\ref{fig:table snip}), with the distributions of age (Figure
+\ref{fig:age plot}), prior charges (Figure \ref{fig:prior plot}), and
+two-year recidivism by race (Figure \ref{fig:recidivism mosaic}) and
+(Table \ref{tab:recid table}) shown below.
 
@@
 
-\caption{Distribution of defendant ages by race.
-The purple curve shows the distribution of the ages of Black defendants, and the green curve shows the distribution of the ages of white defendants.
+\caption{The purple curve shows the distribution of the ages of Black defendants, and the green curve shows the distribution of the ages of white defendants.
 The probability of a defendant's age being between two points on the x-axis is the total shaded area of the curve under the two points.
@@
 
-{\centering \includegraphics[width=1\linewidth]{../images/race_recid_bar_plot_new}
+{\centering \includegraphics[width=1\linewidth]{../images/prior_charges_plot}
 
@@
 
-\caption{This plot shows the distribution of two-year recidivism outcomes by race.
-When we divide the data into Black and white defendants, we can see that Black defendants recidivate more than white defendants and Black defendants are more likely to recidivate than not recidivate.
-We can also see that there are more Black defendants in the dataset overall.}\label{fig:recid race plot}
+\caption{Black defendants, particularly men, are more likely to have a greater count of prior charges than white defendants.
+Male defendants have a higher number of prior charges than do female defendants.
+Though we do not know for sure which information goes into the COMPAS algorithm, it is likely that a defendant with prior charges will be coded as a having a higher risk of recidivism.
+Thus, by looking at the racial discrepancies in prior charges we can already see potential bias in the algorithm.}\label{fig:prior plot}
 \end{figure}
+
+\textbackslash{}begin\{figure\}
+
+\{\centering \includegraphics[width=1\linewidth]{../images/recidivism_mosaic}
+
+\}
+
+\textbackslash{}caption\{When we divide the data into Black and white
+defendants, we can see that Black defendants recidivate more than white
+defendants and Black defendants are more likely to recidivate than not
+recidivate. 39.14\% of white defendants did recidivate within two years
+compared to 52.35\% of black defendants. We can also see that there are
+more Black defendants in the dataset
+overall.\}\label{fig:recidivism mosaic} \textbackslash{}end\{figure\}
 
@@
 \toprule
-Two Year Recidivism by Race & Recidivated & Survived & Total \\
+Two Year Recidivism by Race & Recidivated & Survived &
+Total\tabularnewline
 \midrule
@@
 \toprule
-Two Year Recidivism by Race & Recidivated & Survived & Total \\
+Two Year Recidivism by Race & Recidivated & Survived &
+Total\tabularnewline
 \midrule
 \endhead
-African American & 1661 & 1512 & 3173 \\
-Caucasian & 822 & 1278 & 2100 \\
-Total & 2483 & 2790 & 5273 \\
+African American & 1661 & 1512 & 3173\tabularnewline
+Caucasian & 822 & 1278 & 2100\tabularnewline
+Total & 2483 & 2790 & 5273\tabularnewline
 \bottomrule

### 2022-05-02 Paper_Draft2/Draft2.tex
fixing image bug

@@
 
-\textbackslash{}begin\{figure\}
+\begin{figure}
 
-\{\centering \includegraphics[width=1\linewidth]{../images/recidivism_mosaic}
+{\centering \includegraphics[width=1\linewidth]{../images/recidivism_mosaic}
 
-\}
+}
 
-\textbackslash{}caption\{When we divide the data into Black and white
-defendants, we can see that Black defendants recidivate more than white
-defendants and Black defendants are more likely to recidivate than not
-recidivate. 39.14\% of white defendants did recidivate within two years
-compared to 52.35\% of black defendants. We can also see that there are
-more Black defendants in the dataset
-overall.\}\label{fig:recidivism mosaic} \textbackslash{}end\{figure\}
+\caption{When we divide the data into Black and white defendants, we can see that Black defendants recidivate more than white defendants and Black defendants are more likely to recidivate than not recidivate.
+39.14\% of white defendants did recidivate within two years compared to 52.35\% of Black defendants.
+We can also see that there are more Black defendants in the dataset overall.}\label{fig:recidivism mosaic}
+\end{figure}
 

### 2022-05-03 final_metrics_analysis.Rmd
small edits to interpretations

@@
 # for Amrita
-use_condaenv("anaconda3")
+# use_condaenv("anaconda3")
 ```
@@
 
-The highest percentage in this confusion matrix is for the true negative rate, defendants who were predicted to not recidivate and actually did not, at 36.24%, 0.83% higher than the true negatives in the baseline model.
-This means that the re-weighing model marginally improved the baseline model accuracy in predicting people who did not recidivate.
-However, the true positive rate, or those who were predicted to recidivate and did recidivate, is 24.72%, 1.13% lower than that of the baseline model.
-This means that the re-weighing model slightly lowered the accuracy of the baseline model in predicting people who did recidivate.
+The highest percentage in this confusion matrix is for the percentage of true negatives, defendants who were predicted to not recidivate and actually did not, at 36.24%, 0.83% higher than the true negatives in the baseline model.
+This means that the reweighing model marginally improved the baseline model accuracy in predicting people who did not recidivate.
+However, the percentage of true positives, or those who were predicted to recidivate and did recidivate, is 24.72%, 1.13% lower than that of the baseline model.
+This means that the reweighing model slightly lowered the accuracy of the baseline model in predicting people who did recidivate.
 As a result, the overall reweighing model accuracy remains about the same as the baseline model accuracy.
-The false negative and false positive rates, or the predictions that proved to be incorrect, are 23.20% and 15.85%, respectively.
+The false negative and false positive percentages, or the predictions that proved to be incorrect, are 23.20% and 15.85%, respectively.
 The re-weighing model increased the number of false negatives by 1.14% and decreased the number of false positives by 0.83%.
@@
 The number of true positives, defendants who the model predicted to recidivate and actually recidivated, is lower at 23.88%.
-The percentage of false negatives is lower, whereas the true negative is low at 14.33%.
+The percentage of false negatives is lower, whereas the true negative is lowest at 14.33%.
 This indicates that the model does not do a good job in predicting the defendents who don't recidivate.
@@
 The number of true positives, defendants who the model predicted to recidivate and actually recidivated, is much lower at 3.11%.
-The percentage of false negatives is somewhat high for this baseline model, whereas the false positive rate is very low at 2.05%.
+The percentage of false negatives is somewhat high compared to the baseline model, whereas the percentage of false positives is very low at 2.05%.
 This indicates that the model does a good job in predicting the defendents who don't recidivate.

### 2022-05-03 Paper_Draft2/Draft2.Rmd
fixing image bug

@@
 For the purposes of our project, which endeavors to evaluate the differing effects of the COMPAS algorithm on white defendants versus Black defendants, we have filtered the data to only include individuals whose race is listed as Caucasian or African-American.
-Our data therefore has 5,273 rows (Figure \ref{fig:table snip}), with the distributions of age (Figure \ref{fig:age plot}), prior charges (Figure \ref{fig:prior plot}), and two-year recidivism by race (Figure \ref{fig:recidivism mosaic}) and (Table \ref{tab:recid table}) shown below.
+Our data therefore has 5,273 rows (Figure \ref{fig:table snip}), with the distributions of age (Figure \ref{fig:age plot}), prior charges (Figure \ref{fig:prior plot}), and two-year recidivism by race (Figure \ref{fig:recidivism mosaic} and Table \ref{tab:recid table}) shown below.
 
@@
 ```{r recidivism mosaic, echo=FALSE, fig.align = 'center', fig.cap = 'When we divide the data into Black and white defendants, we can see that Black defendants recidivate more than white defendants and Black defendants are more likely to recidivate than not recidivate.
-39.14% of white defendants did recidivate within two years compared to 52.35% of black defendants.
+39.14\\% of white defendants did recidivate within two years compared to 52.35\\% of Black defendants.
 We can also see that there are more Black defendants in the dataset overall.', out.width= "100%"}

### 2022-05-03 Paper_Draft2/Draft2.Rmd
knit

--- 2022-05-02 Paper_Draft2/Draft2.Rmd
@@
 For example, the COMPAS risk assessment survey asks screeners to speculate if a defendant might be affiliated with a gang.
-It also asks if a defendant has any friends or family members who have been crime victims [@Angwin2016Sample].
+It also asks if a defendant has any friends or family members who have been crime victims.
 Although these questions do not directly ask about race, they do not take into account the pervasive nature of systemic racism that infiltrates every aspect of the lives of marginalized people, thereby indirectly asking about race.
@@
 ## Equal Opportunity Difference
-This metric is computed as the difference of true positive rates between the unprivileged and the privileged groups.
-The true positive rate is the ratio of true positives to the total number of actual positives for a given group [@GoogleDev].
+The [equal opportunity difference](https://developers.google.com/machine-learning/glossary/fairness#e) metric is computed as the difference of true positive rates between the unprivileged and the privileged groups.
+The true positive rate is the ratio of true positives to the total number of actual positives for a given group.
 
@@
 In-processing mitigates bias in classifiers while building a model.
-A classifier “is an algorithm that automatically orders or categorizes data into one or more sets” [@baxter2021what].
+A classifier “is an algorithm that automatically orders or categorizes data into one or more sets” [@baxter2021AI].
 The in-processing technique we use is the prejudice remover algorithm, which accounts for the fairness metric as part of the input and returns a classifier optimized by that particular metric.

### 2022-05-03 Paper_Draft2/Draft2.tex
knit

@@
 if a defendant might be affiliated with a gang. It also asks if a
-defendant has any friends or family members who have been crime victims
-\citep{Angwin2016Sample}. Although these questions do not directly ask
-about race, they do not take into account the pervasive nature of
-systemic racism that infiltrates every aspect of the lives of
-marginalized people, thereby indirectly asking about race.
+defendant has any friends or family members who have been crime victims.
+Although these questions do not directly ask about race, they do not
+take into account the pervasive nature of systemic racism that
+infiltrates every aspect of the lives of marginalized people, thereby
+indirectly asking about race.
 
@@
 
-This metric is computed as the difference of true positive rates between
-the unprivileged and the privileged groups. The true positive rate is
-the ratio of true positives to the total number of actual positives for
-a given group \citep{GoogleDev}.
+The
+\href{https://developers.google.com/machine-learning/glossary/fairness\#e}{equal
+opportunity difference} metric is computed as the difference of true
+positive rates between the unprivileged and the privileged groups. The
+true positive rate is the ratio of true positives to the total number of
+actual positives for a given group.
 
@@
 classifier ``is an algorithm that automatically orders or categorizes
-data into one or more sets'' \citep{baxter2021what}. The in-processing
+data into one or more sets'' \citep{baxter2021AI}. The in-processing
 technique we use is the prejudice remover algorithm, which accounts for

### 2022-05-03 Paper_Draft2/Draft2.Rmd
updating draft with all the parts now!

@@
 COMPAS uses various behavioral and psychological metrics related to recidivism to assist justice systems in assessing a defendant’s potential recidivism risk.
-@angwin2016machine published a ProPublica article in which they conclude that the racial biases in the criminal justice system are reflected in the COMPAS recidivism risk scores.
-In response, @equivant_response_2016 published a rebuttal on behalf of Northpointe defending the COMPAS algorithm and refuting @angwin2016machine’s allegation of racial bias.
+Angwin et al.
+published a ProPublica article in which they conclude that the racial biases in the criminal justice system are reflected in the COMPAS recidivism risk scores.
+In response, Dieterich et al.
+published a rebuttal on behalf of Northpointe defending the COMPAS algorithm and refuting Angwin et al.’s allegation of racial bias.
 Using a human rights framework adopted from the organizations Women at the Table and AI Fairness 360, we use debiasing algorithms and fairness metrics to analyze the argument between Northpointe and ProPublica and determine whether and to what extent there is racial bias in the COMPAS algorithm.
 All four group fairness metrics determine that the COMPAS algorithm favors white defendants over Black defendants.
+Our research found that the pre and post-processing bias mitigation algorithms, specifically reweighing and calibrated equalized odds, are the most effective at improving fairness.
 
@@
 For example, the COMPAS risk assessment survey asks screeners to speculate if a defendant might be affiliated with a gang.
-It also asks if a defendant has any friends or family members who have been crime victims.
+It also asks if a defendant has any friends or family members who have been crime victims [@Angwin2016Sample].
 Although these questions do not directly ask about race, they do not take into account the pervasive nature of systemic racism that infiltrates every aspect of the lives of marginalized people, thereby indirectly asking about race.
@@
 They find that “the algorithm [is] somewhat more accurate than a coin flip,” a worrisome level of accuracy given the potential impact its determinations may have on real people’s lives.
-@angwin2016machine specifically investigate the distribution of COMPAS scores by decile among Black and white defendants.
-They write: “The analysis also [shows] that even when controlling for prior crimes, future recidivism, age, and gender, Black defendants [are] 45 percent more likely to be assigned higher risk scores than white defendants” [@larson2016we].
-After examining the fairness metric statistical parity difference, @angwin2016machine conclude that the algorithm is racially biased [@larson2016we].
+Angwin et al.
+specifically investigate the distribution of COMPAS scores by decile among Black and white defendants.
+They write: “The analysis also [shows] that even when controlling for prior crimes, future recidivism, age, and gender, black defendants [are] 45 percent more likely to be assigned higher risk scores than white defendants” [@larson2016we].
+After examining the fairness metric statistical parity difference, Angwin et al.
+conclude that the algorithm is racially biased [@larson2016we].
 
@@
 
-[Women at the Table](https://www.womenatthetable.net/), the sponsor organization for this project, is “a growing, global gender equality & democracy CSO based in Geneva, Switzerland focused on advancing feminist systems change by using the prism of technology, innovation & AI exercising leverage points in technology, the economy, sustainability & democratic governance.” We are collaborating with the organization on its AI & Equality [@noauthor_ai_nodate] initiative, tasked with de-biasing the COMPAS algorithm [@aif360-oct-2018] and producing a corresponding data story that will be added to its library.
+[Women at the Table](https://www.womenatthetable.net/), the sponsor organization for this project, is “a growing, global gender equality & democracy CSO based in Geneva, Switzerland focused on advancing feminist systems change by using the prism of technology, innovation & AI exercising leverage points in technology, the economy, sustainability & democratic governance.” We are collaborating with the organization on its AI & Equality [@noauthor_ai_nodate] initiative, tasked with debiasing the COMPAS algorithm [@aif360-oct-2018] and producing a corresponding data story that will be added to its library.
 
-Our project builds on Women at the Table’s various de-biasing algorithms used in its AI & Equality Human Rights Toolbox to conduct our own analyses on the COMPAS data set.
+Our project builds on Women at the Table’s various debiasing algorithms used in its AI & Equality Human Rights Toolbox to conduct our own analyses on the COMPAS data set.
 Based on this analysis, we employ a human rights framework to contribute to the ProPublica and Northpointe debate and investigate whether or to what extent there is racial bias in the COMPAS algorithm.
 With a solid understanding of the two sides, we aim to pinpoint the shortcomings of both arguments and correct them in our analyses.
-We will use various de-biasing techniques and fairness metrics to evaluate the level of bias present in the COMPAS data and our algorithm.
+We will use various debiasing techniques and fairness metrics to evaluate the level of bias present in the COMPAS data and our algorithm.
 We will summarize our results using the JupyterNotebook framework from Women at the Table, to be used by members of the organization to teach in a workshop setting.
@@
 
-The data we are using for this project is the COMPAS General Recidivism Risk Scores data set from the AI Fairness 360 (AIF360) toolkit [@aif360-oct-2018], which does the same initial pre-processing as ProPublica.
-The raw data has 6,167 rows and each row represents an arrest charge for a defendant.
-AIF360’s COMPAS data includes the defendant's age, race, sex, prior charges, what they were charged with, and whether or not the defendant ultimately recidivated within a two-year period after their arrest.
-For the purposes of our project, which endeavors to evaluate the differing effects of the COMPAS algorithm on white defendants versus Black defendants, we have filtered the data to only include individuals whose race is listed as Caucasian or African-American.
-Our data therefore has 5,273 rows (Figure \ref{fig:table snip}), with the distributions of age (Figure \ref{fig:age plot}), prior charges (Figure \ref{fig:prior plot}), and two-year recidivism by race (Figure \ref{fig:recidivism mosaic} and Table \ref{tab:recid table}) shown below.
+The data we are using for this library addition is the COMPAS General Recidivism Risk Scores data set from the [AI Fairness 360](https://aif360.readthedocs.io/en/stable/modules/generated/aif360.datasets.CompasDataset.html#aif360.datasets.CompasDataset) (AIF360) toolkit.
+The AIF360 toolkit builds on the dataset released by ProPublica collaborators Larson et al.
+created to examine the racial bias and the true outcomes of the recidivism risk scores in the COMPAS algorithm for the initial ["Machine Bias"](https://www.propublica.org/article/machine-bias-risk-assessments-in-criminal-sentencing) article.
+For this, Larson et al.
+obtained two years worth of COMPAS scores from the Broward County Sheriff’s Office in Florida, as well as the corresponding intake information for each defendant including but not limited to name, sex, race, age, and charge degree and description.
+They also obtained data about whether defendants actually recidivated or not in the two year period following their initial COMPAS score assessment.
+AIF360 then processed the data with the same procedures that Larson et al.
+followed for their analysis.
+
+The raw data that we use from the AIF360 has 6,167 rows, where each row represents an arrest charge for a defendant.
+AIF360’s COMPAS data includes the defendant's age, race, sex, what they were charged with, and whether or not the defendant ultimately recidivated within a two-year period after their arrest.
+
+This exploration aims to evaluate anti-Black algorithmic bias and the differing effects of the COMPAS algorithm between white and Black defendants; as such, we filter the data to only include individuals whose race is listed as Caucasian or African-American.
+Our data therefore has 5,723 rows with information on defendant race, age, gender, prior crimes, and two-year recidivism rate.
+The average African-American defendant in our dataset is 29 years old, male, and has committed two prior crimes.
+The average Caucasian defendant in our dataset is 35 years old, male, and has committed one prior crime.
+The median defendant for both races do not have any juvenile convictions.
 
@@
 The AIF360 toolkit contains various group and individual fairness metrics as well as pre-processing, in-processing, and post-processing algorithms that we used to debias the COMPAS algorithm [@aif360-oct-2018].
+Fairness metrics are mathematical measures of whether an algorithm treats members of different groups (such as racial or gender groups) equally.
+An algorithm has no understanding of the historical oppression of certain groups and how such bias is baked into the data on which the algorithm is trained.
+Thus, fairness metrics provide a method of evaluating an algorithm’s level of bias towards or against the unprivileged group versus the privileged group.
 We researched the definitions and applications of different fairness metrics [@ashokan2021fairness] to determine which metric would be most appropriate for our project.
-We chose to look at four group fairness metrics instead of individual fairness metrics.
+
+Fairness is subjective; one person might consider an algorithm fair if groups are given the same treatment, while someone else might only consider the algorithm fair if groups as a whole receive the same outcomes.
+In our research we choose to use the definition of group fairness as our definition for assessing fairness in the processing approach.
 Group fairness metrics take into account the attributes of a whole group as opposed to just one individual in the group, allowing us to represent systemic issues.
@@
 Group and individual metrics work in opposition of one another, meaning that when group fairness improves, individual fairness gets worse [@kypraiou_what_2021].
+We chose the following four group fairness metrics to evaluate our models.
 
@@
 
-For the next step of our experiment, we need to determine where in the data science pipeline we can mitigate the most bias, using pre-processing, in-processing, and post-processing de-biasing algorithms.
+For the next step of our experiment, we sought to determine where in the data science pipeline we can mitigate the most bias, using pre-processing, in-processing, and post-processing debiasing algorithms.
 These are all based on using predictive models to figure out how we can “fix” the bias that is present.
+
+## Pre-Processing
 
@@
 
+## In-Processing
+
 In-processing mitigates bias in classifiers while building a model.
@@
 In order to do this, we first needed to convert our data frame into a data type called a BinaryLabelDataset.
-Similarly to pre-processing, we compared the results of our in-processing methods with both the baseline and the pre-processing to gauge which method so far has better individual or group fairness.
 
-The prejudice remover is a method for reducing indirect prejudice (i.e.
-how COMPAS is racially biased because it uses proxy variables for race).
-The prejudice remover implements two different regularizers, one to avoid overfitting and one to enforce fair classification.
-The prejudice remover regularizer works by minimizing the prejudice index, a mathematical equation for quantifying fairness defined by Kamishima et.
-al.
+The prejudice remover is a method for reducing indirect prejudice (i.e., how COMPAS is racially biased because it uses proxy variables for race).
+The prejudice remover implements two different regularizers, one to avoid overfitting and one to enforce fair classification [@kamishima2012fairness].
+The prejudice remover regularizer works by minimizing the prejudice index, a mathematical equation for quantifying fairness defined by @kamishima2012fairness.
 This in turn enforces a classifier’s independence from sensitive information (e.g., race).
+Similarly to pre-processing, we compared the results of our in-processing methods with both the baseline and the pre-processing model to gauge which method so far has the better group fairness.
 
-Our last approach, post-processing bias mitigation, is used after training a model.
-Post-processing algorithms equalize the outcomes (i.e., predicted classification values) to mitigate bias instead of adjusting the classifier or the training data [@baxter2021AI].
-We use calibrated equalized odds, which “optimizes over calibrated classifier score outputs to find probabilities with which to change output labels with an equalized odds objective” [insert aif360 website citation here].
-An equalized odds objective constrains classification algorithms such that no error type (false-positive or false-negative) disproportionately affects any population subgroup; both groups, in our case both white and Black defendants, should have the same false-positive and false-negative rates.
+## Post-Processing
+
+Our last approach, post-processing bias mitigation, is implemented after training a model.
+Post-processing algorithms equalize the outcomes (i.e., predicted recidivism values) to mitigate bias instead of adjusting the classifier or the training data [@baxter2021AI].
+We use calibrated equalized odds, which “optimizes over calibrated classifier score outputs to find probabilities with which to change output labels with an equalized odds objective” [@aif360-oct-2018].
+An equalized odds objective constrains classification algorithms such that no error type (false-positive or false-negative) disproportionately affects any population subgroup; both groups, in our case both white and Black defendants, should have the same false-positive and false-negative rates [@pleiss2017fairness].
 Through the calibrated equalized odds method, we want to decrease bias while also maintaining calibration [@pleiss2017fairness].
@@
 The algorithm unfairly benefits white defendants over Black defendants.
+
 Disparate impact ratio is 0.47.
-The ratio of white defendants predicted to not recidivate to the Black defendants predicted to not recidivate is 0.47.
+The ratio of white defendants predicted to not recidivate to Black defendants predicted to not recidivate is 0.47.
 A ratio between 0.8 and 1.25 is considered fair, therefore the algorithm unfairly benefits white defendants.
@@
 The difference of true positive rates between the Black and white groups is -0.41.
-A value less than 0 indicates a benefit to the privileged group, so the algorithm unfairly benefits white defendants.
-The value is substantially less than -0.1, which indicates that the algorithm benefits white defendants.
+A value less than 0 indicates a benefit to the privileged group, so the algorithm benefits white defendants.
+The value is substantially less than -0.1, which indicates that the algorithm is unfair.
 
 All four group fairness metrics determine that the COMPAS algorithm favors white defendants over Black defendants.
-Although the magnitudes of the various fairness metrics are different, all of the fairness metrics are not within their respective fairness thresholds.
+Although the magnitudes of the various fairness metrics are different, none of the metrics are within their respective fairness thresholds.
 Our goal is to use pre-processing, in-processing, and post-processing algorithms in the AIF360 toolkit to see if we can make COMPAS fair at all.
+
+Our confusion matrix for the baseline model indicates that the false positive and false negative percentages are 16.68% and 22.06%, which are the lowest values for this matrix.
+The highest value is the value of true positives, at 35.41%, showing that of the total number of predictions, 35.41% were correct predictions of recidivism.
 
@@
 
-## Pre-Processing Approach
+## Pre-Processing Approach: Reweighing
 
 After running the reweighing algorithm, our fairness metrics are -0.015 for statistical parity difference, 0.015 for equal opportunity difference, 0.014 for average odds difference, and 0.98 for disparate impact ratio.
-Overall, these fairness metrics show that the pre-processing algorithm reweighing improves the bias in the COMPAS algorithm.
+All of these values now fall within the margins of fairness (see reweighing metrics plot).
+These values make sense because the reweighing algorithm chooses different weights depending on whether an attribute is protected.
+This does an effective job of eliminating the bias, shown by each of the values in the graph being within the “fair” range of the particular fairness metric.
+Overall, these fairness metrics show that the reweighing algorithm improves the bias in the COMPAS algorithm.
+
+The highest percentage in this confusion matrix is for the percentage of true negatives, defendants who were predicted to not recidivate and actually did not, at 36.24%, 0.83% higher than the true negatives in the baseline model.
+This means that the reweighing model marginally improved the baseline model accuracy in predicting people who did not recidivate.
+However, the percentage of true positives, or those who were predicted to recidivate and did recidivate, is 24.72%, 1.13% lower than that of the baseline model.
+This means that the reweighing model slightly lowered the accuracy of the baseline model in predicting people who did recidivate.
+As a result, the overall reweighing model accuracy remains about the same as the baseline model accuracy.
+The false negative and false positive percentages, or the predictions that proved to be incorrect, are 23.20% and 15.85%, respectively.
+The re-weighing model increased the number of false negatives by 1.14% and decreased the number of false positives by 0.83%.
+We are most concerned about the false positives, which indicate defendants who are predicted to recidivate and do not actually recidivate.
+While there is a slight improvement in the percent of false positives, it is still relatively high at 15.85%, indicating that we should further precede with additional bias mitigation techniques.
+
+## In-Processing Approach: Prejudice Remover
+
+Our fairness metrics after running the prejudice remover are: 0.367 for statistical parity difference, 0.321 for equal opportunity difference, 0.344 for average odds difference, and 3.3 for disparate impact ratio.
+Like the model performance metrics suggested, the prejudice remover approach resulted in an increased benefit for Black defendants.
+As the orange bars on these plots show, the values of the fairness metrics have reversed from their baseline values.
+Now all four metrics suggest an unfair advantage for Black defendants.
+Thus, this approach removed the model's prejudice against Black people, but it did not result in a "fair" model.
+
+The highest percentage in this confusion matrix is for false positives, defendants who the model predicted to recidivate and actually did not, at 37.76%.
+This false positive percentage concerns us because we do not want defendants who do not recidivate to have unfairly long sentences due to their (incorrectly) predicted recidivism.
+The number of true positives, defendants who the model predicted to recidivate and actually recidivated, is lower at 23.88%.
+The percentage of false negatives is lower, whereas the true negative is low at 14.33%.
+This indicates that the model does not do a good job in predicting the defendants who don't recidivate.
+It is very likely that it will incorrectly predict that someone will recidivate.
+
+After the prejudice remover, the accuracy of the model is fairly similar across the board of races, but the model's accuracy is only 38.21%, which is much lower than the baseline and reweighing models.
+This accuracy score means that the model makes accurate predictions only 38% of the time.
+
+## Post-Processing Approach: Calibrated Equalized Odds
+
+After running the calibrated equalized odds algorithm, our fairness metrics are: 0.134 for statistical parity difference, 0.088 for equal opportunity difference, 0.146 for average odds difference, and 1.155 for disparate impact ratio.
+As we can see from the orange bars, the calibrated equalized odds approach results in all four metrics now suggesting a benefit for the originally unprivileged group, Black defendants.
+Though the values for statistical parity difference and average odds difference are slightly above the range of what is considered fair, the margin is much smaller than the original margin between the value and the range of fairness (as indicated by the blue bars).
+Thus, the calibrated equalized odds approach successfully counteracts the bias against Black defendants and results in a mostly fair model.
+
+The highest percentage in this confusion matrix is for true negatives, defendants who the model predicted to not recidivate and actually did not, at 50.04%.
+The number of true positives, defendants who the model predicted to recidivate and actually recidivated, is much lower at 3.11%.
+The percentage of false negatives is somewhat high for this baseline model, whereas the amount of false positives is very low at 2.05%.
+This indicates that the model does a good job in predicting the defendants who don't recidivate.
+It is very unlikely that it will incorrectly predict that someone will recidivate.
+Though the accuracy score for the model is 0.53, lower than the baseline, this model is much better at identifying true negatives.
+
+# Conclusion
+
+## Limitations
+
+Our project tries to mitigate bias within existing data, rather than within the methods used to collect this data.
+The data are collected using a risk assessment survey, where defendants are asked a series of questions (see introduction) which are supposedly used to determine whether someone will recidivate or not.
+Many of these questions involve proxy variables for race.
+The creation and facilitation of this survey can therefore be assessed for bias mitigation as the methods used may potentially privilege some groups over others.
+As a result, the bias exists even before values are collected, limiting the debiasing work able to be done on the output values.
+
+Furthermore, AIF360 processed the data we used in this project the same way ProPublica processed their data.
+AIF360 simplifies the original defendant data and their COMPAS scores to make the data easier to analyze.
+However, this initial processing may lose some details captured in the original raw data.
+As we have noted throughout this paper, bias can be introduced in any step of the data science pipeline and AIF360’s initial data processing step is no exception.
+
+AIF360 contains an array of processing methods that have been written by various scholars of fairness [@aif360-oct-2018].
+These methods are each distinct and apply a different technique to mitigate bias.
+Our results only contain three processing methods, one for each step in processing.
+This choice was based on time constraints as well as our working environment.
+For example, the code for most of the other in-processing methods would not run for us.
+In-processing on its own works to debias an algorithm, not the input or output values, so some of those methods may not work on certain algorithms.
+We had similar challenges with pre and post-processing, where the methods were challenging to run and would require more time to be spent debugging in order to make them work.
+Had we used alternative processing methods, our results would be quite different, as the method changes apply a different mathematical technique to the input.
+
+## Future Work
+
+In this project, we have only focused on a few select fairness metrics and a few debiasing algorithms.
+We looked at group fairness metrics because we specifically wanted to mitigate systemic racial bias in the COMPAS algorithm, but other researchers could compare how well debiasing algorithms work using individual fairness metrics.
+Future work could also include conducting more tests using other debiasing algorithms to see if the COMPAS algorithm could be made more fair.
+One significant limitation of our work is that some of the in-processing algorithms provided in the AIF360 toolkit were difficult to integrate into our code and we were unable to run many of them.
+The processing methods we chose, though effective, were based on our work environments, and there are other existing algorithms that attempt to mitigate bias.
+
+We only looked at defendants labeled as Caucasian and African-American in our dataset, which diminishes the generalizability and nuance of our results.
+Comparing how COMPAS assesses defendants of other races would be an incredibly relevant extension of our existing work.
+Furthermore, @larson2016we also found that age was the most predictive factor of a higher risk score: “Defendants younger than 25 years old were 2.5 times as likely to get a higher score than middle aged offenders, even when controlling for prior crimes, future criminality, race and gender” [@larson2016we].
+In addition, female defendants were more likely to receive a higher recidivism risk score than male defendants when controlling for the aforementioned factors [@larson2016we].
+Thus, some extensions of our work could include attempting to remove age and gender bias and determining the most effective processing algorithms to do so.
+
+The dataset we are working with only contains defendants’ criminal history records from Broward County, Florida.
+Many other states, including New York and Massachusetts, use COMPAS to calculate recidivism scores, and a further study could apply this paper’s methods to the data coming out of other states.
+By conducting experiments on multiple states the baseline model would be trained on a larger and more diverse set of data, which could improve its accuracy.
+In addition, by adding more geographically diverse data to our set we can start to create more generalizable conclusions about our results.
+However, future researchers must keep in mind the real-life consequences of recidivism risk algorithms like COMPAS and how research on methods of debiasing algorithms could potentially be utilized to justify the continued use of biased algorithms.
+
+## Final Thoughts
+
+Through our methods, we concluded that the pre-processing and post-processing methods mitigate bias and increase fairness most successfully.
+The in-processing technique was effective at mitigating bias, but did not result in a mathematically fair model.
+However, these techniques only treat the symptoms of racial bias in the justice system, as opposed to addressing the root cause, systemic racism.
+We need to question why these flawed algorithms exist and why they can be used to determine someone’s fate in the justice system.
+While we may have methods for reducing bias algorithms such as COMPAS, the focus should be on addressing the over-policing and over-criminalization of Black communities that result in biased data and algorithms.
+
+## Ethical Statement
+
+Throughout this project, we endeavored to debias an incredibly powerful algorithm that can change the course of someone’s life.
+However, we want to recognize that debiasing methods are not the most effective or beneficial way to uproot bias within the American justice system.
+The problems within the justice system are much more complex than an algorithm and are rooted in the United States’s history of racism.
+In addition, algorithmic bias is much more complex than the algorithm itself; algorithmic bias comes from the people who produce the algorithm.
+We all have our own beliefs, biases, and situated knowledge that impact and subsequently limit everything we create.
+
+# Acknowledgements
+
+We would like to thank our sponsor organization Women At The Table and Professor Ben Baumer for introducing us to this incredibly meaningful work, supporting us, and answering the many questions we had throughout this semester-long project.
+We would specifically like to thank Sofia Kypraiou for her mentorship and patience throughout this process.
 

### 2022-05-03 Paper_Draft2/Draft2.tex
updating draft with all the parts now!

@@
 related to recidivism to assist justice systems in assessing a
-defendant's potential recidivism risk. \citet{angwin2016machine}
-published a ProPublica article in which they conclude that the racial
-biases in the criminal justice system are reflected in the COMPAS
-recidivism risk scores. In response, \citet{equivant_response_2016}
-published a rebuttal on behalf of Northpointe defending the COMPAS
-algorithm and refuting \citet{angwin2016machine}'s allegation of racial
-bias. Using a human rights framework adopted from the organizations
-Women at the Table and AI Fairness 360, we use debiasing algorithms and
-fairness metrics to analyze the argument between Northpointe and
-ProPublica and determine whether and to what extent there is racial bias
-in the COMPAS algorithm. All four group fairness metrics determine that
-the COMPAS algorithm favors white defendants over Black defendants.}
+defendant's potential recidivism risk. Angwin et al.~published a
+ProPublica article in which they conclude that the racial biases in the
+criminal justice system are reflected in the COMPAS recidivism risk
+scores. In response, Dieterich et al.~published a rebuttal on behalf of
+Northpointe defending the COMPAS algorithm and refuting Angwin et al.'s
+allegation of racial bias. Using a human rights framework adopted from
+the organizations Women at the Table and AI Fairness 360, we use
+debiasing algorithms and fairness metrics to analyze the argument
+between Northpointe and ProPublica and determine whether and to what
+extent there is racial bias in the COMPAS algorithm. All four group
+fairness metrics determine that the COMPAS algorithm favors white
+defendants over Black defendants. Our research found that the pre and
+post-processing bias mitigation algorithms, specifically reweighing and
+calibrated equalized odds, are the most effective at improving
+fairness.}
 
@@
 if a defendant might be affiliated with a gang. It also asks if a
-defendant has any friends or family members who have been crime victims.
-Although these questions do not directly ask about race, they do not
-take into account the pervasive nature of systemic racism that
-infiltrates every aspect of the lives of marginalized people, thereby
-indirectly asking about race.
+defendant has any friends or family members who have been crime victims
+\citep{Angwin2016Sample}. Although these questions do not directly ask
+about race, they do not take into account the pervasive nature of
+systemic racism that infiltrates every aspect of the lives of
+marginalized people, thereby indirectly asking about race.
 
@@
 a worrisome level of accuracy given the potential impact its
-determinations may have on real people's lives.
-\citet{angwin2016machine} specifically investigate the distribution of
-COMPAS scores by decile among Black and white defendants. They write:
-``The analysis also {[}shows{]} that even when controlling for prior
-crimes, future recidivism, age, and gender, Black defendants {[}are{]}
-45 percent more likely to be assigned higher risk scores than white
-defendants'' \citep{larson2016we}. After examining the fairness metric
-statistical parity difference, \citet{angwin2016machine} conclude that
-the algorithm is racially biased \citep{larson2016we}.
+determinations may have on real people's lives. Angwin et
+al.~specifically investigate the distribution of COMPAS scores by decile
+among Black and white defendants. They write: ``The analysis also
+{[}shows{]} that even when controlling for prior crimes, future
+recidivism, age, and gender, black defendants {[}are{]} 45 percent more
+likely to be assigned higher risk scores than white defendants''
+\citep{larson2016we}. After examining the fairness metric statistical
+parity difference, Angwin et al.~conclude that the algorithm is racially
+biased \citep{larson2016we}.
 
@@
 its AI \& Equality \citep{noauthor_ai_nodate} initiative, tasked with
-de-biasing the COMPAS algorithm \citep{aif360-oct-2018} and producing a
+debiasing the COMPAS algorithm \citep{aif360-oct-2018} and producing a
 corresponding data story that will be added to its library.
 
-Our project builds on Women at the Table's various de-biasing algorithms
+Our project builds on Women at the Table's various debiasing algorithms
 used in its AI \& Equality Human Rights Toolbox to conduct our own
@@
 aim to pinpoint the shortcomings of both arguments and correct them in
-our analyses. We will use various de-biasing techniques and fairness
+our analyses. We will use various debiasing techniques and fairness
 metrics to evaluate the level of bias present in the COMPAS data and our
@@
 
-The data we are using for this project is the COMPAS General Recidivism
-Risk Scores data set from the AI Fairness 360 (AIF360) toolkit
-\citep{aif360-oct-2018}, which does the same initial pre-processing as
-ProPublica. The raw data has 6,167 rows and each row represents an
-arrest charge for a defendant. AIF360's COMPAS data includes the
-defendant's age, race, sex, prior charges, what they were charged with,
+The data we are using for this library addition is the COMPAS General
+Recidivism Risk Scores data set from the
+\href{https://aif360.readthedocs.io/en/stable/modules/generated/aif360.datasets.CompasDataset.html\#aif360.datasets.CompasDataset}{AI
+Fairness 360} (AIF360) toolkit. The AIF360 toolkit builds on the dataset
+released by ProPublica collaborators Larson et al.~created to examine
+the racial bias and the true outcomes of the recidivism risk scores in
+the COMPAS algorithm for the initial
+\href{https://www.propublica.org/article/machine-bias-risk-assessments-in-criminal-sentencing}{``Machine
+Bias''} article. For this, Larson et al.~obtained two years worth of
+COMPAS scores from the Broward County Sheriff's Office in Florida, as
+well as the corresponding intake information for each defendant
+including but not limited to name, sex, race, age, and charge degree and
+description. They also obtained data about whether defendants actually
+recidivated or not in the two year period following their initial COMPAS
+score assessment. AIF360 then processed the data with the same
+procedures that Larson et al.~followed for their analysis.
+
+The raw data that we use from the AIF360 has 6,167 rows, where each row
+represents an arrest charge for a defendant. AIF360's COMPAS data
+includes the defendant's age, race, sex, what they were charged with,
 and whether or not the defendant ultimately recidivated within a
-two-year period after their arrest. For the purposes of our project,
-which endeavors to evaluate the differing effects of the COMPAS
-algorithm on white defendants versus Black defendants, we have filtered
-the data to only include individuals whose race is listed as Caucasian
-or African-American. Our data therefore has 5,273 rows (Figure
-\ref{fig:table snip}), with the distributions of age (Figure
-\ref{fig:age plot}), prior charges (Figure \ref{fig:prior plot}), and
-two-year recidivism by race (Figure \ref{fig:recidivism mosaic}) and
-(Table \ref{tab:recid table}) shown below.
+two-year period after their arrest.
+
+This exploration aims to evaluate anti-Black algorithmic bias and the
+differing effects of the COMPAS algorithm between white and Black
+defendants; as such, we filter the data to only include individuals
+whose race is listed as Caucasian or African-American. Our data
+therefore has 5,723 rows with information on defendant race, age,
+gender, prior crimes, and two-year recidivism rate. The average
+African-American defendant in our dataset is 29 years old, male, and has
+committed two prior crimes. The average Caucasian defendant in our
+dataset is 35 years old, male, and has committed one prior crime. The
+median defendant for both races do not have any juvenile convictions.
 
@@
 algorithms that we used to debias the COMPAS algorithm
-\citep{aif360-oct-2018}. We researched the definitions and applications
-of different fairness metrics \citep{ashokan2021fairness} to determine
-which metric would be most appropriate for our project. We chose to look
-at four group fairness metrics instead of individual fairness metrics.
-Group fairness metrics take into account the attributes of a whole group
-as opposed to just one individual in the group, allowing us to represent
-systemic issues. In general, group fairness metrics require that the
-unprivileged group is treated similarly to the privileged group, whereas
-individual fairness metrics require individuals to be treated
-consistently \citep{kypraiou_what_2021}. Group and individual metrics
-work in opposition of one another, meaning that when group fairness
-improves, individual fairness gets worse \citep{kypraiou_what_2021}.
+\citep{aif360-oct-2018}. Fairness metrics are mathematical measures of
+whether an algorithm treats members of different groups (such as racial
+or gender groups) equally. An algorithm has no understanding of the
+historical oppression of certain groups and how such bias is baked into
+the data on which the algorithm is trained. Thus, fairness metrics
+provide a method of evaluating an algorithm's level of bias towards or
+against the unprivileged group versus the privileged group. We
+researched the definitions and applications of different fairness
+metrics \citep{ashokan2021fairness} to determine which metric would be
+most appropriate for our project.
+
+Fairness is subjective; one person might consider an algorithm fair if
+groups are given the same treatment, while someone else might only
+consider the algorithm fair if groups as a whole receive the same
+outcomes. In our research we choose to use the definition of group
+fairness as our definition for assessing fairness in the processing
+approach. Group fairness metrics take into account the attributes of a
+whole group as opposed to just one individual in the group, allowing us
+to represent systemic issues. In general, group fairness metrics require
+that the unprivileged group is treated similarly to the privileged
+group, whereas individual fairness metrics require individuals to be
+treated consistently \citep{kypraiou_what_2021}. Group and individual
+metrics work in opposition of one another, meaning that when group
+fairness improves, individual fairness gets worse
+\citep{kypraiou_what_2021}. We chose the following four group fairness
+metrics to evaluate our models.
 
@@
 
-For the next step of our experiment, we need to determine where in the
+For the next step of our experiment, we sought to determine where in the
 data science pipeline we can mitigate the most bias, using
-pre-processing, in-processing, and post-processing de-biasing
-algorithms. These are all based on using predictive models to figure out
-how we can ``fix'' the bias that is present.
+pre-processing, in-processing, and post-processing debiasing algorithms.
+These are all based on using predictive models to figure out how we can
+``fix'' the bias that is present.
+
+\hypertarget{pre-processing}{%
+\subsection{Pre-Processing}\label{pre-processing}}
 
@@
 
+\hypertarget{in-processing}{%
+\subsection{In-Processing}\label{in-processing}}
+
 In-processing mitigates bias in classifiers while building a model. A
@@
 needed to convert our data frame into a data type called a
-BinaryLabelDataset. Similarly to pre-processing, we compared the results
-of our in-processing methods with both the baseline and the
-pre-processing to gauge which method so far has better individual or
-group fairness.
+BinaryLabelDataset.
 
-The prejudice remover is a method for reducing indirect prejudice
-(i.e.~how COMPAS is racially biased because it uses proxy variables for
-race). The prejudice remover implements two different regularizers, one
-to avoid overfitting and one to enforce fair classification. The
-prejudice remover regularizer works by minimizing the prejudice index, a
-mathematical equation for quantifying fairness defined by Kamishima et.
-al.~This in turn enforces a classifier's independence from sensitive
-information (e.g., race).
+The prejudice remover is a method for reducing indirect prejudice (i.e.,
+how COMPAS is racially biased because it uses proxy variables for race).
+The prejudice remover implements two different regularizers, one to
+avoid overfitting and one to enforce fair classification
+\citep{kamishima2012fairness}. The prejudice remover regularizer works
+by minimizing the prejudice index, a mathematical equation for
+quantifying fairness defined by \citet{kamishima2012fairness}. This in
+turn enforces a classifier's independence from sensitive information
+(e.g., race). Similarly to pre-processing, we compared the results of
+our in-processing methods with both the baseline and the pre-processing
+model to gauge which method so far has the better group fairness.
 
-Our last approach, post-processing bias mitigation, is used after
+\hypertarget{post-processing}{%
+\subsection{Post-Processing}\label{post-processing}}
+
+Our last approach, post-processing bias mitigation, is implemented after
 training a model. Post-processing algorithms equalize the outcomes
-(i.e., predicted classification values) to mitigate bias instead of
+(i.e., predicted recidivism values) to mitigate bias instead of
 adjusting the classifier or the training data \citep{baxter2021AI}. We
@@
 classifier score outputs to find probabilities with which to change
-output labels with an equalized odds objective'' {[}insert aif360
-website citation here{]}. An equalized odds objective constrains
+output labels with an equalized odds objective''
+\citep{aif360-oct-2018}. An equalized odds objective constrains
 classification algorithms such that no error type (false-positive or
@@
 groups, in our case both white and Black defendants, should have the
-same false-positive and false-negative rates. Through the calibrated
-equalized odds method, we want to decrease bias while also maintaining
-calibration \citep{pleiss2017fairness}.
+same false-positive and false-negative rates \citep{pleiss2017fairness}.
+Through the calibrated equalized odds method, we want to decrease bias
+while also maintaining calibration \citep{pleiss2017fairness}.
 \href{https://medium.com/analytics-vidhya/calibration-in-machine-learning-e7972ac93555}{Calibration}
@@
 or not they recidivate. The algorithm unfairly benefits white defendants
-over Black defendants. Disparate impact ratio is 0.47. The ratio of
-white defendants predicted to not recidivate to the Black defendants
-predicted to not recidivate is 0.47. A ratio between 0.8 and 1.25 is
-considered fair, therefore the algorithm unfairly benefits white
-defendants.
+over Black defendants.
+
+Disparate impact ratio is 0.47. The ratio of white defendants predicted
+to not recidivate to Black defendants predicted to not recidivate is
+0.47. A ratio between 0.8 and 1.25 is considered fair, therefore the
+algorithm unfairly benefits white defendants.
 
@@
 rates between the Black and white groups is -0.41. A value less than 0
-indicates a benefit to the privileged group, so the algorithm unfairly
-benefits white defendants. The value is substantially less than -0.1,
-which indicates that the algorithm benefits white defendants.
+indicates a benefit to the privileged group, so the algorithm benefits
+white defendants. The value is substantially less than -0.1, which
+indicates that the algorithm is unfair.
 
@@
 favors white defendants over Black defendants. Although the magnitudes
-of the various fairness metrics are different, all of the fairness
-metrics are not within their respective fairness thresholds. Our goal is
-to use pre-processing, in-processing, and post-processing algorithms in
-the AIF360 toolkit to see if we can make COMPAS fair at all.
+of the various fairness metrics are different, none of the metrics are
+within their respective fairness thresholds. Our goal is to use
+pre-processing, in-processing, and post-processing algorithms in the
+AIF360 toolkit to see if we can make COMPAS fair at all.
+
+Our confusion matrix for the baseline model indicates that the false
+positive and false negative percentages are 16.68\% and 22.06\%, which
+are the lowest values for this matrix. The highest value is the value of
+true positives, at 35.41\%, showing that of the total number of
+predictions, 35.41\% were correct predictions of recidivism.
 
@@
 
-\hypertarget{pre-processing-approach}{%
-\subsection{Pre-Processing Approach}\label{pre-processing-approach}}
+\hypertarget{pre-processing-approach-reweighing}{%
+\subsection{Pre-Processing Approach:
+Reweighing}\label{pre-processing-approach-reweighing}}
 
@@
 difference, 0.014 for average odds difference, and 0.98 for disparate
-impact ratio. Overall, these fairness metrics show that the
-pre-processing algorithm reweighing improves the bias in the COMPAS
+impact ratio. All of these values now fall within the margins of
+fairness (see reweighing metrics plot). These values make sense because
+the reweighing algorithm chooses different weights depending on whether
+an attribute is protected. This does an effective job of eliminating the
+bias, shown by each of the values in the graph being within the ``fair''
+range of the particular fairness metric. Overall, these fairness metrics
+show that the reweighing algorithm improves the bias in the COMPAS
 algorithm.
+
+The highest percentage in this confusion matrix is for the percentage of
+true negatives, defendants who were predicted to not recidivate and
+actually did not, at 36.24\%, 0.83\% higher than the true negatives in
+the baseline model. This means that the reweighing model marginally
+improved the baseline model accuracy in predicting people who did not
+recidivate. However, the percentage of true positives, or those who were
+predicted to recidivate and did recidivate, is 24.72\%, 1.13\% lower
+than that of the baseline model. This means that the reweighing model
+slightly lowered the accuracy of the baseline model in predicting people
+who did recidivate. As a result, the overall reweighing model accuracy
+remains about the same as the baseline model accuracy. The false
+negative and false positive percentages, or the predictions that proved
+to be incorrect, are 23.20\% and 15.85\%, respectively. The re-weighing
+model increased the number of false negatives by 1.14\% and decreased
+the number of false positives by 0.83\%. We are most concerned about the
+false positives, which indicate defendants who are predicted to
+recidivate and do not actually recidivate. While there is a slight
+improvement in the percent of false positives, it is still relatively
+high at 15.85\%, indicating that we should further precede with
+additional bias mitigation techniques.
+
+\hypertarget{in-processing-approach-prejudice-remover}{%
+\subsection{In-Processing Approach: Prejudice
+Remover}\label{in-processing-approach-prejudice-remover}}
+
+Our fairness metrics after running the prejudice remover are: 0.367 for
+statistical parity difference, 0.321 for equal opportunity difference,
+0.344 for average odds difference, and 3.3 for disparate impact ratio.
+Like the model performance metrics suggested, the prejudice remover
+approach resulted in an increased benefit for Black defendants. As the
+orange bars on these plots show, the values of the fairness metrics have
+reversed from their baseline values. Now all four metrics suggest an
+unfair advantage for Black defendants. Thus, this approach removed the
+model's prejudice against Black people, but it did not result in a
+``fair'' model.
+
+The highest percentage in this confusion matrix is for false positives,
+defendants who the model predicted to recidivate and actually did not,
+at 37.76\%. This false positive percentage concerns us because we do not
+want defendants who do not recidivate to have unfairly long sentences
+due to their (incorrectly) predicted recidivism. The number of true
+positives, defendants who the model predicted to recidivate and actually
+recidivated, is lower at 23.88\%. The percentage of false negatives is
+lower, whereas the true negative is low at 14.33\%. This indicates that
+the model does not do a good job in predicting the defendants who don't
+recidivate. It is very likely that it will incorrectly predict that
+someone will recidivate.
+
+After the prejudice remover, the accuracy of the model is fairly similar
+across the board of races, but the model's accuracy is only 38.21\%,
+which is much lower than the baseline and reweighing models. This
+accuracy score means that the model makes accurate predictions only 38\%
+of the time.
+
+\hypertarget{post-processing-approach-calibrated-equalized-odds}{%
+\subsection{Post-Processing Approach: Calibrated Equalized
+Odds}\label{post-processing-approach-calibrated-equalized-odds}}
+
+After running the calibrated equalized odds algorithm, our fairness
+metrics are: 0.134 for statistical parity difference, 0.088 for equal
+opportunity difference, 0.146 for average odds difference, and 1.155 for
+disparate impact ratio. As we can see from the orange bars, the
+calibrated equalized odds approach results in all four metrics now
+suggesting a benefit for the originally unprivileged group, Black
+defendants. Though the values for statistical parity difference and
+average odds difference are slightly above the range of what is
+considered fair, the margin is much smaller than the original margin
+between the value and the range of fairness (as indicated by the blue
+bars). Thus, the calibrated equalized odds approach successfully
+counteracts the bias against Black defendants and results in a mostly
+fair model.
+
+The highest percentage in this confusion matrix is for true negatives,
+defendants who the model predicted to not recidivate and actually did
+not, at 50.04\%. The number of true positives, defendants who the model
+predicted to recidivate and actually recidivated, is much lower at
+3.11\%. The percentage of false negatives is somewhat high for this
+baseline model, whereas the amount of false positives is very low at
+2.05\%. This indicates that the model does a good job in predicting the
+defendants who don't recidivate. It is very unlikely that it will
+incorrectly predict that someone will recidivate. Though the accuracy
+score for the model is 0.53, lower than the baseline, this model is much
+better at identifying true negatives.
+
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
+
+\hypertarget{limitations}{%
+\subsection{Limitations}\label{limitations}}
+
+Our project tries to mitigate bias within existing data, rather than
+within the methods used to collect this data. The data are collected
+using a risk assessment survey, where defendants are asked a series of
+questions (see introduction) which are supposedly used to determine
+whether someone will recidivate or not. Many of these questions involve
+proxy variables for race. The creation and facilitation of this survey
+can therefore be assessed for bias mitigation as the methods used may
+potentially privilege some groups over others. As a result, the bias
+exists even before values are collected, limiting the debiasing work
+able to be done on the output values.
+
+Furthermore, AIF360 processed the data we used in this project the same
+way ProPublica processed their data. AIF360 simplifies the original
+defendant data and their COMPAS scores to make the data easier to
+analyze. However, this initial processing may lose some details captured
+in the original raw data. As we have noted throughout this paper, bias
+can be introduced in any step of the data science pipeline and AIF360's
+initial data processing step is no exception.
+
+AIF360 contains an array of processing methods that have been written by
+various scholars of fairness \citep{aif360-oct-2018}. These methods are
+each distinct and apply a different technique to mitigate bias. Our
+results only contain three processing methods, one for each step in
+processing. This choice was based on time constraints as well as our
+working environment. For example, the code for most of the other
+in-processing methods would not run for us. In-processing on its own
+works to debias an algorithm, not the input or output values, so some of
+those methods may not work on certain algorithms. We had similar
+challenges with pre and post-processing, where the methods were
+challenging to run and would require more time to be spent debugging in
+order to make them work. Had we used alternative processing methods, our
+results would be quite different, as the method changes apply a
+different mathematical technique to the input.
+
+\hypertarget{future-work}{%
+\subsection{Future Work}\label{future-work}}
+
+In this project, we have only focused on a few select fairness metrics
+and a few debiasing algorithms. We looked at group fairness metrics
+because we specifically wanted to mitigate systemic racial bias in the
+COMPAS algorithm, but other researchers could compare how well debiasing
+algorithms work using individual fairness metrics. Future work could
+also include conducting more tests using other debiasing algorithms to
+see if the COMPAS algorithm could be made more fair. One significant
+limitation of our work is that some of the in-processing algorithms
+provided in the AIF360 toolkit were difficult to integrate into our code
+and we were unable to run many of them. The processing methods we chose,
+though effective, were based on our work environments, and there are
+other existing algorithms that attempt to mitigate bias.
+
+We only looked at defendants labeled as Caucasian and African-American
+in our dataset, which diminishes the generalizability and nuance of our
+results. Comparing how COMPAS assesses defendants of other races would
+be an incredibly relevant extension of our existing work. Furthermore,
+\citet{larson2016we} also found that age was the most predictive factor
+of a higher risk score: ``Defendants younger than 25 years old were 2.5
+times as likely to get a higher score than middle aged offenders, even
+when controlling for prior crimes, future criminality, race and gender''
+\citep{larson2016we}. In addition, female defendants were more likely to
+receive a higher recidivism risk score than male defendants when
+controlling for the aforementioned factors \citep{larson2016we}. Thus,
+some extensions of our work could include attempting to remove age and
+gender bias and determining the most effective processing algorithms to
+do so.
+
+The dataset we are working with only contains defendants' criminal
+history records from Broward County, Florida. Many other states,
+including New York and Massachusetts, use COMPAS to calculate recidivism
+scores, and a further study could apply this paper's methods to the data
+coming out of other states. By conducting experiments on multiple states
+the baseline model would be trained on a larger and more diverse set of
+data, which could improve its accuracy. In addition, by adding more
+geographically diverse data to our set we can start to create more
+generalizable conclusions about our results. However, future researchers
+must keep in mind the real-life consequences of recidivism risk
+algorithms like COMPAS and how research on methods of debiasing
+algorithms could potentially be utilized to justify the continued use of
+biased algorithms.
+
+\hypertarget{final-thoughts}{%
+\subsection{Final Thoughts}\label{final-thoughts}}
+
+Through our methods, we concluded that the pre-processing and
+post-processing methods mitigate bias and increase fairness most
+successfully. The in-processing technique was effective at mitigating
+bias, but did not result in a mathematically fair model. However, these
+techniques only treat the symptoms of racial bias in the justice system,
+as opposed to addressing the root cause, systemic racism. We need to
+question why these flawed algorithms exist and why they can be used to
+determine someone's fate in the justice system. While we may have
+methods for reducing bias algorithms such as COMPAS, the focus should be
+on addressing the over-policing and over-criminalization of Black
+communities that result in biased data and algorithms.
+
+\hypertarget{ethical-statement}{%
+\subsection{Ethical Statement}\label{ethical-statement}}
+
+Throughout this project, we endeavored to debias an incredibly powerful
+algorithm that can change the course of someone's life. However, we want
+to recognize that debiasing methods are not the most effective or
+beneficial way to uproot bias within the American justice system. The
+problems within the justice system are much more complex than an
+algorithm and are rooted in the United States's history of racism. In
+addition, algorithmic bias is much more complex than the algorithm
+itself; algorithmic bias comes from the people who produce the
+algorithm. We all have our own beliefs, biases, and situated knowledge
+that impact and subsequently limit everything we create.
+
+\hypertarget{acknowledgements}{%
+\section{Acknowledgements}\label{acknowledgements}}
+
+We would like to thank our sponsor organization Women At The Table and
+Professor Ben Baumer for introducing us to this incredibly meaningful
+work, supporting us, and answering the many questions we had throughout
+this semester-long project. We would specifically like to thank Sofia
+Kypraiou for her mentorship and patience throughout this process.
 

### 2022-05-03 Paper_Draft2/Draft2.tex
embedding all the images

@@
 % Abstract (Do not insert blank lines, i.e. \\)
-\abstract{A variety of disciplines use risk assessment instruments to
-help humans make data-driven decisions. Northpointe, a software company,
-created an algorithmic risk assessment instrument known as the
-Correctional Offender Management Profiling for Alternative Sanctions
-(COMPAS). COMPAS uses various behavioral and psychological metrics
-related to recidivism to assist justice systems in assessing a
-defendant's potential recidivism risk. Angwin et al.~published a
-ProPublica article in which they conclude that the racial biases in the
-criminal justice system are reflected in the COMPAS recidivism risk
-scores. In response, Dieterich et al.~published a rebuttal on behalf of
-Northpointe defending the COMPAS algorithm and refuting Angwin et al.'s
-allegation of racial bias. Using a human rights framework adopted from
-the organizations Women at the Table and AI Fairness 360, we use
-debiasing algorithms and fairness metrics to analyze the argument
-between Northpointe and ProPublica and determine whether and to what
-extent there is racial bias in the COMPAS algorithm. All four group
-fairness metrics determine that the COMPAS algorithm favors white
-defendants over Black defendants. Our research found that the pre and
-post-processing bias mitigation algorithms, specifically reweighing and
-calibrated equalized odds, are the most effective at improving
-fairness.}
+\abstract{A variety of disciplines use risk assessment instruments to help humans
+make data-driven decisions. Northpointe, a software company, created an
+algorithmic risk assessment instrument known as the Correctional
+Offender Management Profiling for Alternative Sanctions (COMPAS). COMPAS
+uses various behavioral and psychological metrics related to recidivism
+to assist justice systems in assessing a defendant's potential
+recidivism risk. Angwin et al.~published a ProPublica article in which
+they conclude that the racial biases in the criminal justice system are
+reflected in the COMPAS recidivism risk scores. In response, Dieterich
+et al.~published a rebuttal on behalf of Northpointe defending the
+COMPAS algorithm and refuting Angwin et al.'s allegation of racial bias.
+Using a human rights framework adopted from the organizations Women at
+the Table and AI Fairness 360, we use debiasing algorithms and fairness
+metrics to analyze the argument between Northpointe and ProPublica and
+determine whether and to what extent there is racial bias in the COMPAS
+algorithm. All four group fairness metrics determine that the COMPAS
+algorithm favors white defendants over Black defendants. Our research
+found that the pre and post-processing bias mitigation algorithms,
+specifically reweighing and calibrated equalized odds, are the most
+effective at improving fairness.}
 
@@
 therefore has 5,723 rows with information on defendant race, age,
-gender, prior crimes, and two-year recidivism rate. The average
-African-American defendant in our dataset is 29 years old, male, and has
-committed two prior crimes. The average Caucasian defendant in our
-dataset is 35 years old, male, and has committed one prior crime. The
-median defendant for both races do not have any juvenile convictions.
+gender, prior crimes, and two-year recidivism rate (Figure
+\ref{fig:table snip}). The distributions of age (Figure
+\ref{fig:age plot}), prior charges (Figure \ref{fig:prior plot}), and
+recidivism (Figure \ref{fig:recidivism mosaic} and Table
+\ref{tab:recid table}) all vary by race. The average African-American
+defendant in our dataset is 29 years old, male, and has committed two
+prior crimes. The average Caucasian defendant in our dataset is 35 years
+old, male, and has committed one prior crime. The average defendant for
+both races does not have any juvenile convictions.
 
@@
 With our baseline model, we ran the four different group fairness
-metrics we chose and compared the results (pictured in Table
-\ref{tab:baseline metrics table}).
+metrics we chose and compared the results (pictured in Figre
+\ref{fig:baseline metrics}).
 
@@
 
-Our confusion matrix for the baseline model indicates that the false
-positive and false negative percentages are 16.68\% and 22.06\%, which
-are the lowest values for this matrix. The highest value is the value of
-true positives, at 35.41\%, showing that of the total number of
-predictions, 35.41\% were correct predictions of recidivism.
+Our confusion matrix for the baseline model (see Figure
+\ref{fig:baseline matrix}) indicates that the false positive and false
+negative percentages are 16.68\% and 22.06\%, which are the lowest
+values for this matrix. The highest value is the value of true
+positives, at 35.41\%, showing that of the total number of predictions,
+35.41\% were correct predictions of recidivism.
 
-\begin{longtable}[]{@{}llll@{}}
-\caption{Results of Baseline Fairness Metric Analysis
-\label{tab:baseline metrics table}}\tabularnewline
-\toprule
-Fairness Metric & Ideal Value & Baseline Value & Benefited Group \\
-\midrule
-\endfirsthead
-\toprule
-Fairness Metric & Ideal Value & Baseline Value & Benefited Group \\
-\midrule
-\endhead
-Statistical Parity Difference & 0 & -0.14 & White Defendants \\
-Disparate Impact Ratio & 1 & 0.47 & White Defendants \\
-Average Odds Difference & 0 & -0.44 & White Defendants \\
-Equal Opportunity Difference & 0 & -0.41 & White Defendants \\
-\bottomrule
-\end{longtable}
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{../images/baseline_metrics}
+
+}
+
+\caption{Fairness Metrics of Baseline Model.
+For each of the four fairness metrics, the green section represents the range of fair values.
+The blue bars indicate the values of each of the metrics for the baseline model.}\label{fig:baseline metrics}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{../images/baseline_matrix}
+
+}
+
+\caption{Confusion Matrix of the Baseline Model}\label{fig:baseline matrix}
+\end{figure}
 
@@
 impact ratio. All of these values now fall within the margins of
-fairness (see reweighing metrics plot). These values make sense because
-the reweighing algorithm chooses different weights depending on whether
-an attribute is protected. This does an effective job of eliminating the
-bias, shown by each of the values in the graph being within the ``fair''
-range of the particular fairness metric. Overall, these fairness metrics
-show that the reweighing algorithm improves the bias in the COMPAS
-algorithm.
+fairness (see Figure \ref{fig:reweighing metrics}). These values make
+sense because the reweighing algorithm chooses different weights
+depending on whether an attribute is protected. This does an effective
+job of eliminating the bias, shown by each of the values in the graph
+being within the ``fair'' range of the particular fairness metric.
+Overall, these fairness metrics show that the reweighing algorithm
+improves the bias in the COMPAS algorithm.
 
-The highest percentage in this confusion matrix is for the percentage of
-true negatives, defendants who were predicted to not recidivate and
-actually did not, at 36.24\%, 0.83\% higher than the true negatives in
-the baseline model. This means that the reweighing model marginally
-improved the baseline model accuracy in predicting people who did not
-recidivate. However, the percentage of true positives, or those who were
-predicted to recidivate and did recidivate, is 24.72\%, 1.13\% lower
-than that of the baseline model. This means that the reweighing model
-slightly lowered the accuracy of the baseline model in predicting people
-who did recidivate. As a result, the overall reweighing model accuracy
-remains about the same as the baseline model accuracy. The false
-negative and false positive percentages, or the predictions that proved
-to be incorrect, are 23.20\% and 15.85\%, respectively. The re-weighing
-model increased the number of false negatives by 1.14\% and decreased
-the number of false positives by 0.83\%. We are most concerned about the
-false positives, which indicate defendants who are predicted to
-recidivate and do not actually recidivate. While there is a slight
-improvement in the percent of false positives, it is still relatively
-high at 15.85\%, indicating that we should further precede with
-additional bias mitigation techniques.
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{../images/reweighting_metrics}
+
+}
+
+\caption{Fairness Metrics of Reweighing Model. For each of the four fairness metrics, the orange bars indicate the values for the reweighing model.}\label{fig:reweighing metrics}
+\end{figure}
+
+The highest percentage in the reweighing confusion matrix (Figure
+\ref{fig:reweighing matrix}) is for the percentage of true negatives,
+defendants who were predicted to not recidivate and actually did not, at
+36.24\%, 0.83\% higher than the true negatives in the baseline model.
+This means that the reweighing model marginally improved the baseline
+model accuracy in predicting people who did not recidivate. However, the
+percentage of true positives, or those who were predicted to recidivate
+and did recidivate, is 24.72\%, 1.13\% lower than that of the baseline
+model. This means that the reweighing model slightly lowered the
+accuracy of the baseline model in predicting people who did recidivate.
+As a result, the overall reweighing model accuracy remains about the
+same as the baseline model accuracy. The false negative and false
+positive percentages, or the predictions that proved to be incorrect,
+are 23.20\% and 15.85\%, respectively. The re-weighing model increased
+the number of false negatives by 1.14\% and decreased the number of
+false positives by 0.83\%. We are most concerned about the false
+positives, which indicate defendants who are predicted to recidivate and
+do not actually recidivate. While there is a slight improvement in the
+percent of false positives, it is still relatively high at 15.85\%,
+indicating that we should further precede with additional bias
+mitigation techniques.
+
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{../images/reweighing_matrix}
+
+}
+
+\caption{Confusion Matrix of the Reweighing Model}\label{fig:reweighing matrix}
+\end{figure}
 
@@
 
-Our fairness metrics after running the prejudice remover are: 0.367 for
+Our fairness metrics after running the prejudice remover are 0.367 for
 statistical parity difference, 0.321 for equal opportunity difference,
@@
 approach resulted in an increased benefit for Black defendants. As the
-orange bars on these plots show, the values of the fairness metrics have
-reversed from their baseline values. Now all four metrics suggest an
-unfair advantage for Black defendants. Thus, this approach removed the
-model's prejudice against Black people, but it did not result in a
-``fair'' model.
+orange bars on Figure \ref{fig:prejudice metrics} show, the values of
+the fairness metrics have reversed from their baseline values. Now all
+four metrics suggest an unfair advantage for Black defendants. Thus,
+this approach removed the model's prejudice against Black people, but it
+did not result in a ``fair'' model.
 
-The highest percentage in this confusion matrix is for false positives,
-defendants who the model predicted to recidivate and actually did not,
-at 37.76\%. This false positive percentage concerns us because we do not
-want defendants who do not recidivate to have unfairly long sentences
-due to their (incorrectly) predicted recidivism. The number of true
-positives, defendants who the model predicted to recidivate and actually
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{../images/prejudice_remover_metrics}
+
+}
+
+\caption{Fairness Metrics of Prejudice Remover Model. For each of the four fairness metrics, the orange bars indicate the values for the prejudice remover model.}\label{fig:prejudice metrics}
+\end{figure}
+
+The highest percentage in this confusion matrix (Figure
+\ref{fig:prejudice matrix}) is for false positives, defendants who the
+model predicted to recidivate and actually did not, at 37.76\%. This
+false positive percentage concerns us because we do not want defendants
+who do not recidivate to have unfairly long sentences due to their
+(incorrectly) predicted recidivism. The number of true positives,
+defendants who the model predicted to recidivate and actually
 recidivated, is lower at 23.88\%. The percentage of false negatives is
@@
 someone will recidivate.
+
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{../images/prejudice_matrix}
+
+}
+
+\caption{Confusion Matrix of the Prejudice Remover Model}\label{fig:prejudice matrix}
+\end{figure}
 
@@
 After running the calibrated equalized odds algorithm, our fairness
-metrics are: 0.134 for statistical parity difference, 0.088 for equal
+metrics are 0.134 for statistical parity difference, 0.088 for equal
 opportunity difference, 0.146 for average odds difference, and 1.155 for
-disparate impact ratio. As we can see from the orange bars, the
-calibrated equalized odds approach results in all four metrics now
+disparate impact ratio. As we can see in Figure \ref{fig:ceo metrics},
+the calibrated equalized odds approach results in all four metrics now
 suggesting a benefit for the originally unprivileged group, Black
@@
 
-The highest percentage in this confusion matrix is for true negatives,
-defendants who the model predicted to not recidivate and actually did
-not, at 50.04\%. The number of true positives, defendants who the model
-predicted to recidivate and actually recidivated, is much lower at
-3.11\%. The percentage of false negatives is somewhat high for this
-baseline model, whereas the amount of false positives is very low at
-2.05\%. This indicates that the model does a good job in predicting the
-defendants who don't recidivate. It is very unlikely that it will
-incorrectly predict that someone will recidivate. Though the accuracy
-score for the model is 0.53, lower than the baseline, this model is much
-better at identifying true negatives.
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{../images/calibrated_equalized_odds_metrics}
+
+}
+
+\caption{Fairness Metrics of Calibrated Equalized Odds Model.
+For each of the four fairness metrics, the orange bars indicate the values for the calibrated equalized odds model.}\label{fig:ceo metrics}
+\end{figure}
+
+The highest percentage in this confusion matrix (Figure
+\ref{fig:ceo matrix}) is for true negatives, defendants who the model
+predicted to not recidivate and actually did not, at 50.04\%. The number
+of true positives, defendants who the model predicted to recidivate and
+actually recidivated, is much lower at 3.11\%. The percentage of false
+negatives is somewhat high for this baseline model, whereas the amount
+of false positives is very low at 2.05\%. This indicates that the model
+does a good job in predicting the defendants who don't recidivate. It is
+very unlikely that it will incorrectly predict that someone will
+recidivate. Though the accuracy score for the model is 0.53, lower than
+the baseline, this model is much better at identifying true negatives.
+
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{../images/ceo_matrix}
+
+}
+
+\caption{Confusion Matrix of the Calibrated Equalized Odds Model}\label{fig:ceo matrix}
+\end{figure}
 

### 2022-05-03 Paper_Draft2/Draft2.Rmd

@@
 This exploration aims to evaluate anti-Black algorithmic bias and the differing effects of the COMPAS algorithm between white and Black defendants; as such, we filter the data to only include individuals whose race is listed as Caucasian or African-American.
-Our data therefore has 5,723 rows with information on defendant race, age, gender, prior crimes, and two-year recidivism rate.
+Our data therefore has 5,723 rows with information on defendant race, age, gender, prior crimes, and two-year recidivism rate (Figure \ref{fig:table snip}).
+The distributions of age (Figure \ref{fig:age plot}), prior charges (Figure \ref{fig:prior plot}), and recidivism (Figure \ref{fig:recidivism mosaic} and Table \ref{tab:recid table}) all vary by race.
 The average African-American defendant in our dataset is 29 years old, male, and has committed two prior crimes.
 The average Caucasian defendant in our dataset is 35 years old, male, and has committed one prior crime.
-The median defendant for both races do not have any juvenile convictions.
+The average defendant for both races does not have any juvenile convictions.
 
@@
 
-With our baseline model, we ran the four different group fairness metrics we chose and compared the results (pictured in Table \ref{tab:baseline metrics table}).
+With our baseline model, we ran the four different group fairness metrics we chose and compared the results (pictured in Figre \ref{fig:baseline metrics}).
 
@@
 
-Our confusion matrix for the baseline model indicates that the false positive and false negative percentages are 16.68% and 22.06%, which are the lowest values for this matrix.
+Our confusion matrix for the baseline model (see Figure \ref{fig:baseline matrix}) indicates that the false positive and false negative percentages are 16.68% and 22.06%, which are the lowest values for this matrix.
 The highest value is the value of true positives, at 35.41%, showing that of the total number of predictions, 35.41% were correct predictions of recidivism.
 
-```{r, echo = FALSE}
-baseline_metrics <- tribble(~`Fairness Metric`, ~`Ideal Value`, ~ `Baseline Value`, ~`Benefited Group`,
-             "Statistical Parity Difference", "0", "-0.14", "White Defendants",
-             "Disparate Impact Ratio", "1", "0.47", "White Defendants",
-             "Average Odds Difference", "0", "-0.44", "White Defendants",
-             "Equal Opportunity Difference", "0", "-0.41", "White Defendants")
+```{r baseline metrics, echo=FALSE, fig.align = 'center', fig.cap = 'Fairness Metrics of Baseline Model.
+For each of the four fairness metrics, the green section represents the range of fair values.
+The blue bars indicate the values of each of the metrics for the baseline model.', out.width= "100%"}
+knitr::include_graphics("../images/baseline_metrics.png")
 ```
 
-```{r, echo = FALSE}
-knitr::kable(baseline_metrics, caption = "Results of Baseline Fairness Metric Analysis \\label{tab:baseline metrics table}")
+```{r baseline matrix, echo=FALSE, fig.align = 'center', fig.cap = 'Confusion Matrix of the Baseline Model', out.width= "100%"}
+knitr::include_graphics("../images/baseline_matrix.png")
 ```
@@
 After running the reweighing algorithm, our fairness metrics are -0.015 for statistical parity difference, 0.015 for equal opportunity difference, 0.014 for average odds difference, and 0.98 for disparate impact ratio.
-All of these values now fall within the margins of fairness (see reweighing metrics plot).
+All of these values now fall within the margins of fairness (see Figure \ref{fig:reweighing metrics}).
 These values make sense because the reweighing algorithm chooses different weights depending on whether an attribute is protected.
@@
 
-The highest percentage in this confusion matrix is for the percentage of true negatives, defendants who were predicted to not recidivate and actually did not, at 36.24%, 0.83% higher than the true negatives in the baseline model.
+```{r reweighing metrics, echo=FALSE, fig.align = 'center', fig.cap = 'Fairness Metrics of Reweighing Model.
+For each of the four fairness metrics, the orange bars indicate the values for the reweighing model.', out.width= "100%"}
+knitr::include_graphics("../images/reweighting_metrics.png")
+```
+
+The highest percentage in the reweighing confusion matrix (Figure \ref{fig:reweighing matrix}) is for the percentage of true negatives, defendants who were predicted to not recidivate and actually did not, at 36.24%, 0.83% higher than the true negatives in the baseline model.
 This means that the reweighing model marginally improved the baseline model accuracy in predicting people who did not recidivate.
@@
 
+```{r reweighing matrix, echo=FALSE, fig.align = 'center', fig.cap = 'Confusion Matrix of the Reweighing Model', out.width= "100%"}
+knitr::include_graphics("../images/reweighing_matrix.png")
+```
+
 ## In-Processing Approach: Prejudice Remover
 
-Our fairness metrics after running the prejudice remover are: 0.367 for statistical parity difference, 0.321 for equal opportunity difference, 0.344 for average odds difference, and 3.3 for disparate impact ratio.
+Our fairness metrics after running the prejudice remover are 0.367 for statistical parity difference, 0.321 for equal opportunity difference, 0.344 for average odds difference, and 3.3 for disparate impact ratio.
 Like the model performance metrics suggested, the prejudice remover approach resulted in an increased benefit for Black defendants.
-As the orange bars on these plots show, the values of the fairness metrics have reversed from their baseline values.
+As the orange bars on Figure \ref{fig:prejudice metrics} show, the values of the fairness metrics have reversed from their baseline values.
 Now all four metrics suggest an unfair advantage for Black defendants.
@@
 
-The highest percentage in this confusion matrix is for false positives, defendants who the model predicted to recidivate and actually did not, at 37.76%.
+```{r prejudice metrics, echo=FALSE, fig.align = 'center', fig.cap = 'Fairness Metrics of Prejudice Remover Model.
+For each of the four fairness metrics, the orange bars indicate the values for the prejudice remover model.', out.width= "100%"}
+knitr::include_graphics("../images/prejudice_remover_metrics.png")
+```
+
+The highest percentage in this confusion matrix (Figure \ref{fig:prejudice matrix}) is for false positives, defendants who the model predicted to recidivate and actually did not, at 37.76%.
 This false positive percentage concerns us because we do not want defendants who do not recidivate to have unfairly long sentences due to their (incorrectly) predicted recidivism.
@@
 
+```{r prejudice matrix, echo=FALSE, fig.align = 'center', fig.cap = 'Confusion Matrix of the Prejudice Remover Model', out.width= "100%"}
+knitr::include_graphics("../images/prejudice_matrix.png")
+```
+
 After the prejudice remover, the accuracy of the model is fairly similar across the board of races, but the model's accuracy is only 38.21%, which is much lower than the baseline and reweighing models.
@@
 
-After running the calibrated equalized odds algorithm, our fairness metrics are: 0.134 for statistical parity difference, 0.088 for equal opportunity difference, 0.146 for average odds difference, and 1.155 for disparate impact ratio.
-As we can see from the orange bars, the calibrated equalized odds approach results in all four metrics now suggesting a benefit for the originally unprivileged group, Black defendants.
+After running the calibrated equalized odds algorithm, our fairness metrics are 0.134 for statistical parity difference, 0.088 for equal opportunity difference, 0.146 for average odds difference, and 1.155 for disparate impact ratio.
+As we can see in Figure \ref{fig:ceo metrics}, the calibrated equalized odds approach results in all four metrics now suggesting a benefit for the originally unprivileged group, Black defendants.
 Though the values for statistical parity difference and average odds difference are slightly above the range of what is considered fair, the margin is much smaller than the original margin between the value and the range of fairness (as indicated by the blue bars).
@@
 
-The highest percentage in this confusion matrix is for true negatives, defendants who the model predicted to not recidivate and actually did not, at 50.04%.
+```{r ceo metrics, echo=FALSE, fig.align = 'center', fig.cap = 'Fairness Metrics of Calibrated Equalized Odds Model.
+For each of the four fairness metrics, the orange bars indicate the values for the calibrated equalized odds model.', out.width= "100%"}
+knitr::include_graphics("../images/calibrated_equalized_odds_metrics.png")
+```
+
+The highest percentage in this confusion matrix (Figure \ref{fig:ceo matrix}) is for true negatives, defendants who the model predicted to not recidivate and actually did not, at 50.04%.
 The number of true positives, defendants who the model predicted to recidivate and actually recidivated, is much lower at 3.11%.
@@
 Though the accuracy score for the model is 0.53, lower than the baseline, this model is much better at identifying true negatives.
+
+```{r ceo matrix, echo=FALSE, fig.align = 'center', fig.cap = 'Confusion Matrix of the Calibrated Equalized Odds Model', out.width= "100%"}
+knitr::include_graphics("../images/ceo_matrix.png")
+```
 

### 2022-05-04 Paper_Draft2/Draft2.Rmd
Logistic regression model

@@
 We researched the definitions and applications of different fairness metrics [@ashokan2021fairness] to determine which metric would be most appropriate for our project.
+To begin, we implemented a logistic regression model predicting recidivism using the defendant data.
+This analyzes the baseline values without any bias mitigation.
+We use logistic regression, as it is the easiest model to interpret in the given context.
 
@@
 However as this technique is designed to work only with classifiers that can handle row-level weights, this may limit your modeling options” [@Ronaghan2019AI].
-We used a logistic regression model for this algorithm, as it is the easiest to interpret in the given context.
 After running the fairness metrics using the pre-processing algorithm, we were able to compare our results to the baseline metrics from the previous section.

### 2022-05-04 Paper_Draft2/Draft2.tex
merge conflict resolved
delete citation

--- 2022-05-03 Paper_Draft2/Draft2.tex
@@
 metrics \citep{ashokan2021fairness} to determine which metric would be
-most appropriate for our project.
+most appropriate for our project. To begin, we implemented a logistic
+regression model predicting recidivism using the defendant data. This
+analyzes the baseline values without any bias mitigation. We use
+logistic regression, as it is the easiest model to interpret in the
+given context.
 
@@
 row-level weights, this may limit your modeling options''
-\citep{Ronaghan2019AI}. We used a logistic regression model for this
-algorithm, as it is the easiest to interpret in the given context. After
-running the fairness metrics using the pre-processing algorithm, we were
-able to compare our results to the baseline metrics from the previous
-section.
+\citep{Ronaghan2019AI}. After running the fairness metrics using the
+pre-processing algorithm, we were able to compare our results to the
+baseline metrics from the previous section.
 

### 2022-05-04 Paper_Draft2/Draft2.Rmd
delete citation
fixing typos adding wordcount

@@
 
-Our project builds on Women at the Table’s various debiasing algorithms used in its AI & Equality Human Rights Toolbox to conduct our own analyses on the COMPAS data set.
+Our project builds on Women at the Table’s various debiasing algorithms used in its AI & Equality Human Rights Toolbox to conduct our own analyses on the COMPAS dataset.
 Based on this analysis, we employ a human rights framework to contribute to the ProPublica and Northpointe debate and investigate whether or to what extent there is racial bias in the COMPAS algorithm.
@@
 
-The data we are using for this library addition is the COMPAS General Recidivism Risk Scores data set from the [AI Fairness 360](https://aif360.readthedocs.io/en/stable/modules/generated/aif360.datasets.CompasDataset.html#aif360.datasets.CompasDataset) (AIF360) toolkit.
+The data we are using for this library addition is the COMPAS General Recidivism Risk Scores dataset from the [AI Fairness 360](https://aif360.readthedocs.io/en/stable/modules/generated/aif360.datasets.CompasDataset.html#aif360.datasets.CompasDataset) (AIF360) toolkit.
 The AIF360 toolkit builds on the dataset released by ProPublica collaborators Larson et al.
@@
 
-```{r table snip, echo = FALSE, fig.align='center', fig.cap= "A snippet of the data set we will be using, containing information on a defendant's age, sex, race, criminal history, charge degree, charge description, and two-year recidivism outcome.", out.width= "100%"}
+```{r table snip, echo = FALSE, fig.align='center', fig.cap= "A snippet of the dataset we will be using, containing information on a defendant's age, sex, race, criminal history, charge degree, charge description, and two-year recidivism outcome.", out.width= "100%"}
 knitr::include_graphics("../images/table_snippet.png")
@@
 AI Fairness 360 (AIF360) is an open-source Python toolkit that seeks “to help facilitate the transition of fairness research algorithms to use in an industrial setting and to provide a common framework for fairness researchers to share and evaluate algorithms” [@aif360-oct-2018].
-It contains multiple data sets, including the COMPAS data set that accompanied @angwin2016machine.
+It contains multiple datasets, including the COMPAS dataset that accompanied @angwin2016machine.
 
@@
 This metric is best used when it is very important to catch positive outcomes while false positives are not exceptionally problematic [@Cortez2019How].
-This is not the case for the COMPAS data set, as false positives mean extra jail time for someone who will not actually re-offend.
+This is not the case for the COMPAS dataset, as false positives mean extra jail time for someone who will not actually re-offend.
 
@@
 
+This document contains 4793 words
+
 # Bibliography

### 2022-05-04 Paper_Draft2/Draft2.tex
fixed images

@@
 % Abstract (Do not insert blank lines, i.e. \\)
-\abstract{A variety of disciplines use risk assessment instruments to help humans
-make data-driven decisions. Northpointe, a software company, created an
-algorithmic risk assessment instrument known as the Correctional
-Offender Management Profiling for Alternative Sanctions (COMPAS). COMPAS
-uses various behavioral and psychological metrics related to recidivism
-to assist justice systems in assessing a defendant's potential
-recidivism risk. Angwin et al.~published a ProPublica article in which
-they conclude that the racial biases in the criminal justice system are
-reflected in the COMPAS recidivism risk scores. In response, Dieterich
-et al.~published a rebuttal on behalf of Northpointe defending the
-COMPAS algorithm and refuting Angwin et al.'s allegation of racial bias.
-Using a human rights framework adopted from the organizations Women at
-the Table and AI Fairness 360, we use debiasing algorithms and fairness
-metrics to analyze the argument between Northpointe and ProPublica and
-determine whether and to what extent there is racial bias in the COMPAS
-algorithm. All four group fairness metrics determine that the COMPAS
-algorithm favors white defendants over Black defendants. Our research
-found that the pre and post-processing bias mitigation algorithms,
-specifically reweighing and calibrated equalized odds, are the most
-effective at improving fairness.}
+\abstract{A variety of disciplines use risk assessment instruments to
+help humans make data-driven decisions. Northpointe, a software company,
+created an algorithmic risk assessment instrument known as the
+Correctional Offender Management Profiling for Alternative Sanctions
+(COMPAS). COMPAS uses various behavioral and psychological metrics
+related to recidivism to assist justice systems in assessing a
+defendant's potential recidivism risk. Angwin et al.~published a
+ProPublica article in which they conclude that the racial biases in the
+criminal justice system are reflected in the COMPAS recidivism risk
+scores. In response, Dieterich et al.~published a rebuttal on behalf of
+Northpointe defending the COMPAS algorithm and refuting Angwin et al.'s
+allegation of racial bias. Using a human rights framework adopted from
+the organizations Women at the Table and AI Fairness 360, we use
+debiasing algorithms and fairness metrics to analyze the argument
+between Northpointe and ProPublica and determine whether and to what
+extent there is racial bias in the COMPAS algorithm. All four group
+fairness metrics determine that the COMPAS algorithm favors white
+defendants over Black defendants. Our research found that the pre and
+post-processing bias mitigation algorithms, specifically reweighing and
+calibrated equalized odds, are the most effective at improving
+fairness.}
 
@@
 
+This document contains 4793 words
+
 % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2022-05-04 Paper_Draft2/Draft2.Rmd

@@
     email: bbaumer@smith.edu
-
-journal: water
+#journal: water
+#status: submit
 type: article
-status: submit
 bibliography: mybibfile.bib
@@
 
-The AIF360 toolkit contains various group and individual fairness metrics as well as pre-processing, in-processing, and post-processing algorithms that we used to debias the COMPAS algorithm [@aif360-oct-2018].
+The AIF360 toolkit contains various group and individual fairness metrics as well as pre-processing, in-processing, and post-processing algorithms that we use to debias the COMPAS algorithm [@aif360-oct-2018].
 Fairness metrics are mathematical measures of whether an algorithm treats members of different groups (such as racial or gender groups) equally.
@@
 Fairness is subjective; one person might consider an algorithm fair if groups are given the same treatment, while someone else might only consider the algorithm fair if groups as a whole receive the same outcomes.
-In our research we choose to use the definition of group fairness as our definition for assessing fairness in the processing approach.
+In our research we choose to use the definition of group fairness as our definition for assessing fairness in the processing approaches.
 Group fairness metrics take into account the attributes of a whole group as opposed to just one individual in the group, allowing us to represent systemic issues.
@@
 This in turn enforces a classifier’s independence from sensitive information (e.g., race).
-Similarly to pre-processing, we compared the results of our in-processing methods with both the baseline and the pre-processing model to gauge which method so far has the better group fairness.
+Similar to pre-processing, we compare the results of our in-processing methods with both the baseline and the pre-processing model to gauge which method so far has the better group fairness.
 
@@
 
-This document contains 4793 words
+```{r word count, include=F}
+wordcountaddin:::text_stats()
+```
 

### 2022-05-04 Paper_Draft2/Draft2.tex

--- 2022-05-04 Paper_Draft2/Draft2.tex
@@
 %=================================================================
-\documentclass[water,article,submit,moreauthors,pdftex]{mdpi}
+\documentclass[,article,,moreauthors,pdftex]{mdpi}
 
@@
 used in its AI \& Equality Human Rights Toolbox to conduct our own
-analyses on the COMPAS data set. Based on this analysis, we employ a
+analyses on the COMPAS dataset. Based on this analysis, we employ a
 human rights framework to contribute to the ProPublica and Northpointe
@@
 The data we are using for this library addition is the COMPAS General
-Recidivism Risk Scores data set from the
+Recidivism Risk Scores dataset from the
 \href{https://aif360.readthedocs.io/en/stable/modules/generated/aif360.datasets.CompasDataset.html\#aif360.datasets.CompasDataset}{AI
@@
 
-\caption{A snippet of the data set we will be using, containing information on a defendant's age, sex, race, criminal history, charge degree, charge description, and two-year recidivism outcome.}\label{fig:table snip}
+\caption{A snippet of the dataset we will be using, containing information on a defendant's age, sex, race, criminal history, charge degree, charge description, and two-year recidivism outcome.}\label{fig:table snip}
 \end{figure}
@@
 fairness researchers to share and evaluate algorithms''
-\citep{aif360-oct-2018}. It contains multiple data sets, including the
-COMPAS data set that accompanied \citet{angwin2016machine}.
+\citep{aif360-oct-2018}. It contains multiple datasets, including the
+COMPAS dataset that accompanied \citet{angwin2016machine}.
 
@@
 metrics as well as pre-processing, in-processing, and post-processing
-algorithms that we used to debias the COMPAS algorithm
+algorithms that we use to debias the COMPAS algorithm
 \citep{aif360-oct-2018}. Fairness metrics are mathematical measures of
@@
 fairness as our definition for assessing fairness in the processing
-approach. Group fairness metrics take into account the attributes of a
+approaches. Group fairness metrics take into account the attributes of a
 whole group as opposed to just one individual in the group, allowing us
@@
 outcomes while false positives are not exceptionally problematic
-\citep{Cortez2019How}. This is not the case for the COMPAS data set, as
+\citep{Cortez2019How}. This is not the case for the COMPAS dataset, as
 false positives mean extra jail time for someone who will not actually
@@
 turn enforces a classifier's independence from sensitive information
-(e.g., race). Similarly to pre-processing, we compared the results of
-our in-processing methods with both the baseline and the pre-processing
+(e.g., race). Similar to pre-processing, we compare the results of our
+in-processing methods with both the baseline and the pre-processing
 model to gauge which method so far has the better group fairness.

### 2022-05-04 Paper_Draft2/Draft2.Rmd

@@
 
-With our baseline model, we ran the four different group fairness metrics we chose and compared the results (pictured in Figre \ref{fig:baseline metrics}).
+With our baseline model, we ran the four different group fairness metrics we chose and compared the results (pictured in Figure \ref{fig:baseline metrics}).
 
-The statistical parity difference is -0.14.
+Statistical parity difference is -0.14.
 This indicates that there is a large difference between white and Black defendants regarding whether or not they recidivate.
@@
 We need to question why these flawed algorithms exist and why they can be used to determine someone’s fate in the justice system.
-While we may have methods for reducing bias algorithms such as COMPAS, the focus should be on addressing the over-policing and over-criminalization of Black communities that result in biased data and algorithms.
+While we may have methods for reducing biased algorithms such as COMPAS, the focus should be on addressing the over-policing and over-criminalization of Black communities that result in biased data and algorithms.
 
@@
 
-```{r word count, include=F}
+```{r word count, warning=FALSE, message=FALSE, echo=FALSE}
 wordcountaddin:::text_stats()

### 2022-05-04 Paper_Draft2/Draft2.tex

@@
 With our baseline model, we ran the four different group fairness
-metrics we chose and compared the results (pictured in Figre
+metrics we chose and compared the results (pictured in Figure
 \ref{fig:baseline metrics}).
 
-The statistical parity difference is -0.14. This indicates that there is
-a large difference between white and Black defendants regarding whether
-or not they recidivate. The algorithm unfairly benefits white defendants
+Statistical parity difference is -0.14. This indicates that there is a
+large difference between white and Black defendants regarding whether or
+not they recidivate. The algorithm unfairly benefits white defendants
 over Black defendants.
@@
 determine someone's fate in the justice system. While we may have
-methods for reducing bias algorithms such as COMPAS, the focus should be
-on addressing the over-policing and over-criminalization of Black
+methods for reducing biased algorithms such as COMPAS, the focus should
+be on addressing the over-policing and over-criminalization of Black
 communities that result in biased data and algorithms.
@@
 Kypraiou for her mentorship and patience throughout this process.
+
+\begin{longtable}[]{@{}lll@{}}
+\toprule
+Method & koRpus & stringi\tabularnewline
+\midrule
+\endhead
+Word count & 4719 & 4537\tabularnewline
+Character count & 29868 & 29825\tabularnewline
+Sentence count & 280 & Not available\tabularnewline
+Reading time & 23.6 minutes & 22.7 minutes\tabularnewline
+\bottomrule
+\end{longtable}
 
